---
title: "How disrupted spatial fluxes reshape food web structure along the river–sea continuum"
author: "Ismaël Lajaaiti, Eric Edeline, Anik Brind'Amour, Alain Danet"
bibliography: references.bib
prefer-html: true
format:
  html:
    toc: true
    embed-resources: true
  pdf:
    documentclass: article
    toc: true
comments:
    hypothesis: true
crossref:
  custom:
    - kind: float
      reference-prefix: Figure S
      space-before-numbering: false
      key: sfig
      latex-env: sfig
      latex-list-of-description: Supporting Figure
---

```{r}
#| output: false
#| echo: false
library(renv)
renv::load()
renv::restore()
renv::status()

library(here)
library(DiagrammeR)
library(targets)
tar_config_set(
  store = here::here("_targets")
)
```

## Abstract

Anthropic disturbances and ecosystem processes transcend the artificial
boundaries between land, freshwater, and sea.
While these systems are interconnected through species and material fluxes,
they are often studied in isolation.
In response, the concept of metaecosystem has emerged as a powerful framework
to understand how fluxes among ecosystems shape their functions and dynamics.
Disrupting these fluxes can have dramatic consequences that ripple across
entire landscapes.
In aquatic systems, dams and nutrient pollution are known to strongly affect
fish communities. Yet, little is still known about how these effects propagate
and potentially accumulate, through hydrographic basins and ultimately to the
sea.
Using bayesian spatio-temporal statistical models (R-INLA), we aim to analyse
existing data on food-web structure across France, combined with environmental
variables (dams, nutrients, …), to quantify how breaking or increasing fluxes
among systems disturbs food-web structure locally and how these disturbances
spread throughout the metaecosystem.

**Keywords:**
*foodweb ecology,
stream network,
spatial autocorrelation,
bayesian statistics.*


## Introduction

While patterns in community change within river network have been well studied,
these studies have mostly focus on specise richness [@Bonnaffé-etal-2024b], or
roughly defined functional composition [@Vannote-etal-1980].

### River-estuary-sea continuum

- [@Xenopoulos-etal-2017] highlights the importance of accounting for the full
gradient from headwater to the sea. Freshwater and coastal ecosystems are
connected by flux of carbone and nutrients (N, P), as well as, movement of
migratory species. Reviews work connecting river and sea, that said, most of
them focus on chemical fluxes, remain relatively local, and none of them focused
on the food web structure.

- Catchment land-use impacts macro-invertebrates along the river-sea continuum
with no dilution effect [@Bierschenk-Savage-Matthaei-2017].

### Freshwater ecosystems

- The River Continuum Concept has shaped the freshwater ecology researsh by
links physical gradients along rivers to predictable changes in community
structure, energy sources, and ecosystem functioning
[@Vannote-etal-1980; @Doretto-Piano-Larson-2020].

- Serial Discontinuity Concept is a framework to predict the impact of
discontinuities, such as dams, on biotic and abiotic variables along the river
continuum [@Ward-Stanford-1995].

- The Flood Pulse Concept (FPC) complements the RCC as it focuses on the lateral
dimension of rivers and its role in explaining community productivity, while the
RCC focused on the longitudinal aspect. The FPC states that its the rythm of
flood and recession -- that is, the pulse -- which structures ecosystem
productivity, not the flow [@Junk-Bayley-Sparks-1989].
  - Terrestrial organic matter (leaf litter, vegetation, soil nutrients) gets
  dissolved and mobilized into the water
  - Aquatic plants and algae explode in productivity in the shallow, sunlit
  floodwaters
  - Fish and invertebrates move onto the floodplain to feed and reproduce
  - When waters recede, nutrients and organic matter flush back into the main
  channel

- The River Wave concept as unifying framework of different concepts of
freshwater ecology [@Humphries-Keckeis-Finlayson-2014].

- Community richness and hence food web structure is expected to be driven by
nutrient availability [@Ho-Altermatt-Carraro-2023].

### Multiple stressors

- Synergistic and antagonistic effects of multiple stressors
[@Jackson-Pawar-Woodward-2021a; @Orr-etal-2024a].

- Temperature can interact with nutrients and impact food web structure in a 
nonlinear way [@Bonnaffé-etal-2024b, @Binzer-etal-2012].

- [@Danet-etal-2021b; @Bonnaffé-etal-2024b].

### Connectivity

- Small dispersal stabilizes through spatial structure and asynchrony
[@LeCraw-Kratina-Srivastava-2014; @McCann-Rasmussen-Umbanhowar-2005;
@Pillai-Gonzalez-Loreau-2011].

- A metaecosystem model shows that dispersal of living organisms (that consume 
nutrients) is stabilising,while dispersal of non-living compartments such as 
nutrients or detritus is destabilising [@Gounand-etal-2014].

- Prey switching of mobile top predators can stabilize connected food webs
through ‘spatial coupling’
[@LeCraw-Kratina-Srivastava-2014; @McCann-Rasmussen-Umbanhowar-2005;
@Pillai-Gonzalez-Loreau-2011].

- At high dispersal/connectivity the stabilizing effect wanes as strongly 
connected patches behave as a single patch
[@LeCraw-Kratina-Srivastava-2014; @McCann-Rasmussen-Umbanhowar-2005].

- Studies between pairs of the following connectivity, complexity and stability
are common. However, studies investigating how these are interconnected are
rare [@LeCraw-Kratina-Srivastava-2014].

- Connectivity has 4 dimensions: longitudinal, lateral, vertical and temporal
[@Grill-etal-2019].

- We identified five pressure factors to represent the main human
interferences within the four dimensions of fluvial or river connectivity 
[@Grill-etal-2019]:
  (1) river fragmentation (longitudinal);
  (2) flow regulation (lateral and temporal);
  (3) sediment trapping (longitudinal, lateral and vertical);
  (4) water consumption (lateral, vertical and temporal);
  (5) infrastructure development in riparian areas and floodplains
(lateral and longitudinal).

- In freshwater system, fish dispersal is limited. A meta-analysis found that
the mean dispersal kernel is around 10km [@Comte-Olden-2018].

- Neutral metacommunity model explains biodiversity pattern in the Mississipi
river, showing the key role of dispersal in explaining community structure along
the river continuum [@Muneepeerakul-etal-2008].

- Connectivity structures and mititgates effect of drying events in
macroinvertebrates communities [@Chalmandrier-etal-2025b].

- It is key to adopt a network perspective to understand how diversity is
created and maintained within riverine ecosystems [@Altermatt-2013b]. With this
perspective one can account for the dendritic network and so the spatial
processes that shape diversity. 

- From points [@Death-Winterbourn-1995]... to lines (RCC, SDC)... to dendritic
networks [@Shepherd-Ellis-1997].

- River network characteristics, local network properties and their
interaction were important individual contributors to explaining local
species richness of aquatic insects [@Altermatt-Seymour-Martinez-2013].
Effect of connectivity on richness supported experimentally 
[@Carrara-etal-2012].


### Dams

- Decrease river connectivity / reduce migration [@Dean-etal-2023b;
@He-etal-2024; @Pedersen-etal-2012].

- Reservoir, lotic from lentic conditions, can result in hypoxic conditions.
Favor phytoplancktons and macrophytes [@Ding-etal-2026; @Dean-etal-2023b].
Facilitate the spread of invasive species [@Johnson-Olden-VanderZanden-2008b].

- Reservoir increase residence time of nutrients upstream and decrease the
downstream output, which can result in a decrease in primary productivity.

- Hydropeaking - that is, the sudden release of large water to answer the
higher electric demand  - can harm riverine community due to sudden change
in water level and water flow.

- Cumulative impact of dams can reduce the nutrient input in deltas
[@Dean-etal-2023b].

- A review [@Ellis-Jones-2013b] suggests that two recovery gradients exist in
regulated river:
  1) a longer, thermal gradient of around 100 km;
  2) a shorter resource subsidy gradient of around 1-4 km.

## Methods

### Food web inference

To infer local food web structure, we used a metaweb approach. 
That is, we build a web compiling the potential feeding interactions between 
all species of the dataset. Importantly, our metaweb takes into account 
intra-specific variation in species diet. We did so by defining trophic 
interactions at different life stages for each species.

#### Metaweb nodes

The nodes of our metaweb represent a species at a given lifestage. Specifically,
fish species were divided into nine body class sizes evenly distributed. Each
body size class within a given species is what we call a ‘trophic species’
assuming that conspecifics of the same body size class share similar trophic
interactions because trophic interactions are largely determined by
predator–prey body size ratio in freshwater ecosystems
[@Brose-Williams-Martinez-2006].

In addition of fish species, seven basal nodes (resources) were added at the
base of the metaweb: detritus, biofilm, phytoplanckton, zooplanckton,
macrophytes and zoobenthos.
We represent the basal food web structure in the Appendix
(see @sfig-foodweb-base).
We assumed that these resources were present at all sampling sites.

#### Infer missing fish lengths

Most length of fishes are measured but not all.
Yet, we need individual fish length to determine their diet based on their 
ontogeny.
We infer missing fish size using information on the given fishing operation.
When there are more than three observations of the fish species, we use a
truncated normal distribution (package truncdist; @Nadarajah-Kotz-2006).
The mean and variance are given by the modes of the observed distribution of the
fishing operation.
The minimum and maximum of the distribution are given by the minimal and maximal
length observed on the entire data set for that species.
We do so to avoid infering irrealistic lengths.

#### Get fish diet

We use fishbase as a primary source to extract fish diet at their different
life stages, to get on which basal categories species feed on and whether they
are piscivorous or not.
We detail in the Supporting Information how made the correspondance between our
categories and the one of fishbase.
We remove species that have an occurence lower than
`r tar_read(params)$occurence_min` in the entire dataset which comprise
`r tar_read(n_rare)` species.
This enable to avoid retrieving diet information of species that will have a
minimal impact on food web structure due to their rarity.
We add a larvae stage when missing, assuming that fish larvae feed on
zooplankton [@Dabrowski-1984b].
We assume that fish individuals are larvae if they measure less than 2 cm.
To classify individuals as juvenile or adult we use the length at maturity of
the species as a threshold.
When this information was not available we used a single stage (*combined*) by
merging the diets of the two stages.
We did not consider a recruit stage.

#### The metaweb

Predation window are assumed to be 0.03 to 0.45 of individual body size for
piscivorous species.
We then use the package [foodwebbuilder](https://github.com/Fellows-of-the-Fish-Food-Webs-F3W/foodwebbuilder)
to build the metaweb and the local food webs through subsampling.
We investigated the effect of the number of size classes on the structure of the
reconstructed metaweb.
We observed little influence of the number of size classes on the metaweb
connectance @sfig-metaweb-sizeclass.
So, we decided to split each species 5 size classes as in @Bonnaffé-etal-2024b.

::: {.callout-note}
The choice of a constant predation window has an important implication.
It implies that what only matters in terms of ecosystem functioning is the
individual size (once the species is known to be piscivorous).
In this setting, two piscivorous individual of different species but same size
will eat the very same things.
:::

#### Subsampling the metaweb to get local food webs

## Results

## Discussion

### Limitations

Regarding food web structure, we have simplified the base of the web and exclude
macro-invertebrates.
This choice naturally emphasizes the functional role of fishes in shaping food
web structures.
Yet, it would be valuable to include macro-invertebrates within the food web to
get a more exhaustive picture of how aquatic community respond to stressors
along the riversea continuum.

# Supporting Information

## Structure of the food web base

::: {#sfig-foodweb-base}

![](`r riversea::to_rel(targets::tar_read(foodweb_fig))`)

Assumed structure of the food web base.

:::

## Model DAG

Put model DAG here using targets.

## Venn diagramm


::: {#sfig-venn}

![](`r riversea::to_rel(targets::tar_read(venn_diagram))`)

Venn diagram of common species between surveys.

:::

## Number of size classes and metaweb structure

::: {#sfig-metaweb-sizeclass}

```{r}
targets::tar_read(plot_metaweb_connectance)
```

How metaweb connectance varies with the number of size classes.

:::
