---
title: "Exploring coastal data"
format: html
execute:
    message: false
    warning: false
---

# 0. Setup

Load dependencies.

```{r}
#| output: false
library(renv)
renv::load()
renv::restore() # Install missing librairies.
renv::status() # Check the project state.
```

```{r}
library(here)
library(dplyr)
library(ggplot2)
set_theme(theme_minimal())
options(
    ggplot2.continuous.colour = "viridis",
    ggplot2.continuous.fill = "viridis"
)
```

```{r}
data_folder <- here("data", "coast")
list.files(data_folder)
```

We see that we have multiple csv files.
We have abundance and biomass data for the Atlantic (Nurse, Solper), Atlantic and Manche (POMET).

# 1. Pomet campagne

## 1.1. Data cleaning

First let's load files.

```{r}
d.poisson <- read.csv(here(data_folder, "poisson_pomet.csv"), sep = ";")
head(d.poisson)
```

Let's do a bit of cleaning.
Remove irrelevant columns.

```{r}
d.poisson <- d.poisson |> select(-c(Commentaire, Ecologique, Trophique, Engin_peche, Zone))
names(d.poisson)
```

Rename columns for clarity.

```{r}
d.poisson <- d.poisson |> rename(
    zone = ID_Zone,
    year = Annee,
    month = Mois,
    hour = Heure,
    date = Date,
    trait = Trait,
    start_x = Coord_Deb_xmin,
    start_y = Coord_Deb_ymin,
    end_x = Coord_Fin_xmin,
    end_y = Coord_Fin_ymin,
    depth = Prof,
    species = NomScient,
    position = Position,
    abundance = Nind_esp,
    weight_batch = Pds_esp
)
head(d.poisson)
```

We see that the zone ID column also contains zone names which is not necessary.

```{r}
unique(d.poisson$zone)
```

We want to keep only the ID.

```{r}
d.poisson$zone <- substr(d.poisson$zone, 1, 6)
unique(d.poisson$zone)
```

Now let's look at which species are more abundant on average.

```{r}
n <- 10
d <- d.poisson |>
    group_by(species) |>
    summarise(abundance_avg = mean(abundance)) |>
    arrange(desc(abundance_avg)) |>
    slice(1:n)
d
```

Let's pull these species and plot their time series.

::: {.callout-note}
Is it expected that "Spratus" and "Spratus spratus" are two distinct species?
This may be an error when species name was written.
If so species names should be cleaned up.
:::

```{r}
common_sp <- d |> pull(species)

d <- d.poisson |>
    filter(species %in% common_sp) |>
    group_by(species, year) |>
    summarise(abundance_avg = mean(abundance), .groups = "drop")
d$log_abundance <- log(d$abundance_avg)

ggplot(d, aes(x = year, y = log_abundance, color = species)) +
    geom_point() +
    geom_line() +
    labs(x = "Year", y = "Log abundance")
```


## 1.2. Plotting traits

First coordinates need to be formatted.
They are given as strings and should be converted to numeric values.

```{r}
library(stringr)
d.poisson <- d.poisson |>
    mutate(
        start_x = start_x |> str_replace(",", ".") |> as.numeric(),
        start_y = start_y |> str_replace(",", ".") |> as.numeric(),
        end_x = end_x |> str_replace(",", ".") |> as.numeric(),
        end_y = end_y |> str_replace(",", ".") |> as.numeric(),
    )

summary(d.poisson[, c("start_x", "start_y", "end_x", "end_y")])
```

```{r}
ggplot(d.poisson, aes(x = start_x, y = start_y)) +
    geom_segment(aes(xend = end_x, yend = end_y, color = zone)) +
    coord_equal() +
    labs(x = "Longitutde", y = "Latitude")
```

Now let's add a map in the background with `leaflet`.
Because we have many traits, we will focus on a single zone.
First, we need to convert traits to sf object.

```{r}
library(sf)
library(leaflet)

zone_plot <- first(d.poisson$zone)

traits_sf <- d.poisson |>
    filter(zone == zone_plot) |>
    rowwise() |>
    mutate(
        geometry = st_sfc(st_linestring(
            matrix(c(start_x, end_x, start_y, end_y), ncol = 2, byrow = FALSE)
        ), crs = 4326)
    ) |>
    st_as_sf()
```

Then, we can plot them, using year for the color gradient.

```{r}
pal <- colorNumeric(
    palette = "viridis",
    domain = traits_sf$year
)

leaflet(traits_sf) |>
    addTiles() |>
    addPolygons(
        data = traits_sf,
        color = ~ pal(year),
        opacity = 0.7,
        popup = ~ paste("Year: ", year)
    ) |>
    addLegend("bottomright",
        pal = pal, values = ~year,
        title = "Year",
        labFormat = labelFormat(big.mark = ""),
        opacity = 1
    )

```

Most traits follow the river suggesting that we have processed the data correctly.

Now let's focus on how fish sizes is measured.

## 1.3. Fish sizes

### 1.3.1. Preparing size data

First let's read the file containing fish size information.

```{r}
d.size <- read.csv(here(data_folder, "poisson_tailles_pomet.csv"), sep = ";")
head(d.size)
```

As before let's do a bit of cleaning before anything else.

```{r}
d.size <- d.size |> select(-c(
    Commentaire,
    Ecologique,
    Trophique,
    Engin_peche,
    Zone
))

head(d.size)
```

Then let's rename columns.

```{r}
d.size <- d.size |> rename(
    zone = ID_Zone,
    year = Annee,
    month = Mois,
    hour = Heure,
    date = Date,
    trait = Trait,
    start_x = Coord_Deb_xmin,
    start_y = Coord_Deb_ymin,
    end_x = Coord_Fin_xmin,
    end_y = Coord_Fin_ymin,
    depth = Prof,
    species = NomScient,
    position = Position,
    abundance = Nind_esp,
    weight_batch = Pds_esp,
    batch_size = Nind_esp_taille,
    oxygen = Pour_satO2,
    salinity = Salinite,
    length_cm = Longueur_fourche_cm,
    length_mm = Longueur_fourche_mm,
    temperature = Temperature_eau
)
head(d.size)
```

And cleaning zone ID.

```{r}
d.size$zone <- substr(d.size$zone, 1, 6)
head(d.size)
```

Furthermore, we can notice that length in mm are specified as caracters instead of numerics.

```{r}
class(d.size$length_mm)
```

```{r}
d.size$length_mm <- as.numeric(d.size$length_mm)
```

We can check on a subset of our dataframe that length in mm is ten times the length in cm.

```{r}
d <- d.size |>
    filter(zone == zone_plot) |>
    filter(!is.na(length_mm), year == 2014)

ggplot(d, aes(x = length_mm, y = length_cm)) +
    geom_point()
```

So let's remove the length in cm, which is a duplicate and less precise than the length in mm.

```{r}
d.size <- d.size |> select(-length_cm)
head(d.size)
```

### 1.3.2. Full size sampling

Now that the data is ready, we want to see at why frequency all individual fish sizes are measured.

We will begin with rough estimates, then refine these at the species and year levels.

```{r}
d <- d.size |>
    group_by(ID_interne_prelevement, species) |>
    summarise(n_measured = sum(batch_size), .groups = "drop")

d.join <- left_join(d.poisson, d, by = c("ID_interne_prelevement", "species"))
head(d.join)
```

To check that we did well, we verify that for a fishing event for which we know that all fish sizes have been measured that `abundance` = `n_measured`.

```{r}
d.join |> filter(ID_interne_prelevement == 61836630, species == "Abramis brama")
```

First, let's look at rough estimates.

```{r}
n_tot <- nrow(d.join) # 1 row: 1 species & 1 fishing event.
n_full <- nrow(d.join |> filter(abundance == n_measured))
n_full / n_tot
```

When filtering for full size sampling, we keep 83% of the rows of the original dataframe.
This means that most of the time all individual sizes are measured.
But, let's go in further details and how full size sampling depends on year, zone and species.

```{r}
d.sampling <- d.join |>
    filter(!is.na(n_measured)) |>
    group_by(species) |> 
    summarise(measurement_rate = mean(n_measured == abundance)) |>
    arrange(measurement_rate)
head(d.sampling)
```

We have only few species with low full measurement rate.
Even the species with worst full size sampling rate hase a rate of 0.5.
One reason that can explain why some species are more often fully measured than other is because of their abundance.
Measuring many individuals is cumbersome, so we expect a decreasing trend between full measurement rate and average abundance.

```{r}
d.abundance.avg <- d.poisson |>
    filter(!is.na(abundance)) |>
    group_by(species) |>
    summarise(abundance_avg = mean(abundance))

d.sampling <- left_join(d.sampling, d.abundance.avg, by = "species")

ggplot(d.sampling, aes(x = abundance_avg, y = measurement_rate)) +
geom_point() + 
labs(x = "Average abundance", y = "Full measurement rate")
```

Let's look if we have trends per zone.

```{r}
d.sampling <- d.join |>
    filter(!is.na(n_measured)) |>
    group_by(zone) |> 
    summarise(measurement_rate = mean(n_measured == abundance)) |>
    arrange(measurement_rate)
head(d.sampling)
```

There is also high variability between zone, with few zones where the measurement rate is low.
