---
title: "Exploring coastal data"
format: html
execute:
    message: false
    warning: false
---

# 0. Setup

Load dependencies.

```{r}
#| output: false
library(renv)
renv::load()
renv::restore() # Install missing librairies.
renv::status() # Check the project state.
```

```{r}
library(here)
library(dplyr)
library(ggplot2)
set_theme(theme_minimal())
options(
    ggplot2.continuous.colour = "viridis",
    ggplot2.continuous.fill = "viridis"
)
```

```{r}
data_folder <- here("data", "coast")
list.files(data_folder)
```

We see that we have multiple csv files.
We have abundance and biomass data for the Atlantic (Nurse, Solper), Atlantic and Manche (POMET).

# 1. Pomet campagne

## 1.1. Data cleaning

First let's load files.

```{r}
d.poisson <- read.csv(here(data_folder, "poisson_pomet.csv"), sep = ";")
head(d.poisson)

d.size <- read.csv(here(data_folder, "poisson_tailles_pomet.csv"), sep = ";")
head(d.size)
```

Let's do a bit of cleaning.
Remove irrelevant columns.

```{r}
d.poisson <- d.poisson |> select(-c(Commentaire, Ecologique, Trophique, Engin_peche, Zone))
names(d.poisson)
```

Rename columns for clarity.

```{r}
d.poisson <- d.poisson |> rename(
    zone = ID_Zone,
    year = Annee,
    month = Mois,
    hour = Heure,
    date = Date,
    trait = Trait,
    start_x = Coord_Deb_xmin,
    start_y = Coord_Deb_ymin,
    end_x = Coord_Fin_xmin,
    end_y = Coord_Fin_ymin,
    depth = Prof,
    species = NomScient,
    position = Position,
    abundance = Nind_esp,
    weight_batch = Pds_esp
)
head(d.poisson)
```

We see that the zone ID column also contains zone names which is not necessary.

```{r}
unique(d.poisson$zone)
```

We want to keep only the ID.

```{r}
d.poisson$zone <- substr(d.poisson$zone, 1, 6)
unique(d.poisson$zone)
```

Now let's look at which species are more abundant on average.

```{r}
n <- 10
d <- d.poisson |>
    group_by(species) |>
    summarise(abundance_avg = mean(abundance)) |>
    arrange(desc(abundance_avg)) |>
    slice(1:n)
d
```

Let's pull these species and plot their time series.

::: {.callout-note}
Is it expected that "Spratus" and "Spratus spratus" are two distinct species?
This may be an error when species name was written.
If so species names should be cleaned up.
:::

```{r}
common_sp <- d |> pull(species)

d <- d.poisson |>
    filter(species %in% common_sp) |>
    group_by(species, year) |>
    summarise(abundance_avg = mean(abundance), .groups = "drop")
d$log_abundance <- log(d$abundance_avg)

ggplot(d, aes(x = year, y = log_abundance, color = species)) +
    geom_point() +
    geom_line() + 
    labs(x = "Year", y = "Log abundance")
```


## 1.2. Plotting traits

First coordinates need to be formatted.
They are given as strings and should be converted to numeric values.

```{r}
library(stringr)
d.poisson <- d.poisson |>
    mutate(
        start_x = start_x |> str_replace(",", ".") |> as.numeric(),
        start_y = start_y |> str_replace(",", ".") |> as.numeric(),
        end_x = end_x |> str_replace(",", ".") |> as.numeric(),
        end_y = end_y |> str_replace(",", ".") |> as.numeric(),
    )

summary(d.poisson[, c("start_x", "start_y", "end_x", "end_y")])
```

```{r}
ggplot(d.poisson, aes(x = start_x, y = start_y)) +
    geom_segment(aes(xend = end_x, yend = end_y))+
    coord_equal() +
    labs(x = "Longitutde", y = "Latitude")
```

```{r}
library(leaflet)

traits_df <- d.poisson |>
    distinct(trait, .keep_all = TRUE)

leaflet() |>
    addTiles() |>
    addPolylines(
        data = d.poisson |> distinct(trait, .keep_all = TRUE),
        lng0 = ~start_x,
        lat0 = ~start_y,
        lng1 = ~end_x,
        lat1 = ~end_y,
        color = "steelblue",
        opacity = 0.7,
    ) |>
    setView(lng = -2, lat = 47, zoom = 8)

leaflet() |>
    addTiles() |>
    addSegments(
        data = traits_df,
        lng0 = ~start_x, lat0 = ~start_y,
        lng1 = ~end_x, lat1 = ~end_y,
        color = "steelblue",
        opacity = 0.7,
        popup = ~paste("Trait:", trait, "Zone:", zone, "Year:", year)
    ) |>
    setView(lng = -2, lat = 47, zoom = 8)
```

# Solper campagne

```{r}
d <- read.csv(here(data_folder, "captures_SOLPER_2005_2011.csv"), sep = ";")
head(d)
```


```{r}
d.traits <- d |>
    group_by(Annee) |>
    summarise(n_trait = n_distinct(Trait))

d.poids <- d |>
    filter(Espece == "SOLESOL") |>
    group_by(Annee) |>
    summarise(mean_weight = mean(Poids))
```


```{r}
ggplot(d.traits, aes(x = Annee, y = n_trait)) +
    geom_point() +
    labs(x = "Year", y = "Number of traits")


ggplot(d.poids, aes(x = Annee, y = mean_weight)) +
    geom_point() +
    labs(x = "Year", y = "Mean weight")
```

## TODO

- [ ] Seasonality effect
- [ ] Plot traits on a map
- [ ] Investigate whether size measures are complete or not (per trait and species).