---
title: "Define the statistical model"
format: html
execute:
    message: false
    warning: false
---

# Summary

This notebook contains

- the reasonning underlying the model definition that we intend to use in our study
- the list of causal relationships that we assume, illustrated with a DAG
- ecological interpretations relating these causal relationships to ecological processes.

# 1. Write the DAG

## 1.1. Causal dependence and check for bias

First, let's think about the causal relationship between our covariables and the metric we want to predict, that is, food web structure.
For now, we will forget about details of how to capture best dam impact with summary metrics.
Instead, we focus here on the overall causal relationships among our variables.
Directed Acyclic Graph are a common and graphical way to write the causal relationships assumed between the set of variables of interest.

```{r}
#| output: false
library(renv)
renv::load()
renv::restore() # Install missing librairies.
renv::status() # Check the project state.
```

```{r}
library(dagitty)
library(ggdag)
library(ggplot2)
theme_set(theme_dag())
```

To begin with, we consider the following general relationships, where environmental variables are aggregated within a single node ("Env."), these includes: water temperature, nutrient concentrations, salinity, etc.

```{r}
dag <- dagify(
    C ~ S + Comp. + Env.,
    S ~ Env.,
    Comp. ~ Env.,
    latent = "Comp.",
    exposure = "Env.",
    outcome = "C"
    ) 
ggdag(dag)
```

Environmental variables can have an impact on species richness (e.g. nutrients overload can simplify community due to turbidifcation), species composition (e.g. some species may not be suited for a given set of environmental conditions), and directly food web structure (here "C", which can be for example connectance)by affecting species behaviour, density or size.
Moreover, we also assume that richness and composition jointly shape food web structure.

```{r}
ggdag_paths(dag)
```

We see that we have 3 paths from environmental drivers and food web structure. Two of them are mediated by an intermediate variables, respectively composition and richness.
However, as we are not observing species comosition we cannot disantengly between these two paths: Env. -> C and Env. -> Comp. -> C.
That said, we can still interpret the direct effect Env. -> C in our model, as a total effect.

In brief, species composition is treated as a latent variable. As a result, estimated environmental effects on food-web structure should be interpreted as total effects or as effects conditional on species richness, but not as fully direct effects

## 1.2. Write a first model

Now that we have ensure that our model doesn't include statistical bias, we can write down the statistical model.
We will also try to think about the environmental variables that we want to include.
Let's do our list of environmental variables:

- temperature
- nutrient proxy (e.g. BOD)
- distance to the river mouth

For simplicity, we assume no interactions between our environmental variables (for now).

We also let for later the random effects.

### 1.2.1. Model food web connectance

If we are interested in connectance, we write the following model

$$
C_i \sim \text{Beta}(\mu_i, \phi_i),
$$

where we use a beta distribution as a link function to ensure that connectance values are bounded between 0 and 1.
Further, the mean of the distribution is given by

$$
\mu_i = \beta_0 + \beta_T T_i + \beta_N N_i + \beta_D D_i,
$$

where weakly informative prior have to be selected for regularization.

### 1.2.2. Model trophic length

Because trophic length is unbounded we can simply use a gaussian distribution and no link function.

$$
\text{TL}_i \sim \text{Normal}(\mu_i, \sigma_i),
$$

with the mean given by

$$
\mu_i = \beta_0 + \beta_T T_i + \beta_N N_i + \beta_D D_i.
$$

# 2. Capture dam impacts

Next, let's focus on how to best capture dam impact.
We begin with an illustration of how dams can impact a given community at a local scale in @fig-dam.

![Sketch of upstream and downstream dam impacts on a given community. The impact of a upstream dam reflects the decrease of migration (upstream-downstream), and other hydrographic effects of dams such as reduction of flow and sedimentation. The impact of downstream dam reflects the decrease of migration (downstream-upstream), and the transition from lotic to lentic environment.](sketch-dam-impact.png){#fig-dam}

A dam can impact a local community varied ways.
First, depending on whether it is upstream or downstream compared to the site of interest.
A dam downstream can create a reservoir, transitionning the system from lotic to lentic.
It can also prevent fish from migrating upstream.
A dam upstream can interrupt the river flow, change the sedimentation, and overall the hydrogeomorpholical conditions of the system.
It can also prevent migration of fish downstream (although we can expect this effect to be weaker than for the upstream migration).

Furthermore, dams can act at different scale.
There is an obvious local effect of dams, but dams can also have an effect at larger scale ('heritage'). For example, crossing a dam is harder if the fish had to already cross few dams before.
That said, we still can assume that the closest the dam is from a site, the stronger its impact is.

## 2.1. Design metrics

To summarise, we want metrics to capture an upstream and downstream impact of dams, as well as, the scale these impacts operate.

One way to do so, is to design two metrics - one upstream, one downstream - and incorporate the scale within these metrics (fixed for a given model).
Then, we can vary these scale, and assess the scale of dam impacts by comparing model performances (e.g. WAIC).

### 2.1.1. Impact of dam upstream

Here, we assume the impact of dams is proportional to their height ($h_i$). So we want to weight dams impact by their height ($w(h_i)$), as well as accounting for their distance to the site ($d_i$).
Here is the first metric we can think of

$$
\text{Dam}_\text{up} = \sum_i w(h_i) e^{-\frac{d_i}{L}},
$$

where $w(h_i)$ is the scaled dam height (e.g. dam height divided by mean dam height).
$L$ captures the distance at which dam impacts operate.
If $L$ is small only dams close to the site are assumed to have an impact.
On te contrary, if $L$ is large, even far dams can have an impact on the site.

Our idea is to try different $L$, such as, 10km, 50km, 100km. Then, we compare model performances, and from that we can infer the scale of dam impacts.

### 2.1.2. Impact of dam downstream

For downstream dam impact, we could use the very same measure, but instead of summing on dam upstream we could sum on downstream dams.
However, because we assume that one of the main effect of downstream dams is their reservoir effect we can also use their reservoir volume ($v_i$)


$$
\text{Dam}_\text{down} = \sum_i w(v_i) e^{-\frac{d_i}{L}}.
$$

# 3. Finalise the model

## 3.1. Interactions between covariables

First, let's discuss the potential interactions between our covariables that we want to include in our model.
Because interactions are hard to interpret and identify, we want to be as conservative as possible.
The probably most important interactions to include is between nutrient concentrations and $\text{Dam}_\text{down}$ and $\text{Dam}_\text{up}$.
That is because, a downstream dam can increase nutrient impact by preventing water from flowing.
Secondly, an upstream dam can mitigate the impact of nutrient overload by blocking the flux.

::: {.callout-note}
These interactions are to be discussed dependending on the scale at which nutrients concentration is measured.
If it is very local (measured exactly at the site), there are probably no interaction to consider.
However, if the measure is an undirect average around the area of the site, interactions may be relevant.
:::

Lastly, we consider a potential interaction between nutrients and temperature. For example, with increased temperature the metabolic demand of fish increases which could magnify nutrient impacts.

## 3.2. Add random effects

### 3.2.1. Temporal

We expect food web to be correlated in my time. 
Because of legacy effects, species present one year have a higher chance of being present the next year.
This kind of temporal autocorrelation can be captured by AR model.

::: {.callout-note}
The temporal effect sould in any case be compared with and independent random effect.
:::

### 3.2.2. Spatial

For the spatial random effect, we want to account for the fact that site closer to each other looks more alike.
This can be use using the BYM model in INLA, in which we specify which sites are dependent under the form of an adjacency matrix.
An alternative, that can also be model on top, is to use euclidian distance with the "SPDE" method.

### 3.2.3. Measurement

Lastly, we have to add random effects to correct for potential differences between campagne.
Notably, not all compagne use the fishing techniques.

## 3.3. Write the final model

Using everything we have built until there, we can build a first draft of our model.

### 3.3.1. Food web connectance

For connectance, the model writes

$$
C_i \sim \text{Beta}(\mu_i, \phi_i)
$$

where the mean is given by

$$
\begin{align}
\text{logit} (\mu_i) = &\beta_0 + \beta_1 T_i + \beta_2 N_i + \beta_3 D_i + \beta_4 \text{Dam}_\text{up} + \beta_5 \text{Dam}_\text{down} + &\text{[fixed effects]} \\
&\beta_6 (\text{Dam}_\text{up} \times N_i) + \beta_7 (\text{Dam}_\text{down} \times N_i) + \beta_8 (T_i \times N_i) + &\text{[interactions]} \\ 
&\text{AR}(1) + u_i + v_i + \text{Campagne}_i. &\text{[random effects]}
\end{align}
$$

where  $u_i$ is the spatial structured effect, and $v_i$ the unstructured one.

::: {.callout-note}
We begin by assuming that $\phi_i$ is constant and do not dependend on other variables, for sake of simplicity.
:::

### 3.3.2. Trophic length

For trophic length, the model has the same structure.
We simply replace the beta distribution with a gaussian distribution and remove the link function.

$$
\text{TL}_i \sim \text{Normal}(\mu_i, \sigma_i)
$$

where the mean is given by

$$
\begin{align}
\mu_i = &\beta_0 + \beta_1 T_i + \beta_2 N_i + \beta_3 D_i + \beta_4 \text{Dam}_\text{up} + \beta_5 \text{Dam}_\text{down} + &\text{[fixed effects]} \\
&\beta_6 (\text{Dam}_\text{up} \times N_i) + \beta_7 (\text{Dam}_\text{down} \times N_i) + \beta_8 (T_i \times N_i) + &\text{[interactions]} \\ 
&\text{AR}(1) + u_i + v_i + \text{Campagne}_i. &\text{[random effects]}
\end{align}
$$

# 4. Identify ecological mecanisms at different scale

Ecological mecanisms operate at different spatial and temporal scales.
The same environmental driver may affect food-web structure through distinct mecanisms dependending on the scale consider.

## 4.1. Temperature

### 4.1.1 Local scale

At local scales, temperature mainly affect individual physiology through metabolic rates, growth and consumption.
These effects can modify interaction strengths, for example, by increasing predator attack rate and as a consequence reshape the food web.

### 4.1.2. Regional scale

At broader scale, temperature shape species distribution and as a result community composition.
These compositional change in turn affect food web structure.

### 4.1.3. Temporal scale

Interannual variability in temperature can induce population turnover which result in change in food web structure.

## 4.2. Nutrients

### 4.2.1. Local scale
 
Nutrient enrichment can affect primary productivity which propagate throughout the food web. Overall, nutrient concentration affects how energy flows within the community.

### 4.2.2. Intermediate scale

Sustained nutrient inputs can promote community simplification, resulting in decrease of trophic length, and potentially increase in connectance through the loss of specialist species.

### 4.2.3. Long-term scale

Ultimately, sustained high nutrient inputs can induce regime shifts.

## 4.3. Dams

### 4.3.1. Local downstream scale

Immediately downstream of dams, the habitat is altered because of flow regulation, sedimentation retention, and temperature modification.
Reservoirs act as lentic environment modifying the lotic environment, species composition and in turn food web structure.

### 4.3.2. Upstream network scale

At larger spatial scale, dams reduce migration and recolonisation process and alter species composition upstream.

### 4.3.3. Bassin scale

At the scale of the entire river network, the accumulation of dams can modify energy fluxes between the land and the sea, through the loss of connectivity which hinders flows of inorganic and organic matter.
These effects can induce a reorganisation of food web structure from freshwater system to coastal systems.

### 4.3.4. Directionality 

From the perspective of a focal site, dams located downstream can generate backwater effects and act as strong barriers to upstream recolonisation, whereas upstream dams primarily regulate hydrological, sediment, and nutrient fluxes reaching the site.
Because downstream movement across dams is generally less constrained than upstream movement, downstream dams are expected to have a stronger impact on connectivity and recovery processes

For that reason, upstream and downstream dam could act at different scale.
If recolonisation is important, downstream dams may act at intermediate or large scale.
On the other hand, upstream dam may act at more local scale, because the interruption of migration is weaker, but their effect might be more dominated by change in hydrographic condition.

## 4.4. Interaction between nutrients and temperature

Nutrient and warming have been reported to interact in varied ways.
Warming increase metabolic demand of consumers.
This increase can be compensated by an enrichment of the environment, reflecting an antagonistic interaction between nutrients and warming.
However, they are evidence that warming can exacerbate eutrophication symptoms. It is therefore important to assess this interaction.

## 4.5. Interaction between nutrients and dam

Dams can strongly mediate the effects of nutrient inputs by altering nutrient retention, transformation, and transport. 
Reservoirs may trap nutrients and promote primary production locally, while reducing nutrient delivery downstream. 
At larger scales, the interaction between dam-induced fragmentation and nutrient enrichment may lead to spatially heterogeneous food-web responses, with enriched but simplified webs in impounded areas and nutrient-limited systems downstream.




