---
title: "Identify and get missing diet information"
format:
  html:
    embed-resources: true
execute:
    message: false
    warning: false
---

# Summary

In this notebook we

- identify the species for which we have no prior information on their diet
- check for potential taxonomic issues that we will have to deal with
- use fishbase to get diet information

# 0. Setup

We load the environment.

```{r}
#| output: false
library(renv)
renv::load()
renv::restore() # Install missing librairies.
renv::status() # Check the project state.

library(here)
library(stringr)
library(stringdist)
library(dplyr)
library(riversea)
library(rfishbase)
library(ggplot2)
```

Previous information on fish diet can be find on the dataframe hosted at the following address.

```{r}
url <- "https://raw.githubusercontent.com/CamilleLeclerc/FoodWebsLakeStream/refs/tags/v3.0/data/raw-data/fish_diet_shift.csv"
```

We download it and convert it to a dataframe.

```{r}
data_folder <- here("data", "river")
file_name <- "fish_diet.csv"
download.file(url, destfile = here(data_folder, file_name))
```

We can check that the file has been downloaded.

```{r}
list.files(data_folder, pattern = file_name)
```

We then read the file as a dataframe.

```{r}
d.diet <- read.csv(here(data_folder, file_name))
sp_list.diet <- sort(unique(d.diet$species_name))
sp_list.diet <- str_replace(sp_list.diet, "_", " ")
sp_list.diet
```

# 1. Pomet

To identify missing species in this dataframe, we have to load dataframes from sea campagne (solper, pomet, nurse).

```{r}
data_folder <- here("data", "sea", "raw")
pomet <- read.csv(here(data_folder, "poisson_pomet.csv"), sep = ";")
solper <- read.csv(here(data_folder, "captures_SOLPER_2005_2011.csv"), sep = ";")
nurse <- read.csv(here(data_folder, "Captures_Nurse_1980_2023.csv"), sep = ";")

sp_pomet <- unique(pomet$NomScient)
head(sp_pomet)

sp_solper <- unique(solper$Espece)
head(sp_solper)

sp_nurse <- unique(nurse$Espece)
head(sp_nurse)
```

```{r}
reftax_file <- "POMET_RefTax_Sp.csv"
reftax <- read.csv(here(data_folder, reftax_file), sep = ";")
```

## 1.1. Handle taxonomic resolutions: keep only species

```{r}
is_species <- function(name) grepl(" ", name)
species_only <- sp_pomet[is_species(sp_pomet)]
```

## 1.2. Handle different genres: keep only fish

```{r}
remove_non_fish <- function(species_list, reftax) {
    sp_df <- data.frame(nom = species_list)
    sp_df <- left_join(sp_df, reftax, by = "nom")
    sp_df <- sp_df |> filter(phylum == "Chordata" | is.na(phylum))
    sp_df$nom
}

fish_only <- remove_non_fish(species_only, reftax)
species_missing <- setdiff(fish_only, sp_list.diet)
species_missing
```

## 1.3. Handle taxonomic typos

Next, we want to check for mismatch because of typos or alternative taxonomic names.
To do so, we compute a string distance to find species with most similar names between our two lists.
By doing so, we will be able to quickly identify erroneous mismatch.

```{r}

similar_name <- lapply(species_missing, function(name) {
    distances <- stringdist(name, sp_list.diet, method = "jw")
    closest <- sp_list.diet[which.min(distances)]
    data.frame(name = name, closest = closest, distance = min(distances))
})

df_strdist <- do.call(rbind, similar_name) |>
    arrange(distance)
df_strdist
```

We see that there are few issues 

- Typo for the Scardinius erythro. in the diet species list.
- Outdated taxonomic name for Gymnocephalus cernua.
- Subspecies resolution for Gasterosteus aculeatus.
- Two species in one for Abramis brama + Blicca bjoerkna.
- Outdated taxonomic name for Trutta fario.

```{r}
species_to_remove <- c(
    "Abramis brama + Blicca bjoerkna"
)
species_missing <- species_missing[!(species_missing %in% species_to_remove)]
```

Actually, we can use fishbase for taxonomic issues.
Instead of working with raw names, we can ensure that we have validated names, using fishbase functions.

```{r}

name_diet <- data.frame(original = sp_list.diet)
name_diet$validated <- validate_names(sp_list.diet)

name_diet |> filter(original != validated)
```

```{r}
name_missing <- data.frame(original = species_missing)
name_missing$validated <- validate_names(name_missing$original)

name_missing |> filter(original != validated)
```

```{r}
missing_pomet <- setdiff(name_missing$validated, name_diet$validated)
length(missing_pomet)
```


## 1.4. Rare species

Next, we want to compute average abundance for these species to identify potential rare species that wouldn't be interesting to include.

```{r}
abundance_pomet <- pomet |>
    filter(NomScient %in% missing_pomet) |>
    group_by(NomScient) |>
    summarise(n = sum(Nind_esp)) |>
    arrange(n)
head(abundance_pomet)

abundance_rank <- abundance_pomet |>
    group_by(n) |>
    summarise(count = n())
head(abundance_rank)
```

```{r}
n_min <- 20 # Minimal abundance to be included.
rare_pomet <- abundance_pomet |>
    filter(n <= n_min) |>
    pull(NomScient)
length(rare_pomet)

missing_pomet <- name_missing |>
    filter(validated %in% missing_pomet) |>
    filter(!original %in% rare_pomet) |>
    pull(validated)
```

```{r}
set_theme(theme_minimal())
ggplot(abundance_rank, aes(x = log(1 + n), y = count)) +
    geom_line() +
    geom_point() +
    labs(x = "Log(1 + total abundance)", y = "Number of species")
```

# 2. Nurse

```{r}
length(sp_nurse)
```

We see that we have many species.
This probably means that we have many non-fish species that we will have to remove.
First, we remove genus to only keep species names.

```{r}
species_only <- sp_nurse[is_species(sp_nurse)]
length(species_only)
```

Next, we want to remove non-fish species.
We load the corresponding reftax table.

```{r}
list.files(data_folder)
reftax_file <- list.files(data_folder, pattern = "RData")[1]
load(here(data_folder, reftax_file))
reftax <- ESPECE_MAJ
```

```{r}
sp_df <- data.frame(scientificname = species_only)
sp_df <- sp_df |>
    left_join(reftax, by = "scientificname") |>
    filter(phylum == "Chordata" | is.na(phylum)) |>
    filter(class == "Teleostei" | class == "Elasmobranchii" | is.na(class))
fish_only <- sp_df$scientificname

name_nurse <- data.frame(original = fish_only)
name_nurse$validated <- validate_names(name_nurse$original)
name_nurse |> filter(original != validated)

missing_nurse <- setdiff(name_nurse$validated, name_diet$validated)
length(missing_nurse)
```

```{r}
abundance_nurse <- nurse |>
    filter(Espece %in% missing_nurse) |>
    group_by(Espece) |>
    summarise(n = sum(Nombre, na.rm = TRUE)) |>
    arrange(n)
head(abundance_nurse)

rare_nurse <- abundance_nurse |>
    filter(n <= n_min) |>
    pull(Espece)
length(rare_nurse)

missing_nurse <- name_nurse |>
    filter(original %in% missing_nurse) |>
    filter(!validated %in% rare_nurse) |>
    pull(validated)
length(missing_nurse)
```

```{r}
missing_all <- sort(unique(c(missing_nurse, missing_pomet)))
missing_all
```

# 3. Solper

```{r}
length(sp_solper)
```

For Solper, we have code instead of species names.

```{r}
head(sp_solper)
```

::: {.callout-warning}
I cannot find the relation between species code and their scientific name.
:::

# 4. Get diet data from fishbase

## 4.1. Missing species from sea campagnes

Now that we have a species list for which we want to get their diet we can scrap fishbase.

Because this step is crucial and is expected to be repeated we implement it directly in our package `riversea`.

So we can directly obtain diet for all species with a single function call.

```{r}
diet_category <- get_diet_category(missing_all)
head(diet_category)
```

Next, to obtain something similar as the table for riverine fish, we can also extract a wide table.

```{r}
diet_category_wide <- get_diet_category_wide(missing_all)
head(diet_category_wide)
```

## 4.2. Fishbase diet for riverine fish

```{r}
diet_category <- get_diet_category(name_diet$validated)
head(diet_category)
```

```{r}
diet_category_wide <- widen_diet_category(diet_category)
head(diet_category_wide)
```

```{r}
write.csv(diet_category_wide, here("data", "fishbase_diet_river.csv"))
```

Let's focus on a single species *Abramis brama*

```{r}
diet_category_wide |> filter(species == "Abramis brama")
```

and compare with the original table that we have downloaded.

```{r}
d.diet |> filter(species_name == "Abramis_brama")
```

## 4.3. All species

```{r}
name_diet$validated

validate_names("Abramis brama")

name_all <- data.frame(original = name_diet$original, validated = name_diet$validated, campagne = "river")

n_nurse <- unique(nurse$Espece)
name_nurse <- data.frame(original = n_nurse, validate_names(n_nurse), campagne = "nurse")

n_pomet <- unique(pomet$NomScient)
name_pomet <- data.frame(original = n_pomet, validate_names(n_pomet), campagne = "pomet")

unique(nurse$Espece)
```

# 5. AquaticWeb

Let's load the combined dataset.

```{r}
library(tidyr)
tidy_data <- load_raw_sea_data(here("data", "sea", "raw")) |>
    clean_sea_data() |>
    combine_sea_data()

species_names <- tidy_data$catch |>
    pull(species) |>
    unique()
species <- data.frame(name = species_names)

species$name_validated <- validate_names(species$name)

head(species)
```

```{r}
aquatic_web <- read.csv(here("data", "diet", "aquatic-web-extraction-eric.csv"))
head(aquatic_web)

names_aquatic_web <- aquatic_web |>
    pull(predateur) |>
    unique() |>
    validate_names() |>
    na.omit()

species <- species |>
    mutate(in_aquatic_web = name_validated %in% names_aquatic_web)

species |> summarise(
    unique_entry = n(),
    valid_entry = sum(!is.na(name_validated)),
    aquatic_web_entry = sum(in_aquatic_web)
)
```

## TODO

- Do the diet table for fish already documented and compare diet.
