---
title: "Get diet table from fishbase"
format:
  html:
    embed-resources: true
execute:
    message: false
    warning: false
---

# Summary

In this notebook we

- extract a diet table using fishbase

# 0. Setup

```{r}
#| output: false
library(renv)
renv::load()
renv::restore()
renv::status()
```

```{r}
#| output: false
library(here)
library(DiagrammeR)
library(targets)
library(dplyr)
library(rfishbase)
library(riversea)
```

# 1. Extract the diet table

First, we want to get the list of species present in our dataset.

```{r}
data <- tar_read(sea_data_imputed)
species_list <- data$size |>
  pull(species_valid) |>
  unique()
head(species_list)
```

We have `r length(species_list)` species in our data set.
We want to retrive their diet for each of them.

We assume the following food web structure.

```{r}
DiagrammeR::grViz(readLines(here("figures", "foodweb.dot")))
```

Ellipses represent broad basal categories on which fishes feed on.
Our goal here is to determine for each fish species on which of these categories
it feeds, and whether it is piscivorous or not.
To do so we scrap fishbase.

```{r}
items <- rfishbase::fb_tbl("diet_items")
diet_species <- rfishbase::diet(species_list)
diet_category <- diet_species |>
  left_join(items, by = "DietCode") |>
  select(
    Species,
    SampleStage,
    DietCode,
    FoodI,
    FoodII,
    FoodIII,
    ItemName,
    Stage,
    DietPercent
  )
head(diet_category)
```

We see that the food is described by the three columns FoodI, FoodII and 
FoodIII.
The question is how we map our categories using those.

```{r}
diet_category |>
  pull(FoodI) |>
  unique()
```

Here we see that only the "plants" and "others" category cannot be classified
directly.
Nekton correspond to other fishes so it implies piscivory.

We can go further in the classification for the ambiguous categories.

```{r}
diet_category |>
  filter(FoodI == "plants") |>
  select(FoodII, FoodIII) |>
  distinct()
```

So using these subcategories we can categorize fish diet using our categories

- benthic algae > macrophytes;
- periphyton > biofilm;
- terrestrial plants > macrophytes;
- phytoplankton > phytoplantkon.

```{r}
diet_category |>
  filter(FoodI == "others") |>
  select(FoodII, FoodIII) |>
  distinct()
```

These correspond to categories we do not consider in our framework, se we choose
to ignore them.

Like this we can extract a diet table

```{r}
diet_table <- tar_read(diet)
head(diet_table)
```

which for each species, and life stage, tells what it eats.
We can also have the table a in "wide" format, where we have one column for each
basal category.

```{r}
diet_table_wide <- tar_read(diet_wide)
head(diet_table_wide)
```

# 2. Postprocess diet table

## 2.1. Add larvae stage

Next, we want to add a larvae stage for all species.

```{r}
d_larvae <- diet_table_wide |>
  group_by(species) |>
  summarise(has_larvae = "larvae" %in% stage, .groups = "drop")
d_larvae |> count(has_larvae)
```

So we see that for most species, they do not have a larvae stage.
We can add them with a utility function.
We assume that larvae only feed on zooplankton.

```{r}
diet_with_larvae <- add_larvae(diet_table_wide)
diet_with_larvae |> filter(species == sample(species, 1))
```


## 2.2. Remove rare species

To focus on important species for the community we study we filter for very rare
species.
Let's have a look.

```{r}
sea_data <- tar_read(sea_data_imputed)
occurences <- sea_data$catch |>
  group_by(species_valid) |>
  summarise(occurence = n(), .groups = "drop") |>
  arrange(occurence)
```

```{r}
occ_min <- 10
count_occ <- occurences |> count(occurence >= occ_min)
n_remaining <- pull(count_occ, n)[2]
count_occ
```

We see that we have many species have a low occurence.
For example, if you choose a threshold of `r occ_min`, we have only
`r n_remaining` species that remain in the data set.

We created an utility function to remove rare species from the diet table.

```{r}
diet_no_rare <- tar_read(diet_no_rare)
head(diet_no_rare)
```

We can check if we have species with missing diet.

```{r}
diet_no_rare |>
  filter(is.na(stage)) |>
  pull(species)
```

# 3. Life stage lengths

```{r}
library(rfishbase)

species_list <- diet_no_rare |>
  pull(species) |>
  unique()
d_maturity <- maturity(species_list) |>
  select(Species, Lm) |>
  filter(!is.na(Lm)) |>
  group_by(Species) |>
  summarise(
    Lm_avg = mean(Lm),
    .groups = "drop"
  )
head(d_maturity)
```

```{r}
species_missing <- setdiff(species_list, d_maturity$Species)
maturity("Barbus barbus")
```

```{r}
size_extrema <- tar_read(size_extrema) |> rename(species = species_valid)
maturity_length <- tar_read(maturity_length)
sizes <- inner_join(maturity_length, size_extrema, by = "species")
head(sizes)

sizes |> filter(maturity_length > length_max | maturity_length < length_min)
```

Maybe only juveniles are captured?

```{r}
diet_full <- tar_read(diet)
head(diet_full)
```




