---
title: "Identify missing diet information"
format: html
---

# Summary

In this notebook we want to

- identify the species for which we have no prior information on their diet
- check for potential taxonomic issues that we will have to deal with

# 0. Setup

We load the environment.

```{r}
#| output: false
library(renv)
renv::load()
renv::restore() # Install missing librairies.
renv::status() # Check the project state.
```

Previous information on fish diet can be find on the dataframe hosted at the following address.

```{r}
url <- "https://raw.githubusercontent.com/CamilleLeclerc/FoodWebsLakeStream/refs/tags/v3.0/data/raw-data/fish_diet_shift.csv"
```

We download it and convert it to a dataframe.

```{r}
library(here)
data_folder <- here("data", "coast")
file_name <- "fish_diet.csv"
download.file(url, destfile = here(data_folder, file_name))
```

We can check that the file has been downloaded.

```{r}
list.files(data_folder, pattern = file_name)
```

We then read the file as a dataframe.

```{r}
d.diet <- read.csv(here(data_folder, file_name))
sp_list.diet <- unique(d.diet$species_name)
sp_list.diet
```

# 1. Identify species with no diet information

To identify missing species in this dataframe, we have to load dataframes from sea campagne (solper, pomet, nurse).

```{r}
library(stringr)

d.pomet <- read.csv(here(data_folder, "poisson_pomet.csv"), sep = ";")
sp_list.pomet <- unique(d.pomet$NomScient)
sp_list.pomet <- str_replace(sp_list.pomet, " ", "_")
```

```{r}
reftax_file <- list.files(data_folder, pattern = "Reftax")[1]
load(here(data_folder, reftax_file))
head(ESPECE_MAJ)
```


```{r}
library(dplyr)
d.reftax <- ESPECE_MAJ
d.reftax |> filter(grepl("Anguilla", scientificname))
```

To begin with, we remove species that do not have an underscore as it seems to often indicate the name is not resolved at scale of the species.

```{r}
sp_list.pomet <- sp_list.pomet[grepl("_", sp_list.pomet)]
sp_list.pomet <- sp_list.pomet[-3]
```

```{r}
sp_list.pomet[!(sp_list.pomet %in% sp_list.diet)]
```

## TODO

- Handle name resolution (family, genus)
- Remove non-fish species
- Merge species from three campagnes