---
title: "Identify missing diet information"
format: html
execute:
    message: false
    warning: false
---

# Summary

In this notebook we want to

- identify the species for which we have no prior information on their diet
- check for potential taxonomic issues that we will have to deal with

# 0. Setup

We load the environment.

```{r}
#| output: false
library(renv)
renv::load()
renv::restore() # Install missing librairies.
renv::status() # Check the project state.
```

Previous information on fish diet can be find on the dataframe hosted at the following address.

```{r}
url <- "https://raw.githubusercontent.com/CamilleLeclerc/FoodWebsLakeStream/refs/tags/v3.0/data/raw-data/fish_diet_shift.csv"
```

We download it and convert it to a dataframe.

```{r}
library(here)
data_folder <- here("data", "coast")
file_name <- "fish_diet.csv"
download.file(url, destfile = here(data_folder, file_name))
```

We can check that the file has been downloaded.

```{r}
list.files(data_folder, pattern = file_name)
```

We then read the file as a dataframe.

```{r}
library(stringr)

d.diet <- read.csv(here(data_folder, file_name))
sp_list.diet <- sort(unique(d.diet$species_name))
sp_list.diet <- str_replace(sp_list.diet, "_", " ")
sp_list.diet
```

# 1. Pomet

To identify missing species in this dataframe, we have to load dataframes from sea campagne (solper, pomet, nurse).

```{r}
pomet <- read.csv(here(data_folder, "poisson_pomet.csv"), sep = ";")
solper <- read.csv(here(data_folder, "captures_SOLPER_2005_2011.csv"), sep = ";")
nurse <- read.csv(here(data_folder, "Captures_Nurse_1980_2023.csv"), sep = ";")

sp_pomet <- unique(pomet$NomScient)
head(sp_pomet)

sp_solper <- unique(solper$Espece)
head(sp_solper)

sp_nurse <- unique(nurse$Espece)
head(sp_nurse)
```

```{r}
library(dplyr)

reftax_file <- "POMET_RefTax_Sp.csv"
reftax <- read.csv(here(data_folder, reftax_file), sep = ";")
```

## 1.1. Handle taxonomic resolutions: keep only species

```{r}
is_species <- function(name) grepl(" ", name)
species_only <- sp_pomet[is_species(sp_pomet)]
```

## 1.2. Handle different genres: keep only fish

```{r}
remove_non_fish <- function(species_list, reftax) {
    sp_df <- data.frame(nom = species_list)
    sp_df <- left_join(sp_df, reftax, by = "nom")
    sp_df <- sp_df |> filter(phylum == "Chordata" | is.na(phylum))
    sp_df$nom
}

fish_only <- remove_non_fish(species_only, reftax)
species_missing <- setdiff(fish_only, sp_list.diet)
species_missing
```

## 1.3. Handle taxonomic typos

Next, we want to check for mismatch because of typos or alternative taxonomic names.
To do so, we compute a string distance to find species with most similar names between our two lists.
By doing so, we will be able to quickly identify erroneous mismatch.

```{r}
library(stringdist)

similar_name <- lapply(species_missing, function(name) {
    distances <- stringdist(name, sp_list.diet, method = "jw")
    closest <- sp_list.diet[which.min(distances)]
    data.frame(name = name, closest = closest, distance = min(distances))
    })

df_strdist <- do.call(rbind, similar_name) |>
    arrange(distance)
df_strdist
```

We see that there are few issues 

- Typo for the Scardinius erythro. in the diet species list.
- Outdated taxonomic name for Gymnocephalus cernua.
- Subspecies resolution for Gasterosteus aculeatus.
- Two species in one for Abramis brama + Blicca bjoerkna.
- Outdated taxonomic name for Trutta fario.

```{r}
species_to_remove <- c(
    "Abramis brama + Blicca bjoerkna",
    "Gasterosteus aculeatus aculeatus",
    "Salmo trutta fario",
    "Salmo trutta trutta",
    "Gymnocephalus cernuus"
)
missing_pomet <- species_missing[!(species_missing %in% species_to_remove)]
```

# 2. Nurse

```{r}
length(sp_nurse)
```

We see that we have many species.
This probably means that we have many non-fish species that we will have to remove.
First, we remove genus to only keep species names.

```{r}
species_only <- sp_nurse[is_species(sp_nurse)]
length(species_only)
```

Next, we want to remove non-fish species.
We load the corresponding reftax table.

```{r}
list.files(data_folder)
reftax_file <- list.files(data_folder, pattern = "RData")[1]
load(here(data_folder, reftax_file))
reftax <- ESPECE_MAJ
```

```{r}
sp_df <- data.frame(scientificname = species_only)
sp_df <- sp_df |> 
    left_join(reftax, by = "scientificname") |>
    filter(phylum == "Chordata" | is.na(phylum)) |>
    filter(class == "Teleostei" | class == "Elasmobranchii" | is.na(class))
fish_only <- sp_df$scientificname

missing_nurse <- setdiff(fish_only, sp_list.diet)
```

```{r}
similar_name <- lapply(missing_nurse, function(name) {
    distances <- stringdist(name, sp_list.diet, method = "jw")
    closest <- sp_list.diet[which.min(distances)]
    data.frame(name = name, closest = closest, distance = min(distances))
    })

df_strdist <- do.call(rbind, similar_name) |>
    arrange(distance)
df_strdist
```

```{r}
missing_all <- sort(unique(c(missing_nurse, missing_pomet)))
missing_all
```

# 3. Solper

```{r}
length(sp_solper)
```

For Solper, we have code instead of species names.

```{r}
head(sp_solper)
```

::: {.callout-warning}
I cannot find the relation between species code and their scientific name.
:::


## TODO

- Find reference table for species codes in Solper campagne.
- Merge species from three campagnes