---
title: "Build food webs"
format: 
  html:
    embed-resources: true
execute:
    message: false
    warning: false
---

```{r}
#| output: false
library(renv)
renv::load()
renv::restore() # Install missing librairies.
renv::status() # Check the project state.
library(targets)
library(here)
tar_config_set(
  store = here::here("_targets")
)
library(foodwebbuilder)
library(dplyr)
```

## Build the metaweb

To build the metaweb, we need the following.

The diet table for fish species considered.

```{r}
diet_fish <- tar_read(diet_size)
head(diet_fish)
```

The diet for the resources considered.

```{r}
diet_resource <- tar_read(diet_resource)
```

The size of each individual fish.

```{r}
size <- tar_read(sea_data_imputed)$size
head(size)
```

The predation window.

```{r}
predation_window <- tar_read(predation_window)
head(predation_window)
```

From this we can build the metaweb using helper functions of the 
`foodwebbuilder` package.
We wrapped them in a `riversea` function so the metaweb can be build directly.
We just need to specify the number of size class, and the selected resources,
although default values are set if not specified.

We have already computed metaweb for different number of size classes,
so we can check whether this parameter influences the metaweb structure.

```{r}
metaweb_table <- tar_read(metaweb_table)
metaweb_table
```

We can plot the metaweb connectance against the number of size classes.

```{r}
tar_read(plot_metaweb_connectance)
```

We observe little variation indicating that the number of size class does not
change too much the structure of the food web reconstructed.

## Get local food webs

```{r}
web_list <- tar_read(web_list)

head(web_list$metaweb)

length(web_list$local)
```




