---
title: "Build food webs"
format:
  html:
    embed-resources: true
execute:
    message: false
    warning: false
---

```{r}
#| output: false
#| eval: false
library(renv)
renv::load()
renv::restore() # Install missing librairies.
renv::status() # Check the project state.
```
```{r}
#| output: false
library(targets)
library(here)
tar_config_set(
  store = here::here("_targets")
)
library(foodwebbuilder)
library(dplyr)
library(tibble)
devtools::load_all()
library(ggplot2)
set_theme(theme_minimal())
library(INLA)
```



## Build the metaweb

To build the metaweb, we need the following.

The diet table for fish species considered.

```{r}
diet_fish <- tar_read(diet_and_length)
head(diet_fish)
```

The diet for the resources considered.

```{r}
diet_resource <- tar_read(diet_resource)
```

The size of each individual fish.

```{r}
size <- tar_read(sea_data_imputed)$size
head(size)
```

The predation window.

```{r}
predation_window <- tar_read(predation_window)
head(predation_window)
```

From this we can build the metaweb using helper functions of the 
`foodwebbuilder` package.
We wrapped them in a `riversea` function so the metaweb can be build directly.
We just need to specify the number of size class, and the selected resources,
although default values are set if not specified.

For information, here is the structure that we consider between the different
trophic groups on which fish can feed on.

```{r}
tar_read(plot_net_groups)
```

We have already computed metaweb for different number of size classes,
so we can check whether this parameter influences the metaweb structure.

```{r}
metaweb_table <- tar_read(metaweb_table)
metaweb_table
```

We can plot the metaweb connectance against the number of size classes.

```{r}
tar_read(plot_metaweb_connectance)
```

We observe little variation indicating that the number of size class does not
change too much the structure of the food web reconstructed.

## Get local food webs

```{r}
web_list <- tar_read(web_list)

get_connectance(web_list$metaweb)

local_foodwebs <- tar_read(local_foodwebs)
head(local_foodwebs)
```

```{r}
trait_data <- tar_read(sea_data_tidy)$trait
local_foodwebs <- local_foodwebs |>
  left_join(trait_data)
head(local_foodwebs)
```


We have computed the connectance and richness of local metaweb so we can plot
the relationship between these two variables.

```{r}
m1 <- inla(
  connectance ~ 1 + log_trophic_richness + f(survey, model = "iid") + f(year, model = "iid"),
  data = local_foodwebs |> select(-foodweb)
)
summary(m1)

intercept <- m1$summary.fixed["(Intercept)", "mean"]
slope <- m1$summary.fixed["log_trophic_richness", "mean"]

ggplot(local_foodwebs, aes(x = log_trophic_richness, y = connectance, color = survey)) +
  geom_point(alpha  = 0.3) +
  geom_abline(intercept = intercept, slope = slope) +
  labs(x = "Log trophic richness", y = "Connectance", color = "Survey")
```

As expected, we observe a negative trend between connectance and species
richness.
There is also a small random effect attributed to the survey.

Next, we investigate the effect of the year on species richness.

```{r}
ggplot(local_foodwebs, aes(x = year, y = log_trophic_richness, color = survey)) +
  geom_point(alpha  = 0.1) +
  stat_summary(
    fun = mean,
    geom = "line",
    linewidth = 1
  ) +
  labs(x = "Year", y = "Log richness", color = "Survey")
```

As we observe a positive trend between richness and year, and that we have
observed a negative trend between richness and connectance, we expect a negative
trend between connectance and year.

```{r}
ggplot(local_foodwebs, aes(x = year, y = connectance, color = survey)) +
  geom_point(alpha  = 0.1) +
  stat_summary(
    fun = mean,
    geom = "line",
    linewidth = 1
  ) +
  labs(x = "Year", y = "Connectance", color = "Survey")
```

Interestingly, we see that connectance values are in agreement between the two
survey (when there is an overlap), suggesting that it makes sense to merge these
two different data sources.

