---
title: "Validate the diet table"
format:
  html:
    embed-resources: true
execute:
    message: false
    warning: false
---

```{r}
#| output: false
#| code-fold: true

library(renv)
renv::load()
renv::status() # Check the project state.
library(targets)
library(here)
tar_config_set(
  store = here::here("_targets")
)
library(riversea)
```

We now source the diet table for estuarine and coastal fish from a
literature-validated table (`data/diet/marine_diet_validated.csv`) instead of
the fishbase scrape.

We first go over sanity checks to ensure that the diet table is consistent,
and the resulting food webs as well.

## Sanity checks

### Species coverage

We compare the species observed often enough in the sea surveys to the
species covered by the diet table. A species is considered non-rare if it
was caught strictly more than `params$occurence_min` times. The table below
lists commonly observed species missing from the diet table; empty means no
coverage gap.

```{r}
params <- tar_read(params)
diet <- tar_read(diet)
catch <- tar_read(sea_data_tidy)$catch

check_diet_species_coverage(diet, catch, params$occurence_min)
```

### Size coverage per species

For each species, the diet table should define a diet over contiguous
stages covering the whole size range, from 0 to unbounded, otherwise
`foodwebbuilder::build_metaweb()` errors out on a missing diet interval. We
look for three failure modes: gaps between consecutive stages, a first
stage that does not start at 0, and a last stage that does not reach an
unbounded size.

```{r}
check_diet_size_coverage(diet)
```

### Metaweb consistency

We also check the metaweb built from this diet table: every non-resource
(fish) trophic species should have at least one prey item, and no trophic
species should be fully disconnected (neither predator nor prey of
anything).

```{r}
web_list <- tar_read(web_list)
resource_list <- tar_read(resource_list)

check_metaweb_consistency(web_list$metaweb, resource_list)
```

If the three tables above are all empty, the diet table and the metaweb
built from it pass every automated check.
