---
title: "Explore coastal data"
format:
  html:
    embed-resources: true
execute:
    message: false
    warning: false
---

# Summary

In this notebook we

- explore the coastal data of fish catches from different campagnes (Nurse, Pomet and Solper)
- extract general information about each campagne (sampling effort, trait location)
- compute the how often each fish size is measured

# 0. Setup

Load dependencies.

```{r}
#| output: false
library(renv)
renv::load()
renv::restore() # Install missing librairies.
renv::status() # Check the project state.
```

```{r}
library(here)
library(dplyr)
library(ggplot2)
set_theme(theme_minimal())
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)
```

```{r}
data_folder <- here("data", "sea", "raw")
list.files(data_folder)
```

We see that we have multiple csv files.
We have abundance and biomass data for the Atlantic (Nurse, Solper), Atlantic and Manche (POMET).

# 1. Pomet campagne

## 1.1. Clean the data

First let's load files.

```{r}
d.poisson <- read.csv(here(data_folder, "poisson_pomet.csv"), sep = ";")
head(d.poisson)
```

Let's do a bit of cleaning.
Remove irrelevant columns.

```{r}
d.poisson <- d.poisson |> select(-c(Commentaire, Ecologique, Trophique, Engin_peche, Zone))
names(d.poisson)
```

Rename columns for clarity.

```{r}
d.poisson <- d.poisson |> rename(
  zone = ID_Zone,
  year = Annee,
  month = Mois,
  hour = Heure,
  date = Date,
  trait = Trait,
  start_x = Coord_Deb_xmin,
  start_y = Coord_Deb_ymin,
  end_x = Coord_Fin_xmin,
  end_y = Coord_Fin_ymin,
  depth = Prof,
  species = NomScient,
  position = Position,
  abundance = Nind_esp,
  weight_batch = Pds_esp
)
head(d.poisson)
```

We see that the zone ID column also contains zone names which is not necessary.

```{r}
unique(d.poisson$zone)
```

We want to keep only the ID.

```{r}
d.poisson$zone <- substr(d.poisson$zone, 1, 6)
unique(d.poisson$zone)
```

Now let's look at which species are more abundant on average.

```{r}
n <- 10
d <- d.poisson |>
  group_by(species) |>
  summarise(abundance_avg = mean(abundance)) |>
  arrange(desc(abundance_avg)) |>
  slice(1:n)
d
```

Let's pull these species and plot their time series.

::: {.callout-note}
Is it expected that "Spratus" and "Spratus spratus" are two distinct species?
This may be an error when species name was written.
If so species names should be cleaned up.
:::

```{r}
common_sp <- d |> pull(species)

d <- d.poisson |>
  filter(species %in% common_sp) |>
  group_by(species, year) |>
  summarise(abundance_avg = mean(abundance), .groups = "drop")
d$log_abundance <- log(d$abundance_avg)

ggplot(d, aes(x = year, y = log_abundance, color = species)) +
  geom_point() +
  geom_line() +
  labs(x = "Year", y = "Log abundance")
```


## 1.2. Plot traits

First coordinates need to be formatted.
They are given as strings and should be converted to numeric values.

```{r}
library(stringr)
d.poisson <- d.poisson |>
  mutate(
    start_x = start_x |> str_replace(",", ".") |> as.numeric(),
    start_y = start_y |> str_replace(",", ".") |> as.numeric(),
    end_x = end_x |> str_replace(",", ".") |> as.numeric(),
    end_y = end_y |> str_replace(",", ".") |> as.numeric(),
  )

summary(d.poisson[, c("start_x", "start_y", "end_x", "end_y")])
```

```{r}
ggplot(d.poisson, aes(x = start_x, y = start_y)) +
  geom_segment(aes(xend = end_x, yend = end_y, color = zone)) +
  coord_equal() +
  labs(x = "Longitude", y = "Latitude")
```

Now let's add a map in the background with `leaflet`.
Because we have many traits, we will focus on a single zone.
First, we need to convert traits to sf object.

```{r}
library(sf)
library(leaflet)

zone_plot <- first(d.poisson$zone)

traits_sf <- d.poisson |>
  filter(zone == zone_plot) |>
  rowwise() |>
  mutate(
    geometry = st_sfc(st_linestring(
      matrix(c(start_x, end_x, start_y, end_y), ncol = 2, byrow = FALSE)
    ), crs = 4326)
  ) |>
  st_as_sf()
```

Then, we can plot them, using year for the color gradient.

```{r}
pal <- colorNumeric(
  palette = "viridis",
  domain = traits_sf$year
)

leaflet(traits_sf) |>
  addTiles() |>
  addPolygons(
    data = traits_sf,
    color = ~ pal(year),
    opacity = 0.7,
    popup = ~ paste("Year: ", year)
  ) |>
  addLegend("bottomright",
    pal = pal, values = ~year,
    title = "Year",
    labFormat = labelFormat(big.mark = ""),
    opacity = 1
  )
```

Most traits follow the river suggesting that we have processed the data correctly.

Now let's focus on how fish sizes is measured.

## 1.3. Get fish size measurement rates

### 1.3.1. Prepare size data

First let's read the file containing fish size information.

```{r}
d.size <- read.csv(here(data_folder, "poisson_tailles_pomet.csv"), sep = ";")
head(d.size)
```

As before let's do a bit of cleaning before anything else.

```{r}
d.size <- d.size |> select(-c(
  Commentaire,
  Ecologique,
  Trophique,
  Engin_peche,
  Zone
))

head(d.size)
```

Then let's rename columns.

```{r}
d.size <- d.size |> rename(
  zone = ID_Zone,
  year = Annee,
  month = Mois,
  hour = Heure,
  date = Date,
  trait = Trait,
  start_x = Coord_Deb_xmin,
  start_y = Coord_Deb_ymin,
  end_x = Coord_Fin_xmin,
  end_y = Coord_Fin_ymin,
  depth = Prof,
  species = NomScient,
  position = Position,
  abundance = Nind_esp,
  weight_batch = Pds_esp,
  batch_size = Nind_esp_taille,
  oxygen = Pour_satO2,
  salinity = Salinite,
  length_cm = Longueur_fourche_cm,
  length_mm = Longueur_fourche_mm,
  temperature = Temperature_eau
)
head(d.size)
```

And cleaning zone ID.

```{r}
d.size$zone <- substr(d.size$zone, 1, 6)
head(d.size)
```

Furthermore, we can notice that length in mm are specified as caracters instead of numerics.

```{r}
class(d.size$length_mm)
```

```{r}
d.size$length_mm <- as.numeric(d.size$length_mm)
```

We can check on a subset of our dataframe that length in mm is ten times the length in cm.

```{r}
d <- d.size |>
  filter(zone == zone_plot) |>
  filter(!is.na(length_mm), year == 2014)

ggplot(d, aes(x = length_mm, y = length_cm)) +
  geom_point()
```

So let's remove the length in cm, which is a duplicate and less precise than the length in mm.

```{r}
d.size <- d.size |> select(-length_cm)
head(d.size)
```

### 1.3.2. Full size sampling

Now that the data is ready, we want to see at why frequency all individual fish sizes are measured.

We will begin with rough estimates, then refine these at the species and year levels.

```{r}
d <- d.size |>
  group_by(ID_interne_prelevement, species) |>
  summarise(n_measured = sum(batch_size), .groups = "drop")

d.join <- left_join(d.poisson, d, by = c("ID_interne_prelevement", "species"))
head(d.join)
```

To check that we did well, we verify that for a fishing event for which we know that all fish sizes have been measured that `abundance` = `n_measured`.

```{r}
d.join |> filter(ID_interne_prelevement == 61836630, species == "Abramis brama")
```

First, let's look at rough estimates.

```{r}
n_tot <- nrow(d.join) # 1 row: 1 species & 1 fishing event.
n_full <- nrow(d.join |> filter(abundance == n_measured))
n_full / n_tot
```

When filtering for full size sampling, we keep 83% of the rows of the original dataframe.
This means that most of the time all individual sizes are measured.
But, let's go in further details and how full size sampling depends on year, zone and species.

```{r}
d.sampling <- d.join |>
  filter(!is.na(n_measured)) |>
  group_by(species, ID_interne_prelevement) |>
  summarise(measurement_rate = n_measured / abundance) |>
  arrange(measurement_rate)
head(d.sampling)
```

We have only few species with low full measurement rate.
Even the species with worst full size sampling rate hase a rate of 0.5.
One reason that can explain why some species are more often fully measured than other is because of their abundance.
Measuring many individuals is cumbersome, so we expect a decreasing trend between full measurement rate and average abundance.

```{r}
d.abundance.avg <- d.poisson |>
  filter(!is.na(abundance)) |>
  group_by(species) |>
  summarise(abundance_avg = mean(abundance))

d.sampling <- left_join(d.sampling, d.abundance.avg, by = "species")

ggplot(d.sampling, aes(x = abundance_avg, y = measurement_rate)) +
  geom_point() +
  labs(x = "Average abundance", y = "Full measurement rate")
```

Let's look if we have trends per zone.

```{r}
d.sampling <- d.join |>
  filter(!is.na(n_measured)) |>
  group_by(zone) |>
  summarise(measurement_rate = mean(n_measured == abundance)) |>
  arrange(measurement_rate)
head(d.sampling)
```

There is also high variability between zone, with few zones where the measurement rate is low.

For now, we have seen what we wanted.
Overall, fish sizes are most of the time all measured.
This will help food web reconstruction.

# 2. Nurse campagne

Now let's focus on another campagne: Nurse.

## 2.1. Clean the data

Let's load the data.

```{r}
d.nurse.catch <- read.csv(here(data_folder, "Captures_Nurse_1980_2023.csv"), sep = ";")
head(d.nurse.catch)
```

```{r}
d.nurse.size <- read.csv(here(data_folder, "Tailles_Nurse_1980_2023.csv"), sep = ";")
head(d.nurse.size)
```

```{r}
d.nurse.trait <- read.csv(here(data_folder, "Traits_Nurse_1980_2023.csv"), sep = ";")
head(d.nurse.trait)
```

```{r}
d.nurse.strate <- read.csv(here(data_folder, "Strates_Nurse_1980_2023.csv"), sep = ";")
head(d.nurse.strate)
```

We see that we have four files.
The first file contains information about the catches: what, when, where.
The second file contains information about the individual sizes.
The third file contains geographical information about the trait.
I don't kow what contains the fourth file. Should ask Anik.

Let's rename column for consistency with other campagnes.

```{r}
d.nurse.catch <- d.nurse.catch |>
  select(-Campagne) |>
  rename(
    year = Annee,
    trait = Trait,
    species = Espece,
    abundance = Nombre,
    weight_batch = Poids
  )
head(d.nurse.catch)
```

```{r}
d.nurse.size <- d.nurse.size |>
  select(-c(Campagne, Poids, Age, Maturite)) |>
  rename(
    year = Annee,
    trait = Trait,
    species = Espece,
    sex = Sexe,
    length = Longueur,
    batch_size = Nombre
  )
head(d.nurse.size)
```

```{r}
d.nurse.trait <- d.nurse.trait |>
  select(-c(Campagne, ProfMoy)) |>
  rename(
    year = Annee,
    trait = Trait,
    month = Mois,
    strate = Strate,
    surface = SurfaceBalayee,
    lat = Lat,
    long = Long
  )
head(d.nurse.trait)
```

## 2.2. Get general information about the campagne

### 2.2.1. Evolution of the sampling effort

```{r}
trait_counts <- d.nurse.trait |>
  group_by(year) |>
  summarise(n_trait = length(unique(trait)))

ggplot(trait_counts, aes(x = year, y = n_trait)) +
  geom_line() +
  geom_point() +
  labs(x = "Year", y = "Number of traits")
```


```{r}
surface_year <- d.nurse.trait |>
  group_by(year) |>
  summarise(surface_tot = sum(surface))

ggplot(surface_year, aes(x = year, y = surface_tot)) +
  geom_line() +
  geom_point() +
  labs(x = "Year", y = "Surface sampled")
```

We see high variability in the sampling effort across the year.

### 2.2.2. Plot geographical position of traits

```{r}
trait_sf <- st_as_sf(d.nurse.trait, coords = c("long", "lat"), crs = 4326, remove = FALSE)

leaflet(trait_sf) |>
  addTiles() |>
  addCircleMarkers(
    radius = 2, fillOpacity = 0.7, stroke = FALSE,
    popup = ~ paste("Trait:", trait, "Year:", year)
  )
```

### 2.2.3. Fish size measurement rate

```{r}
d.nurse.batch <- d.nurse.size |>
  group_by(trait, species) |>
  summarise(n_measured = sum(batch_size), .groups = "drop")
head(d.nurse.batch)

d.nurse.batch <- left_join(d.nurse.batch, d.nurse.catch, by = c("species", "trait"))
head(d.nurse.batch)
```

```{r}
n_tot <- nrow(d.nurse.batch)
n_full <- nrow(d.nurse.batch |> filter(n_measured > abundance))
n_full #/ n_tot
```

For the Nurse campagne, we also have a very high size measurement rate.
Let's investigate the last campagne: SOLPER.

# 3. Solper campagne

## 3.1. Clean the data

Let's load the data.

```{r}
d.solper.catch <- read.csv(here(data_folder, "captures_SOLPER_2005_2011.csv"), sep = ";")
head(d.solper.catch)

d.solper.size <- read.csv(here(data_folder, "tailles_SOLPER_2005_2011.csv"), sep = ";")
head(d.solper.size)

d.solper.trait <- read.csv(here(data_folder, "traits_SOLPER_2005_2011.csv"), sep = ";")
head(d.solper.trait)
```

As for the Nurse campagne, we have one file for the catches, one for the individual sizes, and one for the trait positions.

As before, we will just rename the column for consistency.

```{r}
d.solper.catch <- d.solper.catch |>
  select(-Campagne) |>
  rename(
    year = Annee,
    trait = Trait,
    species = Espece,
    abundance = Nombre,
    weight_batch = Poids
  )
head(d.solper.catch)

d.solper.size <- d.solper.size |>
  select(-c(Campagne, Age, Maturite)) |>
  rename(
    year = Annee,
    trait = Trait,
    species = Espece,
    batch_size = Nombre,
    sex = Sexe,
    length = Longueur,
    weight = Poids
  )
head(d.solper.size)

d.solper.trait <- d.solper.trait |>
  select(-Campagne) |>
  rename(
    year = Annee,
    trait = Trait,
    month = Mois,
    strate = Strate,
    surface = SurfaceBalayee,
    lat = Lat,
    long = Long,
    depth = ProfMoy
  )
head(d.solper.trait)
```

Now that the data is cleaned let's look at general information about the campagne.

## 3.2. Get general information about the campagne

### 3.2.1. Evolution of sampling effort

```{r}
trait_year <- d.solper.trait |>
  group_by(year) |>
  summarise(n_trait = length(unique(trait)), surf_tot = sum(surface))

ggplot(trait_year, aes(x = year, y = n_trait, color = surf_tot)) +
  geom_line(color = "grey") +
  geom_point() +
  labs(x = "Year", y = "Number of traits", color = "Total surface")
```

The see that the sampling effort has been relatively constant, although decreasing with years.
Furthermore, as expected, the surface sampled is positively correlated with the number of traits.

### 3.2.3. Plot trait locations

```{r}
trait_sf <- st_as_sf(d.solper.trait, coords = c("long", "lat"), crs = 4326, remove = FALSE)

leaflet(trait_sf) |>
  addTiles() |>
  addCircleMarkers(
    radius = 2, fillOpacity = 0.7, stroke = FALSE,
    popup = ~ paste("Trait:", trait, "Year:", year)
  )
```

### 3.2.3. Full size sampling

```{r}
d.solper.batch <- d.solper.size |>
  group_by(species, trait, year) |>
  summarise(n_measured = sum(batch_size), .groups = "drop")

d.solper.batch <- left_join(d.solper.batch, d.solper.catch, by = c("species", "trait", "year"))

n_tot <- nrow(d.solper.batch)
n_full <- nrow(d.solper.batch |> filter(n_measured == abundance))
n_full / n_tot
```

Wow, it seems that for the Solper campagne every single fish has been measured.
It probably means that there has been some filtering about sampled species.
