---
title: "Infer missing fish sizes"
format: html
execute:
    message: false
    warning: false
---

# Summary

In this notebook, we

- we infer missing fish sizes for sea campagnes assuming a normal distribution
- validate our inference using cross-validation

# 0. Setup

```{r}
#| output: false
library(renv)
renv::load()
renv::restore() # Install missing librairies.
renv::status() # Check the project state.

library(targets)
library(here)
tar_config_set(
  store = here::here("_targets")
)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggVennDiagram)
set_theme(theme_minimal())
library(riversea)
```

# 1. Illustrate the inference with synthetic data

The idea of the inference is quite simple.
We observe a given distribution of values, representing fish sizes.
We assume that the underlying distribution is a normal distribution.
This choice can be seen as a choice toward simplicity.
We say that we only know about the mean and the variance of the distribution, and in this case the most probable distribution is a normal distribution.

To begin with, we illustrate the procedure on synthetic data.
So let's simulate values from a normal distribution.


```{r}
n <- 1000 # Sample size.
obs <- rnorm(n)
n_missing <- 100
obs_missing <- data.frame(true = rnorm(n_missing))

infer_missing_sizes <- function(obs, n_missing) {
  mu <- mean(obs)
  sigma <- sd(obs)
  rnorm(n_missing, mu, sigma)
}

obs_missing$pred <- infer_missing_sizes(obs, n_missing)

ggplot(obs_missing, aes(x = true, y = pred)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1)
```

We see that this method is quite bad at predicting fish size at the individual level.
But what really matters is whether the infered distribution of missing sizes corresponds to the true one.

```{r}
obs_missing_long <- obs_missing |>
  pivot_longer(
    cols = c(true, pred),
    names_to = "type",
    values_to = "value"
  )

ggplot(obs_missing_long, aes(x = value, fill = type)) +
  geom_histogram(alpha = 0.5, position = "identity")
```

We see that here it works quite well, because the sample size is large, and we are sampling missing values from a normal distribution, which won't be the case for real data.

# 2. Apply the methods to campagne data

First, we load our data set and summarise the different surveys.

```{r}
tidy_data <- tar_read(sea_data_tidy)
head(tidy_data$size)
```

The first step of the processing is to check the validity of the species name.
For now, we only verify whether the name is recognized by fishbase.
We filter out unrecognized species.

```{r}
valid_data <- filter_valid(tidy_data)
head(valid_data$size)
```

```{r}
tar_read(venn_diagram)
```

Next, we compute the minimal, mean and maximal length for each species.

```{r}
size_extrema <- valid_data$size |>
  group_by(species_valid) |>
  summarise(
    length_min = min(length),
    length_mean = mean(length),
    length_max = max(length),
    n = n(),
    .groups = "drop"
  ) |>
  arrange(desc(n))
head(size_extrema)
```

```{r}
data <- add_observation_number(valid_data)
head(data$catch)
```

```{r}
data_imp <- impute_size(data)
```

For sanity we can check that we have added the right number of sizes to the data frame.

```{r}
head(data_imp$size)
data_imp$catch |>
  mutate(n_miss = abundance_cor - n_measured) |>
  pull(n_miss) |>
  sum()
sum(!data_imp$size$measured)
```

