---
title: "Infer missing fish sizes"
format: html
execute:
    message: false
    warning: false
---

# Summary

In this notebook, we 

- we infer missing fish sizes for sea campagnes assuming a normal distribution
- validate our inference using cross-validation

# 0. Setup

```{r}
#| output: false
library(renv)
renv::load()
renv::restore() # Install missing librairies.
renv::status() # Check the project state.

library(targets)
library(here)
tar_config_set(
    store = here::here("_targets")
)
library(dplyr)
library(tidyr)
library(ggplot2)
set_theme(theme_minimal())
library(riversea)
```

# 1. Illustrate the inference with synthetic data

The idea of the inference is quite simple.
We observe a given distribution of values, representing fish sizes.
We assume that the underlying distribution is a normal distribution.
This choice can be seen as a choice toward simplicity.
We say that we only know about the mean and the variance of the distribution, and in this case the most probable distribution is a normal distribution.

To begin with, we illustrate the procedure on synthetic data.
So let's simulate values from a normal distribution.

```{r}
n <- 1000 # Sample size.
obs <- rnorm(n)
n_missing <- 100
obs_missing <- data.frame(true = rnorm(n_missing))

infer_missing_sizes <- function(obs, n_missing) {
    mu <- mean(obs)
    sigma <- sd(obs)
    return(rnorm(n_missing, mu, sigma))
}

obs_missing$pred <- infer_missing_sizes(obs, n_missing)

ggplot(obs_missing, aes(x = true, y = pred)) +
    geom_point() +
    geom_abline(intercept = 0, slope = 1)
```

We see that this method is quite bad at predicting fish size at the individual level.
But what really matters is whether the infered distribution of missing sizes corresponds to the true one.

```{r}
obs_missing_long <- obs_missing |>
    pivot_longer(
        cols = c(true, pred),
        names_to = "type",
        values_to = "value"
    )

ggplot(obs_missing_long, aes(x = value, fill = type)) +
    geom_histogram(alpha = 0.5, position = "identity")
```

We see that here it works quite well, because the sample size is large, and we are sampling missing values from a normal distribution, which won't be the case for real data.

# 2. Apply the methods to campagne data

First, we load our data set and summarise the different surveys.

```{r}
tidy_data <- tar_read(sea_data_tidy)

tidy_data$catch |>
    group_by(survey) |>
    summarise(
        n_tot = n(),
        year_min = min(year),
        year_max = max(year),
    )
```

```{r}
size_extrema <- tidy_data$size |>
    group_by(species) |>
    summarise(
        length_min = min(length),
        length_mean = mean(length),
        length_max = max(length),
        n = n(),
        .groups = "drop"
    )
head(size_extrema)
```

```{r}
species_params <- tidy_data$size |>
    group_by(species, trait, year) |>
    summarise(
        mu = mean(length),
        sigma = sd(length),
        n_obs = n(),
        .groups = "drop"
    )
head(species_params)

imputation_table <- tidy_data$catch |>
    left_join(species_params, by = c("trait", "species", "year")) |>
    mutate(n_missing = abundance - n_obs) |>
    filter(n_missing > 0)

infer_missing_size <- function(n_missing, mu, sigma, species, trait) {
    if (n_obs < 3) {
        size
    }
}

imputed_sizes <- imputation_table |>
    rowwise() |>
    mutate(
        length = list(rnorm(n_missing, mu, sigma))
    ) |>
    unnest() |> filter(is.na(length))
head(imputed_sizes)

missing_neg <- imputation_table |> filter(n_missing < 0) |> arrange(n_missing)
head(missing_neg)
```

We see that we have negative value for n_missing our imputation table.
These are very few, and probably due to errors.
Here, we choose to correct them, setting the abundance to the number of measured fish.

```{r}

```

```{r}
raw$pomet$size |>
    filter(ID_interne_prelevement == 61836945, NomScient == "Dicentrarchus labrax") |> pull(Nind_esp_taille) |> sum()

raw <- load_raw_sea_data("data/sea/raw")
raw$pomet$size |> filter(ID_interne_prelevement == 61838953, NomScient == "Anguilla anguilla")
```

```{r}
trait_id <- 61836379
tidy_data$catch |> filter(trait == trait_id)
tidy_data$size |>
    filter(trait == trait_id) |>
    uncount(batch_size)
```

## TODO

- Infer missing sizes in the different survey.
- Validate the inference by simulating missing data.