---
title: "Data spatial distribution"
format: 
  html:
    embed-resources: true
execute:
    message: false
    warning: false
---

## Summary

```{r}
#| output: false
library(renv)
renv::load()
renv::restore() # Install missing librairies.
renv::status() # Check the project state.

library(targets)
library(here)
tar_config_set(
  store = here::here("_targets")
)

library(ggplot2)
set_theme(theme_minimal())
library(dplyr)
```

## Sea survey data

Here is the map of the sampling station for the different surveys.

```{r}
tar_read(plot_station)
```

## Sea environmental data

Here is the kind of environment data we have for the sea.

```{r}
env <- tar_read(environment_data)
head(env)
```

We can have a look at the evolution of average temperature with years.

```{r}
env |>
  group_by(year) |>
  summarise(annual_temperature = mean(temp_mean), .groups = "drop") |>
  ggplot(aes(x = year, y = annual_temperature)) +
  geom_point() +
  geom_line() +
  labs(x = "Year", y = "Mean annual temperature")
```

However, this is not very informative because we do not account for the
spatialization of the data.
A better way, to investigate the data is to plot yearly map of the different
environmental variables.
For example, below we plot the distribution of mean oxygen level, chorophyl
concentration, and temperature.

```{r}
#| fig-width: 12
#| fig-height: 8

library(tidyr)
library(sf)
library(rnaturalearth)
library(rnaturalearthhires)

coastline <- ne_coastline(scale = "large", returnclass = "sf")

var_list <- c("o2_mean", "chl_mean", "temp_mean")
year_list <- c(2000, 2005, 2010, 2015, 2020)

env_long <- env |>
  pivot_longer(
    cols = c(
      chl_mean, nppv_mean, o2_mean, o2_cv,
      o2_perc_25, temp_mean, temp_cv, temp_perc_90, sali_mean
    ),
    names_to = "variable",
    values_to = "value"
  ) |>
  group_by(variable) |>
  filter(year %in% year_list & variable %in% var_list) |>
  mutate(value_scaled = scale(value)) |>
  ungroup()

ggplot(env_long, aes(x = x, y = y, fill = value_scaled)) +
  geom_tile() +
  geom_sf(data = coastline, inherit.aes = FALSE, linewidth = 0.3) +
  facet_grid(variable ~ year) +
  scale_fill_viridis_c() +
  coord_sf(xlim = range(env$x), ylim = range(env$y))
```

We can also plot the salinity for a given year, say 2023.
We use a log scale, as salinity changes a lot in the estuary.

```{r}
env |>
  mutate(log_salinity = log10(sali_mean)) |>
  filter(year == 2023) |>
  ggplot(aes(x = x, y = y, fill = log_salinity)) +
  geom_tile() +
  geom_sf(data = coastline, inherit.aes = FALSE, linewidth = 0.3) +
  scale_fill_viridis_c() +
  coord_sf(xlim = range(env$x), ylim = range(env$y))
```

We see that different environmental variables seem correlated.
We can have a look at them more exhaustively with pairwise scatterplot.

```{r}
#| fig-width: 12
#| fig-height: 8

cols <- c(
  "chl_mean", "nppv_mean", "o2_mean", "temp_mean", "sali_mean",
  "bathymetry"
)
year_vals <- c(2000, 2010, 2020)
GGally::ggpairs(
  env |> mutate(year = as.factor(year)) |> filter(year %in% year_vals),
  columns = cols,
  aes(colour = year, alpha = 0.1)
)
```

First, we see that we have some strong correlations between variables, such as,
between oxygen and temperature.
Interestingly, there are also changes between the correlation between the
variables.
In particular, we see that in the recent years, there is now a very strong
correlation between chlorophyll and net primary productivity.

## Food web structure metrics

Next, we investigate the structure of local food webs.
We focus on food web richness, connectance and trophic length.

```{r}
foodweb <- tar_read(local_foodwebs)
head(foodweb, n = 10)

foodweb_long <- foodweb |>
  pivot_longer(
    cols = c(log_species_richness, connectance, trophic_length),
    names_to = "variable",
    values_to = "value"
  ) |>
  filter(year %in% year_list)

vars <- unique(foodweb_long$variable)

plots <- purrr::map(vars, function(var) {
  foodweb_long |>
    filter(variable == var) |>
    ggplot(aes(x = longitude, y = latitude, color = value)) +
    geom_point(size = 2.5) +
    geom_sf(data = coastline, inherit.aes = FALSE, linewidth = 0.3) +
    facet_wrap(~year, nrow = 1) +
    scale_color_viridis_c(name = var) +
    coord_sf(
      xlim = range(foodweb_long$longitude),
      ylim = range(foodweb_long$latitude)
    ) +
    labs(x = NULL, y = NULL)
})

patchwork::wrap_plots(plots, ncol = 1)
```

We can also have a look at the correlations between the different structure
metrics.

```{r}
GGally::ggpairs(
  foodweb |>
    filter(year %in% year_vals) |>
    mutate(year = as.factor(year)),
  columns = c("log_trophic_richness", "connectance", "trophic_length"),
  aes(colour = year)
)
```

We see a strong correlation between richness and connectance and trophic length,
which is respectively negtive and positive.

## Matching food web and environmental data

We match food webs with the closest environmental data available.
We also record the distance between the foodweb and the environmental data,
so we can later filter out food webs for which we don't environmental data
available (e.g. if the distance is greater than 20 km).

```{r}
fw_with_env <- tar_read(fw_with_env)
```

Next, we want to check that the distance to environmental data is consistent
with our expectation, that is, the distance is low for the atlantic coast, and
increases as we go further into the Manche.

```{r}
ggplot(fw_with_env, aes(x = longitude, y = latitude, color = dist_to_env)) +
  geom_point(size = 2.5) +
  geom_sf(data = coastline, inherit.aes = FALSE, linewidth = 0.3) +
  scale_color_viridis_c() +
  coord_sf(
    xlim = range(fw_with_env$longitude),
    ylim = range(fw_with_env$latitude)
  ) +
  labs(x = NULL, y = NULL)
```

```{r}
dist_max <- 20
```

Let's say, that we want to keep only food webs for which with have
environmental variables closer than `r dist_max` km, and map them.

```{r}
ggplot(
  fw_with_env |> filter(dist_to_env < dist_max),
  aes(x = longitude, y = latitude, color = dist_to_env)
) +
  geom_point(size = 2.5) +
  geom_sf(data = coastline, inherit.aes = FALSE, linewidth = 0.3) +
  scale_color_viridis_c() +
  coord_sf(
    xlim = range(fw_with_env$longitude),
    ylim = range(fw_with_env$latitude)
  ) +
  labs(x = NULL, y = NULL)
```

For now we will work with these food webs.

```{r}
fw_with_env <- fw_with_env |>
  filter(dist_to_env < dist_max)
```

We can do a RDA to investigate how environmental variables affect food web
structure metrics.

```{r}
#| fig-width: 12
#| fig-height: 8

response <- fw_with_env |>
  select(
    log_species_richness, log_trophic_richness, connectance, trophic_length
  )

predictor <- fw_with_env |>
  select(
    chl_mean, nppv_mean, o2_mean, temp_mean,
    sali_mean, bathymetry, temp_cv, o2_cv, sali_cv
  ) |>
  scale()

cor(predictor, use = "complete.obs") |> round(2)

m <- vegan::rda(response, predictor)
summary(m)

ggvegan::autoplot(m)
```

Most of the variance is still unexplained by environmental driver.
It may mean two things.
First, we are missing some of the key environmental driver.
Second, there is a high stochaticity (which can be expected).
Interestingly, we see that the salinity is closely related to the richness.

We can have use variance partitioning method to assess what fraction of the
variance is explained by environment, spatial and temporal effects respectively.

```{r}
variance_partitioning <- vegan::varpart(
  response, predictor,
  fw_with_env |> select(longitude, latitude),
  fw_with_env |> select(year)
)

plot(
  variance_partitioning,
  digits = 2,
  Xnames = c("Environment", "Space", "Time")
)
```

## The continuum: map sea and river data


```{r}
load(file = here::here("data", "river", "station.rda"))
web_in_web_loc_path <- here::here("data", "river", "web_in_web.rda")
load(file = web_in_web_loc_path)
plot(riveratlas_station["ord_stra"])

com_metrics_st <- com_metrics |>
  left_join(op_st_filtered$op, by = "opcod") |>
  group_by(station) |>
  summarise(
    begin = min(lubridate::year(date)),
    end = max(lubridate::year(date)),
    nsampling = n(),
    med_richness = median(richness),
    med_biomass = median(biomass),
    med_nind = median(nind),
    .groups = "drop"
  )

net_metrics_st <- network_metrics |>
  group_by(station) |>
  summarise(
    med_connectance = median(connectance),
    med_max_tlvl = median(max_troph_lvl),
    med_nbnode = median(nbnode),
    med_w_avg_tlvl = median(weighted_avg_obs_tlvl),
    .groups = "drop"
  )

st_to_plot <- op_st_filtered$station |>
  select(station, libelle_sandre, town) |>
  left_join(com_metrics_st, by = "station") |>
  left_join(net_metrics_st, by = "station") |>
  mutate(
    longitude = st_coordinates(geometry)[, 1],
    latitude = st_coordinates(geometry)[, 2]
  ) |>
  st_drop_geometry()
```

Is there a continuum?

```{r}
#| fig-width: 12
#| fig-height: 8

river <- st_to_plot |>
  select(longitude, latitude, med_richness, med_connectance, med_max_tlvl) |>
  rename(connectance = med_connectance, trophic_length = med_max_tlvl) |>
  mutate(log_species_richness = log(med_richness), type = "river") |>
  select(-med_richness)

sea <- foodweb |>
  select(
    longitude, latitude, connectance, log_species_richness, trophic_length
  ) |>
  mutate(type = "sea")

river_sea <- rbind(river, sea)

river_sea |>
  ggplot(aes(x = longitude, y = latitude, color = type)) +
  geom_point(size = 2.5) +
  geom_sf(data = coastline, inherit.aes = FALSE, linewidth = 0.3) +
  coord_sf(
    xlim = range(river_sea$longitude),
    ylim = range(river_sea$latitude)
  ) +
  labs(x = NULL, y = NULL)
```

Species richness.

```{r}
#| fig-width: 12
#| fig-height: 8

river_sea |>
  ggplot(aes(x = longitude, y = latitude, color = log_species_richness)) +
  geom_point(size = 2.5) +
  geom_sf(data = coastline, inherit.aes = FALSE, linewidth = 0.3) +
  scale_color_viridis_c() +
  coord_sf(
    xlim = range(river_sea$longitude),
    ylim = range(river_sea$latitude)
  ) +
  labs(x = NULL, y = NULL)
```

Connectance.

```{r}
#| fig-width: 12
#| fig-height: 8

river_sea |>
  ggplot(aes(x = longitude, y = latitude, color = connectance)) +
  geom_point(size = 2.5) +
  geom_sf(data = coastline, inherit.aes = FALSE, linewidth = 0.3) +
  scale_color_viridis_c() +
  coord_sf(
    xlim = range(river_sea$longitude),
    ylim = range(river_sea$latitude)
  ) +
  labs(x = NULL, y = NULL)
```

Trophic length.

```{r}
#| fig-width: 12
#| fig-height: 8

river_sea |>
  ggplot(aes(x = longitude, y = latitude, color = trophic_length)) +
  geom_point(size = 2.5) +
  geom_sf(data = coastline, inherit.aes = FALSE, linewidth = 0.3) +
  scale_color_viridis_c() +
  coord_sf(
    xlim = range(river_sea$longitude),
    ylim = range(river_sea$latitude)
  ) +
  labs(x = NULL, y = NULL)
```

```{r}
#| eval: false

river_sea |> 
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
  mapview::mapview(
    zcol = c("type", "log_species_richness", "connectance", "trophic_length")
  )
```


