---
title: "River model: impact of nutrients and dams on food web structure"
bibliography: ../references.bib
format:
  html:
    theme: cosmo
    code-fold: true
    toc: true
    embed-resources: true
execute:
    message: false
    warning: false
---

## Todo

- [x] Join AMOBIO stressors + distance to mouth to river food web structure
- [x] Checkpoint: coverage and missingness of covariates
- [x] Checkpoint: spatial/temporal coverage and response distributions
- [x] Decide how to handle nutrient / dam-distance-downstream missingness
- [ ] Build SPDE mesh over river station locations (Lambert-93)
- [ ] Fit richness ~ nutrients + dams * distance to mouth + year + SPDE
  (path *a*)
- [ ] Checkpoint: richness model diagnostics and predictions
- [ ] Fit connectance / trophic length / diet overlap ~ richness + nutrients +
  dams * distance to mouth + year + SPDE (paths *b* and *c'*)
- [ ] Checkpoint each model
- [ ] Direct/indirect effect decomposition (product of coefficients, via
  posterior samples)
- [ ] Swap Euclidean SPDE for a `MetricGraph` hydrographic-distance spatial
  term [@Bolin-Simas-Wallin-2023]

```{r setup}
library(renv)
renv::load()
library(targets)
library(here)
tar_config_set(
  store = here::here("_targets")
)
library(ggplot2)
library(dplyr)
library(sf)
devtools::load_all()
nice_theme()
knitr::opts_chunk$set(autodep = TRUE)
```

## Data preparation

We join two dataframes:

- AMOBIO environmental stressors (nutrients and dams), already matched with
food web structure (`amobio_aspe_data`);

- and hydrographic distance to river mouth (`inland_distance_to_mouth`)

```{r data-overview}
amobio <- tar_read(amobio_aspe_data) |> st_drop_geometry()
head(amobio)

dist_mouth <- tar_read(inland_distance_to_mouth) |>
  select(operation_id, district, dist_mouth_m)
head(dist_mouth)

river_width <- tar_read(river_width)
head(river_width)
```

`amobio` and `dist_mouth` both carry `operation_id`, drawn from the same
ASPE ID space, so they join directly. `river_width` is station-level
(Naiades hydromorphology survey, averaged across all available visit
dates per station), so it joins on `sandre_code`.

```{r prepare-data}
river_data <- amobio |>
  inner_join(dist_mouth, by = "operation_id") |>
  left_join(river_width, by = "sandre_code") |>
  select(
    operation_id, sandre_code, year, survey, district, longitude, latitude,
    log_species_richness, connectance, trophic_length, diet_overlap,
    elevation, river_width,
    ammonium, nitrite, nitrate, phosphate,
    dam_distance_upstream, dam_distance_downstream,
    barrier_upstream, barrier_downstream,
    dist_mouth_m
  )

head(river_data)
```

We go from `r nrow(amobio)` river operations with AMOBIO data, and
`r nrow(dist_mouth)` with a computed distance to mouth, down to
`r nrow(river_data)` once both are required simultaneously. Most droppped
rows are because of the filtering on hydrographic basin kept, and a few
stations that are not present in AMOBIO dataset.

## Data overview

### Missing values

```{r missing-values}
vars <- c(
  "ammonium", "nitrite", "nitrate", "phosphate",
  "dam_distance_upstream", "dam_distance_downstream",
  "barrier_upstream", "barrier_downstream", "dist_mouth_m",
  "elevation", "river_width",
  "log_species_richness", "connectance", "trophic_length", "diet_overlap"
)

nutrient_na <- river_data |>
  select(ammonium, nitrite, nitrate, phosphate) |>
  mutate(
    one_is_na = if_any(c(ammonium, nitrite, nitrate, phosphate), is.na),
    all_are_na = if_all(c(ammonium, nitrite, nitrate, phosphate), is.na)
  )

n_one_na <- sum(nutrient_na$one_is_na)
n_all_na <- sum(nutrient_na$all_are_na)
n_mismatch <- sum(nutrient_na$one_is_na & !nutrient_na$all_are_na)

tibble(
  variable = vars,
  n_na = sapply(river_data[vars], \(x) sum(is.na(x))),
  pct_na = round(100 * n_na / nrow(river_data), 1)
) |>
  arrange(desc(n_na))
```

Two things stand out

- nutrients are missing together, not independently: `r n_all_na` of the
  `r n_one_na` rows missing any nutrient are missing *all four* at once, and
  only `r n_mismatch` rows show a partial mismatch;

- dam upstream distance is missing far more of than the downstream distance
  (`dam_distance_upstream`/`dam_distance_downstream` were swapped in
  `rename_amobio()` relative to AMOBIO's own scale codes - fixed in
  `R/amobio.R`, pipeline rebuilt).

Coverage for `dam_distance_upstream` is too low, so we drop it from the
model. For the other covariates we'll use complete cases per model.

### Spatial coverage

```{r spatial-coverage}
#| fig-width: 9
#| fig-height: 8

basin <- tar_read(basin)
france <- rnaturalearth::ne_countries(
  geounit = "france", type = "map_units", scale = "medium", returnclass = "sf"
)

ggplot() +
  geom_sf(data = france, fill = NA, color = "grey80", linewidth = 0.25) +
  geom_sf(
    data = basin |> filter(district %in% unique(river_data$district)),
    fill = "grey97", color = "grey40", linewidth = 0.4
  ) +
  geom_point(
    data = river_data, aes(longitude, latitude, color = district),
    size = 0.6, alpha = 0.7
  ) +
  coord_sf(xlim = c(-5.3, 8.3), ylim = c(41.2, 51.2), expand = FALSE) +
  labs(x = "Longitude", y = "Latitude", color = "Catchment") +
  theme(panel.grid = element_blank())
```

Good coverage in all four basin although they vary in sizes.

### Temporal coverage

```{r temporal-coverage}
#| fig-width: 7
#| fig-height: 3.5

ggplot(river_data, aes(year)) +
  geom_bar(fill = "steelblue") +
  labs(x = NULL, y = "Number of operations")
```

Mostly even coverage, but for 2018 and 2021 which have far less fishing
operations.

## Response distributions

Let's see the distributions of our response variable so we can choose the
relevant reponse distributions.

```{r response-distributions}
#| fig-width: 9
#| fig-height: 6

river_data |>
  select(log_species_richness, connectance, trophic_length, diet_overlap) |>
  tidyr::pivot_longer(everything(), names_to = "metric", values_to = "value") |>
  ggplot(aes(value)) +
  geom_histogram(bins = 40, fill = "steelblue") +
  facet_wrap(~metric, scales = "free") +
  labs(x = NULL, y = NULL)
```

Connectance and diet overlap are bounded between 0 and 1; they will be properly
modeled by a beta distribution. Log species richness might be handle well by a
gaussian distribution. The real thinking should be done for the trophic level
which is clearly multimodal. Let's try to investigate why.

We split each local food web by the predation depth of its top species: does
it eat no fish at all, a fish that itself only eats invertebrates, or a fish
that is itself piscivorous.


```{r trophic-length-modes}
#| cache: true
#| fig-width: 9
#| fig-height: 4

river_foodweb <- tar_read(river_foodweb_structure) |>
  filter(operation_id %in% river_data$operation_id)
resource_list <- tar_read(resource_list)

classify_predation_depth <- function(web, resource_list) {
  nodes <- data.frame(node = colnames(web))
  links <- web |>
    as.data.frame() |>
    mutate(resource = rownames(web)) |>
    tidyr::pivot_longer(-resource, names_to = "consumer") |>
    filter(value == 1) |>
    select(-value)
  community <- cheddar::Community(
    nodes = nodes, properties = list(title = "Community"), trophic.links = links
  )
  tl <- cheddar::TrophicLevels(community) |>
    as.data.frame() |>
    tibble::rownames_to_column("node") |>
    rename(TL = ChainAveragedTL)
  top_nodes <- tl$node[!is.na(tl$TL) & tl$TL == max(tl$TL, na.rm = TRUE)]

  fish_prey_of <- function(n) {
    prey <- rownames(web)[web[, n] == 1]
    prey[!(prey %in% resource_list)]
  }
  top_fish_prey <- unique(unlist(lapply(top_nodes, fish_prey_of)))

  if (length(top_fish_prey) == 0) {
    "no fish eaten"
  } else if (any(sapply(top_fish_prey, \(p) length(fish_prey_of(p)) > 0))) {
    "eats a piscivorous fish"
  } else {
    "eats a non-piscivorous fish"
  }
}

river_foodweb <- river_foodweb |>
  mutate(
    predation_depth = sapply(
      foodweb, classify_predation_depth, resource_list = resource_list
    ),
    predation_depth = factor(
      predation_depth,
      levels = c(
        "no fish eaten", "eats a non-piscivorous fish",
        "eats a piscivorous fish"
      )
    )
  )

tl_range <- range(river_foodweb$trophic_length)
normal_overlay <- river_foodweb |>
  group_by(predation_depth) |>
  summarise(
    mean_tl = mean(trophic_length), sd_tl = sd(trophic_length), .groups = "drop"
  ) |>
  rowwise() |>
  reframe(
    predation_depth = predation_depth,
    x = seq(tl_range[1], tl_range[2], length.out = 200),
    y = dnorm(x, mean_tl, sd_tl)
  )

ggplot(river_foodweb, aes(trophic_length)) +
  geom_histogram(
    aes(y = after_stat(density)), bins = 30, fill = "steelblue", alpha = 0.7
  ) +
  geom_line(
    data = normal_overlay, aes(x, y), color = "firebrick", linewidth = 0.8
  ) +
  facet_wrap(~predation_depth, scales = "free") +
  labs(x = "Trophic length", y = "Density")
```

We see that the 'predation depth' cleanly separates our three modes of the
trophic level distribution. So we will go for an ordinal model the classify
these modes and model each of the mode individually as a gaussian.

## Building the mesh

We build a mesh that covers all fishing stations.

```{r river-mesh}
library(INLA)
library(fmesher)

stations <- river_data |>
  distinct(sandre_code, longitude, latitude) |>
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
  st_transform(2154) # Lambert-93, coordinates in meters.

coords <- st_coordinates(stations)

mesh <- fm_mesh_2d_inla(
  loc = coords,
  max.edge = c(15000, 60000),
  offset = c(15000, 100000),
  cutoff = 5000
)
plot(mesh, edge.color = rgb(0.6, 0.6, 0.6), lwd = 0.5)
points(coords, col = "steelblue", pch = 16, cex = 0.4)
```

```{r spde-setup}
spde <- inla.spde2.pcmatern(
  mesh = mesh,
  prior.range = c(50000, 0.5),
  prior.sigma = c(1, 0.01)
)

A <- inla.spde.make.A(mesh = mesh, loc = coords)
spatial_index <- inla.spde.make.index("spatial.field", n.spde = spde$n.spde)

site_idx <- match(river_data$sandre_code, stations$sandre_code)
A_expanded <- A[site_idx, ]
```

### Collinearity check

Before fitting, we check pairwise correlation between covariates, on the
same complete-case rows used for fitting, log-transforming the skewed ones
first.

```{r collinearity}
#| fig-width: 12
#| fig-height: 10

candidate_vars <- c(
  "log_species_richness", "elevation", "river_width",
  "ammonium", "nitrite", "nitrate", "phosphate",
  "dam_distance_downstream", "barrier_upstream", "barrier_downstream",
  "dist_mouth_m", "year", "sandre_code", "district"
)
river_model_data_raw <- river_data |>
  filter(complete.cases(river_data[candidate_vars]))

river_model_data_raw <- river_model_data_raw |>
  mutate(
    log_ammonium = log(ammonium),
    log_nitrite = log(nitrite),
    log_nitrate = log(nitrate),
    log_phosphate = log(phosphate),
    log_dam_distance_downstream = log(dam_distance_downstream),
    log_barrier_upstream = log1p(barrier_upstream),
    log_barrier_downstream = log1p(barrier_downstream),
    log_river_width = log(river_width)
  )

river_model_data_raw |>
  select(
    log_ammonium, log_nitrite, log_nitrate, log_phosphate,
    log_dam_distance_downstream, log_barrier_upstream,
    log_barrier_downstream, dist_mouth_m, elevation, log_river_width
  ) |>
  GGally::ggpairs()
```

We observe a strong correlation between ammonium and the other nutrients, so we
decide to drop this variable to ensure model identifiability.

```{r}
river_model_data_raw <- river_model_data_raw |>
  select(-log_ammonium)
```

Shared term ordering/typing and plotting helper for the coefficient plots
of all three structure models below (richness, connectance, predation
depth), so it isn't redefined per model. Also defines the small helpers
used to pull named components (fixed effects, random effect vectors,
the spatial field) out of an `inla.posterior.sample()` draw, reused by
every AME/interaction computation later in the notebook.

```{r coefficient-plot-helper}
get_val <- function(s, name) {
  unname(s$latent[rownames(s$latent) == paste0(name, ":1"), 1])
}
get_re_vec <- function(s, prefix, k) {
  unname(sapply(
    seq_len(k), \(i) s$latent[rownames(s$latent) == paste0(prefix, ":", i), 1]
  ))
}
get_spatial_vec <- function(s) {
  unname(s$latent[grep("^spatial.field:", rownames(s$latent)), 1])
}

term_order <- c(
  "log_species_richness",
  "dist_mouth_m", "elevation", "river_width",
  "nitrate", "nitrate2", "phosphate", "phosphate2",
  "barrier_upstream", "barrier_downstream",
  "barrier_upstream:dist_mouth_m", "barrier_downstream:dist_mouth_m"
)
term_type <- c(
  log_species_richness = "richness",
  dist_mouth_m = "physical", elevation = "physical", river_width = "physical",
  nitrate = "nutrient", nitrate2 = "nutrient",
  phosphate = "nutrient", phosphate2 = "nutrient",
  barrier_upstream = "connectivity", barrier_downstream = "connectivity",
  `barrier_upstream:dist_mouth_m` = "interaction",
  `barrier_downstream:dist_mouth_m` = "interaction"
)

plot_coefficients <- function(fit, x_label) {
  fit$summary.fixed |>
    tibble::rownames_to_column("term") |>
    filter(term %in% term_order) |>
    mutate(
      term = factor(term, levels = rev(term_order)),
      type = term_type[as.character(term)]
    ) |>
    ggplot(aes(mean, term, color = type)) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
    geom_pointrange(aes(xmin = `0.025quant`, xmax = `0.975quant`)) +
    labs(x = x_label, y = NULL, color = "Term type")
}
```

## Richness model

$$
\log S_i \sim \mathcal{N}(\mu_i, \sigma_i)
$$
$$
\mu_i = \mathbf{x}_i^\top \boldsymbol\beta +
\text{year}_i + \text{district}_i +
\text{SPDE}_i
$$

where $\mathbf{x}$ is the vector of covariates (nutrients, dams, etc.) and
$\beta$ their fixed effect.


```{r richness-model}
#| cache: true

river_model_data <- river_model_data_raw |>
  mutate(
    nitrate = log(nitrate),
    phosphate = log(phosphate),
    barrier_upstream = log1p(barrier_upstream),
    barrier_downstream = log1p(barrier_downstream),
    river_width = log(river_width),
    district_id = as.integer(factor(district))
  ) |>
  mutate(across(
    c(
      nitrate, phosphate,
      barrier_upstream, barrier_downstream, dist_mouth_m,
      elevation, river_width
    ),
    \(x) as.numeric(scale(x))
  )) |>
  mutate(nitrate2 = nitrate^2, phosphate2 = phosphate^2)

site_idx_model <- match(river_model_data$sandre_code, stations$sandre_code)
A_model <- A[site_idx_model, ]

richness_stack <- inla.stack(
  data = list(y = river_model_data$log_species_richness),
  A = list(A_model, 1),
  effects = list(
    spatial.field = spatial_index,
    data.frame(
      intercept = 1,
      elevation = river_model_data$elevation,
      river_width = river_model_data$river_width,
      nitrate = river_model_data$nitrate,
      nitrate2 = river_model_data$nitrate2,
      phosphate = river_model_data$phosphate,
      phosphate2 = river_model_data$phosphate2,
      barrier_upstream = river_model_data$barrier_upstream,
      barrier_downstream = river_model_data$barrier_downstream,
      dist_mouth_m = river_model_data$dist_mouth_m,
      year = river_model_data$year,
      district_id = river_model_data$district_id
    )
  ),
  tag = "est"
)

fit_richness <- inla(
  y ~ -1 + intercept +
    elevation + river_width +
    nitrate + nitrate2 + phosphate + phosphate2 +
    barrier_upstream + barrier_downstream +
    dist_mouth_m +
    barrier_upstream:dist_mouth_m +
    barrier_downstream:dist_mouth_m +
    f(year, model = "iid") +
    f(district_id, model = "iid") +
    f(spatial.field, model = spde),
  family = "gaussian",
  data = inla.stack.data(richness_stack),
  control.predictor = list(A = inla.stack.A(richness_stack), compute = TRUE),
  control.compute = list(waic = TRUE, config = TRUE)
)

summary(fit_richness)
```

All covariates are standardized, so posterior means/CIs are directly
comparable across terms.

```{r coefficient-plot}
#| fig-width: 9
#| fig-height: 6

plot_coefficients(
  fit_richness, "Standardized coefficient (posterior mean, 95% CI)"
)
```

## Connectance model

$$
\text{connectance}_i \sim \text{Beta}(\mu_i, \phi)
$$
$$
\text{logit}(\mu_i) = \gamma_r \, \text{richness}_i +
\mathbf{x}_i^\top \boldsymbol\gamma +
\text{year}_i + \text{district}_i + \text{SPDE}_i
$$

where $\mu_i \in (0, 1)$ is the mean and $\phi$ a shared precision
hyperparameter. We include richness in the model as mediator variable, so we can
disentangle direct and richness-mediated effect of stressors on food web
structure.

We first prepare the data for the model.

```{r connectance-data}
connectance_vars <- c(
  "connectance", "log_species_richness", "elevation", "river_width",
  "nitrate", "phosphate",
  "barrier_upstream", "barrier_downstream",
  "dist_mouth_m", "year", "sandre_code", "district"
)
river_connectance_data <- river_data |>
  filter(complete.cases(river_data[connectance_vars]))

river_connectance_data <- river_connectance_data |>
  mutate(
    nitrate = log(nitrate),
    phosphate = log(phosphate),
    barrier_upstream = log1p(barrier_upstream),
    barrier_downstream = log1p(barrier_downstream),
    river_width = log(river_width),
    district_id = as.integer(factor(district))
  ) |>
  mutate(across(
    c(
      log_species_richness, nitrate, phosphate,
      barrier_upstream, barrier_downstream,
      dist_mouth_m, elevation, river_width
    ),
    \(x) as.numeric(scale(x))
  )) |>
  mutate(nitrate2 = nitrate^2, phosphate2 = phosphate^2)

site_idx_connectance <- match(
  river_connectance_data$sandre_code, stations$sandre_code
)
A_connectance <- A[site_idx_connectance, ]
```

Then, we define and fit the model.

```{r connectance-model}
#| cache: true

connectance_stack <- inla.stack(
  data = list(y = river_connectance_data$connectance),
  A = list(A_connectance, 1),
  effects = list(
    spatial.field = spatial_index,
    data.frame(
      intercept = 1,
      log_species_richness = river_connectance_data$log_species_richness,
      elevation = river_connectance_data$elevation,
      river_width = river_connectance_data$river_width,
      nitrate = river_connectance_data$nitrate,
      nitrate2 = river_connectance_data$nitrate2,
      phosphate = river_connectance_data$phosphate,
      phosphate2 = river_connectance_data$phosphate2,
      barrier_upstream = river_connectance_data$barrier_upstream,
      barrier_downstream = river_connectance_data$barrier_downstream,
      dist_mouth_m = river_connectance_data$dist_mouth_m,
      year = river_connectance_data$year,
      district_id = river_connectance_data$district_id
    )
  ),
  tag = "est"
)

fit_connectance <- inla(
  y ~ -1 + intercept + log_species_richness + elevation + river_width +
    nitrate + nitrate2 + phosphate + phosphate2 +
    barrier_upstream + barrier_downstream +
    dist_mouth_m +
    barrier_upstream:dist_mouth_m +
    barrier_downstream:dist_mouth_m +
    f(year, model = "iid") +
    f(district_id, model = "iid") +
    f(spatial.field, model = spde),
  family = "beta",
  data = inla.stack.data(connectance_stack),
  control.predictor = list(A = inla.stack.A(connectance_stack), compute = TRUE),
  control.compute = list(waic = TRUE, config = TRUE, cpo = TRUE)
)
```

We show the model fixed effects below.

```{r connectance-coefficient-plot}
#| fig-width: 9
#| fig-height: 6

plot_coefficients(
  fit_connectance, "Standardized coefficient (posterior mean, 95% CI)"
)
```

Richness as large negative effect on connectance, which strong reflects
topological constraint. Beyond this effect, we observe direct effects of
stressors on food web connectance. Nitrate increases connectance, while
phosphate has the opposite effect. This oppposite effect of nutrients is
interesting, but further analysis are required to better understand these
effects. The distance to downstream has a positive effect on connectance. We
discuss the dam impact along the river continuum further below.

### Sanity check

Before that we run a quick sanity check to ensure that our gaussian random field
capture properly the residual spatial structure, and so that no large spatial
struture (e.g. basin scale) remains.

```{r connectance-spatial-residual}
#| fig-width: 9
#| fig-height: 8

proj <- inla.mesh.projector(mesh, dims = c(300, 300))
field_grid <- inla.mesh.project(
  proj, fit_connectance$summary.random$spatial.field$mean
)
grid_df <- expand.grid(x = proj$x, y = proj$y)
grid_df$z <- as.vector(field_grid)
grid_df <- grid_df |> filter(!is.na(z))

basin_connectance <- basin |>
  filter(district %in% unique(river_connectance_data$district)) |>
  st_transform(2154)
stations_connectance <- stations |>
  filter(sandre_code %in% river_connectance_data$sandre_code)

ggplot() +
  geom_raster(data = grid_df, aes(x, y, fill = z)) +
  scale_fill_gradient2(
    low = "firebrick", mid = "white", high = "steelblue", midpoint = 0
  ) +
  geom_sf(data = basin_connectance, fill = NA, color = "grey20", linewidth = 0.4) +
  geom_sf(data = stations_connectance, size = 0.4, alpha = 0.5, color = "black") +
  coord_sf(crs = 2154, datum = NA) +
  labs(
    x = NULL, y = NULL, fill = "spatial\neffect"
  )
```

We observe no large spatial structure, meaning that the spatial random field does the
work.

### Interaction AME setup

Reused later, both by the nutrient curvature section above and the
combined dam/barrier interaction figure at the end of the notebook -
no standalone result here.

```{r connectance-interaction-setup}
conn_main_vars <- c(
  "log_species_richness", "elevation", "river_width",
  "nitrate", "phosphate",
  "barrier_upstream", "barrier_downstream",
  "dist_mouth_m"
)
conn_inter_pairs <- list(
  c("barrier_upstream", "dist_mouth_m"),
  c("barrier_downstream", "dist_mouth_m")
)
conn_inter_names <- sapply(conn_inter_pairs, \(p) paste0(p[1], ":", p[2]))
conn_quad_vars <- c("nitrate", "phosphate")
conn_quad_names <- paste0(conn_quad_vars, "2")

X_conn <- as.matrix(river_connectance_data[conn_main_vars])
colnames(X_conn) <- conn_main_vars
year_levels_conn <- sort(unique(river_connectance_data$year))
year_idx_conn <- match(river_connectance_data$year, year_levels_conn)
district_idx_conn <- river_connectance_data$district_id

extract_conn <- function(s) {
  list(
    intercept = get_val(s, "intercept"),
    betas = setNames(sapply(conn_main_vars, \(v) get_val(s, v)), conn_main_vars),
    inter_betas = setNames(
      sapply(conn_inter_pairs, \(p) get_val(s, paste0(p[1], ":", p[2]))), conn_inter_names
    ),
    quad_betas = setNames(sapply(conn_quad_names, \(v) get_val(s, v)), conn_quad_names),
    re_offset = get_re_vec(s, "year", length(year_levels_conn))[year_idx_conn] +
      get_re_vec(s, "district_id", 4)[district_idx_conn] +
      as.numeric(A_connectance %*% get_spatial_vec(s))
  )
}

linpred_conn <- function(Xmat, betas, inter_betas, quad_betas) {
  out <- as.numeric(Xmat %*% betas[conn_main_vars])
  for (k in seq_along(conn_inter_pairs)) {
    p <- conn_inter_pairs[[k]]
    out <- out + (Xmat[, p[1]] * Xmat[, p[2]]) * inter_betas[k]
  }
  for (k in seq_along(conn_quad_vars)) {
    out <- out + (Xmat[, conn_quad_vars[k]]^2) * quad_betas[k]
  }
  out
}

predicted_conn <- function(Xmat, cd) {
  eta <- cd$intercept + linpred_conn(Xmat, cd$betas, cd$inter_betas, cd$quad_betas) + cd$re_offset
  plogis(eta)
}
```

```{r connectance-interaction-compute}
#| cache: true

n_ame_draws_conn <- 300
ame_eps_conn <- 0.02
set.seed(1)
ps_conn <- inla.posterior.sample(n_ame_draws_conn, fit_connectance)
cd_list <- lapply(ps_conn, extract_conn)

dist_mouth_levels_conn <- quantile(river_connectance_data$dist_mouth_m, probs = c(0.1, 0.5, 0.9))
level_labels <- c("low", "median", "high")

rows <- list()
for (pair_idx in seq_along(conn_inter_pairs)) {
  var <- conn_inter_pairs[[pair_idx]][1]
  for (lvl_idx in seq_along(dist_mouth_levels_conn)) {
    lvl <- dist_mouth_levels_conn[lvl_idx]
    draws <- numeric(n_ame_draws_conn)
    for (m in seq_len(n_ame_draws_conn)) {
      cd <- cd_list[[m]]
      X_plus <- X_conn; X_plus[, var] <- X_plus[, var] + ame_eps_conn
      X_plus[, "dist_mouth_m"] <- lvl
      X_minus <- X_conn; X_minus[, var] <- X_minus[, var] - ame_eps_conn
      X_minus[, "dist_mouth_m"] <- lvl
      pred_plus <- predicted_conn(X_plus, cd)
      pred_minus <- predicted_conn(X_minus, cd)
      draws[m] <- mean((pred_plus - pred_minus) / (2 * ame_eps_conn))
    }
    rows[[length(rows) + 1]] <- tibble(
      var = var, dist_mouth_level = level_labels[lvl_idx],
      dist_mouth_value = as.numeric(lvl), mean = mean(draws),
      `0.025quant` = quantile(draws, 0.025),
      `0.975quant` = quantile(draws, 0.975)
    )
  }
}
interaction_ame_conn <- bind_rows(rows) |>
  mutate(dist_mouth_level = factor(dist_mouth_level, levels = level_labels))
```

## Trophic length model

We model trophic lengths in two distinct parts as it is trimodal.
The ordinal part models the `predation_depth`, that is, whether the community
has:
1) no species eating fish species;
2) a fish species eating a non-piscivorous;
3) a fish species eating a piscivorous species.
Then, knowing this category, we model trophic length as a conditional gaussian
distribution.

### Ordinal part

With $D_i \in \{1, 2, 3\}$ the predation-depth category (no fish eaten /
eats a non-piscivorous fish / eats a piscivorous fish) of operation $i$,
a proportional-odds (cumulative logit) model:

$$
\text{logit}\left[P(D_i \le k)\right] = \theta_k - \eta_i,
\quad k \in \{1, 2\}
$$
$$
\eta_i = \mathbf{x}_i^\top \boldsymbol\beta +
\text{year}_i + \text{district}_i +
\text{SPDE}_i
$$

where $\mathbf{x}_i$ is the same covariate vector as the richness model
(nutrients, dam/barrier variables and their interaction with distance to
mouth, elevation, river width) and $\theta_1 < \theta_2$ are estimated
thresholds. A positive $\beta$ pushes towards a higher (more
piscivorous) category.

We first prepare the data,

```{r trophic-length-data}
tl_vars <- c(
  "log_species_richness", "elevation", "river_width",
  "nitrate", "phosphate",
  "barrier_upstream", "barrier_downstream",
  "dist_mouth_m", "year", "sandre_code", "district"
)
river_tl_foodweb <- river_foodweb |>
  st_drop_geometry() |>
  select(operation_id, predation_depth)

river_tl_data <- river_data |>
  filter(complete.cases(river_data[tl_vars])) |>
  left_join(river_tl_foodweb, by = "operation_id")

river_tl_data <- river_tl_data |>
  mutate(
    nitrate = log(nitrate),
    phosphate = log(phosphate),
    barrier_upstream = log1p(barrier_upstream),
    barrier_downstream = log1p(barrier_downstream),
    river_width = log(river_width),
    district_id = as.integer(factor(district)),
    predation_depth_int = as.integer(predation_depth)
  ) |>
  mutate(across(
    c(
      log_species_richness, nitrate, phosphate,
      barrier_upstream, barrier_downstream,
      dist_mouth_m, elevation, river_width
    ),
    \(x) as.numeric(scale(x))
  )) |>
  mutate(nitrate2 = nitrate^2, phosphate2 = phosphate^2)

site_idx_tl <- match(river_tl_data$sandre_code, stations$sandre_code)
A_tl <- A[site_idx_tl, ]

tl_covariate_df <- data.frame(
  log_species_richness = river_tl_data$log_species_richness,
  elevation = river_tl_data$elevation,
  river_width = river_tl_data$river_width,
  nitrate = river_tl_data$nitrate,
  nitrate2 = river_tl_data$nitrate2,
  phosphate = river_tl_data$phosphate,
  phosphate2 = river_tl_data$phosphate2,
  barrier_upstream = river_tl_data$barrier_upstream,
  barrier_downstream = river_tl_data$barrier_downstream,
  dist_mouth_m = river_tl_data$dist_mouth_m,
  year = river_tl_data$year,
  district_id = river_tl_data$district_id
)
```

then define and fit the model,

```{r predation-depth-model}
#| cache: true

stk_ordinal <- inla.stack(
  data = list(y = river_tl_data$predation_depth_int),
  A = list(A_tl, 1),
  effects = list(spatial.field = spatial_index, tl_covariate_df),
  tag = "est"
)

fit_predation_depth <- inla(
  y ~ log_species_richness + elevation + river_width +
    nitrate + nitrate2 + phosphate + phosphate2 +
    barrier_upstream + barrier_downstream +
    dist_mouth_m +
    barrier_upstream:dist_mouth_m +
    barrier_downstream:dist_mouth_m +
    f(year, model = "iid") +
    f(district_id, model = "iid") +
    f(spatial.field, model = spde),
  family = "pom",
  data = inla.stack.data(stk_ordinal),
  control.predictor = list(A = inla.stack.A(stk_ordinal), compute = TRUE),
  control.compute = list(waic = TRUE, config = TRUE)
)
```

and plot its fixed effects.

```{r predation-depth-coefficient-plot}
#| fig-width: 9
#| fig-height: 6

plot_coefficients(
  fit_predation_depth, "Cumulative-logit coefficient (posterior mean, 95% CI)"
)
```

For nutrients, we observe a positive effect of nitrate which is consistent with
its impact on connectance. Phosphate has a humped-shape effect on trophic
length. We observe a negative of barrier downstream, consistent with the
trapping hypothesis discussed in the connectance model. We expect this effect to
decreases as we move upstream.

But before that, we have a look at the model parameters to ensure that the
category inferred make sense. First, the mean and standard deviation of $\theta$.
Second, the probability of belonging to each category $p$ (on average), which
should roughly correspond to what is observed in the data.

```{r predation-depth-thresholds}
thresholds <- fit_predation_depth$summary.hyperpar |>
  tibble::rownames_to_column("term") |>
  filter(grepl("POM", term)) |>
  select(term, mean, sd, `0.025quant`, `0.975quant`)
thresholds |> knitr::kable(digits = 3)

# implied category split at eta = 0 (an "average-covariate" site)
p_le1 <- plogis(thresholds$mean[1])
p_le2 <- plogis(thresholds$mean[2])
tibble(
  predation_depth = levels(river_tl_data$predation_depth),
  p = c(p_le1, p_le2 - p_le1, 1 - p_le2)
) |>
  knitr::kable(digits = 3)
```

$\theta_1$ is poorly constrained (95% CI on the probability scale: 0.6% to
41%) because only ~4% of rows fall in `no fish eaten` - $\theta_2$, between
the two well-populated categories, is tight (83.6% to 85.6%). The implied
split at $\eta_i = 0$ (6% / 79% / 15%) differs from the raw sample split
(4% / 42% / 54%): this is expected, not a bug - $\eta_i=0$ means every
covariate at its mean simultaneously, and because the logistic link is
non-linear and `log_species_richness` has a large, widely-spread effect,
averaging predictions *across* sites is not the same as predicting *at*
the average site.

Next, we have a look at the gaussian part of the model, conditioned on the
predation depth category.

### Conditional gaussian part

Conditional on the mode $D_i$, a Gaussian for the continuous variation
within a mode, with a separate intercept per category (no shared global
intercept, since the three category dummies already span that) and the
same covariates and random-effect structure:

$$
\text{trophic\_length}_i \mid D_i \sim \mathcal{N}(\mu_i, \sigma^2)
$$
$$
\mu_i = \alpha_{D_i} + \mathbf{x}_i^\top \boldsymbol\gamma +
\text{year}_i + \text{district}_i + \text{SPDE}_i
$$

We fit the model,

```{r trophic-length-model}
#| cache: true

tl_covariate_df$predation_depth <- river_tl_data$predation_depth

stk_tl <- inla.stack(
  data = list(y = river_tl_data$trophic_length),
  A = list(A_tl, 1),
  effects = list(spatial.field = spatial_index, tl_covariate_df),
  tag = "est"
)

fit_trophic_length <- inla(
  y ~ -1 + predation_depth + log_species_richness + elevation +
    river_width +
    nitrate + nitrate2 + phosphate + phosphate2 +
    barrier_upstream + barrier_downstream +
    dist_mouth_m +
    barrier_upstream:dist_mouth_m +
    barrier_downstream:dist_mouth_m +
    f(year, model = "iid") +
    f(district_id, model = "iid") +
    f(spatial.field, model = spde),
  family = "gaussian",
  data = inla.stack.data(stk_tl),
  control.predictor = list(A = inla.stack.A(stk_tl), compute = TRUE),
  control.compute = list(waic = TRUE, config = TRUE)
)
```

and plot its fixed effects.

```{r trophic-length-coefficient-plot}
#| fig-width: 9
#| fig-height: 6

plot_coefficients(
  fit_trophic_length, "Standardized coefficient (posterior mean, 95% CI)"
)
```

`plot_coefficients()` drops the three `predation_depth` category
intercepts (not in `term_order` - they're baseline levels, not effects
on the same scale as the covariates), so they're reported separately:

```{r trophic-length-category-intercepts}
fit_trophic_length$summary.fixed |>
  tibble::rownames_to_column("term") |>
  filter(grepl("^predation_depth", term)) |>
  select(term, mean, sd, `0.025quant`, `0.975quant`) |>
  knitr::kable(digits = 3)
```

*To discuss: the three category intercepts should closely match the
histogram modes from the checkpoint above, a good consistency check;
which covariates have a credible effect on trophic_length once the mode
itself is accounted for, versus the ordinal model's list above - several
covariates that look credible for shifting the mode may not have an
additional within-mode effect, or vice versa.*

### Average marginal effect

Neither model alone gives a covariate's *total* effect on
`trophic_length`: the ordinal model gives its effect on which mode a web
reaches, the Gaussian gives its effect within a mode. Combining them
requires simulation, not a coefficient product, because the mode-shift
component is a non-linear function of $x_i$ itself:

$$
E[\text{trophic\_length}_i \mid x_i] =
\sum_{k=1}^{3} P(D_i = k \mid x_i) \cdot \mu_{i,k}
$$

We compute the average marginal effect (AME) of each covariate $x_j$ as
the average, over all sites and over posterior draws sampled
independently from both models, of the numerical derivative of this
mixture expectation with respect to $x_j$ (central difference, small
step on the standardized scale). Treating the two models' posteriors as
independent is an approximation - it ignores any correlation between
their uncertainty, since they were not fit jointly.

```{r ame-setup}
main_vars <- c(
  "log_species_richness", "elevation", "river_width",
  "nitrate", "phosphate",
  "barrier_upstream", "barrier_downstream",
  "dist_mouth_m"
)
inter_pairs <- list(
  c("barrier_upstream", "dist_mouth_m"),
  c("barrier_downstream", "dist_mouth_m")
)
inter_names <- sapply(inter_pairs, \(p) paste0(p[1], ":", p[2]))
# quadratic terms are derived from a main_vars column (x^2), not
# independently perturbable - handled like interactions, not like main_vars
quad_vars <- c("nitrate", "phosphate")
quad_names <- paste0(quad_vars, "2")

X_tl <- as.matrix(river_tl_data[main_vars])
colnames(X_tl) <- main_vars
year_levels_tl <- sort(unique(river_tl_data$year))
year_idx_tl <- match(river_tl_data$year, year_levels_tl)
district_idx_tl <- river_tl_data$district_id

# everything x-independent from one ordinal posterior draw
extract_pd <- function(s) {
  list(
    intercept = get_val(s, "(Intercept)"),
    betas = setNames(sapply(main_vars, \(v) get_val(s, v)), main_vars),
    inter_betas = setNames(
      sapply(inter_pairs, \(p) get_val(s, paste0(p[1], ":", p[2]))), inter_names
    ),
    quad_betas = setNames(sapply(quad_names, \(v) get_val(s, v)), quad_names),
    re_offset = get_re_vec(s, "year", length(year_levels_tl))[year_idx_tl] +
      get_re_vec(s, "district_id", 4)[district_idx_tl] +
      as.numeric(A_tl %*% get_spatial_vec(s)),
    theta1 = s$hyperpar["theta1 for POM"],
    theta2 = s$hyperpar["theta2 for POM"]
  )
}
# everything x-independent from one conditional-gaussian posterior draw
extract_tl <- function(s) {
  list(
    alpha = c(
      get_val(s, "predation_depthno fish eaten"),
      get_val(s, "predation_deptheats a non-piscivorous fish"),
      get_val(s, "predation_deptheats a piscivorous fish")
    ),
    betas = setNames(sapply(main_vars, \(v) get_val(s, v)), main_vars),
    inter_betas = setNames(
      sapply(inter_pairs, \(p) get_val(s, paste0(p[1], ":", p[2]))), inter_names
    ),
    quad_betas = setNames(sapply(quad_names, \(v) get_val(s, v)), quad_names),
    re_offset = get_re_vec(s, "year", length(year_levels_tl))[year_idx_tl] +
      get_re_vec(s, "district_id", 4)[district_idx_tl] +
      as.numeric(A_tl %*% get_spatial_vec(s))
  )
}

linpred_main <- function(Xmat, betas, inter_betas, quad_betas) {
  out <- as.numeric(Xmat %*% betas[main_vars])
  for (k in seq_along(inter_pairs)) {
    p <- inter_pairs[[k]]
    out <- out + (Xmat[, p[1]] * Xmat[, p[2]]) * inter_betas[k]
  }
  for (k in seq_along(quad_vars)) {
    out <- out + (Xmat[, quad_vars[k]]^2) * quad_betas[k]
  }
  out
}

predicted_tl <- function(Xmat, pd, tl) {
  eta <- pd$intercept +
    linpred_main(Xmat, pd$betas, pd$inter_betas, pd$quad_betas) + pd$re_offset
  p_le1 <- plogis(pd$theta1 - eta)
  p_le2 <- plogis(pd$theta2 - eta)
  p1 <- p_le1
  p2 <- p_le2 - p_le1
  p3 <- 1 - p_le2
  shared <- linpred_main(Xmat, tl$betas, tl$inter_betas, tl$quad_betas) +
    tl$re_offset
  p1 * (tl$alpha[1] + shared) +
    p2 * (tl$alpha[2] + shared) +
    p3 * (tl$alpha[3] + shared)
}
```

```{r ame-compute}
#| cache: true

n_ame_draws <- 300
ame_eps <- 0.02
set.seed(1)
ps_pd <- inla.posterior.sample(n_ame_draws, fit_predation_depth)
ps_tl <- inla.posterior.sample(n_ame_draws, fit_trophic_length)

ame_draws <- matrix(
  NA, nrow = n_ame_draws, ncol = length(main_vars),
  dimnames = list(NULL, main_vars)
)
for (m in seq_len(n_ame_draws)) {
  pd <- extract_pd(ps_pd[[m]])
  tl <- extract_tl(ps_tl[[m]])
  for (j in main_vars) {
    X_plus <- X_tl; X_plus[, j] <- X_plus[, j] + ame_eps
    X_minus <- X_tl; X_minus[, j] <- X_minus[, j] - ame_eps
    pred_plus <- predicted_tl(X_plus, pd, tl)
    pred_minus <- predicted_tl(X_minus, pd, tl)
    ame_draws[m, j] <- mean((pred_plus - pred_minus) / (2 * ame_eps))
  }
}

ame_summary <- tibble(
  term = main_vars,
  mean = apply(ame_draws, 2, mean),
  `0.025quant` = apply(ame_draws, 2, quantile, 0.025),
  `0.975quant` = apply(ame_draws, 2, quantile, 0.975)
)
ame_summary |> knitr::kable(digits = 3)
```

```{r ame-plot}
#| fig-width: 9
#| fig-height: 6

ame_fit <- list(summary.fixed = data.frame(
  mean = ame_summary$mean,
  `0.025quant` = ame_summary$`0.025quant`,
  `0.975quant` = ame_summary$`0.975quant`,
  row.names = ame_summary$term,
  check.names = FALSE
))
plot_coefficients(
  ame_fit,
  "Average marginal effect on trophic_length (posterior mean, 95% CI)"
)
```

*To discuss: log_species_richness's total effect (0.19) is much larger
than its within-mode-only effect from the conditional Gaussian (0.08),
because it also strongly shifts the mode (ordinal coefficient 1.57) -
similarly nitrate and barrier_downstream are credible here despite
having no within-mode effect, driven entirely by mode-shifting; whether
this AME is the number worth reporting as "the" effect of a covariate on
trophic_length, versus keeping the two mechanisms separate for
interpretability.*

### Interaction AME setup

Reused later by the combined dam/barrier interaction figure at the end
of the notebook - no standalone result here.

```{r tl-interaction-compute}
#| cache: true

dist_mouth_levels <- quantile(river_tl_data$dist_mouth_m, probs = c(0.1, 0.5, 0.9))
level_labels <- c("low", "median", "high")

pd_list <- lapply(ps_pd, extract_pd)
tl_list <- lapply(ps_tl, extract_tl)

rows <- list()
for (pair_idx in seq_along(inter_pairs)) {
  var <- inter_pairs[[pair_idx]][1]
  for (lvl_idx in seq_along(dist_mouth_levels)) {
    lvl <- dist_mouth_levels[lvl_idx]
    draws <- numeric(n_ame_draws)
    for (m in seq_len(n_ame_draws)) {
      pd <- pd_list[[m]]; tl <- tl_list[[m]]
      X_plus <- X_tl; X_plus[, var] <- X_plus[, var] + ame_eps
      X_plus[, "dist_mouth_m"] <- lvl
      X_minus <- X_tl; X_minus[, var] <- X_minus[, var] - ame_eps
      X_minus[, "dist_mouth_m"] <- lvl
      pred_plus <- predicted_tl(X_plus, pd, tl)
      pred_minus <- predicted_tl(X_minus, pd, tl)
      draws[m] <- mean((pred_plus - pred_minus) / (2 * ame_eps))
    }
    rows[[length(rows) + 1]] <- tibble(
      var = var, dist_mouth_level = level_labels[lvl_idx],
      dist_mouth_value = as.numeric(lvl), mean = mean(draws),
      `0.025quant` = quantile(draws, 0.025),
      `0.975quant` = quantile(draws, 0.975)
    )
  }
}
interaction_ame <- bind_rows(rows) |>
  mutate(dist_mouth_level = factor(dist_mouth_level, levels = level_labels))
```

## Nutrient curvature effect

Both `nitrate`/`nitrate2` and `phosphate`/`phosphate2` mean a single
coefficient doesn't capture their effect - the same reasoning as the
dam/barrier interactions above, just a self-interaction (curvature)
instead of a two-way one, so there's no second variable to hold fixed.
Instead, for richness, connectance and trophic_length alike, we
discretize the dose-response curve into the AME of each nutrient
evaluated locally at its own 10th/50th/90th percentile - does the
*slope* of the nutrient's effect change across its own range.

```{r nutrient-curvature-setup}
own_curvature_ame <- function(v, X_base, predicted_fn, draw_list, eps = 0.02,
                               probs = c(0.1, 0.5, 0.9),
                               level_labels = c("low", "median", "high")) {
  levels_v <- quantile(X_base[, v], probs = probs)
  rows <- list()
  for (lvl_idx in seq_along(levels_v)) {
    lvl <- levels_v[lvl_idx]
    draws <- sapply(draw_list, \(obj) {
      X_plus <- X_base; X_plus[, v] <- lvl + eps
      X_minus <- X_base; X_minus[, v] <- lvl - eps
      mean((predicted_fn(X_plus, obj) - predicted_fn(X_minus, obj)) / (2 * eps))
    })
    rows[[lvl_idx]] <- tibble(
      var = v, level = level_labels[lvl_idx], value = as.numeric(lvl),
      mean = mean(draws),
      `0.025quant` = quantile(draws, 0.025),
      `0.975quant` = quantile(draws, 0.975)
    )
  }
  bind_rows(rows) |> mutate(level = factor(level, levels = level_labels))
}

rich_main_vars <- c(
  "elevation", "river_width", "nitrate", "phosphate",
  "barrier_upstream", "barrier_downstream",
  "dist_mouth_m"
)
rich_inter_pairs <- list(
  c("barrier_upstream", "dist_mouth_m"),
  c("barrier_downstream", "dist_mouth_m")
)
rich_inter_names <- sapply(rich_inter_pairs, \(p) paste0(p[1], ":", p[2]))
rich_quad_vars <- c("nitrate", "phosphate")
rich_quad_names <- paste0(rich_quad_vars, "2")

X_rich <- as.matrix(river_model_data[rich_main_vars])
colnames(X_rich) <- rich_main_vars
year_levels_rich <- sort(unique(river_model_data$year))
year_idx_rich <- match(river_model_data$year, year_levels_rich)
district_idx_rich <- river_model_data$district_id

extract_rich <- function(s) {
  list(
    intercept = get_val(s, "intercept"),
    betas = setNames(sapply(rich_main_vars, \(v) get_val(s, v)), rich_main_vars),
    inter_betas = setNames(
      sapply(rich_inter_pairs, \(p) get_val(s, paste0(p[1], ":", p[2]))), rich_inter_names
    ),
    quad_betas = setNames(sapply(rich_quad_names, \(v) get_val(s, v)), rich_quad_names),
    re_offset = get_re_vec(s, "year", length(year_levels_rich))[year_idx_rich] +
      get_re_vec(s, "district_id", 4)[district_idx_rich] +
      as.numeric(A_model %*% get_spatial_vec(s))
  )
}
linpred_rich <- function(Xmat, betas, inter_betas, quad_betas) {
  out <- as.numeric(Xmat %*% betas[rich_main_vars])
  for (k in seq_along(rich_inter_pairs)) {
    p <- rich_inter_pairs[[k]]
    out <- out + (Xmat[, p[1]] * Xmat[, p[2]]) * inter_betas[k]
  }
  for (k in seq_along(rich_quad_vars)) out <- out + (Xmat[, rich_quad_vars[k]]^2) * quad_betas[k]
  out
}
predicted_rich <- function(Xmat, rd) {
  rd$intercept + linpred_rich(Xmat, rd$betas, rd$inter_betas, rd$quad_betas) + rd$re_offset
}

n_rich_draws <- 300
set.seed(1)
ps_richness <- inla.posterior.sample(n_rich_draws, fit_richness)
rd_list <- lapply(ps_richness, extract_rich)
```

```{r nutrient-curvature-compute}
#| cache: true

nutrient_vars <- c("nitrate", "phosphate")

predicted_tl_paired <- function(Xmat, obj) predicted_tl(Xmat, obj$pd, obj$tl)
tl_draw_pairs <- Map(function(pd, tl) list(pd = pd, tl = tl), pd_list, tl_list)

curvature_for <- function(X_base, predicted_fn, draw_list, outcome_name) {
  bind_rows(lapply(
    nutrient_vars, own_curvature_ame,
    X_base = X_base, predicted_fn = predicted_fn, draw_list = draw_list
  )) |>
    mutate(outcome = outcome_name)
}

nutrient_curvature <- bind_rows(
  curvature_for(X_rich, predicted_rich, rd_list, "richness"),
  curvature_for(X_conn, predicted_conn, cd_list, "connectance"),
  curvature_for(X_tl, predicted_tl_paired, tl_draw_pairs, "trophic_length")
) |>
  mutate(outcome = factor(
    outcome, levels = c("richness", "connectance", "trophic_length")
  ))
```

```{r nutrient-curvature-plot}
#| fig-width: 9
#| fig-height: 3.5

nutrient_colors <- c(nitrate = "steelblue", phosphate = "darkorange")

ggplot(nutrient_curvature, aes(level, mean, color = var)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_pointrange(
    aes(ymin = `0.025quant`, ymax = `0.975quant`),
    position = position_dodge(width = 0.3)
  ) +
  scale_color_manual(values = nutrient_colors) +
  facet_wrap(~outcome, scales = "free_y") +
  labs(x = "nutrient level (own percentile)", y = "AME (local slope)", color = NULL)
```

## Dam interaction effect

Same simple-slopes approach as the nutrient curvature figure, but for
the two barrier variables against `dist_mouth_m`: the AME of each,
with `dist_mouth_m` held fixed at its 10th/50th/90th percentile instead
of averaged over sites. Richness didn't have this built yet, so we
compute it here too, reusing the same setup as the nutrient curvature
section above.

```{r dam-interaction-richness}
#| cache: true

dist_mouth_levels_rich <- quantile(river_model_data$dist_mouth_m, probs = c(0.1, 0.5, 0.9))
level_labels <- c("low", "median", "high")

rows <- list()
for (pair_idx in seq_along(rich_inter_pairs)) {
  var <- rich_inter_pairs[[pair_idx]][1]
  for (lvl_idx in seq_along(dist_mouth_levels_rich)) {
    lvl <- dist_mouth_levels_rich[lvl_idx]
    draws <- sapply(rd_list, \(rd) {
      X_plus <- X_rich; X_plus[, var] <- X_plus[, var] + 0.02
      X_plus[, "dist_mouth_m"] <- lvl
      X_minus <- X_rich; X_minus[, var] <- X_minus[, var] - 0.02
      X_minus[, "dist_mouth_m"] <- lvl
      mean((predicted_rich(X_plus, rd) - predicted_rich(X_minus, rd)) / (2 * 0.02))
    })
    rows[[length(rows) + 1]] <- tibble(
      var = var, level = level_labels[lvl_idx],
      mean = mean(draws),
      `0.025quant` = quantile(draws, 0.025),
      `0.975quant` = quantile(draws, 0.975)
    )
  }
}
dam_interaction_rich <- bind_rows(rows) |>
  mutate(level = factor(level, levels = level_labels))
```

```{r dam-interaction-combine}
dam_interaction_all <- bind_rows(
  dam_interaction_rich |> mutate(outcome = "richness"),
  interaction_ame_conn |>
    rename(level = dist_mouth_level) |>
    select(var, level, mean, `0.025quant`, `0.975quant`) |>
    mutate(outcome = "connectance"),
  interaction_ame |>
    rename(level = dist_mouth_level) |>
    select(var, level, mean, `0.025quant`, `0.975quant`) |>
    mutate(outcome = "trophic_length")
) |>
  mutate(outcome = factor(outcome, levels = c("richness", "connectance", "trophic_length")))
```

```{r dam-interaction-plot}
#| fig-width: 10
#| fig-height: 4

dam_colors <- c(
  barrier_upstream = "#1b9e77",
  barrier_downstream = "#d95f02"
)

ggplot(dam_interaction_all, aes(level, mean, color = var)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_pointrange(
    aes(ymin = `0.025quant`, ymax = `0.975quant`),
    position = position_dodge(width = 0.4)
  ) +
  scale_color_manual(values = dam_colors) +
  facet_wrap(~outcome, scales = "free_y") +
  labs(x = "distance to river mouth", y = "AME (local slope)", color = NULL)
```

We observe that downstream barrier effect has a strong negative impact on
connectance, and this effect vanishes at high distances. By constrast, the
barrier upstream has a strong postive impact near the river mouth and this
effect vanishes upstream. These patterns are constitent with the hypothesis that
first dams from the river mouth trap migratory generalist predator, whose
presence increases the connectance locally.

Both credible interactions found for trophic_length point the same
direction, and match the same "trapping" hypothesis: dams filter out
large, upstream-migrating predators, so they accumulate below the first
dam they meet coming from the sea. Taller dams downstream of a site
lower its trophic length (harder for predators from the sea to reach
it), taller dams upstream raise it (predators blocked from going
further accumulate in the reach below, including the site) - and both
effects fade with distance to the mouth, consistent with dams mattering
most where they sit closest to the sea-derived predator pool.

Still a hypothesis, not a conclusion: the data can't rule out other
differences correlated with tall dams (e.g. altered habitat immediately
around them) as the actual driver.
