---
title: "River model: impact of nutrients and dams on food web structure"
bibliography: ../references.bib
format:
  html:
    theme: cosmo
    code-fold: true
    toc: true
    embed-resources: true
execute:
    message: false
    warning: false
---

## Todo

- [x] Join AMOBIO stressors + distance to mouth to river food web structure
- [x] Checkpoint: coverage and missingness of covariates
- [x] Checkpoint: spatial/temporal coverage and response distributions
- [ ] Decide how to handle nutrient / dam-distance-downstream missingness
- [ ] Build SPDE mesh over river station locations (Lambert-93)
- [ ] Fit richness ~ nutrients + dams * distance to mouth + year + SPDE (path *a*)
- [ ] Checkpoint: richness model diagnostics and predictions
- [ ] Fit connectance / trophic length / diet overlap ~ richness + nutrients +
  dams * distance to mouth + year + SPDE (paths *b* and *c'*)
- [ ] Checkpoint each model
- [ ] Direct/indirect effect decomposition (product of coefficients, via
  posterior samples)
- [ ] Swap Euclidean SPDE for a `MetricGraph` hydrographic-distance spatial
  term [@Bolin-Simas-Wallin-2023]

## Setup

```{r}
library(renv)
renv::load()
library(targets)
library(here)
tar_config_set(
  store = here::here("_targets")
)
library(ggplot2)
library(dplyr)
library(sf)
devtools::load_all()
nice_theme()
```

## 1. Data preparation

We join two existing pipeline targets:

- `amobio_aspe_data`: AMOBIO environmental stressors (nutrients, dams) already
  matched to river food web structure metrics (`connectance`, `trophic_length`,
  `diet_overlap`, `log_species_richness`) via `sandre_code`/`date`.
- `inland_distance_to_mouth`: hydrographic distance to the nearest river mouth,
  per operation.

Both carry `operation_id`, drawn from the same ASPE ID space, so they join
directly.

```{r}
amobio <- tar_read(amobio_aspe_data) |> st_drop_geometry()
dist_mouth <- tar_read(inland_distance_to_mouth) |>
  select(operation_id, district, dist_mouth_m)

river_data <- amobio |>
  inner_join(dist_mouth, by = "operation_id") |>
  select(
    operation_id, year, survey, district, longitude, latitude,
    # Responses.
    log_species_richness, connectance, trophic_length, diet_overlap,
    # Nutrients.
    ammonium, nitrite, nitrate, phosphate,
    # Dams.
    dam_distance_upstream, dam_distance_downstream,
    barrier_upstream, barrier_downstream,
    # Position along the continuum.
    dist_mouth_m
  )

nrow(river_data)
```

We go from 8372 river operations with AMOBIO data, and 19045 with a computed
distance to mouth, down to `r nrow(river_data)` once both are required
simultaneously.

## 2. Checkpoint: missingness

```{r}
vars <- c(
  "ammonium", "nitrite", "nitrate", "phosphate",
  "dam_distance_upstream", "dam_distance_downstream",
  "barrier_upstream", "barrier_downstream", "dist_mouth_m",
  "log_species_richness", "connectance", "trophic_length", "diet_overlap"
)

tibble(
  variable = vars,
  n_na = sapply(river_data[vars], \(x) sum(is.na(x))),
  pct_na = round(100 * n_na / nrow(river_data), 1)
) |>
  arrange(desc(n_na))
```

Two things stand out:

- **Nutrients are missing together, not independently**: `r sum(rowSums(is.na(river_data[c("ammonium", "nitrite", "nitrate", "phosphate")])) == 4)`
  of the ~1930 rows missing any nutrient are missing *all four* at once. This
  points to whole sampling dates lacking a water-quality sample, rather than
  per-parameter gaps - so there's little to gain from imputing per nutrient
  independently.
- **`dam_distance_downstream` is missing far more often (~40%) than
  `dam_distance_upstream` (~6%)**, while the cumulative height indices
  (`barrier_upstream`/`barrier_downstream`) are always present. This asymmetry
  is worth understanding before we decide how to handle it - it may be
  structural (no downstream dam within AMOBIO's search radius, which would
  make the missingness itself informative) rather than a data gap. We haven't
  resolved this yet (see Todo).

Requiring complete cases on every covariate at once keeps only
`r sum(complete.cases(river_data[vars]))` of `r nrow(river_data)` operations
(`r round(100 * sum(complete.cases(river_data[vars])) / nrow(river_data))`%).
Since each of the four response models won't necessarily use every covariate
at once (and since missingness clusters so unevenly across variables), we'll
likely want to handle this per-model rather than with one global complete-case
filter - to be decided once we see how much it affects power for each model.

## 3. Checkpoint: spatial and temporal coverage

```{r}
#| fig-width: 9
#| fig-height: 8

basin <- tar_read(basin)
france <- rnaturalearth::ne_countries(
  geounit = "france", type = "map_units", scale = "medium", returnclass = "sf"
)

ggplot() +
  geom_sf(data = france, fill = NA, color = "grey80", linewidth = 0.25) +
  geom_sf(
    data = basin |> filter(district %in% unique(river_data$district)),
    fill = "grey97", color = "grey40", linewidth = 0.4
  ) +
  geom_point(
    data = river_data, aes(longitude, latitude, color = district),
    size = 0.6, alpha = 0.7
  ) +
  coord_sf(xlim = c(-5.3, 8.3), ylim = c(41.2, 51.2), expand = FALSE) +
  labs(x = "Longitude", y = "Latitude", color = "Bassin versant") +
  theme(panel.grid = element_blank())
```

```{r}
#| fig-width: 7
#| fig-height: 3.5

ggplot(river_data, aes(year)) +
  geom_bar(fill = "steelblue") +
  labs(x = NULL, y = "Nombre d'opérations")
```

Coverage is very uneven across the four basins (Loire and Seine dominate;
Escaut-Somme is thin), and 2020 has a visible spike. Both are things to keep
in mind when we get to the spatial/temporal random effects - a `year` random
effect should absorb the 2020 spike, but the basin imbalance means the SPDE
field will be much better informed in some catchments than others.

## 4. Checkpoint: response distributions

```{r}
#| fig-width: 9
#| fig-height: 6

river_data |>
  select(log_species_richness, connectance, trophic_length, diet_overlap) |>
  tidyr::pivot_longer(everything(), names_to = "metric", values_to = "value") |>
  ggplot(aes(value)) +
  geom_histogram(bins = 40, fill = "steelblue") +
  facet_wrap(~metric, scales = "free") +
  labs(x = NULL, y = NULL)
```

`connectance` and `diet_overlap` both stay strictly within (0, 1) (no exact 0
or 1 values), which is what a Beta likelihood needs, and both look reasonably
unimodal. `log_species_richness` is right-skewed with a shoulder but not
obviously multimodal - plausible for a Gaussian-on-log approximation, though
worth watching in residuals.

`trophic_length` is not what we expected: it's clearly **bimodal**, with a
narrow spike around 3.9-4.0 separated from a second, wider mode around
4.5-5.7. A single Gaussian isn't going to fit this well as-is. This may be
related to an earlier finding (see the metaweb investigation in `D9`): some
species classified as piscivorous never actually realize a fish-eating edge
in this dataset, because their size-class midpoints don't cross into the diet
table's "adult"/piscivorous stage. If so, this could be the same mechanism
showing up here: sites where the food chain tops out on a species that
genuinely eats fish (higher trophic length) versus sites where it tops out on
a species classified as piscivorous but that never actually predates on fish
in practice (trophic length capped around 4). Worth confirming directly (e.g.
splitting the histogram by whether the local web's top species has a realized
fish-eating edge) before we commit to a Gaussian likelihood for this response
- a mixture, or a covariate that captures this split, might be needed
instead.
