---
title: "River model: impact of nutrients and dams on food web structure"
bibliography: ../references.bib
format:
  html:
    theme: cosmo
    code-fold: true
    toc: true
    embed-resources: true
execute:
    message: false
    warning: false
---

## Todo

- [x] Join AMOBIO stressors + distance to mouth to river food web structure
- [x] Checkpoint: coverage and missingness of covariates
- [x] Checkpoint: spatial/temporal coverage and response distributions
- [x] Decide how to handle nutrient / dam-distance-downstream missingness
- [ ] Build SPDE mesh over river station locations (Lambert-93)
- [ ] Fit richness ~ nutrients + dams * distance to mouth + year + SPDE
  (path *a*)
- [ ] Checkpoint: richness model diagnostics and predictions
- [ ] Fit connectance / trophic length / diet overlap ~ richness + nutrients +
  dams * distance to mouth + year + SPDE (paths *b* and *c'*)
- [ ] Checkpoint each model
- [ ] Direct/indirect effect decomposition (product of coefficients, via
  posterior samples)
- [ ] Swap Euclidean SPDE for a `MetricGraph` hydrographic-distance spatial
  term [@Bolin-Simas-Wallin-2023]

```{r setup}
library(renv)
renv::load()
library(targets)
library(here)
tar_config_set(
  store = here::here("_targets")
)
library(ggplot2)
library(dplyr)
library(sf)
devtools::load_all()
nice_theme()
```

## Data preparation

We join two dataframes:

- AMOBIO environmental stressors (nutrients and dams), already matched with
food web structure (`amobio_aspe_data`);

- and hydrographic distance to river mouth (`inland_distance_to_mouth`)

```{r data-overview}
amobio <- tar_read(amobio_aspe_data) |> st_drop_geometry()
head(amobio)

dist_mouth <- tar_read(inland_distance_to_mouth) |>
  select(operation_id, district, dist_mouth_m)
head(dist_mouth)

river_width <- tar_read(river_width)
head(river_width)
```

`amobio` and `dist_mouth` both carry `operation_id`, drawn from the same
ASPE ID space, so they join directly. `river_width` is station-level
(Naiades hydromorphology survey, averaged across all available visit
dates per station), so it joins on `sandre_code`.

```{r prepare-data}
river_data <- amobio |>
  inner_join(dist_mouth, by = "operation_id") |>
  left_join(river_width, by = "sandre_code") |>
  select(
    operation_id, sandre_code, year, survey, district, longitude, latitude,
    log_species_richness, connectance, trophic_length, diet_overlap,
    elevation, river_width,
    ammonium, nitrite, nitrate, phosphate,
    dam_distance_upstream, dam_distance_downstream,
    barrier_upstream, barrier_downstream,
    dist_mouth_m
  )

head(river_data)
```

We go from `r nrow(amobio)` river operations with AMOBIO data, and
`r nrow(dist_mouth)` with a computed distance to mouth, down to
`r nrow(river_data)` once both are required simultaneously. Most droppped
rows are because of the filtering on hydrographic basin kept, and a few
stations that are not present in AMOBIO dataset.

## Data overview

### Missing values

```{r missing-values}
vars <- c(
  "ammonium", "nitrite", "nitrate", "phosphate",
  "dam_distance_upstream", "dam_distance_downstream",
  "barrier_upstream", "barrier_downstream", "dist_mouth_m",
  "elevation", "river_width",
  "log_species_richness", "connectance", "trophic_length", "diet_overlap"
)

nutrient_na <- river_data |>
  select(ammonium, nitrite, nitrate, phosphate) |>
  mutate(
    one_is_na = if_any(c(ammonium, nitrite, nitrate, phosphate), is.na),
    all_are_na = if_all(c(ammonium, nitrite, nitrate, phosphate), is.na)
  )

n_one_na <- sum(nutrient_na$one_is_na)
n_all_na <- sum(nutrient_na$all_are_na)
n_mismatch <- sum(nutrient_na$one_is_na & !nutrient_na$all_are_na)

tibble(
  variable = vars,
  n_na = sapply(river_data[vars], \(x) sum(is.na(x))),
  pct_na = round(100 * n_na / nrow(river_data), 1)
) |>
  arrange(desc(n_na))
```

Two things stand out

- nutrients are missing together, not independently: `r n_all_na` of the
  `r n_one_na` rows missing any nutrient are missing *all four* at once, and
  only `r n_mismatch` rows show a partial mismatch;

- dam upstream distance is missing far more of than the downstream distance
  (`dam_distance_upstream`/`dam_distance_downstream` were swapped in
  `rename_amobio()` relative to AMOBIO's own scale codes - fixed in
  `R/amobio.R`, pipeline rebuilt).

Coverage for `dam_distance_upstream` is too low, so we drop it from the
model. For the other covariates we'll use complete cases per model.

### Spatial coverage

```{r spatial-coverage}
#| fig-width: 9
#| fig-height: 8

basin <- tar_read(basin)
france <- rnaturalearth::ne_countries(
  geounit = "france", type = "map_units", scale = "medium", returnclass = "sf"
)

ggplot() +
  geom_sf(data = france, fill = NA, color = "grey80", linewidth = 0.25) +
  geom_sf(
    data = basin |> filter(district %in% unique(river_data$district)),
    fill = "grey97", color = "grey40", linewidth = 0.4
  ) +
  geom_point(
    data = river_data, aes(longitude, latitude, color = district),
    size = 0.6, alpha = 0.7
  ) +
  coord_sf(xlim = c(-5.3, 8.3), ylim = c(41.2, 51.2), expand = FALSE) +
  labs(x = "Longitude", y = "Latitude", color = "Catchment") +
  theme(panel.grid = element_blank())
```

Good coverage in all four basin although they vary in sizes.

### Temporal coverage

```{r temporal-coverage}
#| fig-width: 7
#| fig-height: 3.5

ggplot(river_data, aes(year)) +
  geom_bar(fill = "steelblue") +
  labs(x = NULL, y = "Number of operations")
```

Mostly even coverage, but for 2018 and 2021 which have far less fishing
operations.

## Response distributions

Let's see the distributions of our response variable so we can choose the
relevant reponse distributions.

```{r response-distributions}
#| fig-width: 9
#| fig-height: 6

river_data |>
  select(log_species_richness, connectance, trophic_length, diet_overlap) |>
  tidyr::pivot_longer(everything(), names_to = "metric", values_to = "value") |>
  ggplot(aes(value)) +
  geom_histogram(bins = 40, fill = "steelblue") +
  facet_wrap(~metric, scales = "free") +
  labs(x = NULL, y = NULL)
```

Connectance and diet overlap are bounded between 0 and 1; they will be properly
modeled by a beta distribution. Log species richness might be handle well by a
gaussian distribution. The real thinking should be done for the trophic level
which is clearly multimodal. Let's try to investigate why.

We split each local food web by the predation depth of its top species: does
it eat no fish at all, a fish that itself only eats invertebrates, or a fish
that is itself piscivorous.

```{r trophic-length-modes}
#| fig-width: 9
#| fig-height: 4

river_foodweb <- tar_read(river_foodweb_structure) |>
  filter(operation_id %in% river_data$operation_id)
resource_list <- tar_read(resource_list)

classify_predation_depth <- function(web, resource_list) {
  nodes <- data.frame(node = colnames(web))
  links <- web |>
    as.data.frame() |>
    mutate(resource = rownames(web)) |>
    tidyr::pivot_longer(-resource, names_to = "consumer") |>
    filter(value == 1) |>
    select(-value)
  community <- cheddar::Community(
    nodes = nodes, properties = list(title = "Community"), trophic.links = links
  )
  tl <- cheddar::TrophicLevels(community) |>
    as.data.frame() |>
    tibble::rownames_to_column("node") |>
    rename(TL = ChainAveragedTL)
  top_nodes <- tl$node[!is.na(tl$TL) & tl$TL == max(tl$TL, na.rm = TRUE)]

  fish_prey_of <- function(n) {
    prey <- rownames(web)[web[, n] == 1]
    prey[!(prey %in% resource_list)]
  }
  top_fish_prey <- unique(unlist(lapply(top_nodes, fish_prey_of)))

  if (length(top_fish_prey) == 0) {
    "no fish eaten"
  } else if (any(sapply(top_fish_prey, \(p) length(fish_prey_of(p)) > 0))) {
    "eats a piscivorous fish"
  } else {
    "eats a non-piscivorous fish"
  }
}

river_foodweb <- river_foodweb |>
  mutate(
    predation_depth = sapply(
      foodweb, classify_predation_depth, resource_list = resource_list
    ),
    predation_depth = factor(
      predation_depth,
      levels = c(
        "no fish eaten", "eats a non-piscivorous fish",
        "eats a piscivorous fish"
      )
    )
  )

tl_range <- range(river_foodweb$trophic_length)
normal_overlay <- river_foodweb |>
  group_by(predation_depth) |>
  summarise(
    mean_tl = mean(trophic_length), sd_tl = sd(trophic_length), .groups = "drop"
  ) |>
  rowwise() |>
  reframe(
    predation_depth = predation_depth,
    x = seq(tl_range[1], tl_range[2], length.out = 200),
    y = dnorm(x, mean_tl, sd_tl)
  )

ggplot(river_foodweb, aes(trophic_length)) +
  geom_histogram(
    aes(y = after_stat(density)), bins = 30, fill = "steelblue", alpha = 0.7
  ) +
  geom_line(
    data = normal_overlay, aes(x, y), color = "firebrick", linewidth = 0.8
  ) +
  facet_wrap(~predation_depth, scales = "free") +
  labs(x = "Trophic length", y = "Density")
```

We see that the 'predation depth' cleanly separates our three modes of the
trophic level distribution. So we will go for an ordinal model the classify
these modes and model each of the mode individually as a gaussian.

## Building the mesh

We build a mesh that covers all fishing stations.

```{r river-mesh}
library(INLA)
library(fmesher)

stations <- river_data |>
  distinct(sandre_code, longitude, latitude) |>
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
  st_transform(2154) # Lambert-93, coordinates in meters.

coords <- st_coordinates(stations)

mesh <- fm_mesh_2d_inla(
  loc = coords,
  max.edge = c(15000, 60000),
  offset = c(15000, 100000),
  cutoff = 5000
)
plot(mesh, edge.color = rgb(0.6, 0.6, 0.6), lwd = 0.5)
points(coords, col = "steelblue", pch = 16, cex = 0.4)
```

```{r spde-setup}
spde <- inla.spde2.pcmatern(
  mesh = mesh,
  prior.range = c(50000, 0.5),
  prior.sigma = c(1, 0.01)
)

A <- inla.spde.make.A(mesh = mesh, loc = coords)
spatial_index <- inla.spde.make.index("spatial.field", n.spde = spde$n.spde)

site_idx <- match(river_data$sandre_code, stations$sandre_code)
A_expanded <- A[site_idx, ]
```

## Richness model

Path *a*: `log_species_richness ~ nutrients + dams * distance to mouth +
elevation + river_width + year + district + SPDE`, restricted to complete
cases on the covariates used here (`dam_distance_upstream` excluded, see
above). `ammonium` is dropped (collinear with `nitrite`, r=0.79 on log
scale, see below) and we keep `nitrite` only. Nitrite, nitrate, phosphate,
`dam_distance_downstream` and `river_width` are log-transformed;
`barrier_upstream`/`barrier_downstream` are log1p-transformed (both have
exact zeros). We also add an IID random effect for `district` (hydrographic
basin) alongside the SPDE field.

`elevation` and `river_width` were added after noticing `barrier_upstream`
had a counterintuitive positive effect on richness: `elevation` is a real
confound for the whole dam/distance story (see checkpoint below), and
`river_width` (Naiades hydromorphology data, joined on `sandre_code`)
turned out to fully explain `barrier_upstream`'s effect - larger rivers
carry both more historic barrier infrastructure and higher richness.
`river_width` has substantial missingness (~35% of stations were never
surveyed for hydromorphology at all), which is the dominant cost of adding
it here.

```{r richness-data}
richness_vars <- c(
  "log_species_richness", "elevation", "river_width",
  "nitrite", "nitrate", "phosphate",
  "dam_distance_downstream", "barrier_upstream", "barrier_downstream",
  "dist_mouth_m", "year", "sandre_code", "district"
)
river_model_data_raw <- river_data |>
  filter(complete.cases(river_data[richness_vars]))
nrow(river_model_data_raw)
```

### Collinearity check

Before fitting, we check pairwise correlation between covariates, on the
same complete-case rows used for fitting, log-transforming the skewed ones
first.

```{r collinearity}
#| fig-width: 12
#| fig-height: 10

river_model_data_raw <- river_model_data_raw |>
  mutate(
    log_nitrite = log(nitrite),
    log_nitrate = log(nitrate),
    log_phosphate = log(phosphate),
    log_dam_distance_downstream = log(dam_distance_downstream),
    log_barrier_upstream = log1p(barrier_upstream),
    log_barrier_downstream = log1p(barrier_downstream),
    log_river_width = log(river_width)
  )

river_model_data_raw |>
  select(
    log_nitrite, log_nitrate, log_phosphate,
    log_dam_distance_downstream, log_barrier_upstream,
    log_barrier_downstream, dist_mouth_m, elevation, log_river_width
  ) |>
  GGally::ggpairs()
```

*To discuss: whether dropping ammonium and keeping nitrite resolved the
worst collinearity, whether log1p on the barrier variables changed their
correlation structure the way plain log did for nutrients and
dam_distance_downstream, and whether any remaining pair is still high
enough to worry about before fitting.*

### Model fit

```{r richness-model}
river_model_data <- river_model_data_raw |>
  mutate(
    nitrite = log(nitrite),
    nitrate = log(nitrate),
    phosphate = log(phosphate),
    dam_distance_downstream = log(dam_distance_downstream),
    barrier_upstream = log1p(barrier_upstream),
    barrier_downstream = log1p(barrier_downstream),
    river_width = log(river_width),
    district_id = as.integer(factor(district))
  ) |>
  mutate(across(
    c(
      nitrite, nitrate, phosphate, dam_distance_downstream,
      barrier_upstream, barrier_downstream, dist_mouth_m,
      elevation, river_width
    ),
    \(x) as.numeric(scale(x))
  ))

site_idx_model <- match(river_model_data$sandre_code, stations$sandre_code)
A_model <- A[site_idx_model, ]

richness_stack <- inla.stack(
  data = list(y = river_model_data$log_species_richness),
  A = list(A_model, 1),
  effects = list(
    spatial.field = spatial_index,
    data.frame(
      intercept = 1,
      elevation = river_model_data$elevation,
      river_width = river_model_data$river_width,
      nitrite = river_model_data$nitrite,
      nitrate = river_model_data$nitrate,
      phosphate = river_model_data$phosphate,
      dam_distance_downstream = river_model_data$dam_distance_downstream,
      barrier_upstream = river_model_data$barrier_upstream,
      barrier_downstream = river_model_data$barrier_downstream,
      dist_mouth_m = river_model_data$dist_mouth_m,
      year = river_model_data$year,
      district_id = river_model_data$district_id
    )
  ),
  tag = "est"
)

fit_richness <- inla(
  y ~ -1 + intercept +
    elevation + river_width +
    nitrite + nitrate + phosphate +
    dam_distance_downstream + barrier_upstream + barrier_downstream +
    dist_mouth_m +
    dam_distance_downstream:dist_mouth_m +
    barrier_upstream:dist_mouth_m +
    barrier_downstream:dist_mouth_m +
    f(year, model = "iid") +
    f(district_id, model = "iid") +
    f(spatial.field, model = spde),
  family = "gaussian",
  data = inla.stack.data(richness_stack),
  control.predictor = list(A = inla.stack.A(richness_stack), compute = TRUE),
  control.compute = list(waic = TRUE, config = TRUE)
)

summary(fit_richness)
```

*To discuss: which fixed effects have credible intervals excluding zero
now that elevation and river_width are in the model, whether
barrier_upstream's effect has vanished the way it did in the smaller
robustness checks, how much the district IID effect's variance compares to
the SPDE field's, whether the spatial range and dist_mouth_m's sign are
stable now versus the elevation-only checkpoint, and whether the ~35% drop
in coverage from river_width's missingness is an acceptable cost for this
model.*

## Coefficient plot

All covariates are standardized, so posterior means/CIs are directly
comparable across terms.

```{r coefficient-plot}
#| fig-width: 9
#| fig-height: 6

term_order <- c(
  "elevation", "river_width",
  "nitrite", "nitrate", "phosphate",
  "dam_distance_downstream", "barrier_upstream", "barrier_downstream",
  "dist_mouth_m",
  "dam_distance_downstream:dist_mouth_m", "barrier_upstream:dist_mouth_m",
  "barrier_downstream:dist_mouth_m"
)
term_type <- c(
  elevation = "physical", river_width = "physical",
  nitrite = "nutrient", nitrate = "nutrient", phosphate = "nutrient",
  dam_distance_downstream = "connectivity", barrier_upstream = "connectivity",
  barrier_downstream = "connectivity", dist_mouth_m = "distance",
  `dam_distance_downstream:dist_mouth_m` = "interaction",
  `barrier_upstream:dist_mouth_m` = "interaction",
  `barrier_downstream:dist_mouth_m` = "interaction"
)

coef_df <- fit_richness$summary.fixed |>
  tibble::rownames_to_column("term") |>
  filter(term != "intercept") |>
  mutate(
    term = factor(term, levels = rev(term_order)),
    type = term_type[as.character(term)]
  )

ggplot(coef_df, aes(x = mean, y = term, color = type)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_pointrange(aes(xmin = `0.025quant`, xmax = `0.975quant`)) +
  labs(
    x = "Standardized coefficient (posterior mean, 95% CI)", y = NULL,
    color = "Term type"
  )
```

## Visualizing interactions

A coefficient bar for an interaction term only tells us it's non-zero, not
what it means for the underlying relationship. Instead, for each dam
variable we plot its *marginal effect* on richness (main effect +
interaction x distance) as a function of distance to the river mouth,
across the observed range, with a credible band. Since the main effect and
interaction coefficients are correlated in the posterior, we use posterior
samples (not the two terms' separate CIs, which would ignore that
covariance) to build the band correctly.

```{r interaction-plot}
#| fig-width: 10
#| fig-height: 4

set.seed(1)
n_samples <- 1000
post_samples <- inla.posterior.sample(n_samples, fit_richness)

get_coef <- function(name) {
  sapply(post_samples, function(s) {
    s$latent[rownames(s$latent) == paste0(name, ":1"), 1]
  })
}

dam_vars <- c(
  "dam_distance_downstream", "barrier_upstream", "barrier_downstream"
)
z_grid <- seq(-2, 3, length.out = 50)

marginal_df <- purrr::map_dfr(dam_vars, function(v) {
  main <- get_coef(v)
  inter <- get_coef(paste0(v, ":dist_mouth_m"))
  eff <- sapply(z_grid, function(z) main + inter * z)
  tibble::tibble(
    variable = v,
    z = z_grid,
    mean = apply(eff, 2, mean),
    lower = apply(eff, 2, quantile, 0.025),
    upper = apply(eff, 2, quantile, 0.975)
  )
})

ggplot(marginal_df, aes(x = z, y = mean)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_ribbon(
    aes(ymin = lower, ymax = upper), alpha = 0.2, fill = "steelblue"
  ) +
  geom_line(color = "steelblue") +
  facet_wrap(~variable) +
  labs(
    x = "Distance to river mouth (standardized)",
    y = "Estimated effect on log richness"
  )
```

*To discuss: barrier_downstream's effect crossing from ~0 near the mouth to
clearly negative far upstream, barrier_upstream's effect crossing from ~0
to clearly positive far upstream, and dam_distance_downstream staying flat
near zero throughout (consistent with its non-credible interaction) -
whether this is worth reporting as "connectivity effects only bite far
from the sea" or needs more caution given the earlier confound flags on
barrier_upstream.*
