---
title: "Community structure along the river continuum"
bibliography: ../references.bib
format:
  html:
    theme: cosmo
    code-fold: true
    toc: true
    embed-resources: true
execute:
    message: false
    warning: false
---

The goal of this notebook is to describe how fish communities change from
upstream to downstream along the river continuum in diversity and food web
structure.

## Setup

```{r}
library(renv)
renv::load()
library(targets)
library(here)
tar_config_set(
  store = here::here("_targets")
)
library(ggplot2)
library(dplyr)
devtools::load_all()
nice_theme()
```

## Continuum variables: the RHT network

The theoretical hydrographic network [RHT, @Pella-etal-2012] covers mainland
France, one row per reach, with topographic, hydrologic and climatic
attributes. Each river (ASPE) operation is matched to its nearest reach
(`inland_operations_rht`). Estuarine survey operations (Pomet, Nurse) inside
the river basins are left out: they belong to the estuary-sea side of the
continuum.

```{r}
rht <- tar_read(rht)
rht_operation <- tar_read(inland_operations_rht)
basin <- tar_read(basin)
district_kept <- tar_read(params)$district_kept
```

```{r}
#| fig-width: 10
#| fig-height: 10

basin_kept <- basin |>
  filter(district %in% district_kept) |>
  sf::st_transform(sf::st_crs(rht))
rht_kept <- rht |>
  sf::st_filter(basin_kept)

ggplot() +
  geom_sf(data = basin_kept, fill = "grey97", color = "grey40") +
  geom_sf(
    data = rht_kept |> arrange(strahler),
    aes(color = strahler, linewidth = strahler)
  ) +
  scale_color_viridis_c(name = "Strahler order", direction = -1) +
  scale_linewidth(range = c(0.05, 1), guide = "none") +
  theme_void()
```

```{r}
cat(sprintf(
  "%d reaches in total, %d in the kept watersheds\n",
  nrow(rht), nrow(rht_kept)
))
```

Variables of the RHT, described from the user guide shipped with the data
(`rht_user_2020.pdf`).

```{r}
rht_description <- c(
  id_drain = "Reach identifier (numbering follows the large basins)",
  tjan_moy = "Mean January air temperature, 1960-1990 (°C)",
  tjuil_moy = "Mean July air temperature, 1960-1990 (°C)",
  noncalc = "1 if the models cannot be computed for the reach",
  region_csp = "CSP region abbreviation",
  d_source = "Distance to source (km)",
  altitude = "Mean reach altitude (m)",
  sfbvu = "Area of the reach's own (unit) catchment (km²)",
  long_ = "Reach length (m)",
  strahler = "Strahler order",
  surf_bv = "Upstream catchment area (km²)",
  pente = "Reach slope (‰)",
  minqmx = "Natural minimum mean monthly discharge (m³/s)",
  module_ = "Natural mean annual discharge (m³/s)",
  ad = "Width hydraulic geometry: Width = ad * Module^bd * (Q/Module)^b",
  bd = "Width hydraulic geometry exponent (downstream)",
  b = "Width hydraulic geometry exponent (at-a-station)",
  cd = "Depth hydraulic geometry: Depth = cd * Module^fd * (Q/Module)^f",
  fd = "Depth hydraulic geometry exponent (downstream)",
  f = "Depth hydraulic geometry exponent (at-a-station)",
  l = "Modelled width at mean annual discharge (m)",
  h = "Modelled depth at mean annual discharge (m)",
  h50_Morel = "Depth at median discharge, Morel et al. 2019 (m)",
  l50_Morel = "Width at median discharge, Morel et al. 2019 (m)",
  b_Morel = "Width exponent, Morel et al. 2019",
  f_Morel = "Depth exponent, Morel et al. 2019",
  phi = "Mean sediment size, -log2(D)",
  proba_interm = "Probability of flow intermittence",
  qmna5_rht = "Dry-year minimum monthly discharge, 5-year return period (m³/s)",
  qixa_10 = "Annual maximum peak discharge, 10-year return period (m³/s)"
)
rht_species <- c(
  abl = "bleak", ang = "eel", baf = "barbel", bam = "Mediterranean barbel",
  bln = "blageon", bou = "bitterling", brb = "white bream", bro = "pike",
  cas = "crucian carp", cco = "common carp", cha = "bullhead", che = "chub",
  epi = "three-spined stickleback", ept = "nine-spined stickleback",
  gar = "roach", gou = "gudgeon", gre = "ruffe", hot = "nase",
  lof = "stone loach", lot = "burbot", lpp = "brook lamprey",
  obr = "grayling", pch = "black bullhead", per = "perch",
  pes = "pumpkinseed", rot = "rudd", san = "pikeperch",
  sat = "Atlantic salmon", spi = "schneider", tan = "tench",
  tox = "toxostome", trf = "brown trout", vai = "minnow", van = "dace"
)

describe_rht_variable <- function(variable) {
  base <- sub("(icinf|icsup|_inf|_sup)$", "", variable)
  bound <- c(icinf = "Lower", `_inf` = "Lower", icsup = "Upper", `_sup` = "Upper")[
    sub(paste0("^", base), "", variable)
  ]
  case_when(
    variable %in% names(rht_description) ~ rht_description[variable],
    base != variable ~ paste(bound, "bound of the confidence interval of", base),
    grepl("^pesp_", variable) ~ paste(
      "Presence probability:", rht_species[sub("^pesp_", "", variable)]
    ),
    grepl("^q[0-9]+_rht$", variable) ~ paste(
      "Flow duration curve, quantile", sub("_rht$", "", variable), "(m³/s)"
    )
  )
}

rht_variables <- tibble(variable = names(sf::st_drop_geometry(rht))) |>
  mutate(
    description = vapply(variable, describe_rht_variable, character(1)),
    danet = variable %in% c("altitude", "pente", "d_source", "l", "h", "strahler"),
    pct_na_operation = round(
      100 * colMeans(is.na(rht_operation[variable])), 1
    )
  )
knitr::kable(rht_variables)
```

The six site variables used by @Danet-etal-2021 are: altitude,
slope, distance from source, width, depth, Strahler order.

### Snapping operations to the RHT

Distance between each operation and its matched reach (distances below 1 m
shown at 1 m):

```{r}
#| fig-width: 7
#| fig-height: 4

ggplot(rht_operation, aes(pmax(snap_dist_m, 1))) +
  geom_histogram(bins = 60) +
  scale_x_log10() +
  labs(x = "Distance to the nearest RHT reach (m)", y = "Operations")
```

Distance alone does not tell whether an operation was matched to the right
reach. We check it against an independent value: the upstream catchment area
of the station in AMOBIO (`drained_area`, constant per station so joined on
`sandre_code`), compared with the area of the matched RHT reach (`surf_bv`).

```{r}
rht_area_check <- rht_operation |>
  mutate(small_catchment = drained_area <= 10)

cat(sprintf(
  "%d / %d operations have an AMOBIO drained area\n",
  sum(!is.na(rht_area_check$area_ratio)), nrow(rht_area_check)
))
```

```{r}
#| fig-width: 9
#| fig-height: 5

rht_area_check |>
  filter(!is.na(area_ratio), area_ratio > 0) |>
  ggplot(aes(pmax(snap_dist_m, 1), area_ratio, color = small_catchment)) +
  annotate("rect",
    xmin = 0, xmax = Inf, ymin = 1 / 2, ymax = 2,
    fill = "grey85", alpha = 0.5
  ) +
  geom_point(alpha = 0.3, size = 0.6) +
  geom_vline(xintercept = 50, linetype = "dashed") +
  scale_x_log10() +
  scale_y_log10() +
  scale_color_manual(
    values = c(`FALSE` = "grey30", `TRUE` = "tomato"),
    labels = c(`FALSE` = "> 10 km²", `TRUE` = "<= 10 km²"),
    name = "AMOBIO drained area"
  ) +
  labs(
    x = "Distance to the nearest RHT reach (m)",
    y = "RHT / AMOBIO catchment area"
  )
```

Grey band: areas within a factor 2. Share of operations outside it, by
snapping distance and catchment size:

```{r}
rht_area_check |>
  filter(!is.na(area_ratio), area_ratio > 0) |>
  mutate(
    distance = cut(
      snap_dist_m, c(0, 50, 100, 200, Inf),
      include.lowest = TRUE
    ),
    mismatch = area_ratio > 2 | area_ratio < 1 / 2
  ) |>
  summarise(
    n = n(),
    pct_mismatch = round(100 * mean(mismatch), 1),
    .by = c(distance, small_catchment)
  ) |>
  arrange(distance, small_catchment)
```

Rule to keep an operation: RHT and AMOBIO areas within a factor 2 when the
AMOBIO area is known, snapping distance <= 50 m otherwise

We remove operations with Strahler order 0 (2 operations), these are marked with
NA.

### Pairwise relationships between continuum variables

The site variables of @Danet-etal-2021, from the RHT, plus the
hydrographic distance to the river mouth (D8). Depth is left out: in the
RHT, width and depth are both modelled from mean annual discharge and are
perfectly correlated (Spearman 1), so only width is kept. Operations kept by
the snapping rule only, one point per sampling location (repeated visits
share the same site variables). Upper panels: Spearman correlations,
unaffected by the log transforms of the lower panels. Slope, distance from
source and altitude contain zeros (altitude also slightly negative values,
down to -0.6 m), hence `log(x + 1)`.

```{r}
continuum_operation <- tar_read(continuum_operation)
continuum_pca <- tar_read(continuum_pca)

continuum_labels <- c(
  log_altitude = "log(altitude + 1)",
  log_slope = "log(slope + 1)",
  log_dist_source = "log(dist. source + 1)",
  log_width = "log(width)",
  strahler = "strahler",
  log_dist_mouth = "log(dist. mouth)"
)
continuum_location <- continuum_operation |>
  distinct(longitude, latitude, .keep_all = TRUE)

cat(sprintf("%d sampling locations\n", nrow(continuum_location)))
```

```{r}
#| fig-width: 12
#| fig-height: 12

GGally::ggpairs(
  continuum_location,
  columns = names(continuum_labels),
  columnLabels = unname(continuum_labels),
  upper = list(continuous = GGally::wrap("cor", method = "spearman")),
  lower = list(continuous = GGally::wrap("points", alpha = 0.1, size = 0.3)),
  diag = list(continuous = GGally::wrap("densityDiag"))
)
```

### PCA of the continuum variables

PCA on the standardised variables above, one row per sampling location
(`continuum_pca`). PC1 is oriented so that its scores increase with river
width (downstream), PC2 so that they increase with distance to mouth. Every
operation gets the scores of its location (`continuum_operation`).

```{r}
round(summary(continuum_pca)$importance, 3)
```

```{r}
round(continuum_pca$rotation, 2)
```

```{r}
#| fig-width: 9
#| fig-height: 7

arrow_scale <- 0.9 * max(abs(continuum_pca$x[, 1:2])) /
  max(abs(continuum_pca$rotation[, 1:2]))
loadings <- as_tibble(continuum_pca$rotation[, 1:2], rownames = "variable") |>
  mutate(
    across(c(PC1, PC2), \(x) x * arrow_scale),
    variable = continuum_labels[variable]
  )
pct_var <- round(100 * summary(continuum_pca)$importance[2, 1:2], 1)

ggplot(continuum_location, aes(PC1, PC2)) +
  geom_point(aes(color = district), alpha = 0.3, size = 0.6) +
  geom_segment(
    data = loadings,
    aes(x = 0, y = 0, xend = PC1, yend = PC2),
    arrow = arrow(length = unit(0.2, "cm"))
  ) +
  geom_label(data = loadings, aes(label = variable), size = 3) +
  labs(
    x = sprintf("PC1 (%.1f%%)", pct_var[1]),
    y = sprintf("PC2 (%.1f%%)", pct_var[2]),
    color = NULL
  )
```

```{r}
#| fig-width: 12
#| fig-height: 6

continuum_location |>
  tidyr::pivot_longer(c(PC1, PC2), names_to = "axis", values_to = "score") |>
  ggplot() +
  geom_sf(data = basin_kept, fill = "grey97", color = "grey40") +
  geom_point(
    aes(longitude, latitude, color = score),
    size = 0.6
  ) +
  scale_color_viridis_c() +
  facet_wrap(~axis) +
  coord_sf(crs = 4326) +
  theme_void()
```

PC1 (58.6% of the variance) is a river size gradient: distance from source,
width and Strahler order load positively, slope and altitude negatively,
while distance to mouth barely contributes. Low scores are small, steep
headwater streams; high scores are large, low-gradient rivers. PC2 (28.4%)
is mostly driven by distance to mouth and altitude (the size variables load
weakly and positively): it separates sites far inland and at higher
altitude (high scores) from lowland sites close to the sea (low scores), and
is uncorrelated with PC1.

### Food web structure along PC1

River (ASPE) food webs (`foodweb_structure`) with a PC1 score. Diet overlap
is the mean Jaccard similarity of prey sets over all pairs of consumers,
including the invertebrate groups; the fish-only version compares fish
consumers only (`fish_diet_overlap`).

```{r}
fw_continuum <- tar_read(foodweb_structure) |>
  select(
    operation_id, log_species_richness, connectance, trophic_length,
    diet_overlap, fish_diet_overlap
  ) |>
  inner_join(
    continuum_operation |> select(operation_id, district, PC1, PC2),
    by = "operation_id"
  ) |>
  mutate(species_richness = exp(log_species_richness))

cat(sprintf("%d river food webs\n", nrow(fw_continuum)))
```

```{r}
plot_fw_vs_axis <- function(axis, xlab) {
  fw_continuum |>
    tidyr::pivot_longer(
      c(
        species_richness, connectance, trophic_length, diet_overlap,
        fish_diet_overlap
      ),
      names_to = "metric", values_to = "value"
    ) |>
    mutate(metric = factor(
      metric,
      levels = c(
        "species_richness", "connectance", "trophic_length", "diet_overlap",
        "fish_diet_overlap"
      ),
      labels = c(
        "Species richness", "Connectance", "Trophic length", "Diet overlap",
        "Diet overlap (fish only)"
      )
    )) |>
    ggplot(aes(.data[[axis]], value)) +
    geom_point(alpha = 0.1, size = 0.5) +
    geom_smooth(aes(color = district), se = TRUE) +
    facet_wrap(~metric, scales = "free_y", ncol = 2) +
    labs(x = xlab, y = NULL, color = NULL)
}
```

```{r}
#| fig-width: 10
#| fig-height: 11

plot_fw_vs_axis(
  "PC1", sprintf("PC1 (%.1f%%), upstream to downstream", pct_var[1])
)
```

Signal is clear and goes as expected. Downstream web are longer, wider and
richer. Diet overlap should be computed only on fish species (excluding resource
nodes) so we can observe a clear signal.

### Food web structure along PC2

Same metrics along PC2, which carries distance to mouth and altitude.

```{r}
#| fig-width: 10
#| fig-height: 11

plot_fw_vs_axis(
  "PC2", sprintf("PC2 (%.1f%%), close to far from the sea", pct_var[2])
)
```

Here the pattern is less much clear, showing that distance to the river mouth is
not the only the relevant variable for food web structure (because a river can
be thin, but close to the sea).

## Beta diversity along the continuum

### One species list per site

Fish species lists come from the river (ASPE) operations with a food web and
a PC1 score (`size_year_filtered`). Sites are visited repeatedly, and pooling
visits inflates richness with the number of visits, so each site keeps a
single visit: the most recent one. A randomly chosen visit is used as a
check.

```{r}
site_visit <- tar_read(size_year_filtered) |>
  filter(survey == "river") |>
  distinct(operation_id, species_valid) |>
  inner_join(
    continuum_operation |> select(operation_id, district, PC1),
    by = "operation_id"
  ) |>
  left_join(
    tar_read(river_operation) |> distinct(operation_id, sandre_code, date),
    by = "operation_id"
  )
visit <- site_visit |> distinct(operation_id, sandre_code, date)

visit |>
  count(sandre_code, name = "visits") |>
  summarise(
    sites = n(),
    median_visits = median(visits),
    max_visits = max(visits)
  )
```

```{r}
richness_by_visits <- site_visit |>
  distinct(sandre_code, species_valid) |>
  count(sandre_code, name = "pooled") |>
  inner_join(
    site_visit |>
      semi_join(
        visit |> slice_max(date, n = 1, with_ties = FALSE, by = sandre_code),
        by = "operation_id"
      ) |>
      count(sandre_code, name = "last_visit"),
    by = "sandre_code"
  ) |>
  inner_join(visit |> count(sandre_code, name = "visits"), by = "sandre_code")

richness_by_visits |>
  mutate(visits = cut(visits, c(0, 1, 3, 6, 10, Inf))) |>
  summarise(
    sites = n(),
    richness_last_visit = median(last_visit),
    richness_pooled = median(pooled),
    .by = visits
  ) |>
  arrange(visits)
```

```{r}
site_matrix <- function(selected_visit) {
  presence <- site_visit |>
    semi_join(selected_visit, by = "operation_id") |>
    distinct(sandre_code, district, PC1, species_valid) |>
    mutate(present = 1L) |>
    tidyr::pivot_wider(
      names_from = species_valid, values_from = present, values_fill = 0L
    ) |>
    arrange(sandre_code)
  species <- presence |>
    select(-sandre_code, -district, -PC1) |>
    as.matrix()
  list(
    site = presence |> select(sandre_code, district, PC1),
    presence = species[, sort(colnames(species))]
  )
}

last_visit <- visit |>
  slice_max(date, n = 1, with_ties = FALSE, by = sandre_code)
set.seed(1)
random_visit <- visit |>
  slice_sample(n = 1, by = sandre_code)
community <- list(
  last = site_matrix(last_visit),
  random = site_matrix(random_visit)
)
sorensen <- lapply(
  community, \(x) vegan::vegdist(x$presence, method = "bray", binary = TRUE)
)
stopifnot(identical(
  community$last$site$sandre_code, community$random$site$sandre_code
))

cat(sprintf(
  "%d sites, %d species\n",
  nrow(community$last$presence), ncol(community$last$presence)
))
```

### Composition along PC1

Principal coordinates analysis (PCoA) of the Sørensen dissimilarities
(square-rooted, to make them Euclidean), last visit per site.

```{r}
#| fig-width: 13
#| fig-height: 6

pcoa <- cmdscale(sqrt(sorensen$last), k = 2, eig = TRUE)
pcoa_pct <- round(100 * pcoa$eig[1:2] / sum(pcoa$eig[pcoa$eig > 0]), 1)
pcoa_scores <- community$last$site |>
  mutate(axis1 = pcoa$points[, 1], axis2 = pcoa$points[, 2])

pcoa_plot <- function(color_var) {
  ggplot(pcoa_scores, aes(axis1, axis2, color = .data[[color_var]])) +
    geom_point(size = 0.7, alpha = 0.6) +
    coord_equal() +
    labs(
      x = sprintf("PCoA 1 (%.1f%%)", pcoa_pct[1]),
      y = sprintf("PCoA 2 (%.1f%%)", pcoa_pct[2])
    )
}
patchwork::wrap_plots(
  pcoa_plot("PC1") + scale_color_viridis_c(),
  pcoa_plot("district") + labs(color = NULL)
)
```

### Species along PC1

Association of each species with PC1: Spearman correlation between presence
and PC1 across sites (species present at >= 5% of sites). Species with
|rho| >= 0.2 are shown, as their frequency of occurrence in tenths of the PC1
range (equal number of sites per tenth), ordered by their mean PC1 where
present.

```{r}
presence_long <- community$last$presence |>
  as_tibble() |>
  mutate(sandre_code = community$last$site$sandre_code) |>
  tidyr::pivot_longer(-sandre_code, names_to = "species", values_to = "present") |>
  left_join(community$last$site, by = "sandre_code") |>
  mutate(pc1_bin = ntile(PC1, 10), .by = species)

species_pc1 <- presence_long |>
  summarise(
    prevalence = mean(present),
    rho = cor(present, PC1, method = "spearman"),
    mean_pc1 = sum(PC1 * present) / sum(present),
    .by = species
  ) |>
  filter(prevalence >= 0.05) |>
  arrange(desc(abs(rho)))
```

```{r}
#| fig-width: 9
#| fig-height: 7

species_shown <- species_pc1 |>
  filter(abs(rho) >= 0.2) |>
  arrange(mean_pc1)
bin_pc1 <- presence_long |>
  distinct(sandre_code, PC1, pc1_bin) |>
  summarise(pc1 = median(PC1), .by = pc1_bin) |>
  arrange(pc1_bin)

presence_long |>
  filter(species %in% species_shown$species) |>
  summarise(frequency = mean(present), .by = c(species, pc1_bin)) |>
  mutate(species = factor(species, levels = species_shown$species)) |>
  ggplot(aes(factor(pc1_bin), species, fill = frequency)) +
  geom_tile() +
  scale_x_discrete(labels = sprintf("%.1f", bin_pc1$pc1)) +
  scale_fill_viridis_c(name = "Frequency of\noccurrence") +
  labs(x = "PC1, median of each tenth (upstream to downstream)", y = NULL) +
  theme(axis.text.y = element_text(face = "italic"))
```

### Turnover and nestedness along PC1

For every pair of sites in the same watershed, the Sørensen dissimilarity
is partitioned into turnover (species replacement, Simpson dissimilarity)
and nestedness (species loss, the remainder) [@Baselga-2010], and related to
the distance between the two sites along PC1. With `a` the number of shared
species and `b`, `c` the species found at only one of the two sites:
Sørensen = (b + c) / (2a + b + c), turnover = min(b, c) / (a + min(b, c)).

```{r}
beta_pairs <- function(presence, pc1) {
  shared <- tcrossprod(presence)
  only_i <- rowSums(presence) - shared
  only_j <- t(only_i)
  sorensen <- (only_i + only_j) / (2 * shared + only_i + only_j)
  min_only <- pmin(only_i, only_j)
  turnover <- min_only / (shared + min_only)
  upper <- upper.tri(shared)
  sorensen <- sorensen[upper]
  turnover <- turnover[upper]
  tibble(
    distance_pc1 = abs(outer(pc1, pc1, "-"))[upper],
    total = sorensen,
    turnover = turnover,
    nestedness = sorensen - turnover
  )
}

beta_partition <- lapply(
  split(seq_len(nrow(community$last$site)), community$last$site$district),
  \(i) beta_pairs(community$last$presence[i, ], community$last$site$PC1[i])
) |>
  bind_rows(.id = "district")

beta_partition |>
  summarise(
    pairs = n(),
    across(c(total, turnover, nestedness), \(x) round(mean(x), 3)),
    .by = district
  )
```

```{r}
#| fig-width: 10
#| fig-height: 7

beta_partition |>
  mutate(distance_bin = cut_width(distance_pc1, 0.5, boundary = 0)) |>
  summarise(
    n = n(),
    distance_pc1 = mean(distance_pc1),
    across(c(total, turnover, nestedness), mean),
    .by = c(district, distance_bin)
  ) |>
  filter(n >= 30) |>
  tidyr::pivot_longer(
    c(total, turnover, nestedness),
    names_to = "component", values_to = "dissimilarity"
  ) |>
  mutate(component = factor(
    component,
    levels = c("total", "turnover", "nestedness"),
    labels = c("Total (Sørensen)", "Turnover", "Nestedness")
  )) |>
  ggplot(aes(distance_pc1, dissimilarity, color = component)) +
  geom_line() +
  geom_point(size = 0.8) +
  facet_wrap(~district) +
  labs(
    x = "Distance along PC1 between the two sites",
    y = "Mean dissimilarity (bins of 0.5, >= 30 pairs)",
    color = NULL
  )
```

Sites at the same position along PC1 already differ substantially (total
dissimilarity about 0.47-0.58), mostly through turnover (0.33-0.44). As the
distance along PC1 increases, total dissimilarity rises to 0.75-0.88, with
the two components taking over in turn. For small to moderate distances (up
to about 5 PC1 units), the rise comes mostly from nestedness (0.15 to
0.25-0.30) while turnover stays flat: pairs differ mainly because the
downstream site holds additional species, consistent with the species
gained downstream (cyprinids, percids) on top of widespread headwater species
(bullhead, minnow, stone loach). For the largest distances, turnover rises
sharply and nestedness declines, especially in the Seine and Loire: the most
upstream and most downstream sites share few species (brown trout-dominated
headwaters against cyprinid- and percid-rich large rivers).

### Beta diversity within PC1 bins

Do sites at the same position along PC1 differ more from each other upstream
than downstream? Sites are grouped by watershed and tenth of PC1 (the same
tenths as above). Two measures per group:

- mean dissimilarity between pairs of sites of the group, partitioned as
  above; as a mean over pairs it does not depend on the number of sites;
- Whittaker's beta (total number of species of the group / mean number of
  species per site), on 200 random draws of 15 sites, since the total grows
  with the number of sites.

The median distance between the sites of each pair is reported to check that
upstream groups are not simply more spread out.

```{r}
site_bin <- community$last$site |>
  left_join(
    last_visit |>
      left_join(
        continuum_operation |> select(operation_id, longitude, latitude),
        by = "operation_id"
      ) |>
      select(sandre_code, longitude, latitude),
    by = "sandre_code"
  ) |>
  mutate(pc1_bin = ntile(PC1, 10))
site_km <- site_bin |>
  sf::st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
  sf::st_transform(2154) |>
  sf::st_coordinates() / 1000

bin_groups <- split(
  seq_len(nrow(site_bin)),
  list(site_bin$district, site_bin$pc1_bin),
  drop = TRUE
)
bin_groups <- bin_groups[lengths(bin_groups) >= 5]

beta_within_bin <- lapply(bin_groups, \(i) {
  upper <- upper.tri(diag(length(i)))
  beta_pairs(community$last$presence[i, ], site_bin$PC1[i]) |>
    mutate(km = as.matrix(dist(site_km[i, ]))[upper])
}) |>
  bind_rows(.id = "group") |>
  tidyr::separate(
    group, c("district", "pc1_bin"), sep = "\\.", convert = TRUE
  ) |>
  summarise(
    across(c(total, turnover, nestedness), mean),
    median_km = median(km),
    .by = c(district, pc1_bin)
  ) |>
  left_join(
    site_bin |>
      summarise(sites = n(), pc1 = median(PC1), .by = c(district, pc1_bin)),
    by = c("district", "pc1_bin")
  ) |>
  arrange(district, pc1_bin)
```

```{r}
#| fig-width: 10
#| fig-height: 7

beta_within_bin |>
  tidyr::pivot_longer(
    c(total, turnover, nestedness),
    names_to = "component", values_to = "dissimilarity"
  ) |>
  mutate(component = factor(
    component,
    levels = c("total", "turnover", "nestedness"),
    labels = c("Total (Sørensen)", "Turnover", "Nestedness")
  )) |>
  ggplot(aes(pc1, dissimilarity, color = component)) +
  geom_line() +
  geom_point(aes(size = sites)) +
  scale_size_area(max_size = 3) +
  facet_wrap(~district) +
  labs(
    x = "PC1, median of each tenth (upstream to downstream)",
    y = "Mean dissimilarity between sites of the same tenth",
    color = NULL
  )
```

```{r}
#| fig-width: 9
#| fig-height: 4.5

n_draw_sites <- 15
set.seed(1)
whittaker <- lapply(bin_groups[lengths(bin_groups) >= n_draw_sites], \(i) {
  draws <- replicate(200, {
    drawn <- community$last$presence[sample(i, n_draw_sites), ]
    c(
      gamma = sum(colSums(drawn) > 0),
      alpha = mean(rowSums(drawn))
    )
  })
  tibble(
    gamma = mean(draws["gamma", ]),
    alpha = mean(draws["alpha", ]),
    whittaker = mean(draws["gamma", ] / draws["alpha", ])
  )
}) |>
  bind_rows(.id = "group") |>
  tidyr::separate(
    group, c("district", "pc1_bin"), sep = "\\.", convert = TRUE
  ) |>
  left_join(
    beta_within_bin |> select(district, pc1_bin, pc1),
    by = c("district", "pc1_bin")
  )

whittaker |>
  tidyr::pivot_longer(
    c(alpha, whittaker),
    names_to = "measure", values_to = "value"
  ) |>
  mutate(measure = factor(
    measure,
    levels = c("alpha", "whittaker"),
    labels = c("Species per site", "Whittaker's beta")
  )) |>
  ggplot(aes(pc1, value, color = district)) +
  geom_line() +
  geom_point(size = 1) +
  facet_wrap(~measure, scales = "free_y") +
  labs(
    x = "PC1, median of each tenth (upstream to downstream)",
    y = NULL, color = NULL
  )
```

Sites at the same position along PC1 are more different from each other
upstream than downstream. Mean pairwise dissimilarity falls from 0.63-0.65 in
the most upstream tenth to 0.38-0.45 in the most downstream one (Adour-Garonne,
Loire, Seine), and Whittaker's beta from about 4.5-5 to 2.2-2.5. Downstream
tenths hold more species in total (about 22-24 against 36-40 species in 15
sites; `whittaker$gamma`), but each site holds a larger share of them (about 5
species per site upstream against 15-18 downstream). Both components
decrease, nestedness more consistently than turnover. The distance between
sites does not explain the pattern in the Loire and Adour-Garonne (upstream
pairs are as close as or closer than downstream ones); in the Seine, upstream
pairs are farther apart, which may contribute. With about 5 species per site
upstream, a species missed at a single visit weighs heavily on dissimilarity,
so part of the high upstream beta, especially its nestedness part, may come
from imperfect detection.

## Species contribution to food web structure

Each fish species is removed in turn from each local food web, with its
size classes and any consumer left without prey (secondary extinctions),
and the metrics are recomputed (`species_contribution_operation`). The
change is the metric of the full web minus the metric without the species
(positive: the species raises the metric). Every removal lowers richness by
one, so the change is expressed relative to the mean change over the fish
species of the same web: this keeps the effect of the species' identity and
removes the effect of losing any one species. Webs with a single fish species
have no relative contribution.

```{r}
contribution <- tar_read(species_contribution_operation) |>
  inner_join(
    continuum_operation |> select(operation_id, district, PC1),
    by = "operation_id"
  ) |>
  tidyr::pivot_longer(
    starts_with("relative_"),
    names_to = "metric", names_prefix = "relative_",
    values_to = "relative"
  ) |>
  filter(!is.na(relative)) |>
  mutate(metric = factor(
    metric,
    levels = c("trophic_length", "connectance", "fish_diet_overlap"),
    labels = c("Trophic length", "Connectance", "Fish diet overlap")
  ))

contribution |>
  distinct(operation_id, species, n_secondary) |>
  summarise(
    webs = n_distinct(operation_id),
    removals = n(),
    with_secondary_extinctions = mean(n_secondary > 0)
  )
```

### Mean contribution per species

Mean relative contribution over the webs where the species occurs (species
in >= 50 webs), with an approximate 95% interval (mean +/- 2 standard
errors). Species are ordered and coloured by their mean trophic level across
the webs where they occur (prey-averaged, mean over their size classes;
highest at the top).

```{r}
#| fig-width: 12
#| fig-height: 9

species_order <- contribution |>
  distinct(operation_id, species, trophic_level) |>
  summarise(webs = n(), trophic_level = mean(trophic_level), .by = species) |>
  filter(webs >= 50) |>
  arrange(trophic_level)

contribution |>
  filter(species %in% species_order$species) |>
  summarise(
    mean = mean(relative),
    se = sd(relative) / sqrt(n()),
    .by = c(species, metric)
  ) |>
  left_join(species_order |> select(species, trophic_level), by = "species") |>
  mutate(species = factor(species, levels = species_order$species)) |>
  ggplot(aes(mean, species, color = trophic_level)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_pointrange(aes(xmin = mean - 2 * se, xmax = mean + 2 * se), size = 0.3) +
  scale_color_viridis_c(name = "Trophic\nlevel") +
  facet_wrap(~metric, scales = "free_x") +
  labs(x = "Mean relative contribution", y = NULL) +
  theme(axis.text.y = element_text(face = "italic"))
```

Contributions to trophic length follow trophic level: piscivores raise it
(pike, brown trout, pikeperch, perch, eel) and low-level species lower it
(brook lamprey, roach, nase). Connectance is raised most by generalists (chub,
eel, wels catfish). Fish diet overlap is lowered by species with distinctive
diets (brook lamprey, black bullhead, pike, nase) and raised by common
invertebrate feeders (grayling, stickleback, bullhead, stone loach, minnow).
These contributions follow from the diet links of the metaweb.
