---
title: "How eutrophisation affect food web structures along the estuary-sea
   continuum"
format:
  html:
    embed-resources: true
execute:
    message: false
    warning: false
---

## Summary

```{r}
#| output: false
library(renv)
renv::load()
library(targets)
library(here)
tar_config_set(
  store = here::here("_targets")
)

library(ggplot2)
set_theme(theme_minimal())
library(dplyr)
library(riversea)
library(sf)
```

## Estuary classification

First, we have different zones that we can download from the Sandre database.
They take the form of polygons.

```{r}
area <- tar_read(hydrographic_area)
head(area)
```

::: {.callout-note}
By investigating with `mapview`, we can see that there is a zone missing near Le
Havre.
:::

We can match food webs to zone either if they fit within a polygon, or if they
are very close to it (e.g. less than 5km away).

To classify food webs in different bassin, we use the define of the hydrographic
district given by the DCE. It is a very broad geographical unit, but we will
start from there.

```{r}
fw_with_district <- tar_read(fw_with_district)

library(rnaturalearth)
library(rnaturalearthhires)
coastline <- ne_coastline(scale = "large", returnclass = "sf")

ggplot(fw_with_district, aes(x = longitude, y = latitude, color = district)) +
  geom_point(size = 2.5) +
  geom_sf(data = coastline, inherit.aes = FALSE, linewidth = 0.3) +
  scale_color_viridis_d() +
  coord_sf(
    xlim = range(fw_with_district$longitude),
    ylim = range(fw_with_district$latitude)
  ) +
  labs(x = NULL, y = NULL, color = "Hydrographic district")
```

We will later this classification to model it as a random effect to take into
account the differences that can be due to the hydrographic bassin.

## First models

### Format the data

```{r}
# Food web with zone information (DCE hydrographic district).
district <- tar_read(fw_with_district) |>
  select(trait, district, zone_id)

# Food web with environment information.
environment <- tar_read(fw_with_env) |>
  filter(dist_to_env < 5) |>
  select(
    trait,
    log_species_richness,
    log_trophic_richness,
    trophic_length,
    sali_mean,
    chl_mean,
    year,
    survey,
    longitude,
    latitude
  ) |>
  mutate(
    salinity = scale(log10(sali_mean)),
    chlorophyll = scale(chl_mean)
  )

data <- inner_join(district, environment, by = join_by(trait)) |>
  mutate(log_trophic_richness = scale(log_trophic_richness))
```

### Fit the models

We fit three models to begin with that consider trophic length is shaped by
1. salinity and chlorophyll
2. salinity, chlorophyll and their interactions
3. salinity, chlorophyll and trophic richness (number of nodes in the local web)

```{r}
library(INLA)

m0 <- inla(
  trophic_length ~ 1 + salinity +
    f(year, model = "iid") + f(survey, model = "iid") +
    f(zone_id, model = "iid"),
  data = data
)
summary(m0)

m1 <- inla(
  trophic_length ~ 1 + salinity + chlorophyll +
    f(year, model = "iid") + f(survey, model = "iid") +
    f(zone_id, model = "iid"),
  data = data
)
summary(m1)

m2 <- inla(
  trophic_length ~ 1 + salinity * chlorophyll +
    f(year, model = "iid") + f(survey, model = "iid") +
    f(zone_id, model = "iid"),
  data = data
)
summary(m2)

m3 <- inla(
  trophic_length ~ 1 + salinity + chlorophyll + log_trophic_richness +
    f(year, model = "iid") + f(survey, model = "iid") +
    f(zone_id, model = "iid"),
  data = data
)
summary(m3)
```

### Visualise fixed effects

```{r}
extract_fixed <- function(model, model_name) {
  model$summary.fixed |>
    tibble::rownames_to_column("term") |>
    filter(term != "(Intercept)") |>
    mutate(model = model_name)
}

fixed <- bind_rows(
  extract_fixed(m0, "Salinity"),
  extract_fixed(m1, "Salinity + Chlorophyll"),
  extract_fixed(m2, "Salinity * Chlorophyll"),
  extract_fixed(m3, "Salinity + Chlorophyll + Richness")
)

ggplot(fixed, aes(x = mean, y = term, color = model)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey60") +
  geom_linerange(aes(xmin = `0.025quant`, xmax = `0.975quant`),
    position = position_dodge(width = 0.5)
  ) +
  geom_point(aes(x = `0.5quant`),
    size = 3,
    position = position_dodge(width = 0.5)
  ) +
  labs(x = "Effect on trophic length", y = NULL, color = NULL) +
  theme(legend.position = "bottom")
```

We see that the main driver of trophic length is clearly trophic richness.
The effects of salinity and chlorophyll in the two models that do not include
richness appear to be a consequence of their correlation of with richness and
not a direct effect.

It is partly expected. Trophic structure can be shaped by three factors: trophic
richness, individual sizes, and their predation window.
As we have for now taken the same predation window for all species, only effect
of environmental drivers on indivdual sizes could explain variation in trophic
structure.
Probably by refining predation window and species diet we will be able to
capture an effect of environmental driver beyond pure richness effects.

If we further compare these two models, we see that the interaction between
salinity and chlorophyll overlaps with zero, indicating that it does not
influence community trophic length.

```{r}
fixed |>
  filter_out(model == "With richness") |>
  ggplot(aes(x = mean, y = term, color = model)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey60") +
  geom_linerange(aes(xmin = `0.025quant`, xmax = `0.975quant`),
    position = position_dodge(width = 0.5)
  ) +
  geom_point(aes(x = `0.5quant`),
    size = 3,
    position = position_dodge(width = 0.5)
  ) +
  labs(x = "Effect on trophic length", y = NULL, color = NULL) +
  theme(legend.position = "bottom")
```

### Visualise marginal effets

```{r}
salinity_vals <- seq(min(data$salinity), max(data$salinity), length.out = 100)

prediction_salinity <- data.frame(
  salinity    = salinity_vals,
  chlorophyll = 0, # held at mean (scaled)
  year        = NA,
  survey      = NA,
  district    = NA
)

data_pred <- bind_rows(data, prediction_salinity)
m_pred_salinity <- inla(
  trophic_length ~ 1 + salinity + chlorophyll +
    f(year, model = "iid") + f(survey, model = "iid") +
    f(zone_id, model = "iid"),
  data = data_pred,
  control.predictor = list(compute = TRUE, link = 1)
)

# extract predictions for the NA rows
n_obs <- nrow(data)
preds <- m_pred_salinity$summary.fitted.values[(n_obs + 1):(n_obs + 100), ]

prediction_salinity |>
  bind_cols(preds) |>
  ggplot(aes(x = salinity_vals)) +
  geom_ribbon(aes(ymin = `0.025quant`, ymax = `0.975quant`), alpha = 0.2) +
  geom_line(aes(y = mean)) +
  geom_point(data = data, aes(x = salinity, y = trophic_length), alpha = 0.1) +
  labs(x = "Salinity (scaled)", y = "Trophic length")
```


```{r}
chlorophyll_vals <- seq(min(data$chlorophyll), max(data$chlorophyll), length.out = 100)

prediction_chlorophyll <- data.frame(
  chlorophyll = chlorophyll_vals,
  salinity = 0, # held at mean (scaled)
  year = NA,
  survey = NA,
  district = NA
)

data_pred <- bind_rows(data, prediction_chlorophyll)
m_pred_chlorophyll <- inla(
  trophic_length ~ 1 + salinity + chlorophyll +
    f(year, model = "iid") + f(survey, model = "iid") +
    f(zone_id, model = "iid"),
  data = data_pred,
  control.predictor = list(compute = TRUE, link = 1)
)

# extract predictions for the NA rows
n_obs <- nrow(data)
preds <- m_pred_chlorophyll$summary.fitted.values[(n_obs + 1):(n_obs + 100), ]

prediction_chlorophyll |>
  bind_cols(preds) |>
  ggplot(aes(x = chlorophyll_vals)) +
  geom_ribbon(aes(ymin = `0.025quant`, ymax = `0.975quant`), alpha = 0.2) +
  geom_line(aes(y = mean)) +
  geom_point(data = data, aes(x = chlorophyll, y = trophic_length), alpha = 0.1) +
  labs(x = "Chlorophyll (scaled)", y = "Trophic length")
```


```{r}
richness_vals <- seq(min(data$log_trophic_richness),
  max(data$log_trophic_richness),
  length.out = 100
)

prediction <- data.frame(
  log_trophic_richness = richness_vals,
  chlorophyll = 0,
  salinity = 0,
  year = NA,
  survey = NA,
  district = NA
)

data_pred <- bind_rows(data, prediction)
m_pred <- inla(
  trophic_length ~ 1 + salinity + chlorophyll + log_trophic_richness +
    f(year, model = "iid") + f(survey, model = "iid") +
    f(zone_id, model = "iid"),
  data = data_pred,
  control.predictor = list(compute = TRUE, link = 1)
)

# extract predictions for the NA rows
n_obs <- nrow(data)
preds <- m_pred$summary.fitted.values[(n_obs + 1):(n_obs + 100), ]

prediction |>
  bind_cols(preds) |>
  ggplot(aes(x = richness_vals)) +
  geom_ribbon(aes(ymin = `0.025quant`, ymax = `0.975quant`), alpha = 0.2) +
  geom_line(aes(y = mean)) +
  geom_point(data = data, aes(x = log_trophic_richness, y = trophic_length), alpha = 0.1) +
  labs(x = "Log trophic richness (scaled)", y = "Trophic length")
```


### Visualise random effects

#### Temporal

```{r}
year_re <- m3$summary.random$year |>
  as_tibble() |>
  rename(year = ID)

ggplot(year_re, aes(x = year, y = mean)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey60") +
  geom_ribbon(aes(ymin = `0.025quant`, ymax = `0.975quant`), alpha = 0.2) +
  geom_line() +
  geom_point() +
  labs(x = "Year", y = "Year random effect on trophic length") +
  theme_minimal()
```

#### Spatial

```{r}
area <- tar_read(hydrographic_area)
spatial_re <- m3$summary.random$zone_id |>
  as_tibble() |>
  rename(zone_id = ID) |>
  left_join(area, by = join_by(zone_id)) |>
  st_as_sf()

bbox <- st_bbox(spatial_re)

ggplot(spatial_re) +
  geom_sf(aes(fill = mean)) +
  geom_sf(data = coastline) +
  scale_fill_gradient2(
    low = "steelblue", mid = "white", high = "tomato",
    midpoint = 0, name = "RE"
  ) +
  coord_sf(
    xlim = c(bbox["xmin"], bbox["xmax"]),
    ylim = c(bbox["ymin"], bbox["ymax"])
  ) +
  labs(x = NULL, y = NULL)
```


