---
title: "How eutrophisation affect food web structures along the estuary-sea
   continuum"
format:
  html:
    embed-resources: true
execute:
    message: false
    warning: false
---

## Summary

```{r}
#| output: false
library(renv)
renv::load()
library(targets)
library(here)
tar_config_set(
  store = here::here("_targets")
)

library(ggplot2)
set_theme(theme_minimal())
library(dplyr)
library(riversea)
library(sf)
```

## Estuary classification

First, we have different zones that we can download from the Sandre database.
They take the form of polygons.

```{r}
area <- tar_read(hydrographic_area)
head(area)
```

::: {.callout-note}
By investigating with `mapview`, we can see that there is a zone missing near Le
Havre.
:::

We can match food webs to zone either if they fit within a polygon, or if they
are very close to it (e.g. less than 5km away).

To classify food webs in different bassin, we use the define of the hydrographic
district given by the DCE. It is a very broad geographical unit, but we will
start from there.

```{r}
fw_with_district <- tar_read(fw_with_district)

library(rnaturalearth)
library(rnaturalearthhires)
coastline <- ne_coastline(scale = "large", returnclass = "sf")

ggplot(fw_with_district, aes(x = longitude, y = latitude, color = district)) +
  geom_point(size = 2.5) +
  geom_sf(data = coastline, inherit.aes = FALSE, linewidth = 0.3) +
  scale_color_viridis_d() +
  coord_sf(
    xlim = range(fw_with_district$longitude),
    ylim = range(fw_with_district$latitude)
  ) +
  labs(x = NULL, y = NULL, color = "Hydrographic district")
```

We will later this classification to model it as a random effect to take into
account the differences that can be due to the hydrographic bassin.

## First model

```{r}
library(INLA)

fw_with_env <- tar_read(fw_with_env) |>
  select(
    log_species_richness,
    trophic_length,
    sali_mean,
    chl_mean,
    year
  ) |>
  mutate(
    salinity = scale(log10(sali_mean)),
    chlorophyll = scale(chl_mean)
  )

m <- inla(
  trophic_length ~ 1 + salinity + chlorophyll + f(year, model = "iid"),
  data = fw_with_env
)
summary(m)
```


