---
title: "How community richness changes along the estuary-sea continuum"
format:
  html:
    embed-resources: true
execute:
    message: false
    warning: false
---

We want to model how species richness changes from the estuary to the sea.
To do so, we will quantify how the species richness changes along the salinity
gradient.

```{r}
#| echo: false

library(renv)
renv::load()
library(targets)
library(here)
tar_config_set(
  store = here::here("_targets")
)

library(ggplot2)
library(hrbrthemes)
theme_set(theme_ipsum())
library(dplyr)
library(riversea)
library(sf)
```

## Preparing the data

We begin by loading our dataset.

```{r}
data <- tar_read(fw_with_env) |>
  filter(dist_to_env < 5) |>
  mutate(
    salinity = sali_mean |> as.numeric()
  ) |>
  select(
    year,
    survey,
    log_species_richness,
    log_trophic_richness,
    longitude,
    latitude,
    salinity
  ) |>
  mutate(is_nurse = if_else(survey == "nurse", 1, 0))

plot_data_on_map(data)
```

Here is the map of our sampling points.

## Modeling alpha-diversity with salinity

Next, we want to model how species richness changes with salinity.
To do that properly with to account for spatial autocorrelation.

```{r}
library(INLA)
library(fmesher)

coords <- cbind(data$longitude, data$latitude)

mesh <- fmesher::fm_mesh_2d_inla(
  loc       = coords,
  max.edge  = c(0.1, 0.5),
  offset    = c(0.1, 0.3),
  cutoff    = 0.05
)
plot(mesh)
points(coords, col = "red", cex = 0.1, pch = 3)
```

Using a Spatial Gaussian Random Field on our mesh, we can fit a model that
accounts for spatial autocorrelation.

```{r}
spde <- inla.spde2.pcmatern(
  mesh = mesh,
  prior.range = c(0.5, 0.5),
  prior.sigma = c(1, 0.01)
)

A <- inla.spde.make.A(mesh = mesh, loc = coords)

stk <- inla.stack(
  data = list(y = data$log_species_richness),
  A = list(A, 1),
  effects = list(
    list(spatial.field = 1:spde$n.spde),
    list(
      intercept = rep(1, nrow(data)),
      salinity = data$salinity,
      is_nurse = data$is_nurse,
      year = data$year
    )
  ),
  tag = "obs"
)

m1 <- inla(
  y ~ -1 +
    intercept +
    salinity +
    is_nurse +
    f(year, model = "iid") + f(spatial.field, model = spde),
  data = inla.stack.data(stk),
  control.predictor = list(
    A = inla.stack.A(stk),
    compute = TRUE
  ),
  control.compute = list(
    waic = TRUE,
    config = TRUE
  )
)

summary(m1)
```

Next, we can investigate our model.

```{r}
spde.result <- inla.spde2.result(m1, "spatial.field", spde, do.transf = TRUE)

inla.emarginal(function(x) x, spde.result$marginals.range.nominal[[1]])
```

This captures the caracteristic range of the spatial autocorrelation.
It is quite small, meaning that it acts at a small spatial scale, and our data
is not too sensitive to this biais. This may be due to the act that our data is
quite noisy due to fishing operations that may often miss many species. So even
at the same place, two communities may not be too strongly correlated due to
the sampling noise.

```{r}
sal_marginal <- m1$marginals.fixed$salinity

# Convert to a data frame for plotting the full posterior density
sal_df <- as.data.frame(inla.smarginal(sal_marginal))

ggplot(sal_df, aes(x = x, y = y)) +
  geom_line(linewidth = 1, colour = "#2166ac") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey40") +
  # 95% credible interval shading
  geom_ribbon(
    data = subset(sal_df, x >= inla.hpdmarginal(0.95, sal_marginal)[1] &
      x <= inla.hpdmarginal(0.95, sal_marginal)[2]),
    aes(ymin = 0, ymax = y), fill = "#2166ac", alpha = 0.2
  ) +
  labs(
    title = "Posterior distribution — salinity effect",
    x     = "Effect size (log species richness per salinity unit)",
    y     = "Density"
  )
```

```{r}
n_samples <- 100
samples <- inla.posterior.sample(n_samples, m1)
beta_int_samples <- sapply(samples, function(s) s$latent["intercept:1", ])
beta_sal_samples <- sapply(samples, function(s) s$latent["salinity:1", ])
sal_seq <- seq(min(data$salinity), max(data$salinity), length.out = 200)

pred_matrix <- outer(rep(1, 200), beta_int_samples) +
  outer(sal_seq, beta_sal_samples)

pred_df <- data.frame(
  salinity = sal_seq,
  fit      = apply(pred_matrix, 1, median),
  lwr      = apply(pred_matrix, 1, quantile, probs = 0.045),
  upr      = apply(pred_matrix, 1, quantile, probs = 0.955)
)

zones <- data.frame(
  xmin  = c(-Inf, 0.5, 5, 18, 30),
  xmax  = c(0.5, 5, 18, 30, Inf),
  label = c("Freshwater", "Oligohaline", "Mesohaline", "Polyhaline", "Euhaline"),
  fill  = c("#a8d5f5", "#d4edda", "#fff3b0", "#f4a96a", "#e07b7b")
)

# Clip zones to your actual data range
x_min_data <- min(data$salinity)
x_max_data <- max(data$salinity)

zones$xmin <- pmax(zones$xmin, x_min_data)
zones$xmax <- pmin(zones$xmax, x_max_data)

# Drop zones outside your data range
zones <- zones[zones$xmin < zones$xmax, ]

ggplot(pred_df, aes(x = salinity)) +
  geom_point(
    data = data, aes(x = salinity, y = log_species_richness),
    alpha = 0.1
  ) +
  geom_vline(
    xintercept = c(5, 18, 30),
    linetype = "dashed", colour = "grey40", linewidth = 0.5
  ) +
  geom_text(
    data = zones,
    aes(x = (xmin + xmax) / 2, y = Inf, label = label),
    vjust = 1.5, size = 3.2, colour = "grey30",
    inherit.aes = FALSE
  ) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), fill = "lightblue", alpha = 0.2) +
  geom_line(aes(y = fit), colour = "lightblue", linewidth = 1.2) +
  labs(
    x = "Salinity",
    y = "Log Species Richness"
  )
```

## Modeling alpha, beta and gamma diversity with salinity

Next, we focus on the different aspects of species diversity

- alpha-diversity, that is, the diversity of local communities
- beta-diversity, that is, how species composition changes
- gamma-diversity, that is, the aggregated diversity

Here, for simplicity we compute beta diversity as the ratio of gamma to alpha
diversity.

First, we will focus only on the Garonne estuary, to compare sites that are
comparable.

```{r}
data <- tar_read(fw_with_env) |>
  filter(dist_to_env < 5)

district <- tar_read(fw_with_district) |>
  select(trait, year, zone_id, district)

data <- data |>
  left_join(district, by = join_by(trait, year)) |>
  mutate(
    salinity = sali_mean |> as.numeric()
  ) |>
  select(
    trait,
    year,
    survey,
    log_species_richness,
    log_trophic_richness,
    longitude,
    latitude,
    salinity,
    foodweb, #  Now kept to extract species composition.
    district,
    zone_id,
    substrat,
    bathymetry,
    sali_perc_90,
    chl_mean,
    nppv_mean,
    substrat,
    temp_mean,
    temp_cv,
    o2_mean,
    o2_cv,
    wave_mean
  ) |>
  mutate(is_nurse = if_else(survey == "nurse", 1, 0))

data <- data |>
  filter(district == "Ardour-Garonne")
```

First let's plot our sampling points and the salinity gradient on a map.

```{r}
environment_data <- tar_read(environment_data)

coastline <- rnaturalearth::ne_coastline(scale = "large", returnclass = "sf")

e <- c(-0.1, 0.1)
environment_data |>
  group_by(year, x, y) |>
  summarise(salinity = mean(sali_mean), .groups = "drop") |>
  ggplot(aes(x = x, y = y, fill = salinity)) +
  geom_tile() +
  geom_sf(data = coastline, inherit.aes = FALSE, linewidth = 0.3) +
  scale_fill_viridis_c() +
  geom_point(data, mapping = aes(longitude, latitude), alpha = 0.3) +
  coord_sf(xlim = range(data$longitude) + e, ylim = range(data$latitude) + e) +
  labs(x = NULL, y = NULL, fill = "Salinity [PSU]")
```

We then compute species composition for each local community to compute the
three measure of diversity.

```{r}
resource_list <- tar_read(resource_list)

get_species <- function(web) {
  sub("_.*", "", colnames(web)) |>
    unique() |>
    setdiff(resource_list)
}

data <- data |>
  mutate(
    species = purrr::map(foodweb, get_species),
    species_richness = purrr::map_dbl(species, length)
  ) |>
  arrange(salinity)

plot_rolling_window <- function(data, window) {
  diversity_df <- purrr::map_dfr(
    seq_len(nrow(data) - window + 1),
    function(i) {
      w <- data[i:(i + window - 1), ]
      tibble(
        salinity = mean(w$salinity),
        gamma = n_distinct(unlist(w$species)),
        alpha = mean(w$species_richness)
      )
    }
  ) |>
    mutate(beta = gamma / alpha) |>
    tidyr::pivot_longer(cols = c(alpha, beta, gamma), names_to = "diversity")
  ggplot(diversity_df, aes(salinity, value)) +
    geom_point() +
    geom_line() +
    geom_vline(
      xintercept = c(5, 18, 30),
      linetype = "dashed", colour = "grey40", linewidth = 0.5
    ) +
    facet_wrap("diversity", scales = "free_y") +
    labs(x = "Salinity [PSU]", y = "Community diversity")
}

plot_rolling_window(data, 200)
```

First, we see that both alpha and gamma diversity increases with salinity,
suggesting that communities become richer through the apparition of marine
species, and that the brackish diversity is lower. 

We observe two peaks in beta diversitt, respectively around 10 and 30 PSU. Those
could signals the shift in community due to the transition caused by the
presence brackish and marine species respectively.

## Salinity niches

Here we are interested in caracterising species salinity niches, that is,
the salinity range where they can live. To avoid having to many species, we
filter the more rare species.

```{r}
n_op <- data |>
  pull(trait) |>
  unique() |>
  length()

threshold <- 0.01

occurence <- data |>
  tidyr::unnest(species) |>
  group_by(species) |>
  summarise(n = n(), .groups = "drop")

rare_species <- occurence |>
  filter(n < threshold * n_op) |>
  pull(species)

niche <- data |>
  select(species, salinity) |>
  tidyr::unnest(species) |>
  filter_out(species %in% rare_species) |>
  group_by(species) |>
  summarise(
    salinity_med = median(salinity),
    salinity_q10 = quantile(salinity, 0.1),
    salinity_q90 = quantile(salinity, 0.9),
    salinity_min = min(salinity),
    salinity_max = max(salinity),
    .groups = "drop"
  )
```


```{r}
niche <- niche |>
  arrange(salinity_med) |>
  mutate(
    species = factor(species, levels = species)
  )

ggplot(niche, aes(y = species)) +
  geom_segment(
    aes(x = salinity_q10, xend = salinity_q90, yend = species),
    linewidth = 1.5,
  ) +
  geom_segment(
    aes(x = salinity_min, xend = salinity_max, yend = species),
    linewidth = 0.5,
  ) +
  geom_point(aes(x = salinity_med), size = 2) +
  labs(x = "Salinity [PSU]", y = NULL)
```

We see that most species have wide salinity niche, meaning that they can endure
high variability in salinity, these species notably comprise migratory species
such as the Anguilla or the Alosa.

We find that most species have a wide salinity niche, which is not surprising
for species found in estuaries. That said, a few species seem to be marine such
as

- [Arnoglossus laterna](https://www.fishbase.se/summary/Arnoglossus-laterna.html)
- [Buglossidium luteum](https://fishbase.se/summary/Buglossidium-luteum)

which are both marine and dermersal species that live on sandy bottoms.

This analysis invites to couple our food web analysis with species traits to
better understand what drives food web structure changes. In this regard,
information about species migratory behaviour and whether they are benthic or
pelagic. For example, change in food web structure could signal the apparition
of a bentho-pelagic coupling.

## Community composition ordination

### Identify the salinity gradient as the main structuring axis

To ordinate communities along the salinity gradient, we first need to build the
community matrix which indicate presence and absence of the different species
for each site (or here fishing operation).

```{r}
community_matrix <- data |>
  filter(year > 2000) |>
  select(trait, species, year) |>
  tidyr::unnest(species) |>
  filter_out(species %in% rare_species) |>
  mutate(value = 1) |>
  tidyr::pivot_wider(
    names_from = species,
    values_fill = 0
  )
```

Then, we use the package `vegan` to do an NMDS using the Jaccard distance to
measure dissimilarity between communities. Here, we use three dimensions so we
have a stress inferior to 0.2, meaning that our ordination is somewhat
acceptable.

```{r}
#| include: false

library(vegan)

com_std <- decostand(community_matrix |> select(-trait, year), method = "pa")
nmds <- metaMDS(
  com_std,
  distance = "jaccard",
  binary = TRUE,
  k = 3
)
```

Next, we want to plot our ordinated communities along the salinity gradient, and
visualise which species are driving the change in community composition.
Typically, we would expect freshwater and marine species to drive the two ends
of the gradient.

```{r}
scores_df <- as.data.frame(scores(nmds, display = "sites"))

scores_df$salinity <- community_matrix |>
  left_join(data, by = join_by(trait)) |>
  pull(salinity)

scores_df$salinity <- community_matrix |>
  left_join(data, by = join_by(trait)) |>
  pull(salinity)

sp_scores <- as.data.frame(scores(nmds, display = "species"))
sp_scores$species <- rownames(sp_scores)
sp_scores <- sp_scores |>
  mutate(r2 = NMDS1^2 + NMDS2^2) |>
  filter(r2 > quantile(r2, 0.8))

ggplot(scores_df, aes(x = NMDS1, y = NMDS2)) +
  geom_point(aes(colour = salinity), size = 2) +
  scale_colour_viridis_c() +
  geom_segment(
    data = sp_scores,
    aes(x = 0, y = 0, xend = NMDS1, yend = NMDS2),
    arrow = arrow(length = unit(0.2, "cm")),
    colour = "grey40", linewidth = 0.5
  ) +
  geom_text(
    data = sp_scores,
    aes(label = species), size = 3, colour = "grey20",
    nudge_y = 0.02
  ) +
  annotate("text",
    x = Inf, y = -Inf,
    label = paste("Stress =", round(nmds$stress, 3)),
    hjust = 1.1, vjust = -0.5, size = 3.5, colour = "grey30"
  )
```

We see that the main structuring axis of the community is salinity as
could be expected. Second, we see that marine species (pointing to the right)
seems to be the more structuring ones, which are the one we identified
previously with our salinity niches. What is less clear is what is the other
axis of variation. We could expect the habitat, and so, for example the presence
of benthic species to be another structuring axis.

We can plot the our community on the map, coloring with NMDS first axis to see
how are structure are communities in space.

```{r}
scores_df$latitude <- community_matrix |>
  left_join(data, by = join_by(trait)) |>
  pull(latitude)
scores_df$longitude <- community_matrix |>
  left_join(data, by = join_by(trait)) |>
  pull(longitude)

ggplot(scores_df, aes(x = longitude, y = latitude, color = NMDS1)) +
  geom_point() +
  scale_colour_viridis_c() +
  geom_sf(data = coastline, inherit.aes = FALSE, linewidth = 0.3) +
  coord_sf(xlim = range(data$longitude) + e, ylim = range(data$latitude) + e) +
  labs(x = NULL, y = NULL)
```

They are struture along the estuary gradient. In particular, to visualise how
salinity structure communities, we can plot the first axis of the NMDS against
salinity.

```{r}
scores_df$log_species_richness <- community_matrix |>
  left_join(data, by = join_by(trait)) |>
  pull(log_species_richness)

ggplot(scores_df, aes(x = salinity, y = NMDS1)) +
  geom_point(aes(colour = log_species_richness), size = 2) +
  scale_colour_viridis_c() +
  labs(x = "Salinity", colour = "Log Species\nRichness")
```

We see a positive association. As suggested by the relative high stress
of the NMDS (~0.18), the community are difficult to ordinate. This may be due to
to the noise related to the sampling process, sometimes we do not fish all
species that are there. It may also be due to the fact that we aggregate points
over the year and so we have differencies from one year to another.

### Identify structuring variables beyond salinity

We want to do a more general analysis to find which environmental variables
structure our communities.

```{r}
env <- data |>
  select(
    trait, salinity, bathymetry, o2_mean, chl_mean, nppv_mean,
    temp_mean, year
  ) |>
  right_join(community_matrix |> select(trait, year), by = join_by(trait, year)) |>
  select(-c(trait, year))

fit <- envfit(nmds, env)

fit_df <- as.data.frame(scores(fit, display = "vectors"))
fit_df$variable <- rownames(fit_df)
fit_df$r2 <- fit$vectors$r
fit_df$p <- fit$vectors$pvals
fit_df <- fit_df |> filter(p <= 0.05)

ggplot(scores_df, aes(x = NMDS1, y = NMDS2)) +
  geom_point(aes(colour = log_species_richness), size = 2, alpha = 0.7) +
  scale_colour_viridis_c() +
  geom_segment(
    data = fit_df,
    aes(x = 0, y = 0, xend = NMDS1, yend = NMDS2),
    arrow = arrow(length = unit(0.25, "cm")),
    colour = "black", linewidth = 0.7
  ) +
  geom_label(
    data = fit_df,
    aes(
      x = NMDS1 * 1.15, y = NMDS2 * 1.15,
      label = paste0(variable, "\nR²=", round(r2, 2))
    ),
    size = 3, fill = "white", label.size = 0.2
  ) +
  labs(colour = "Log Species Richness")
```

We recover that most of the structure is driven by the salinity gradient and the
correlated variable.


