---
title: "How community richness changes along the estuary-sea continuum"
format:
  html:
    embed-resources: true
execute:
    message: false
    warning: false
---

We want to model how species richness changes from the estuary to the sea.
To do so, we will quantify how the species richness changes along the salinity
gradient.

```{r}
#| echo: false

library(renv)
renv::load()
library(targets)
library(here)
tar_config_set(
  store = here::here("_targets")
)

library(ggplot2)
set_theme(theme_minimal())
library(dplyr)
library(riversea)
library(sf)
```

## Preparing the data

We begin by loading our dataset.

```{r}
data <- tar_read(fw_with_env) |>
  filter(dist_to_env < 5) |>
  mutate(
    salinity = sali_mean |> as.numeric()
  ) |>
  select(
    year,
    survey,
    log_species_richness,
    log_trophic_richness,
    longitude,
    latitude,
    salinity
  ) |>
  mutate(is_nurse = if_else(survey == "nurse", 1, 0))

plot_data_on_map(data)
```

Here is the map of our sampling points.

## Modeling alpha-diversity with salinity

Next, we want to model how species richness changes with salinity.
To do that properly with to account for spatial autocorrelation.

```{r}
library(INLA)
library(fmesher)

coords <- cbind(data$longitude, data$latitude)

mesh <- fmesher::fm_mesh_2d_inla(
  loc       = coords,
  max.edge  = c(0.1, 0.5), # in units of your CRS — adjust!
  offset    = c(0.1, 0.3),
  cutoff    = 0.05
)
plot(mesh)
points(coords, col = "red", cex = 0.1, pch = 3) # sanity check: points inside mesh
```

Using a Spatial Gaussian Random Field on our mesh, we can fit a model that
accounts for spatial autocorrelation.

```{r}
spde <- inla.spde2.pcmatern(
  mesh = mesh,
  prior.range = c(0.5, 0.5),
  prior.sigma = c(1, 0.01)
)

A <- inla.spde.make.A(mesh = mesh, loc = coords)

stk <- inla.stack(
  data = list(y = data$log_species_richness),
  A = list(A, 1),
  effects = list(
    list(spatial.field = 1:spde$n.spde),
    list(
      intercept = rep(1, nrow(data)),
      salinity = data$salinity,
      is_nurse = data$is_nurse,
      year = data$year
    )
  ),
  tag = "obs"
)

m1 <- inla(
  y ~ -1 +
    intercept +
    salinity +
    is_nurse +
    f(year, model = "iid") + f(spatial.field, model = spde),
  data = inla.stack.data(stk),
  control.predictor = list(
    A = inla.stack.A(stk),
    compute = TRUE
  ),
  control.compute = list(
    waic = TRUE,
    config = TRUE
  )
)

summary(m1)
```

Next, we can investigate our model.

```{r}
spde.result <- inla.spde2.result(m1, "spatial.field", spde, do.transf = TRUE)

# Posterior mean of spatial range
inla.emarginal(function(x) x, spde.result$marginals.range.nominal[[1]])

# Posterior mean of spatial std. deviation
inla.emarginal(function(x) x, spde.result$marginals.variance.nominal[[1]])
```

```{r}
sal_marginal <- m1$marginals.fixed$salinity

# Convert to a data frame for plotting the full posterior density
sal_df <- as.data.frame(inla.smarginal(sal_marginal))

ggplot(sal_df, aes(x = x, y = y)) +
  geom_line(linewidth = 1, colour = "#2166ac") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey40") +
  # 95% credible interval shading
  geom_ribbon(
    data = subset(sal_df, x >= inla.hpdmarginal(0.95, sal_marginal)[1] &
      x <= inla.hpdmarginal(0.95, sal_marginal)[2]),
    aes(ymin = 0, ymax = y), fill = "#2166ac", alpha = 0.2
  ) +
  labs(
    title = "Posterior distribution — salinity effect",
    x     = "Effect size (log species richness per salinity unit)",
    y     = "Density"
  )
```

```{r}
n_samples <- 100
samples <- inla.posterior.sample(n_samples, m1)
beta_int_samples <- sapply(samples, function(s) s$latent["intercept:1", ])
beta_sal_samples <- sapply(samples, function(s) s$latent["salinity:1", ])
sal_seq <- seq(min(data$salinity), max(data$salinity), length.out = 200)

pred_matrix <- outer(rep(1, 200), beta_int_samples) +
  outer(sal_seq, beta_sal_samples)

pred_df <- data.frame(
  salinity = sal_seq,
  fit      = apply(pred_matrix, 1, median),
  lwr      = apply(pred_matrix, 1, quantile, probs = 0.045),
  upr      = apply(pred_matrix, 1, quantile, probs = 0.955)
)

zones <- data.frame(
  xmin  = c(-Inf, 0.5, 5, 18, 30),
  xmax  = c(0.5, 5, 18, 30, Inf),
  label = c("Freshwater", "Oligohaline", "Mesohaline", "Polyhaline", "Euhaline"),
  fill  = c("#a8d5f5", "#d4edda", "#fff3b0", "#f4a96a", "#e07b7b")
)

# Clip zones to your actual data range
x_min_data <- min(data$salinity)
x_max_data <- max(data$salinity)

zones$xmin <- pmax(zones$xmin, x_min_data)
zones$xmax <- pmin(zones$xmax, x_max_data)

# Drop zones outside your data range
zones <- zones[zones$xmin < zones$xmax, ]

ggplot(pred_df, aes(x = salinity)) +
  geom_point(
    data = data, aes(x = salinity, y = log_species_richness),
    alpha = 0.1
  ) +
  geom_vline(
    xintercept = c(5, 18, 30),
    linetype = "dashed", colour = "grey40", linewidth = 0.5
  ) +
  geom_text(
    data = zones,
    aes(x = (xmin + xmax) / 2, y = Inf, label = label),
    vjust = 1.5, size = 3.2, colour = "grey30",
    inherit.aes = FALSE
  ) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), fill = "lightblue", alpha = 0.2) +
  geom_line(aes(y = fit), colour = "lightblue", linewidth = 1.2) +
  labs(
    x = "Salinity",
    y = "Log Species Richness"
  )
```

## Modeling gamma-diversity with salinity

Next, we focus on other aspects of species diversity:

- beta-diversity that is how species composition changes
- gamma-diversity that is the aggregated diversity

To do so, we need to extract species composition for each community.

```{r}
data <- tar_read(fw_with_env) |>
  filter(dist_to_env < 5) |>
  mutate(
    salinity = sali_mean |> as.numeric()
  ) |>
  select(
    year,
    survey,
    log_species_richness,
    log_trophic_richness,
    longitude,
    latitude,
    salinity,
    foodweb #  Now kept to extract species composition.
  ) |>
  mutate(is_nurse = if_else(survey == "nurse", 1, 0))

resources <- c(
  "biofilm",
  "detritus",
  "detritus",
  "phytoplankton",
  "macrophyte",
  "worm",
  "crustacean",
  "mollusk",
  "echinoderm",
  "insect",
  "zooplankton"
)

get_species <- function(web) {
  sub("_.*", "", colnames(web)) |>
    unique() |>
    setdiff(resources)
}

data <- data |>
  mutate(
    species = purrr::map(foodweb, get_species),
    species_richness = purrr::map_dbl(species, length)
  )

data <- data |>
  arrange(salinity)

window <- 100
diversity_df <- purrr::map_dfr(
  seq_len(nrow(data) - window + 1),
  function(i) {
    w <- data[i:(i + window - 1), ]
    tibble(
      salinity = mean(w$salinity),
      gamma = n_distinct(unlist(w$species)),
      alpha = mean(w$species_richness)
    )
  }
) |>
  mutate(beta = gamma / alpha) |>
  tidyr::pivot_longer(cols = c(alpha, beta, gamma), names_to = "diversity")


ggplot(diversity_df, aes(salinity, value)) +
  geom_point() +
  geom_vline(
    xintercept = c(5, 18, 30),
    linetype = "dashed", colour = "grey40", linewidth = 0.5
  ) +
  facet_wrap("diversity", scale = "free") +
  labs(x = "Salinity [PSU]", y = "Species diversity")
```


```{r}
ggplot(data, aes(x = salinity)) +
  geom_histogram()
```




