---
title: "Overview of AMOBIO data"
format:
  html:
    theme: cosmo
    code-fold: true     # hidden but clickable to reveal, best of both worlds
    toc: true
    embed-resources: true
execute:
    message: false
    warning: false
---

Here, we investigate what is in the data of the AMOBIO project.

## Setup

```{r}
library(renv)
renv::load()
library(targets)
library(here)
tar_config_set(
  store = here::here("_targets")
)

library(ggplot2)
library(hrbrthemes)
theme_set(
  theme_ipsum() +
    theme(
      plot.background = element_rect(fill = "#FAFAF8", color = NA),
      panel.background = element_rect(fill = "#FAFAF8", color = NA),
      legend.position = "bottom"
    )
)
library(dplyr)
library(riversea)
library(sf)
```

## Stations mismatch

We begin by loading the files containing the information.

First, we have a file containing many metrics measured at many different points
that is, at different nodes.

```{r}
metric_file <- here("data", "amobio", "METRICS_AMOBIO.Rdata")
base::load(metric_file)
```

Second, we have a file containing geographical information about the nodes,
where they are and how they are connected to one another.

To begin with, we want to focus on the nodes corresponding to station where fish
were sampled.

```{r}
network_file <- here("data", "amobio", "NETWORK_AMOBIO.Rdata")
base::load(network_file)
nodes <- network_AMOBIO_uptate |>
  sfnetworks::activate("nodes") |>
  as_tibble() |>
  as_tibble()
```

```{r}
nodes <- nodes |>
  filter(IS_FISH == "FISH") |>
  select(node_id, geometry)

df_metric <- WP3_MATRIX_FULL |>
  mutate(node_id = node_id |> as.character() |> as.integer()) |>
  select(node_id, ID, POE_DB1_L4) |>
  mutate(metric = POE_DB1_L4, sandre_code = ID) |>
  select(-POE_DB1_L4) |>
  filter_out(is.na(metric))

base::load( # Load station_aspe object.
  here("data", "amobio", "station_operation_filtered_aspe.rda")
)

df_metric <- df_metric |>
  mutate(
    sandre_code = as.character(sandre_code)
    |> stringr::str_pad(8, pad = "0")
  )

sandre_code_amobio <- df_metric |>
  pull(sandre_code) |>
  unique()

station_aspe <- station_aspe |>
  mutate(in_amobio = sandre_code %in% sandre_code_amobio)
```

Then, we plot the stations on a map of France.

```{r}
crs <- st_crs(station_aspe)

coastline <- rnaturalearth::ne_coastline(scale = "large", returnclass = "sf") |>
  st_transform(crs)

boundaries <- st_bbox(st_as_sf(nodes))

ggplot() +
  geom_sf(data = coastline) +
  geom_sf(data = station_aspe, size = 0.8, color = "black", alpha = 0.7) +
  geom_sf(
    data = station_aspe |> mutate(in_amobio = if_else(in_amobio, "Yes", "No")),
    aes(color = in_amobio),
    size = 0.5
  ) +
  coord_sf(
    xlim = boundaries[c("xmin", "xmax")],
    ylim = boundaries[c("ymin", "ymax")]
  ) +
  labs(x = "Latitude", y = "Longitude", color = "In AMOBIO database?")
```


```{r}
base::load(here("data", "river", "station.rda"))
base::load(here("data", "river", "op.rda"))
head(op)

op |>
  mutate(id = station) |>
  left_join(
    station |> select(id, sandre_code),
    by = join_by(id)
  ) |>
  filter(sandre_code %in% sandre_code_amobio)

station |> filter(sandre_code %in% sandre_code_amobio)
station |> filter_out(sandre_code %in% sandre_code_amobio)
```



