---
title: "Overview of AMOBIO data"
format:
  html:
    embed-resources: true
execute:
    message: false
    warning: false
---

Here, we investigate what is in the data of the AMOBIO project.

```{r}
#| echo: false

library(renv)
renv::load()
library(targets)
library(here)
tar_config_set(
  store = here::here("_targets")
)

library(ggplot2)
library(hrbrthemes)
theme_set(theme_ipsum())
library(dplyr)
library(riversea)
library(sf)
```

We begin by loading the files containing the information.

First, we have a file containing many metrics measured at many different points
that is, at different nodes.

```{r}
metric_file <- here("data", "amobio", "METRICS_AMOBIO.Rdata")
base::load(metric_file)
df_metric <- WP3_MATRIX_FULL
```

Second, we have a file containing geographical information about the nodes,
where they are and how they are connected to one another.

To begin with, we want to focus on the nodes corresponding to station where fish
were sampled.

```{r}
network_file <- here("data", "amobio", "NETWORK_AMOBIO.Rdata")
base::load(network_file)
nodes <- network_AMOBIO_uptate |>
  sfnetworks::activate("nodes") |>
  as_tibble() |>
  as_tibble()


```

```{r}
df_metric |> 
  mutate(node_id = node_id |> as.character() |> as.integer()) |> 
  filter(node_id == 27)
```




```{r}
node_oe <- left_join(
  nodes,
  df_metric |> mutate(node_id = as.integer(as.character(node_id))), by = join_by(node_id)
) |>
  select(node_id, POE_DDAM_L2_DAM) |>
  mutate(metric = POE_DDAM_L2_DAM) |>
  filter_out(is.na(metric)) |>
  distinct()

str(node_oe)
```



Then, we plot the stations on a map of France.

```{r}
crs <- st_crs(node_oe)

coastline <- rnaturalearth::ne_coastline(scale = "large", returnclass = "sf") |>
  st_transform(crs)

boundaries <- st_bbox(node_oe)

ggplot() +
  geom_sf(data = coastline) +
  geom_sf(data = node_oe, aes(color = metric), size = 2) +
  coord_sf(
    xlim = boundaries[c("xmin", "xmax")],
    ylim = boundaries[c("ymin", "ymax")]
  ) +
  labs(x = NULL, y = NULL)
```

```{r}
fish_metrics <- nodes |>
  st_drop_geometry() |>
  left_join(
    df_metric |> mutate(node_id = as.integer(node_id)),
    by = join_by(node_id)
  )
```

```{r}
dam_nodes <- network_AMOBIO_uptate |>
  sfnetworks::activate("nodes") |>
  sf::st_as_sf() |>
  select(node_id, geometry, IS_OE) |>
  filter(IS_OE == "OE") |>
  select(-IS_OE)

ggplot() +
  geom_sf(data = coastline) +
  geom_sf(data = dam_nodes, size = 0.2, alpha  = 0.5) +
  coord_sf(
    xlim = boundaries[c("xmin", "xmax")],
    ylim = boundaries[c("ymin", "ymax")]
  ) +
  labs(x = NULL, y = NULL)
```


```{r}
dam_metrics <- dam_nodes |>
  st_drop_geometry() |>
  left_join(
    df_metric |> mutate(node_id = as.integer(node_id)),
    by = join_by(node_id)
  ) |>
  filter_out(is.na(POE_BH1_L4))
```

## Station mismatch

```{r}
station_bastien <- nodes |>
  mutate(sandre_code = FISH_ID)

base::load( # Load station_aspe object.
  here("data", "amobio", "station_operation_filtered_aspe.rda")
)

station_bastien |>
  st_drop_geometry() |>
  inner_join(
    station_aspe |> st_drop_geometry(),
    by = join_by(sandre_code)
  )
```

I wanted to identify the station that do not match between Alain's and Bastien's
data but I failed to retrieve the Sandre code in Bastien data set.
