---
title: "Overview of AMOBIO data"
format:
  html:
    theme: cosmo
    code-fold: true
    toc: true
    embed-resources: true
execute:
    message: false
    warning: false
---

Let's get an overview of the AMOBIO dataset.

## Setup

```{r}
library(renv)
renv::load()
library(targets)
library(here)
tar_config_set(
  store = here::here("_targets")
)

library(ggplot2)
library(hrbrthemes)
theme_set(
  theme_ipsum() +
    theme(
      plot.background = element_rect(fill = "#FAFAF8", color = NA),
      panel.background = element_rect(fill = "#FAFAF8", color = NA),
      legend.position = "bottom"
    )
)
library(dplyr)
library(riversea)
library(sf)
```

We have cleaned and filtered variables of interest in a single object that
contains our biological stations, their ID, geographical positions and the
metrics measuring the impacts of dam and nutrients.

```{r}
amobio_data <- tar_read(combined_amobio_data)
```

## Elevation

To begin with, let's plot our points and their elevation.

```{r}
#| fig-width: 5
#| fig-height: 7

riversea::plot_amobio_metric(
  amobio_data,
  "elevation",
  title = "Station elevations",
  subtitle = "Of stations from AMOBIO database.",
  legend = "Elevation [m]"
)
```


## Dam impacts

We can also visualise the impact of dams across our stations.

The impact of dam can be quantified with the distance to the closest dam.

```{r}
#| fig-width: 5
#| fig-height: 7

riversea::plot_amobio_metric(
  amobio_data,
  "poe_ddam_l2_dam",
  title = "Distance of closest dam",
  subtitle = "First upstream obstacle.",
  legend = "Distance [m]",
  f_transform = log10,
)
```

Or a measure of the barrier effect accounting for dam heights.

```{r}
#| fig-width: 5
#| fig-height: 7

riversea::plot_amobio_metric(
  amobio_data,
  "poe_bh5_l4",
  title = "Mean height",
  subtitle = "Of closest dams.",
  legend = "Height [m]",
  f_transform = log10,
)
```

## Nutrient concentrations

```{r}
#| fig-width: 5
#| fig-height: 7

riversea::plot_amobio_metric(
  amobio_data,
  "pc_n_ptotal_s1_1y",
  title = "Phosphorus concentration",
  subtitle = "Year 2019. Total phosporus concentration averaged\nover one year.",
  legend = "Concentration\n[microgram/L]",
  f_transform = log10,
  year = 2019
)
```


```{r}
#| fig-width: 5
#| fig-height: 7

riversea::plot_amobio_metric(
  amobio_data,
  "pc_n_nh4_s1_1y",
  title = "Ammonium concentration",
  subtitle = "Year 2019. Ammonium concentration averaged\nover one year.",
  legend = "Concentration\n[microgram/L]",
  f_transform = log10,
  year = 2019
)
```

```{r}
#| fig-width: 5
#| fig-height: 7

riversea::plot_amobio_metric(
  amobio_data,
  "pc_n_no2_s1_1y",
  title = "Nitrite concentration",
  subtitle = "Year 2019. Nitrite concentration averaged\nover one year.",
  legend = "Concentration\n[microgram/L]",
  f_transform = log10,
  year = 2019
)
```

```{r}
#| fig-width: 5
#| fig-height: 7

riversea::plot_amobio_metric(
  amobio_data,
  "pc_n_nkj_s1_1y",
  title = "Kjedahl nitrogen",
  subtitle = "Year 2019. Kjedahl nitrogen concentration averaged\nover one year.",
  legend = "Concentration\n[microgram/L]",
  f_transform = log10,
  year = 2019
)
```

```{r}
#| fig-width: 5
#| fig-height: 7

riversea::plot_amobio_metric(
  amobio_data,
  "pc_n_no3_s1_1y",
  title = "Nitrate concentration",
  subtitle = "Year 2019. Nitrate concentration averaged\nover one year.",
  legend = "Concentration\n[microgram/L]",
  f_transform = log10,
  year = 2019
)
```
