---
title: "Merge AMOBIO and ASPE data"
format:
  html:
    theme: cosmo
    code-fold: true
    toc: true
    embed-resources: true
execute:
    message: false
    warning: false
---

The goal is to merge our two datasets. On one side, we have the AMOBIO dataset
that contains information about the impact of dams and nutrients for the
biological stations. On the other hand, we have the ASPE dataset that contains
the information about the state of the fish community at the stations.

## Todo


- [x] Check join between AMOBIO and ASPE data
- [ ] Make a table of the number of points (once merged), and NAs per
environmental variables
- [ ] Check colinarity between variables
- [ ] Check for extreme values in environmental variables (in particular,
nutrients)
- [ ] Compute the distance to river mouth

## Setup

```{r}
library(renv)
renv::load()
library(targets)
library(here)
tar_config_set(
  store = here::here("_targets")
)
library(ggplot2)
library(hrbrthemes)
library(dplyr)
library(riversea)
library(sf)
riversea::nice_theme()
```

From ASPE data, we want to extract information about the food web structure, and
the fishing operation (ID, date, sandre code). Here is the structure of the
dataframe we get.

```{r}
web <- get(base::load(here("data", "river", "output_size2webs.rda")))
metrics <- web$tab_local_foodwebs_summary_metrics |>
  mutate(operation_id = as.integer(operation_id))

size <- get(base::load(here("data", "river", "output_individual_fish.rda")))
operation_sandre <- left_join(
  size$fishing_operation |> select(operation_id, site_id, date),
  size$station |> select(site_id, sandre_code),
  by = join_by(site_id)
)

aspe_data <- metrics |>
  left_join(operation_sandre, by = join_by(operation_id))

str(aspe_data)
```

## Map trophic structure

```{r}
#| fig-width: 5
#| fig-height: 7

amobio_data <- tar_read(combined_amobio_data)

data <- amobio_data |>
  inner_join(aspe_data, by = join_by(sandre_code, date))

riversea::plot_amobio_metric(
  data,
  "S",
  year = 2019,
  title = "Species richness",
  subtitle = "Year 2019.",
  legend = "S"
)

riversea::plot_amobio_metric(
  data,
  "C",
  year = 2019,
  title = "Connectance",
  subtitle = "Year 2019.",
  legend = "C"
)

riversea::plot_amobio_metric(
  data,
  "maxTL",
  year = 2019,
  title = "Max. trophic level",
  subtitle = "Year 2019.",
  legend = "max TL"
)
```

## Link food web structure to environment


```{r}
#| fig-width: 5
#| fig-height: 4

ggplot(data, aes(x = elevation, y = S)) +
  geom_point(alpha = 0.01) +
  labs(
    x = "Elevation [m]",
    y = "Species richness",
    title = "Richness and elevation",
    subtitle = "All years put together."
  )

ggplot(data, aes(x = log10(poe_ddam_l3_dam), y = S)) +
  geom_point(alpha = 0.05) +
  geom_smooth() +
  labs(
    x = "Distance to closest dam [km]",
    y = "Species richness",
    title = "Dam proximity and richness",
    subtitle = "Downstream distance. All years put together."
  )

ggplot(data, aes(x = log10(poe_ddam_l2_dam), y = S)) +
  geom_point(alpha = 0.05) +
  geom_smooth() +
  labs(
    x = "Distance to closest dam [km]",
    y = "Species richness",
    title = "Dam proximity and richness",
    subtitle = "Upstream distance. All years put together."
  )
```

## Statistical models


```{r}
library(INLA)

data_inla <- data |>
  st_drop_geometry() |>
  filter_out(poe_ddam_l2_dam == 0) |>
  mutate(
    year = as.integer(lubridate::year(date)),
    site_id = as.integer(sandre_code),
    log_dist_dam_up_std = scale(log10(poe_ddam_l3_dam)),
    log_dist_dam_down_std = scale(log10(poe_ddam_l2_dam)),
    barrier_up_std = scale(poe_bh5_l6),
    barrier_down_std = scale(poe_bh5_l5),
    log_phosphorus_std = scale(log10(pc_n_ptotal_s1_1y)),
    elevation_std = scale(elevation)
  ) |>
  select(
    S,
    year,
    site_id,
    log_dist_dam_up_std,
    log_dist_dam_down_std,
    barrier_up_std,
    barrier_down_std,
    log_phosphorus_std,
    elevation_std
  )


m_S <- inla(
  S ~ log_dist_dam_up_std + log_dist_dam_down_std +
    barrier_up_std + barrier_down_std +
    log_phosphorus_std +
    elevation_std +
    f(site_id, model = "iid") +
    f(year, model = "iid"),
  family = "nbinomial",
  data = data_inla
)

extract_fixed <- function(model, model_name) {
  model$summary.fixed |>
    tibble::rownames_to_column("term") |>
    filter(term != "(Intercept)") |>
    mutate(model = model_name)
}

fixed <- extract_fixed(m_S, "Richness")

ggplot(fixed, aes(x = mean, y = term)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey60") +
  geom_linerange(aes(xmin = `0.025quant`, xmax = `0.975quant`),
    position = position_dodge(width = 0.5)
  ) +
  geom_point(aes(x = `0.5quant`),
    size = 3,
    position = position_dodge(width = 0.5)
  ) +
  labs(x = "Effect size", y = NULL, color = NULL) +
  theme(legend.position = "bottom")
```


