---
title: "Merge AMOBIO and ASPE data"
format:
  html:
    theme: cosmo
    code-fold: true
    toc: true
    embed-resources: true
execute:
    message: false
    warning: false
---

We merge the AMOBIO and ASPE datasets. AMOBIO contains information about
environmental stressors, and in particular, nutrients and dam impacts. ASPE
contains information about local food web structure. By merging these two
datasets we can relate food web structure to environmental stressors.

## Todo

- [x] Check join between AMOBIO and ASPE data
- [x] Select variable of interest
- [x] Function to join AMOBIO and ASPE data from files
- [x] Improve figure theme
- [x] Make a table of the number of points (once merged), and NAs per
environmental variables
- [x] Check colinarity between variables
- [x] Check for extreme values in environmental variables (in particular,
nutrients)
- [x] Add regulatory nutrient thresholds
- [ ] Make an synthetic variabel for chemical compounds (if relevant) that would
be a proxy of eutrophication
- [ ] Update and improve spatial visualisation of structure metrics
- [ ] Update and improve statistical model

## Setup

```{r}
#| output: false
#| eval: false
library(renv)
renv::load()
```

```{r}
library(targets)
library(here)
tar_config_set(
  store = here::here("_targets")
)
library(ggplot2)
library(hrbrthemes)
library(dplyr)
library(sf)
devtools::load_all()
riversea::nice_theme()
```



Here is the merged dataframe, which contains information food web structure and
environmental stressors.

```{r}
data <- tar_read(amobio_aspe_data)
str(data)
```

## Data completeness

### Missing values

The merged dataset has `r nrow(data)` points. Here is the number and share of
missing values for each environmental variable, along with the range of
observed values.

```{r}
riversea::summarise_env_na(data) |>
  knitr::kable(digits = 2)
```

### Temporal coverage

Here is the the number of observation (removing NAs) per variable, by year.

```{r}
#| fig-width: 7
#| fig-height: 4

riversea::plot_env_temporal_coverage(data)
```

We see that we have good temporal coverage, in particular since 2005. We can
also note that we have a lower coverage in 2018.

## Collinearity between environmental variables

Before fitting statistical models, we check whether environmental variables
are correlated with each other, which would make it hard to separate their
individual effects. We subsample points to keep the pair plot readable and
fast to render.

```{r}
#| fig-width: 12
#| fig-height: 10

env_vars <- setdiff(names(riversea::rename_amobio()), c("date", "sandre_code"))

data |>
  st_drop_geometry() |>
  select(all_of(env_vars)) |>
  slice_sample(n = 5000) |>
  GGally::ggpairs(columns = env_vars)
```

## Extreme values

We check for extreme values in the nutrient variables (phosphorus and nitrogen
derivatives). Values are flagged as extreme using Tukey's fences (below `Q1 -
1.5 * IQR` or above `Q3 + 1.5 * IQR`), computed on the log10 scale.

```{r}
riversea::summarise_env_outliers(data) |>
  knitr::kable(digits = 3)
```

We have also extracted threshold data to have an idea when the nutrient
concentration is harmful for living organisms. The data is taken from
[here](https://rhin-meuse.eaufrance.fr/IMG/pdf/grilles-seq-eau-v2.pdf).

```{r}
seq_eau_file <- here("data", "river", "seq_eau_nutrient_thresholds.csv")
data_seq_eau <- readr::read_csv(seq_eau_file)
data_seq_eau |>
  knitr::kable(digits = 3)
```

Next, we plot the distribution of the nutrient concentrations and overlay the
threshold values corresponding to a "mauvais" water health state.

```{r}
#| fig-width: 7
#| fig-height: 5

riversea::plot_env_boxplot(data, thresholds = data_seq_eau)
```


## Map trophic structure

```{r}
#| fig-width: 5
#| fig-height: 7

riversea::plot_amobio_metric(
  data,
  "S",
  year = 2019,
  title = "Species richness",
  subtitle = "Year 2019",
  legend = "S"
)

riversea::plot_amobio_metric(
  data,
  "C",
  year = 2019,
  title = "Connectance",
  subtitle = "Year 2019.",
  legend = "C"
)

riversea::plot_amobio_metric(
  data,
  "maxTL",
  year = 2019,
  title = "Max. trophic level",
  subtitle = "Year 2019.",
  legend = "max TL"
)
```

## Link food web structure to environment


```{r}
#| fig-width: 5
#| fig-height: 4

ggplot(data, aes(x = elevation, y = S)) +
  geom_point(alpha = 0.01) +
  labs(
    x = "Elevation [m]",
    y = "Species richness",
    title = "Richness and elevation",
    subtitle = "All years put together."
  )

ggplot(data, aes(x = log10(dam_distance_downstream), y = S)) +
  geom_point(alpha = 0.05) +
  geom_smooth() +
  labs(
    x = "Distance to closest dam [km]",
    y = "Species richness",
    title = "Dam proximity and richness",
    subtitle = "Downstream distance. All years put together."
  )

ggplot(data, aes(x = log10(dam_distance_upstream), y = S)) +
  geom_point(alpha = 0.05) +
  geom_smooth() +
  labs(
    x = "Distance to closest dam [km]",
    y = "Species richness",
    title = "Dam proximity and richness",
    subtitle = "Upstream distance. All years put together."
  )
```

## Statistical models


```{r}
library(INLA)

data_inla <- data |>
  st_drop_geometry() |>
  filter_out(dam_distance_downstream == 0) |>
  mutate(
    year = as.integer(lubridate::year(date)),
    site_id = as.integer(sandre_code),
    log_dist_dam_up_std = scale(log10(dam_distance_upstream)),
    log_dist_dam_down_std = scale(log10(dam_distance_downstream)),
    barrier_up_std = scale(barrier_upstream),
    barrier_down_std = scale(barrier_downstream),
    log_phosphorus_std = scale(log10(total_phosphorus)),
    elevation_std = scale(elevation)
  ) |>
  select(
    S,
    year,
    site_id,
    log_dist_dam_up_std,
    log_dist_dam_down_std,
    barrier_up_std,
    barrier_down_std,
    log_phosphorus_std,
    elevation_std
  )


m_S <- inla(
  S ~ log_dist_dam_up_std + log_dist_dam_down_std +
    barrier_up_std + barrier_down_std +
    log_phosphorus_std +
    elevation_std +
    f(site_id, model = "iid") +
    f(year, model = "iid"),
  family = "nbinomial",
  data = data_inla
)

extract_fixed <- function(model, model_name) {
  model$summary.fixed |>
    tibble::rownames_to_column("term") |>
    filter(term != "(Intercept)") |>
    mutate(model = model_name)
}

fixed <- extract_fixed(m_S, "Richness")

ggplot(fixed, aes(x = mean, y = term)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey60") +
  geom_linerange(aes(xmin = `0.025quant`, xmax = `0.975quant`),
    position = position_dodge(width = 0.5)
  ) +
  geom_point(aes(x = `0.5quant`),
    size = 3,
    position = position_dodge(width = 0.5)
  ) +
  labs(x = "Effect size", y = NULL, color = NULL) +
  theme(legend.position = "bottom")
```


