---
title: "Overview of REPHY data"
format:
  html:
    theme: cosmo
    code-fold: true
    toc: true
    embed-resources: true
execute:
    message: false
    warning: false
---

We investigate the [REPHY database](https://www.seanoe.org/data/00361/47248/).
See [user manual](https://archimer.ifremer.fr/doc/00409/52017/52675.pdf).

## Todo

- [x] Download REPHY data
- [x] Read REPHY data
- [x] Clean and extract nutrients
- [x] Get an overview of the data - spatial and temporal coverage
- [x] Fix figure size
- [x] Fix NA coverage (either go wide, either remove na cols)
- [x] Plot time series of nutrient grouped by areas/regions
- [x] Plot seasonality of different nutrient variables
- [x] Check sampling methods
- [x] Plot raw time series of different nutrients
- [x] Check "niveau de qualité"
- [x] Remove "douteux" samples
- [x] Check levels in the water column
- [x] Filter only surface depth sampling

## Setup

```{r}
library(renv)
renv::load()
library(targets)
library(here)
tar_config_set(
  store = here::here("_targets")
)
library(ggplot2)
library(hrbrthemes)
library(dplyr)
library(sf)
devtools::load_all()
riversea::nice_theme()
```

We begin by reading the `.csv` file, which is not an easy step because of the
formatting issue. One it's done we can have a look at the data structure.

```{r}
raw_data <- tar_read(rephy_raw_data)
str(raw_data)
```

In particular, we can look at the what parameters are recorded.

```{r}
raw_data |>
  select("Résultat : Code paramètre", "Résultat : Libellé paramètre") |>
  distinct() |>
  print(n = 40)
```

Here are the parameters that are relevant for our study.

```{r}
raw_data |>
  select("Résultat : Code paramètre", "Résultat : Libellé paramètre") |>
  distinct() |>
  filter(`Résultat : Code paramètre` %in% rephy_nutrient_codes())
```

Next, we clean the table. Rename and select the relevant columns. Filter for the
parameters above.

```{r}
data <- tar_read(rephy_data)
str(data)
```

## Data completeness

For each nutrient parameter, here is the number of observations,
the number of distinct sampling sites, and the covered date range.

```{r}
riversea::summarise_rephy_coverage(data) |>
  knitr::kable(digits = 1)
```

We notice that we have more than twice as many points for the oxygen than for
the nutrients. The number of points for the other nutrients are closed to one
another, suggesting that when one nutrient is measured the others are too.

## Spatial coverage

We map the distinct sampling sites to get a sense of the spatial resolution
of the REPHY network.

```{r}
#| fig-width: 6
#| fig-height: 9.15

riversea::plot_rephy_map(data, facet = FALSE)
```

Grouping by parameter shows that spatial coverage is fairly consistent
across the four nutrients: sites are not more likely to be missing one
nutrient than another.

```{r}
#| fig-width: 8
#| fig-height: 13

riversea::plot_rephy_map(data, facet = TRUE)
```

Interstingly we see that concentration of nitrogen derivatives, ammonium and
nitrite + nitrate, are higher on Britany coast. This could be expected because
of agricultural nutrient runoff.

## Temporal coverage

Here is the number of observations per parameter, by year.

```{r}
#| fig-width: 7
#| fig-height: 3

riversea::plot_rephy_temporal_coverage(data)
```

Nitrate + nitrite (`NO3+NO2`) is only reported from 1998 onwards, and
sampling effort increases markedly from 2009 for all parameters. We should
keep this in mind when comparing periods.



## Yearly trends by district

To plot a yearly time series per nutrient we need a geographic grouping.
We reuse the DCE hydrographic districts already used for the river food web
(`coastal_water`, built from the SANDRE coastal/transitional water body
shapefiles), so that river and coastal analyses stay on the same spatial
units. Each site is joined to the district it falls in, dropping the few
sites too far from any polygon to be matched reliably.

```{r}
district_data <- tar_read(rephy_district_data)

sites <- data |>
  distinct(site_id, longitude, latitude) |>
  mutate(matched = site_id %in% district_data$site_id)

coastal_water <- tar_read(coastal_water) |>
  filter(district %in% unique(district_data$district))
```

```{r}
#| fig-width: 6
#| fig-height: 9

ggplot() +
  geom_sf(data = coastal_water, aes(fill = district), alpha = 0.5) +
  geom_point(data = sites, aes(x = longitude, y = latitude, color = matched)) +
  coord_sf(xlim = range(sites$longitude), ylim = range(sites$latitude)) +
  scale_color_manual(values = c(`TRUE` = "grey20", `FALSE` = "red")) +
  labs(fill = "District", color = "Matched", x = "Longitude", y = "Latitude") +
  theme(legend.box = "vertical")
```

`{r} nrow(data) - nrow(district_data)` points out of `{r} nrow(data)` could not
be matched to a district and were dropped. We then average nutrient
concentrations by year and district.

```{r}
yearly <- tar_read(rephy_yearly)
str(yearly)
```

```{r}
#| fig-width: 9
#| fig-height: 6

riversea::plot_rephy_temporal_trend(yearly |> filter(year >= 2000))
```

The Loire district stands out with clearly higher nitrate + nitrite and
phosphate concentrations, consistent with agricultural nutrient runoff from
the Loire basin. We filter early years because we have very few observations.

## Seasonal trends

### Distribution of overall sampling effort

```{r}
site_sampling_effort <- data |>
  group_by(site_id) |>
  summarise(sampling_effort = n()) |>
  arrange(desc(sampling_effort))

ggplot(site_sampling_effort, aes(x = sampling_effort)) +
  geom_histogram()
```

### Station most sampled

We see clear seasonal pattern with a clear peak of nitrites and nitrates after
summer that decreases in autumn. We also notice that ammonium seems to peak just
before, or long after nitrites and nitrates. This could be explained by the
lifecycle of nitrogen compounds (check references).

```{r}
site_most_point <- site_sampling_effort |>
  slice(1) |>
  pull(site_id) |>
  first()

data_most_point <- data |>
  filter(site_id == site_most_point) |>
  filter(lubridate::year(date) > 2015)

ggplot(data_most_point, aes(x = date, y = value, color = parameter_code)) +
  geom_point() +
  geom_line()
```

### Station sampled randomly

We sample random stations and plot their 2020 seasonal trend,
to get an idea of the sampling effort during the year, per variable.

```{r}
data_2020 <- data |>
  filter(lubridate::year(date) == 2020)

selected_sites <- data_2020 |>
  pull(site_id) |>
  unique() |>
  sample(9)

selected_data <- data_2020 |>
  filter(site_id %in% selected_sites)

ggplot(selected_data, aes(x = date, y = value, color = parameter_code)) +
  geom_point() +
  geom_line() +
  facet_wrap(~site_id, scales = "free")

ggplot(
  selected_data |> filter(parameter_code %in% c("NH4", "PO4")),
  aes(x = date, y = value, color = parameter_code)
) +
  geom_point() +
  geom_line() +
  facet_wrap(~site_id, scales = "free")
```

We can also have an idea of the average sampling effort per month per variable,
to assess whether some nutrients are sampled during specific period each year,
or not.

```{r}
sampling_per_month <- data |>
  mutate(month = as.factor(lubridate::month(date, label = TRUE))) |>
  left_join(
    data |>
      group_by(parameter_code) |>
      summarise(n_tot = n(), .groups = "drop"),
    by = join_by(parameter_code)
  ) |>
  group_by(parameter_code, month, n_tot) |>
  summarise(sampling_effort = n(), .groups = "drop") |>
  mutate(
    rel_sampling_effort = sampling_effort / n_tot
  )

ggplot(
  sampling_per_month,
  aes(x = month, y = rel_sampling_effort, color = parameter_code, group = parameter_code)
) +
  geom_point() +
  geom_line() +
  labs(color = "Parameter", x = "Month", y = "Average sampling effort")
```

The average sampling effort appears to be relatively even for nutrients.

## Sampling methods

### Overview over all stations

Raw yearly counts per method are dominated by the sampling effort trend
above, which makes method changes hard to spot. We instead look at the
share of observations per method each year, for each parameter.

```{r}
#| fig-width: 9
#| fig-height: 6

method_share <- riversea::summarise_rephy_method_share(raw_data) |>
  mutate(method = stringr::str_replace(
    method,
    "Spectrophotométrie",
    "Sp."
  )) |>
  filter(year >= 2000)

riversea::plot_rephy_method_share(
  method_share |>
    filter(parameter_code == "NH4")
)

riversea::plot_rephy_method_share(
  method_share |>
    filter(parameter_code == "NO3+NO2")
)

riversea::plot_rephy_method_share(
  method_share |>
    filter(parameter_code == "OXYGENE")
)

riversea::plot_rephy_method_share(
  method_share |>
    filter(parameter_code == "PO4")
)
```

We notice a change in measurement methods for all parameters around 2006.

### Focus on one station

We can try to see if this change of method is reflected in nutrient
concentration by looking at a station where the change occurred.

```{r}
site <- site_most_point

data_site <- raw_data |>
  filter(`Lieu de surveillance : Identifiant` == site)

method_share_site <- riversea::summarise_rephy_method_share(data_site) |>
  mutate(method = stringr::str_replace(
    method,
    "Spectrophotométrie",
    "Sp."
  )) |>
  filter(year >= 2000)

riversea::plot_rephy_method_share(method_share_site)
```

So for this station there is a clear change in sampling method. Let's see if it
is reflected in nutrient values.

```{r}
data_most_point <- data |>
  filter(site_id == site_most_point) |>
  filter(lubridate::year(date) >= 2003 & lubridate::year(date) <= 2012)

ggplot(data_most_point, aes(x = date, y = value, color = parameter_code)) +
  geom_point() +
  geom_line()
```

Luckily, we see that the change of methods doesn't seem to cause a change in
nutrient values.

## Sampling quality

We inspect that quality of the data collected across the year.

```{r}
raw_data |>
  summarise_rephy_quality_share() |>
  plot_rephy_quality_share()
```

We see that most has good quality. We remove samples whose quality is "douteux".

## Depth

Now we investigate the distrubtion of the depths at which variable are measured.

```{r}
data |>
  filter_out(is.na(level)) |>
  group_by(parameter_code, level) |>
  summarise(n = n()) |>
  arrange(parameter_code, desc(n)) |>
  print(n = 30)

ggplot(data |> filter_out(is.na(depth)), aes(x = depth)) +
  geom_histogram() +
  facet_wrap(~level)
```

We see that apart from oxygene variables are mainly measured at the surface.
That means we can focus on this depth, and remove other depths.

For more details, we can check how variable change with depth.

```{r}
ggplot(data, aes(x = depth, y = value)) +
  geom_point(alpha = 0.1) +
  facet_wrap(~parameter_code, scales = "free_y")
```

We see that oxygen is relatively constant meaning that using the surface
concentration is representative of the rest of the water column. We keep only
surface measurements in the following dataframe.

```{r}
str(tar_read(rephy_data_surface))
```

