---
title: "Overview of REPHY data"
format:
  html:
    theme: cosmo
    code-fold: true
    toc: true
    embed-resources: true
execute:
    message: false
    warning: false
---

We investigate the [REPHY database](https://www.seanoe.org/data/00361/47248/).
See [user manual](https://archimer.ifremer.fr/doc/00409/52017/52675.pdf).

## Todo

- [x] Download REPHY data
- [x] Read REPHY data
- [x] Clean and extract nutrients
- [x] Get an overview of the data - spatial and temporal coverage
- [x] Fix figure size
- [x] Fix NA coverage (either go wide, either remove na cols)
- [x] Plot time series of nutrient grouped by areas/regions
- [ ] Convert units so they are consistent between REPHY and AMOBIO
- [ ] Compute NO3+NO2 in AMOBIO
- [ ] Check for consistency between datasets in nutrient and oxygen value
- [ ] Also extract O2 in AMOBIO
- [ ] Interpolate + yearly average at fishing sites

## Setup

```{r}
library(renv)
renv::load()
library(targets)
library(here)
tar_config_set(
  store = here::here("_targets")
)
library(ggplot2)
library(hrbrthemes)
library(dplyr)
library(sf)
devtools::load_all()
riversea::nice_theme()
```

We begin by reading the `.csv` file, which is not an easy step because of the
formatting issue. One it's done we can have a look at the data structure.

```{r}
raw_data <- tar_read(rephy_raw_data)
str(raw_data)
```

In particular, we can look at the what parameters are recorded.

```{r}
raw_data |>
  select("Résultat : Code paramètre", "Résultat : Libellé paramètre") |>
  distinct() |>
  print(n = 40)
```

Here are the parameters that are relevant for our study.

```{r}
raw_data |>
  select("Résultat : Code paramètre", "Résultat : Libellé paramètre") |>
  distinct() |>
  filter(`Résultat : Code paramètre` %in% rephy_nutrient_codes())
```

Next, we clean the table. Rename and select the relevant columns. Filter for the
parameters above.

```{r}
data <- tar_read(rephy_data)
str(data)
```

## Data completeness

For each nutrient parameter, here is the number of observations,
the number of distinct sampling sites, and the covered date range.

```{r}
riversea::summarise_rephy_coverage(data) |>
  knitr::kable(digits = 1)
```

We notice that we have more than twice as many points for the oxygen than for
the nutrients. The number of points for the other nutrients are closed to one
another, suggesting that when one nutrient is measured the others are too.

## Spatial coverage

We map the distinct sampling sites to get a sense of the spatial resolution
of the REPHY network.

```{r}
#| fig-width: 6
#| fig-height: 9.15

riversea::plot_rephy_map(data, facet = FALSE)
```

Grouping by parameter shows that spatial coverage is fairly consistent
across the four nutrients: sites are not more likely to be missing one
nutrient than another.

```{r}
#| fig-width: 8
#| fig-height: 13

riversea::plot_rephy_map(data, facet = TRUE)
```

Interstingly we see that concentration of nitrogen derivatives, ammonium and
nitrite + nitrate, are higher on Britany coast. This could be expected because
of agricultural nutrient runoff.

## Temporal coverage

Here is the number of observations per parameter, by year.

```{r}
#| fig-width: 7
#| fig-height: 3

riversea::plot_rephy_temporal_coverage(data)
```

Nitrate + nitrite (`NO3+NO2`) is only reported from 1998 onwards, and
sampling effort increases markedly from 2009 for all parameters. We should
keep this in mind when comparing periods.

## Temporal trends by district

To plot a yearly time series per nutrient we need a geographic grouping.
We reuse the DCE hydrographic districts already used for the river food web
(`hydrographic_area`, built from the SANDRE coastal/transitional water body
shapefiles), so that river and coastal analyses stay on the same spatial
units. Each site is joined to the district it falls in, dropping the few
sites too far from any polygon to be matched reliably.

```{r}
district_data <- tar_read(rephy_district_data)

sites <- data |>
  distinct(site_id, longitude, latitude) |>
  mutate(matched = site_id %in% district_data$site_id)

hydro_area <- tar_read(hydrographic_area) |>
  filter(district %in% unique(district_data$district))
```

```{r}
#| fig-width: 6
#| fig-height: 9

ggplot() +
  geom_sf(data = hydro_area, aes(fill = district),  alpha = 0.5) +
  geom_point(data = sites, aes(x = longitude, y = latitude, color = matched)) +
  coord_sf(xlim = range(sites$longitude), ylim = range(sites$latitude)) +
  scale_color_manual(values = c(`TRUE` = "grey20", `FALSE` = "red")) +
  labs(fill = "District", color = "Matched", x = "Longitude", y = "Latitude") +
  theme(legend.box = "vertical")
```

`{r} nrow(data) - nrow(district_data)` points out of `{r} nrow(data)` could not
be matched to a district and were dropped. We then average nutrient
concentrations by year and district.

```{r}
yearly <- tar_read(rephy_yearly)
str(yearly)
```

```{r}
#| fig-width: 9
#| fig-height: 6

riversea::plot_rephy_temporal_trend(yearly |> filter(year >= 2000))
```

The Loire district stands out with clearly higher nitrate + nitrite and
phosphate concentrations, consistent with agricultural nutrient runoff from
the Loire basin. We filter early years because we have very few observations.

