---
title: "Interpolation of REPHY data"
bibliography: ../references.bib
format:
  html:
    theme: cosmo
    code-fold: true
    toc: true
    embed-resources: true
execute:
    message: false
    warning: false
---

The goal is to interpolate REPHY data -- nutrient and oxygen concentrations --
when and where fishing operation occured. It means interpolate the variables of
interest in space and time. We intend to do so using the Bayesian framework
INLA.

## Todo

- [x] Setup notebook
- [x] Plot REPHY stations and fishing operations together (space and time)
- [x] Fit additive spatio-temporal SPDE model (space + year + cyclic month) for
one parameter
- [x] Leave-station-out cross-validation for PO4
- [x] Diagnose limited CV skill (outliers? under-sampled folds?)
- [x] Write interpolation model
- [x] Extend to the other three parameters
- [x] Diagnose model performence on oxygen
- [x] Diagnose model performence on nitrite and nitrate
- [x] Diagnose model performence on ammonium
- [x] Compare interpolation methods with cross-validation
- [x] Plot model performence on yearly average
- [x] Check spatial structure of residuals
- [x] Add salinity as a covariable
- [ ] Check interpolation performence without log scale
- [x] Inteprolate value at fishing operations
- [ ] Remove REPHY stations too uncertain to interpolate reliably from

## Setup

```{r setup}
library(renv)
renv::load()
library(targets)
library(here)
tar_config_set(
  store = here::here("_targets")
)
library(ggplot2)
library(dplyr)
library(sf)
library(patchwork)
devtools::load_all()
riversea::nice_theme()
knitr::opts_chunk$set(autodep = TRUE)
```

## Presentation of the data

Here is the REPHY data that we have collected and cleaned. 

```{r rephy-data}
rephy_data <- tar_read(rephy_data_surface)
str(rephy_data)
```

For more information see [Overview of REPHY data](D6-rephy.qmd). And, here is
the fishing operation data.

```{r foodwebs-data}
foodwebs <- tar_read(foodweb_structure) |>
  dplyr::filter(survey %in% c("nurse", "pomet"))
```

This interpolation exercise is only concerned with sea-survey operations
(`nurse` and `pomet`), not river-survey ones -- REPHY monitors marine and
estuarine chemistry, which has no bearing hundreds of kilometers up a river.

Before interpolating anything, we start with a simple picture of where and when
the two datasets overlap: REPHY nutrient and oxygen stations on one side, fishing
operations from the food web data on the other.

### Spatial coverage

```{r spatial-coverage-plot}
#| fig-width: 6
#| fig-height: 9

sites <- dplyr::bind_rows(
  rephy_data |>
    dplyr::distinct(longitude, latitude) |>
    dplyr::mutate(source = "REPHY station"),
  foodwebs |>
    dplyr::distinct(longitude, latitude) |>
    dplyr::mutate(source = "Fishing operation")
)

ggplot(sites, aes(x = longitude, y = latitude, color = source)) +
  geom_point() +
  geom_sf(data = coastline(), inherit.aes = FALSE, linewidth = 0.3) +
  coord_sf(xlim = range(sites$longitude), ylim = range(sites$latitude)) +
  labs(x = NULL, y = NULL, color = NULL) +
  theme(panel.grid = element_blank())
```

Fishing operations are much more numerous and spatially diffuse than REPHY
stations, which are fixed monitoring points. Interpolation will need to be based
on REPHY stations that are reasonably close to fishing operations.

### Temporal coverage

```{r temporal-coverage-plot}
#| fig-width: 7
#| fig-height: 3

yearly_counts <- dplyr::bind_rows(
  rephy_data |>
    dplyr::mutate(year = lubridate::year(date)) |>
    dplyr::count(year, name = "n") |>
    dplyr::mutate(source = "REPHY station"),
  foodwebs |>
    dplyr::count(year, name = "n") |>
    dplyr::mutate(source = "Fishing operation")
)

ggplot(yearly_counts, aes(x = year, y = n, fill = source)) +
  geom_col(position = "identity") +
  scale_x_continuous(breaks = scales::breaks_pretty()) +
  labs(x = NULL, y = "Number of observations", fill = NULL)
```

Both datasets have good overlap from 2000 onwards, which is where we should
expect the interpolation to be best supported.

## Interpolation model

Each REPHY parameter is interpolated with an additive spatio-temporal model,
fit with INLA: a spatial Matérn field (via the SPDE approximation), a smooth
year trend, and a cyclic month effect to capture seasonality. Values are
modelled on the log scale, since concentrations are strictly positive and
right-skewed. 

$$
\log(\text{nutrient}) = \text{SPDE}_\text{2D} + \text{year}_\text{RW} 
                       + \text{month}_\text{RW}
$$

We restrict the data to 2000 onward.

```{r rephy-filtered}
rephy_filtered <- tar_read(rephy_data_sample)
```

### Spatial mesh

We build a mesh around REPHY stations and fishing operations. Its boundary
follows the coastline (a non-convex hull) rather than the convex hull of the
points, which would otherwise includes the land between the Atlantic and Channel
clusters. It's a `targets` pipeline target (`rephy_mesh`), not built here, so
it's reusable from other notebooks without recomputing it.

```{r mesh-plot}
mesh <- tar_read(rephy_mesh)
plot(mesh, main = "")
plot(sf::st_geometry(coastline()), add = TRUE, col = "grey60")
points(
  foodwebs$longitude,
  foodwebs$latitude,
  col = "coral",
  pch = 19,
  cex = 0.5
)
points(
  rephy_filtered$longitude,
  rephy_filtered$latitude,
  col = "cyan3",
  pch = 19,
  cex = 0.5
)
legend(
  "bottomright",
  legend = c("REPHY station", "Fishing operation"),
  col = c("cyan3", "coral"),
  pch = 19,
  bty = "n"
)
```

This is the mesh the model is going to work with. Next, we fit first model and
evaluate them with cross-validation.

### Cross-validation

```{r n-fold}
n_fold <- 5
```

We evaluate the model on each of the four parameters with leave-station-out
k-fold cross-validation. Stations are split into `r n_fold` groups, and each
group is predicted from a model trained on the other stations only. This
tests spatial generalization and doesn't let a station's other-year data leak
into the test fold.

We compare our model performence (RMSE, MAE and correlation) against a naive
baseline prediction: the training fold's mean.

```{r cv-baseline}
#| cache: true

set.seed(1)
cv_all <- purrr::map_dfr(unique(rephy_filtered$parameter_code), \(parameter) {
  d_param <- rephy_filtered |> dplyr::filter(parameter_code == parameter)
  cv <- cv_rephy_spde(d_param, mesh, k = n_fold)
  cv$parameter_code <- parameter
  cv
})

cv_all |>
  dplyr::group_by(parameter_code) |>
  dplyr::group_modify(\(cv, ...) {
    metrics <- rephy_cv_metrics(cv)
    baseline <- rephy_cv_metrics(cv, pred_col = "pred_baseline")
    data.frame(
      model = c("SPDE", "Naive baseline"),
      rmse = c(metrics$rmse, baseline$rmse),
      mae = c(metrics$mae, baseline$mae),
      cor = c(metrics$cor, baseline$cor)
    )
  }) |>
  dplyr::ungroup() |>
  knitr::kable(digits = 2)
```

Across parameters, the SPDE model has lower log-scale RMSE than the naive
baseline and correlates positively with observed values, where the baseline
is essentially uncorrelated by construction. The spatio-temporal structure is
clearly adding information, even if there's room to improve.

```{r cv-baseline-plot}
#| fig-width: 8
#| fig-height: 8

cv_all <- cv_all |>
  dplyr::mutate(
    highlight = dplyr::case_when(
      parameter_code == "PO4" & site_id == 22044027 & year %in% 2011:2013 ~
        "Pont St Maurice (2011-2013)",
      parameter_code == "OXYGENE" & value < 2 * 1.42903 ~ "Hypoxia (< 2 ml/L)",
      TRUE ~ "Other"
    )
  )

ggplot(cv_all, aes(x = value, y = pred, color = highlight)) +
  geom_point(alpha = 0.4) +
  geom_abline(slope = 1, intercept = 0) +
  scale_x_log10() +
  scale_y_log10() +
  scale_color_manual(
    values = c(
      "Other" = "grey40",
      "Pont St Maurice (2011-2013)" = "#d95f02",
      "Hypoxia (< 2 ml/L)" = "#7570b3"
    )
  ) +
  facet_wrap(~parameter_code, scales = "free") +
  labs(x = "Observed", y = "Predicted (held-out station)", color = NULL)
```

The highlighted points, on the PO4 panel, show an anomaly at the station
Pont St Maurice, that cannot be predicted by its nearest neighbour. This
shows that this point cannot be predicted by a smooth model as they are very
localized in space and time. Excluding these, we see that the predictions of
our model are quite reasonnable.

On the OXYGENE panel, the highlighted points are hypoxia events (< 2 ml/L, i.e.
< 2.86 mg/L @Diaz-Rosenberg-2008) unpredictable from neighbouring stations when
the station itself is held out.

We will use yearly averages rather than raw values in the next modelling steps,
we also plot the same predictions averaged per year (observed and predicted).

```{r cv-yearly-plot}
#| fig-width: 8
#| fig-height: 8

cv_yearly <- cv_all |>
  dplyr::summarise(
    value = mean(value),
    pred = mean(pred),
    .by = c(parameter_code, site_id, site_label, year)
  ) |>
  dplyr::mutate(
    highlight = dplyr::case_when(
      parameter_code == "PO4" & site_id == 22044027 & year %in% 2011:2013 ~
        "Pont St Maurice (2011-2013)",
      parameter_code == "OXYGENE" & value < 2 * 1.42903 ~ "Hypoxia (< 2 ml/L)",
      TRUE ~ "Other"
    )
  )

ggplot(cv_yearly, aes(x = value, y = pred, color = highlight)) +
  geom_point(alpha = 0.4) +
  geom_abline(slope = 1, intercept = 0) +
  scale_x_log10() +
  scale_y_log10() +
  scale_color_manual(
    values = c(
      "Other" = "grey40",
      "Pont St Maurice (2011-2013)" = "#d95f02",
      "Hypoxia (< 2 ml/L)" = "#7570b3"
    )
  ) +
  facet_wrap(~parameter_code, scales = "free") +
  labs(
    x = "Observed (yearly mean)", 
    y = "Predicted (yearly mean, held-out station)", 
    color = NULL
  )
```

We see that mechanically averaging per year remove most of the noise. On yearly
average our model performs more than fine so we will use this interpolation
model to get environmental values at our fishing operations. But before doing
so, we will check the structure of the model residuals.

### Residual structure

#### Temporal structure

We check whether the CV residuals still carry spatial or temporal structure. 

```{r residual-temporal-plot}
#| fig-width: 8
#| fig-height: 4

cv_resid <- cv_all |>
  dplyr::filter(value > 0) |>
  dplyr::mutate(resid = log(pred) - log(value))

year_resid <- cv_resid |>
  dplyr::summarise(resid = mean(resid), .by = c(parameter_code, year))
month_resid <- cv_resid |>
  dplyr::summarise(resid = mean(resid), .by = c(parameter_code, month))

ggplot(year_resid, aes(x = year, y = resid)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_line() +
  geom_point() +
  facet_wrap(~parameter_code, scales = "free_y") +
  labs(x = NULL, y = "Mean log-residual")

ggplot(month_resid, aes(x = month, y = resid)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = 1:12) +
  facet_wrap(~parameter_code, scales = "free_y") +
  labs(x = "Month", y = "Mean log-residual")
```

All four parameters show a large residual change in 2000-2003 that then flattens
out. it looks like a coverage artifact of the year random walk rather than a
missed trend. Month residuals are flat everywhere: the cyclic month term
captures seasonality well.

#### Spatial structure

```{r residual-spatial-plot}
#| fig-width: 10
#| fig-height: 8

station_resid <- cv_resid |>
  dplyr::summarise(
    resid = mean(resid),
    n = dplyr::n(),
    .by = c(parameter_code, site_id, longitude, latitude)
  )

ggplot(station_resid, aes(x = longitude, y = latitude, color = resid, size = n)) +
  geom_point(alpha = 1) +
  geom_sf(data = coastline(), inherit.aes = FALSE, linewidth = 0.3) +
  coord_sf(
    xlim = range(station_resid$longitude), ylim = range(station_resid$latitude)
  ) +
  scale_color_gradient2(
    low = "steelblue",
    mid = "grey90",
    high = "firebrick",
    midpoint = 0
  ) +
  facet_wrap(~parameter_code) +
  labs(x = NULL, y = NULL, color = "Mean log-resid", size = "n") +
  theme(panel.grid = element_blank())
```

There is no broad geographic gradient, but a set of estuarine stations are
consistently under-predicted. This point to an unmodeled driver of local
nutrient-loading that could bias our analysis afterwards. For that reason, we
should certainly include distance to the river mouth as covariate to better
predict nutrient concentrations at those sites.

## Refining the model

### Salinity

REPHY and some fishing operations also measures salinity at most stations, so we
can use it to inform our predictions. In particular this may help with the
understimation issue we observe in estuaries.

```{r salinity-coverage-table}
rephy_filtered |>
  dplyr::summarise(
    n = dplyr::n(),
    n_salinity = sum(!is.na(salinity)),
    pct_salinity = round(100 * mean(!is.na(salinity)), 1),
    .by = parameter_code
  ) |>
  knitr::kable()
```

Coverage is close to complete for all four parameters. We next plot the
relationship between our parameters and salinity.

```{r salinity-scatter-plot}
#| fig-width: 8
#| fig-height: 8

sal_data <- rephy_filtered |> dplyr::filter(!is.na(salinity), value > 0)

ggplot(sal_data, aes(x = salinity, y = value)) +
  geom_point(alpha = 0.2, size = 0.5) +
  scale_y_log10() +
  facet_wrap(~parameter_code, scales = "free_y") +
  labs(x = "Salinity", y = "Value (log scale)")
```

We see a strong relationship, and it's clearly not a straight line on the
log scale -- so we compare the baseline model against three ways of adding
salinity: a linear term, a quadratic term, and a smooth term (a P-spline:
a cubic B-spline basis with an RW2 smoothness penalty on its coefficients)
to capture that curvature.

$$
\log(\text{nutrient}) = \text{SPDE}_\text{2D} + \text{year}_\text{RW}
                       + \text{month}_\text{RW} + \beta \cdot \text{salinity}
$$

$$
\log(\text{nutrient}) = \text{SPDE}_\text{2D} + \text{year}_\text{RW}
                       + \text{month}_\text{RW} + \beta_1 \cdot \text{salinity}
                       + \beta_2 \cdot \text{salinity}^2
$$

$$
\log(\text{nutrient}) = \text{SPDE}_\text{2D} + \text{year}_\text{RW}
                       + \text{month}_\text{RW} + \text{salinity}_\text{P-spline}
$$

::: {.callout-note}
We use linear scale for salinity in the model, not logged. We tested
`log(salinity)` and found that raw salinity performed better on every parameter.
:::

We compare all four models with the same leave-station-out cross-validation
as before, on the same data (samples with a matched salinity value), so the
only thing that differs between them is the covariate.

```{r cv-salinity-comparison}
#| cache: true

rephy_sal <- rephy_filtered |> dplyr::filter(!is.na(salinity))

set.seed(1)
cv_base_sal <- purrr::map_dfr(unique(rephy_sal$parameter_code), \(parameter) {
  d_param <- rephy_sal |>
    dplyr::filter(parameter_code == parameter)
  cv <- cv_rephy_spde(d_param, mesh, k = n_fold)
  cv$parameter_code <- parameter
  cv
})

set.seed(1)
cv_salinity <- purrr::map_dfr(unique(rephy_sal$parameter_code), \(parameter) {
  d_param <- rephy_sal |>
    dplyr::filter(parameter_code == parameter)
  cv <- cv_rephy_spde(
    d_param, mesh,
    predict_fn = predict_rephy_spde_salinity, k = n_fold
  )
  cv$parameter_code <- parameter
  cv
})
```

```{r cv-salinity-quadratic}
#| cache: true

set.seed(1)
cv_salinity_quadratic <- purrr::map_dfr(unique(rephy_sal$parameter_code), \(parameter) {
  d_param <- rephy_sal |>
    dplyr::filter(parameter_code == parameter)
  cv <- cv_rephy_spde(
    d_param, mesh,
    predict_fn = predict_rephy_spde_salinity_quadratic, k = n_fold
  )
  cv$parameter_code <- parameter
  cv
})
```

```{r cv-salinity-smooth}
#| cache: true

set.seed(1)
cv_salinity_smooth <- purrr::map_dfr(unique(rephy_sal$parameter_code), \(parameter) {
  d_param <- rephy_sal |>
    dplyr::filter(parameter_code == parameter)
  cv <- cv_rephy_spde(
    d_param, mesh,
    predict_fn = predict_rephy_spde_salinity_smooth, k = n_fold
  )
  cv$parameter_code <- parameter
  cv
})
```

To check that the smooth term isn't just overfitting we also compute WAIC from a
single full-data fit per model.

```{r salinity-criteria}
#| cache: true

rephy_criteria <- purrr::map_dfr(unique(rephy_sal$parameter_code), \(parameter) {
  d_param <- rephy_sal |> dplyr::filter(parameter_code == parameter)
  dplyr::bind_rows(
    rephy_model_criteria(predict_rephy_spde(d_param, d_param, mesh)) |>
      dplyr::mutate(model = "Baseline"),
    rephy_model_criteria(predict_rephy_spde_salinity(d_param, d_param, mesh)) |>
      dplyr::mutate(model = "Salinity (linear)"),
    rephy_model_criteria(predict_rephy_spde_salinity_quadratic(d_param, d_param, mesh)) |>
      dplyr::mutate(model = "Salinity (quadratic)"),
    rephy_model_criteria(predict_rephy_spde_salinity_smooth(d_param, d_param, mesh)) |>
      dplyr::mutate(model = "Salinity (smooth)")
  ) |>
    dplyr::mutate(parameter_code = parameter)
})
```

```{r salinity-comparison-table}
dplyr::bind_rows(
  cv_base_sal |>
    dplyr::group_by(parameter_code) |>
    dplyr::group_modify(\(cv, ...) as.data.frame(rephy_cv_metrics(cv))) |>
    dplyr::mutate(model = "Baseline"),
  cv_salinity |>
    dplyr::group_by(parameter_code) |>
    dplyr::group_modify(\(cv, ...) as.data.frame(rephy_cv_metrics(cv))) |>
    dplyr::mutate(model = "Salinity (linear)"),
  cv_salinity_quadratic |>
    dplyr::group_by(parameter_code) |>
    dplyr::group_modify(\(cv, ...) as.data.frame(rephy_cv_metrics(cv))) |>
    dplyr::mutate(model = "Salinity (quadratic)"),
  cv_salinity_smooth |>
    dplyr::group_by(parameter_code) |>
    dplyr::group_modify(\(cv, ...) as.data.frame(rephy_cv_metrics(cv))) |>
    dplyr::mutate(model = "Salinity (smooth)")
) |>
  dplyr::ungroup() |>
  dplyr::left_join(
    rephy_criteria |> dplyr::select(parameter_code, model, waic),
    by = c("parameter_code", "model")
  ) |>
  dplyr::mutate(is_best = waic == min(waic), .by = parameter_code) |>
  dplyr::arrange(parameter_code, model) |>
  dplyr::transmute(
    parameter_code,
    model,
    rmse = sprintf("%.2f", rmse),
    mae = sprintf("%.2f", mae),
    waic = sprintf("%.1f", waic),
    is_best
  ) |>
  dplyr::mutate(
    dplyr::across(
      c(model, rmse, mae, waic),
      \(x) dplyr::if_else(is_best, paste0("<strong>", x, "</strong>"), x)
    )
  ) |>
  dplyr::select(-is_best) |>
  knitr::kable(escape = FALSE)
```

Smooth salinity is the best model (lowest WAIC) for NH4, NO3+NO2 and PO4,
beating linear and quadratic salinity as well as the baseline For oxygen,
salinity doesn't help in any form, so we drop it there and keep the baseline.

#### Marginal salinity effect

To check that the P-spline is actually behaving like a smooth function of
salinity (and not just overfitting), we plot its marginal effect.

```{r salinity-marginal-plot}
#| fig-width: 10
#| fig-height: 4

spline_curve <- function(fit_obj) {
  boundary <- attr(fit_obj$bs_basis, "Boundary.knots")
  salinity <- seq(boundary[1], boundary[2], length.out = 100)
  mean <- predict(fit_obj$bs_basis, salinity) %*% fit_obj$fit$summary.random$spline$mean
  data.frame(salinity = salinity, mean = as.vector(mean))
}

salinity_marginal <- dplyr::bind_rows(
  "NH4" = spline_curve(tar_read(rephy_fit_nh4)),
  "NO3+NO2" = spline_curve(tar_read(rephy_fit_no3_no2)),
  "PO4" = spline_curve(tar_read(rephy_fit_po4)),
  .id = "parameter_code"
)

ggplot(salinity_marginal, aes(x = salinity, y = mean)) +
  geom_line() +
  facet_wrap(~parameter_code, scales = "free_y") +
  labs(x = "Salinity", y = "Effect (log scale)")
```

The curves are smooth throughout -- an earlier RW1-over-binned-salinity version
of this same model produced a visibly noisier curve, which is why we moved to
the P-spline.

### Spatial residuals

The residual-structure map above showed a set of estuarine stations
consistently under-predicted. We rebuild the same map, baseline vs. smooth
salinity (the model we're keeping), on the same station set so the two
rows are directly comparable.

```{r cv-salinity-residual-map}
#| fig-width: 10
#| fig-height: 10

resid_compare <- dplyr::bind_rows(
  cv_base_sal |> dplyr::mutate(model = "Baseline"),
  cv_salinity_smooth |> dplyr::mutate(model = "+ Salinity (smooth)")
) |>
  dplyr::filter(value > 0) |>
  dplyr::mutate(resid = log(pred) - log(value)) |>
  dplyr::summarise(
    resid = mean(resid),
    n = dplyr::n(),
    .by = c(model, parameter_code, site_id, longitude, latitude)
  )

ggplot(resid_compare, aes(x = longitude, y = latitude, color = resid, size = n)) +
  geom_point(alpha = 1) +
  geom_sf(data = coastline(), inherit.aes = FALSE, linewidth = 0.3) +
  coord_sf(
    xlim = range(resid_compare$longitude), ylim = range(resid_compare$latitude)
  ) +
  scale_color_gradient2(
    low = "steelblue",
    mid = "grey90",
    high = "firebrick",
    midpoint = 0
  ) +
  facet_grid(model ~ parameter_code) +
  labs(x = NULL, y = NULL, color = "Mean log-resid", size = "n") +
  theme(panel.grid = element_blank())
```

We see a clear improvement of the model in the estuaries. Next, we move to
predicting nutrient values at our fishing operations.

## Predicting at fishing operations

This is our goal. Predict nutrient values at every `nurse` and
`pomet` fishing operation. We use the smooth-salinity model for NH4, NO3+NO2 and
PO4 -- OXYGENE is dropped. Each parameter is fit once on the full
salinity-complete REPHY dataset, then predicted at every operation. `pomet`
operations use their own measured salinity where available; the rest (all
`nurse` and 4% of `pomet`) fall back to a space-time-only prediction for that
operation.

```{r foodweb-predictions}
#| cache: true

foodweb_predictions <- purrr::map_dfr(c("NH4", "NO3+NO2", "PO4"), \(parameter) {
  train <- rephy_sal |> dplyr::filter(parameter_code == parameter)
  pred <- predict_rephy_spde_salinity_smooth(train, foodwebs, mesh)
  foodwebs |>
    dplyr::transmute(operation_id, survey, longitude, latitude, year, month, salinity) |>
    dplyr::mutate(parameter_code = parameter, pred = pred, sd = attr(pred, "sd"))
})
```

```{r foodweb-predictions-map}
#| fig-width: 12
#| fig-height: 8

plot_operation_map <- function(parameter, value_col, legend) {
  d <- foodweb_predictions |> dplyr::filter(parameter_code == parameter)
  ggplot(d, aes(longitude, latitude, color = .data[[value_col]])) +
    geom_point(size = 0.6, alpha = 0.7) +
    geom_sf(data = coastline(), inherit.aes = FALSE, linewidth = 0.3) +
    coord_sf(
      xlim = range(foodweb_predictions$longitude),
      ylim = range(foodweb_predictions$latitude)
    ) +
    scale_color_viridis_c(trans = if (value_col == "pred") "log10" else "identity") +
    labs(x = NULL, y = NULL, color = legend, title = parameter) +
    theme(panel.grid = element_blank())
}

patchwork::wrap_plots(
  lapply(c("NH4", "NO3+NO2", "PO4"), plot_operation_map, value_col = "pred", legend = NULL),
  nrow = 1
)
```

```{r foodweb-predictions-table}
foodweb_predictions |>
  dplyr::summarise(
    n = dplyr::n(),
    median = round(median(pred), 2),
    p10 = round(quantile(pred, 0.1), 2),
    p90 = round(quantile(pred, 0.9), 2),
    .by = c(parameter_code, survey)
  ) |>
  dplyr::arrange(parameter_code, survey) |>
  knitr::kable()
```

Predictions are consistently higher at `pomet` operations than `nurse` across
all three parameters (e.g. NO3+NO2 median 22.2 vs. 3.5) -- exactly what we'd
expect given `pomet` reaches deep into estuaries while `nurse` stays coastal.

### Prediction uncertainty

We next investigate the uncertainty of our predictions.

```{r foodweb-predictions-uncertainty-map}
#| fig-width: 12
#| fig-height: 8

patchwork::wrap_plots(
  lapply(
    c("NH4", "NO3+NO2", "PO4"), plot_operation_map,
    value_col = "sd", legend = "log-scale SD"
  ),
  nrow = 1
)
```

We note that uncertainty is higher when reaching far upstream, as the model as
less data points there to learn, and so make more uncertain predictions. The
good things with bayesian model is that we can actually propagate this
uncertainty. Instead of making a producing a point-estimate, we will sample the
posterior probability repeatidly and feed this to the downstream model that will
actually work on food web structure.

A quick demonstration, 200 draws for PO4 at 5 operations, from the cached fit:

```{r posterior-draws-demo}
#| cache: true
#| fig-width: 5
#| fig-height: 3.5

fit_po4 <- tar_read(rephy_fit_po4)
draws_demo <- sample_rephy_spde_smooth(
  fit_po4,
  foodwebs |> dplyr::slice_sample(n = 5),
  n_draws = 200
)

ggplot(data.frame(draw = draws_demo[1, ]), aes(draw)) +
  geom_histogram(bins = 30) +
  labs(x = "PO4, posterior draws for one operation", y = NULL) +
  theme(panel.grid = element_blank())
```

