---
title: "Interpolation of REPHY data"
format:
  html:
    theme: cosmo
    code-fold: true
    toc: true
    embed-resources: true
execute:
    message: false
    warning: false
---

The goal is to interpolate REPHY data -- nutrient and oxygen concentrations --
when and where fishing operation occured. It means interpolate the variables of
interest in space and time. We intend to do so using the Bayesian framework
INLA.

## Todo

- [x] Setup notebook
- [x] Plot REPHY stations and fishing operations together (space and time)
- [x] Fit additive spatio-temporal SPDE model (space + year + cyclic month) for
one parameter
- [x] Leave-station-out cross-validation for PO4
- [x] Diagnose limited CV skill (outliers? under-sampled folds?)
- [ ] Extend to the other three parameters
- [ ] Compare interpolation methods with cross-validation

## Setup

```{r}
library(renv)
renv::load()
library(targets)
library(here)
tar_config_set(
  store = here::here("_targets")
)
library(ggplot2)
library(hrbrthemes)
library(dplyr)
library(sf)
devtools::load_all()
riversea::nice_theme()
```

## Presentation of the data

Here is the REPHY data that we have collected and cleaned. 

```{r}
rephy_data <- tar_read(rephy_data_surface)
str(rephy_data)
```

For more information see [Overview of REPHY data](D6-rephy.qmd). And, here is
the fishing operation data.

```{r}
foodwebs <- tar_read(local_foodwebs)
```

Before interpolating anything, we start with a simple picture of where and when
the two datasets overlap: REPHY nutrient and oxygen stations on one side, fishing
operations from the food web data on the other.

### Spatial coverage

```{r}
#| fig-width: 6
#| fig-height: 9

sites <- dplyr::bind_rows(
  rephy_data |>
    dplyr::distinct(longitude, latitude) |>
    dplyr::mutate(source = "REPHY station"),
  foodwebs |>
    dplyr::distinct(longitude, latitude) |>
    dplyr::mutate(source = "Fishing operation")
)

ggplot(sites, aes(x = longitude, y = latitude, color = source)) +
  geom_point() +
  geom_sf(data = coastline(), inherit.aes = FALSE, linewidth = 0.3) +
  coord_sf(xlim = range(sites$longitude), ylim = range(sites$latitude)) +
  labs(x = NULL, y = NULL, color = NULL)
```

Fishing operations are much more numerous and spatially diffuse than REPHY
stations, which are fixed monitoring points. Interpolation will need to be based
on REPHY stations that are reasonably close to fishing operations.

### Temporal coverage

```{r}
#| fig-width: 7
#| fig-height: 3

yearly_counts <- dplyr::bind_rows(
  rephy_data |>
    dplyr::mutate(year = lubridate::year(date)) |>
    dplyr::count(year, name = "n") |>
    dplyr::mutate(source = "REPHY station"),
  foodwebs |>
    dplyr::count(year, name = "n") |>
    dplyr::mutate(source = "Fishing operation")
)

ggplot(yearly_counts, aes(x = year, y = n, fill = source)) +
  geom_col(position = "identity") +
  scale_x_continuous(breaks = scales::breaks_pretty()) +
  labs(x = NULL, y = "Number of observations", fill = NULL)
```

Both datasets have good overlap from 2000 onwarrds, which is where we should
expect the interpolation to be best supported.

## Interpolation model

Each REPHY parameter is interpolated with an additive spatio-temporal model,
fit with INLA: a spatial Matérn field (via the SPDE approximation), a smooth
year trend, and a cyclic month effect to capture seasonality. Values are
modelled on the log scale, since concentrations are strictly positive and
right-skewed. We restrict to 2000 onward.

```{r}
rephy_filtered <- tar_read(rephy_data_recent)
```

### Spatial mesh

We build a single mesh, shared across parameters, from all filtered REPHY
stations. Its boundary follows the coastline (a non-convex hull) rather than
the convex hull of the stations, which would otherwise span the land between
the Atlantic and Channel clusters.

```{r}
mesh <- build_rephy_mesh(rephy_filtered)
plot(mesh, main = "")
points(
  foodwebs$longitude,
  foodwebs$latitude,
  col = "coral",
  pch = 19,
  cex = 0.5
)
points(
  rephy_filtered$longitude,
  rephy_filtered$latitude,
  col = "cyan3",
  pch = 19,
  cex = 0.5
)
legend(
  "bottomright",
  legend = c("REPHY station", "Fishing operation"),
  col = c("coral", "cyan3"),
  pch = 19,
  bty = "n"
)
```

### Cross-validation: phosphate

We evaluate the model on phosphate with leave-station-out k-fold
cross-validation: stations are split into 10 groups, and each group is predicted
from a model trained on the other stations only. This tests spatial
generalization and doesn't let a station's other-year data leak into its own
training set.

RMSE and MAE in isolation don't mean much. They need a reference to be
interpretable. For that reason, we compare our model resuts against a naive
baseline prediction: the training fold's geometric mean.

```{r}
po4 <- rephy_filtered |> dplyr::filter(parameter_code == "PO4")

set.seed(1)
cv <- cv_rephy_spde(po4, mesh, k = 10)
metrics <- rephy_cv_metrics(cv)
baseline <- rephy_cv_metrics(cv, pred_col = "pred_baseline")

data.frame(
  model = c("SPDE", "Naive baseline"),
  rmse = c(metrics$rmse, baseline$rmse),
  mae = c(metrics$mae, baseline$mae),
  cor = c(metrics$cor, baseline$cor)
) |>
  knitr::kable(digits = 2)
```

The SPDE model cuts log-scale RMSE by about a quarter relative to the naive
baseline, and correlates with observed values (`{r} round(metrics$cor, 2)`)
where the baseline is essentially uncorrelated (`{r} round(baseline$cor, 2)`).
The spatio-temporal structure is clearly adding information, even
if there's room to improve.

```{r}
#| fig-width: 6
#| fig-height: 6

cv <- cv |>
  dplyr::mutate(
    highlight = ifelse(
      site_id == 22044027 & year %in% 2011:2013,
      "Pont St Maurice (2011-2013)", "Other"
    )
  )

ggplot(cv, aes(x = value, y = pred, color = highlight)) +
  geom_point(alpha = 0.4) +
  geom_abline(slope = 1, intercept = 0) +
  scale_x_log10() +
  scale_y_log10() +
  scale_color_manual(
    values = c("Other" = "grey40", "Pont St Maurice (2011-2013)" = "#d95f02")
  ) +
  labs(x = "Observed PO4", y = "Predicted PO4 (held-out station)", color = NULL)
```

The highlighted points show an anomaly at the station Pont St Maurice, that
cannot be predicted by its nearest neighbour. This shows that this point cannot
be predicted by a smooth model as they are very localized in space and time.
Excluding these, we see that the predictions of our model are quite reasonnable.
