---
title: "Interpolation of REPHY data"
bibliography: ../references.bib
format:
  html:
    theme: cosmo
    code-fold: true
    toc: true
    embed-resources: true
execute:
    message: false
    warning: false
---

The goal is to interpolate REPHY data -- nutrient and oxygen concentrations --
when and where fishing operation occured. It means interpolate the variables of
interest in space and time. We intend to do so using the Bayesian framework
INLA.

## Todo

- [x] Setup notebook
- [x] Plot REPHY stations and fishing operations together (space and time)
- [x] Fit additive spatio-temporal SPDE model (space + year + cyclic month) for
one parameter
- [x] Leave-station-out cross-validation for PO4
- [x] Diagnose limited CV skill (outliers? under-sampled folds?)
- [x] Write interpolation model
- [x] Extend to the other three parameters
- [x] Diagnose model performence on oxygen
- [x] Diagnose model performence on nitrite and nitrate
- [x] Diagnose model performence on ammonium
- [x] Compare interpolation methods with cross-validation
- [x] Plot model performence on yearly average
- [x] Check spatial structure of residuals
- [ ] Add salinity as a covariable
- [ ] Do a multivariate interpolation instead of predicting variables separately
- [ ] Check interpolation performence without log scale
- [ ] **(waiting for distance-to-river mouth)** Inteprolate value at fishing
operations, remove stations too uncertain

## Setup

```{r}
library(renv)
renv::load()
library(targets)
library(here)
tar_config_set(
  store = here::here("_targets")
)
library(ggplot2)
library(dplyr)
library(sf)
devtools::load_all()
riversea::nice_theme()
```

## Presentation of the data

Here is the REPHY data that we have collected and cleaned. 

```{r}
rephy_data <- tar_read(rephy_data_surface)
str(rephy_data)
```

For more information see [Overview of REPHY data](D6-rephy.qmd). And, here is
the fishing operation data.

```{r}
foodwebs <- tar_read(local_foodwebs)
```

Before interpolating anything, we start with a simple picture of where and when
the two datasets overlap: REPHY nutrient and oxygen stations on one side, fishing
operations from the food web data on the other.

### Spatial coverage

```{r}
#| fig-width: 6
#| fig-height: 9

sites <- dplyr::bind_rows(
  rephy_data |>
    dplyr::distinct(longitude, latitude) |>
    dplyr::mutate(source = "REPHY station"),
  foodwebs |>
    dplyr::distinct(longitude, latitude) |>
    dplyr::mutate(source = "Fishing operation")
)

ggplot(sites, aes(x = longitude, y = latitude, color = source)) +
  geom_point() +
  geom_sf(data = coastline(), inherit.aes = FALSE, linewidth = 0.3) +
  coord_sf(xlim = range(sites$longitude), ylim = range(sites$latitude)) +
  labs(x = NULL, y = NULL, color = NULL)
```

Fishing operations are much more numerous and spatially diffuse than REPHY
stations, which are fixed monitoring points. Interpolation will need to be based
on REPHY stations that are reasonably close to fishing operations.

### Temporal coverage

```{r}
#| fig-width: 7
#| fig-height: 3

yearly_counts <- dplyr::bind_rows(
  rephy_data |>
    dplyr::mutate(year = lubridate::year(date)) |>
    dplyr::count(year, name = "n") |>
    dplyr::mutate(source = "REPHY station"),
  foodwebs |>
    dplyr::count(year, name = "n") |>
    dplyr::mutate(source = "Fishing operation")
)

ggplot(yearly_counts, aes(x = year, y = n, fill = source)) +
  geom_col(position = "identity") +
  scale_x_continuous(breaks = scales::breaks_pretty()) +
  labs(x = NULL, y = "Number of observations", fill = NULL)
```

Both datasets have good overlap from 2000 onwarrds, which is where we should
expect the interpolation to be best supported.

## Interpolation model

Each REPHY parameter is interpolated with an additive spatio-temporal model,
fit with INLA: a spatial Matérn field (via the SPDE approximation), a smooth
year trend, and a cyclic month effect to capture seasonality. Values are
modelled on the log scale, since concentrations are strictly positive and
right-skewed. 

$$
\log(\text{nutrient}) = \text{SPDE}_\text{2D} + \text{year}_\text{RW} 
                       + \text{month}_\text{RW}
$$

We restrict the data to 2000 onward.

```{r}
rephy_filtered <- tar_read(rephy_data_sample)
```

### Spatial mesh

We build a mesh, shared across parameters, from all filtered REPHY stations. Its
boundary follows the coastline (a non-convex hull) rather than the convex hull
of the stations, which would otherwise span the land between the Atlantic and
Channel clusters.

```{r}
mesh <- build_rephy_mesh(rephy_filtered)
plot(mesh, main = "")
points(
  foodwebs$longitude,
  foodwebs$latitude,
  col = "coral",
  pch = 19,
  cex = 0.5
)
points(
  rephy_filtered$longitude,
  rephy_filtered$latitude,
  col = "cyan3",
  pch = 19,
  cex = 0.5
)
legend(
  "bottomright",
  legend = c("REPHY station", "Fishing operation"),
  col = c("coral", "cyan3"),
  pch = 19,
  bty = "n"
)
```

### Cross-validation

```{r}
n_fold <- 5
```

We evaluate the model on each of the four parameters with leave-station-out
k-fold cross-validation. Stations are split into `r n_fold` groups, and each
group is predicted from a model trained on the other stations only. This
tests spatial generalization and doesn't let a station's other-year data leak
into the test fold.

We compare our model performence (RMSE, MAE and correlation) against a naive
baseline prediction: the training fold's mean.

```{r}
set.seed(1)
cv_all <- purrr::map_dfr(unique(rephy_filtered$parameter_code), \(parameter) {
  d_param <- rephy_filtered |> dplyr::filter(parameter_code == parameter)
  cv <- cv_rephy_spde(d_param, mesh, k = n_fold)
  cv$parameter_code <- parameter
  cv
})

cv_all |>
  dplyr::group_by(parameter_code) |>
  dplyr::group_modify(\(cv, ...) {
    metrics <- rephy_cv_metrics(cv)
    baseline <- rephy_cv_metrics(cv, pred_col = "pred_baseline")
    data.frame(
      model = c("SPDE", "Naive baseline"),
      rmse = c(metrics$rmse, baseline$rmse),
      mae = c(metrics$mae, baseline$mae),
      cor = c(metrics$cor, baseline$cor)
    )
  }) |>
  dplyr::ungroup() |>
  knitr::kable(digits = 2)
```

Across parameters, the SPDE model has lower log-scale RMSE than the naive
baseline and correlates positively with observed values, where the baseline
is essentially uncorrelated by construction. The spatio-temporal structure is
clearly adding information, even if there's room to improve.

```{r}
#| fig-width: 8
#| fig-height: 8

cv_all <- cv_all |>
  dplyr::mutate(
    highlight = dplyr::case_when(
      parameter_code == "PO4" & site_id == 22044027 & year %in% 2011:2013 ~
        "Pont St Maurice (2011-2013)",
      parameter_code == "OXYGENE" & value < 2 * 1.42903 ~ "Hypoxia (< 2 ml/L)",
      TRUE ~ "Other"
    )
  )

ggplot(cv_all, aes(x = value, y = pred, color = highlight)) +
  geom_point(alpha = 0.4) +
  geom_abline(slope = 1, intercept = 0) +
  scale_x_log10() +
  scale_y_log10() +
  scale_color_manual(
    values = c(
      "Other" = "grey40",
      "Pont St Maurice (2011-2013)" = "#d95f02",
      "Hypoxia (< 2 ml/L)" = "#7570b3"
    )
  ) +
  facet_wrap(~parameter_code, scales = "free") +
  labs(x = "Observed", y = "Predicted (held-out station)", color = NULL)
```

The highlighted points, on the PO4 panel, show an anomaly at the station
Pont St Maurice, that cannot be predicted by its nearest neighbour. This
shows that this point cannot be predicted by a smooth model as they are very
localized in space and time. Excluding these, we see that the predictions of
our model are quite reasonnable.

On the OXYGENE panel, the highlighted points are hypoxia events (< 2 ml/L, i.e.
< 2.86 mg/L @Diaz-Rosenberg-2008) unpredictable from neighbouring stations when
the station itself is held out.

We will use yearly averages rather than raw values in the next modelling steps,
we also plot the same predictions averaged per year (observed and predicted).

```{r}
#| fig-width: 8
#| fig-height: 8

cv_yearly <- cv_all |>
  dplyr::summarise(
    value = mean(value),
    pred = mean(pred),
    .by = c(parameter_code, site_id, site_label, year)
  ) |>
  dplyr::mutate(
    highlight = dplyr::case_when(
      parameter_code == "PO4" & site_id == 22044027 & year %in% 2011:2013 ~
        "Pont St Maurice (2011-2013)",
      parameter_code == "OXYGENE" & value < 2 * 1.42903 ~ "Hypoxia (< 2 ml/L)",
      TRUE ~ "Other"
    )
  )

ggplot(cv_yearly, aes(x = value, y = pred, color = highlight)) +
  geom_point(alpha = 0.4) +
  geom_abline(slope = 1, intercept = 0) +
  scale_x_log10() +
  scale_y_log10() +
  scale_color_manual(
    values = c(
      "Other" = "grey40",
      "Pont St Maurice (2011-2013)" = "#d95f02",
      "Hypoxia (< 2 ml/L)" = "#7570b3"
    )
  ) +
  facet_wrap(~parameter_code, scales = "free") +
  labs(
    x = "Observed (yearly mean)", 
    y = "Predicted (yearly mean, held-out station)", 
    color = NULL
  )
```

We see that mechanically averaging per year remove most of the noise. On yearly
average our model performs more than fine so we will use this interpolation
model to get environmental values at our fishing operations. But before doing
so, we will check the structure of the model residuals.

### Residual structure

#### Temporal structure

We check whether the CV residuals still carry spatial or temporal structure. 

```{r}
#| fig-width: 8
#| fig-height: 4

cv_resid <- cv_all |>
  dplyr::filter(value > 0) |>
  dplyr::mutate(resid = log(pred) - log(value))

year_resid <- cv_resid |> dplyr::summarise(resid = mean(resid), .by = c(parameter_code, year))
month_resid <- cv_resid |> dplyr::summarise(resid = mean(resid), .by = c(parameter_code, month))

ggplot(year_resid, aes(x = year, y = resid)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_line() +
  geom_point() +
  facet_wrap(~parameter_code, scales = "free_y") +
  labs(x = NULL, y = "Mean log-residual")

ggplot(month_resid, aes(x = month, y = resid)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = 1:12) +
  facet_wrap(~parameter_code, scales = "free_y") +
  labs(x = "Month", y = "Mean log-residual")
```

All four parameters show a large residual change in 2000-2003 that then flattens
out. it looks like a coverage artifact of the year random walk rather than a
missed trend. Month residuals are flat everywhere: the cyclic month term
captures seasonality well.

#### Spatial structure

```{r}
#| fig-width: 10
#| fig-height: 8

station_resid <- cv_resid |>
  dplyr::summarise(
    resid = mean(resid),
    n = dplyr::n(),
    .by = c(parameter_code, site_id, longitude, latitude)
  )

ggplot(station_resid, aes(x = longitude, y = latitude, color = resid, size = n)) +
  geom_point(alpha = 1) +
  geom_sf(data = coastline(), inherit.aes = FALSE, linewidth = 0.3) +
  coord_sf(
    xlim = range(station_resid$longitude), ylim = range(station_resid$latitude)
  ) +
  scale_color_gradient2(
    low = "steelblue",
    mid = "grey90",
    high = "firebrick",
    midpoint = 0
  ) +
  facet_wrap(~parameter_code) +
  labs(x = NULL, y = NULL, color = "Mean log-resid", size = "n")
```

There is no broad geographic gradient, but a set of estuarine stations are
consistently under-predicted. This point to an unmodeled driver of local
nutrient-loading that could bias our analysis afterwards. For that reason, we
should certainly include distance to the river mouth as covariate to better
predict nutrient concentrations at those sites.
