---
title: "Compute distance to river mouth"
bibliography: ../references.bib
format:
  html:
    theme: cosmo
    code-fold: true
    toc: true
    embed-resources: true
execute:
    message: false
    warning: false
---

The goal is to compute the distance to river mouth REPHY data and fishing
operations at sea.

## Todo

- [x] Extract river mouth points
- [x] Plot river mouth points
- [x] Simplify AMOBIO network
- [x] Filter river mouths
- [x] Restrict river network
- [x] Check rephy data overlap with restricted river network
- [x] Reconstruct full river network for `river` survey stations
- [x] Snap `river` survey stations to the network
- [x] Classify `river` survey stations by true catchment
- [x] Close the `Ardour-Garonne` under-classification gap
- [x] Investigate why the Vendée cluster is disconnected
- [x] Patch the missing link (Nantes/Sèvre Nantaise) and confirm the
cluster recovers
- [x] Snap directly against the restricted network now that it's fixed
- [x] Check quality of snapping, how many non-snapped stations and where?
- [x] Find about Somme classification, as it doesn't appear in the current
cateogries
- [x] Keep edge linestring geometry in `extract_river_network()`
(currently dropped for straight node-to-node lines) - check if it
recovers some non-snapped stations
- [ ] Clean notebook, reorganise, snapping and rephy are mixed
- [ ] Plot catchments
- [ ] Merge catchments (if needed) to cover land and coast
- [ ] Check connected components vs. catchments - is one catchment one
connected component?
- [ ] Compute distance to river mouth for `river` survey stations

## Setup

```{r}
library(renv)
renv::load()
library(targets)
library(here)
tar_config_set(
  store = here::here("_targets")
)
library(ggplot2)
library(dplyr)
library(sf)
library(patchwork)
devtools::load_all()
riversea::nice_theme()
```

```{r}
river_mouth <- tar_read(amobio_river_mouth)
```

## Snapping river survey stations to the network

### Catchments

We focus on four catchments: Loire, Ardour-Garonne, Seine and Escaut-Somme. We
attribute stations to catchment using the BD Topage's polygones delimiting
hydrographic basins.

```{r}
network_river <- tar_read(river_network)
river_stations <- tar_read(inland_operations_snapped) |>
  filter(survey == "river")
basin <- tar_read(basin)
district_kept <- c("Loire", "Ardour-Garonne", "Seine", "Escaut-Somme")

cols <- c(
  Loire = "steelblue",
  "Ardour-Garonne" = "firebrick",
  Seine = "darkgreen",
  "Escaut-Somme" = "orange"
)

basin_kept <- basin |> filter(district %in% district_kept)
in_scope <- river_stations |> filter(district %in% district_kept)
river_coords <- network_river$nodes |>
  sf::st_as_sf() |>
  sf::st_transform(4326) |>
  sf::st_coordinates()
```

We can visualise the catchments of interest as wells the corresponding stations.

```{r}
#| fig-width: 7
#| fig-height: 9

plot(st_geometry(basin), col = NA, border = "grey80")
plot(
  st_geometry(basin_kept),
  col = scales::alpha(cols[basin_kept$district], 0.15),
  border = "grey70",
  add = TRUE
)
points(
  river_coords[, "X"],
  river_coords[, "Y"],
  col = scales::alpha("grey60", 0.25),
  pch = ".",
  cex = 0.5
)
points(
  in_scope$longitude,
  in_scope$latitude,
  col = cols[in_scope$district],
  pch = 19,
  cex = 0.3
)
legend("bottomright", legend = names(cols), col = cols, pch = 19, bty = "n")
```

We can have a look at how many stations we have per catchments

```{r}
count(river_stations, district) |>
  filter(district %in% district_kept) |>
  arrange(n) |>
  knitr::kable()
```

Next, we evaluate the snapping quality for those catchments.

### Evaluate snapping quality

We snap fish-monitoring stations to the network hydrographic network. Because
snapping requires precision, we use the network with line geometry
`river_network_geo`.  We can evaluate the snapping quality through the snapping
distance, that is, how far the station was from the closest point in the river
network. Short distances indicate confident snapping.

```{r}
stations_selected <- river_stations |>
  filter(district %in% district_kept)

ggplot(stations_selected, aes(x = snap_dist_m)) +
  geom_histogram(fill = "#a9c9e0", color = "white", bins = 40) +
  geom_vline(xintercept = 150) +
  annotate(
    "text",
    x = 150, y = Inf, label = "150m", hjust = -0.2, vjust = 1.5
  ) +
  scale_x_continuous(trans = "log10", labels = scales::label_number()) +
  labs(x = "Snapping distance (m)", y = "Number of stations")
```

```{r}
threshold <- 150
n_bad_snap <- sum(stations_selected$snap_dist_m > threshold)
```

We see that the vast majority of stations are well-snapped to the river network.
In particular, we have only with `r n_bad_snap` a snapping distance greater than
`r threshold` meters.

### Adding a network missing edge (Nantes)

While investigating, one component near the Vendée coast appeared disconnected
from the rest of the Loire basin, despite sitting inside it. That led this
component to be disconnected from any river mouth. After checking, we found
two points that were unlinked but suspiciously close to one another, so we
decided to investigate where they were.

```{r}
river_network_full <- tar_read(river_network_full)
network_before <- river_network_full
network_before$edges <- network_before$edges |>
  filter(!(from == 614458 & to == 614552))

comp_before <- igraph::components(igraph::graph_from_edgelist(
  as.matrix(network_before$edges[c("from", "to")]),
  directed = FALSE
))
vendee_comp_nodes <- which(
  comp_before$membership == comp_before$membership[595001]
)
main_sf <- river_network_full$nodes |>
  filter(
    node_id %in% which(comp_before$membership == which.max(comp_before$csize))
  ) |>
  sf::st_as_sf()
vendee_sf <- river_network_full$nodes |>
  filter(node_id %in% vendee_comp_nodes) |>
  sf::st_as_sf()

gap_dist_m <- as.numeric(sf::st_distance(
  vendee_sf, main_sf[sf::st_nearest_feature(vendee_sf, main_sf), ],
  by_element = TRUE
))

gap_pts <- river_network_full$nodes |>
  filter(node_id %in% c(614458, 614552)) |>
  sf::st_as_sf() |>
  sf::st_transform(4326) |>
  mutate(label = c("Vendée-side node", "Main-network node"))

leaflet::leaflet(gap_pts) |>
  leaflet::addProviderTiles(leaflet::providers$Esri.WorldTopoMap) |>
  leaflet::addCircleMarkers(
    color = c("firebrick", "steelblue"),
    radius = 8,
    label = ~label
  ) |>
  leaflet::setView(
    lng = mean(sf::st_coordinates(gap_pts)[, 1]),
    lat = mean(sf::st_coordinates(gap_pts)[, 2]),
    zoom = 15
  )
```


These points represent the confluence between the Loire and the Sèvre Nantaise.
Restoring this missing link allowed us to recover the entire connected
component. This operation is documented in `patch_river_network()`.

## Distance to river mouth

Here we have everything to compute the distance to the river mouth for river
stations at least.

### River operations

We group our stations per catchment, and identify for each catchment transitions
point between freshwater and marine areas.

```{r}
river_mouth_district <- tar_read(river_mouth_district)
mouths_kept <- river_mouth_district |>
  filter(matched, district %in% district_kept)
mouth_coords <- sf::st_coordinates(mouths_kept)
```

```{r}
#| fig-width: 7
#| fig-height: 9

plot(st_geometry(basin), col = NA, border = "grey80")
plot(
  st_geometry(basin_kept),
  col = scales::alpha(cols[basin_kept$district], 0.15),
  border = "grey70",
  add = TRUE
)
points(
  river_coords[, "X"],
  river_coords[, "Y"],
  col = scales::alpha("grey60", 0.25),
  pch = ".",
  cex = 0.5
)
points(
  in_scope$longitude,
  in_scope$latitude,
  col = cols[in_scope$district],
  pch = 19,
  cex = 0.3
)
points(
  mouth_coords[, "X"], mouth_coords[, "Y"],
  col = "black", pch = 17, cex = 0.8
)
legend(
  "bottomright",
  legend = c(names(cols), "River mouth"),
  col = c(cols, "black"),
  pch = c(rep(19, length(cols)), 17),
  bty = "n"
)
```

We then compute for each station the channel-length distance to the closest
river mouth of its own catchment (so a station can only be matched to a mouth of
its own catchment - see `compute_distance_to_mouth()`).

```{r}
river_distance <- tar_read(inland_distance_to_mouth) |>
  filter(survey == "river")

ggplot(river_distance, aes(x = dist_mouth_m / 1000, fill = district)) +
  geom_histogram(bins = 30, color = "white") +
  scale_fill_manual(values = cols, guide = "none") +
  facet_wrap(~district, scales = "free_y") +
  labs(x = "Distance to river mouth (km)", y = "Number of stations")
```

```{r}
river_distance |>
  group_by(district) |>
  summarise(
    n = n(),
    median_km = round(median(dist_mouth_m) / 1000, 1),
    max_km = round(max(dist_mouth_m) / 1000, 1)
  ) |>
  arrange(median_km) |>
  knitr::kable()
```

```{r}
#| fig-width: 7
#| fig-height: 9.5

river_coords_df <- as.data.frame(river_coords)
mouth_coords_df <- as.data.frame(mouth_coords)

ggplot() +
  geom_sf(data = basin, fill = NA, color = "grey80") +
  geom_sf(
    data = basin_kept, aes(fill = district), color = "grey70", alpha = 0.10
  ) +
  scale_fill_manual(values = cols, name = NULL) +
  geom_point(
    data = river_distance,
    aes(longitude, latitude, color = dist_mouth_m / 1000),
    size = 0.4
  ) +
  scale_color_viridis_c(name = "Distance to river mouth (km)") +
  geom_point(data = mouth_coords_df, aes(X, Y), shape = 17, size = 1) +
  coord_sf() +
  labs(x = NULL, y = NULL) +
  guides(
    fill = guide_legend(nrow = 1, override.aes = list(alpha = 0.3)),
    color = guide_colorbar(barwidth = 10)
  ) +
  theme(legend.box = "vertical", panel.grid = element_blank())
```

Distances range from a few tens of meters (stations right at the mouth) to
under 950km. Escaut-Somme has the shortest distances overall (small,
coastal catchment), while Seine stations sit farthest inland on average.

### Sea operations

#### Inland vs. sea-side

For sea operations, we first need to know whether an operation sits
upstream or downstream of the river mouth before a distance is even
meaningful. One idea: use `basin` (land catchment) vs.
`coastal_water` (coastal/transitional water body) - if a point falls
in the basin polygon, it's inland; if in the coastal water polygon,
sea-side.

To validate our choice let's have a look at the Gironde and Loire estuaries.

```{r}
#| fig-width: 12
#| fig-height: 7

coastal_water <- tar_read(coastal_water)

plot_estuary <- function(name, xlim, ylim) {
  bbox <- st_as_sfc(st_bbox(
    c(xmin = xlim[1], xmax = xlim[2], ymin = ylim[1], ymax = ylim[2]),
    crs = 4326
  ))
  basin_e <- st_intersection(st_make_valid(basin), bbox)
  area_e <- st_intersection(st_make_valid(coastal_water), bbox)

  ggplot() +
    geom_sf(
      data = basin_e, aes(fill = "Basin (catchment)"), color = NA, alpha = 0.4
    ) +
    geom_sf(
      data = area_e, aes(fill = "Area (coastal water body)"),
      color = NA, alpha = 0.4
    ) +
    geom_point(data = mouth_coords_df, aes(X, Y), shape = 17, size = 1) +
    scale_fill_manual(
      name = NULL,
      values = c(
        "Basin (catchment)" = "steelblue",
        "Area (coastal water body)" = "firebrick"
      )
    ) +
    coord_sf(xlim = xlim, ylim = ylim) +
    labs(x = NULL, y = NULL, title = name) +
    theme(panel.grid = element_blank())
}

p_gironde <- plot_estuary("Gironde", c(-1.3, -0.3), c(44.8, 45.7))
p_loire <- plot_estuary("Loire", c(-2.3, -1.0), c(47.0, 47.5))

p_gironde + p_loire +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

From what we see there, it is quite clear that the natural delimitation is
the basin polygon, whose boundaries fit well with our river mouths. So, to
infer whether an operation is inland or sea-side, we check whether it falls
within a hydrographic basin: if so, it is inland; if not, it is sea-side.

Let's see, per survey, how many inland and sea-side operations we have.

```{r}
operation_location <- tar_read(operation_location)

operation_location |>
  count(survey, inland) |>
  tidyr::pivot_wider(
    names_from = inland,
    values_from = n,
    values_fill = 0,
    names_prefix = "inland_"
  ) |>
  rename(
    inland = inland_TRUE,
    "sea-side" = inland_FALSE,
  ) |>
  knitr::kable()
```

#### Catchment attribution

Next, we want to attribute a catchment to the sea operations. We proceed as
follows: if the operation is inland, we use the district of the basin it
falls in; if the operation is sea-side, we use the district of the closest
coastal water body. Here, the question is to define a threshold beyond which
we consider that a coastal water body cannot be reliably attributed to an
operation. To help us, let's plot, in the Gironde estuary, the operations
farthest from the closest water body.

```{r}
operation_year_filtered <- tar_read(operation_year_filtered)

outside_sea <- operation_year_filtered |>
  filter(survey != "river") |>
  sf::st_as_sf(
    coords = c("longitude", "latitude"), crs = 4326, remove = FALSE
  ) |>
  sf::st_join(basin["district"], join = sf::st_within) |>
  filter(is.na(district)) |>
  select(-district)

nearest_idx <- sf::st_nearest_feature(outside_sea, coastal_water)
outside_sea <- outside_sea |>
  mutate(dist_km = as.numeric(sf::st_distance(
    outside_sea, coastal_water[nearest_idx, ],
    by_element = TRUE
  )) / 1000)

thresholds <- c(1, 2, 5, 10, 15, 20)
tibble::tibble(threshold_km = thresholds) |>
  mutate(
    pct_matched = round(
      100 * purrr::map_dbl(threshold_km, \(t) mean(outside_sea$dist_km <= t)),
      1
    ),
    n_unmatched = purrr::map_int(
      threshold_km, \(t) sum(outside_sea$dist_km > t)
    )
  ) |>
  knitr::kable()
```

Here, a threshold of 10km leaves 7 operations unmatched; however, they can
all be clearly attributed to the Gironde district. Raising the threshold to
15km enables us to attribute all operations correctly, so this is the value
we select. Raising the threshold further leads to no change, as every
operation is already attributed to a district.


```{r}
#| fig-width: 7
#| fig-height: 9

threshold_candidate <- 10
outside_sea <- outside_sea |>
  mutate(matched = dist_km <= threshold_candidate)

ggplot() +
  geom_sf(data = basin, fill = NA, color = "grey80") +
  geom_sf(data = coastal_water, fill = "firebrick", alpha = 0.1, color = NA) +
  geom_sf(data = outside_sea, aes(color = matched), size = 1.3) +
  scale_color_manual(
    values = c(`TRUE` = "steelblue", `FALSE` = "firebrick"),
    name = paste0("Matched within ", threshold_candidate, "km")
  ) +
  coord_sf(xlim = c(-1.45, -0.9), ylim = c(45.4, 45.75)) +
  theme(panel.grid = element_blank())
```

We can now confidently attribute a district to all sea operations, whether
they are inland or sea-side. Let's do it, and investigate, per district
(corresponding to the catchments), how many inland and sea-side operations
we have.

```{r}
operation_location |>
  mutate(district = tidyr::replace_na(district, "(sea, no catchment)")) |>
  count(district, inland) |>
  tidyr::pivot_wider(
    names_from = inland,
    values_from = n,
    values_fill = 0,
    names_prefix = "inland_"
  ) |>
  rename(
    inland = inland_TRUE,
    "sea-side" = inland_FALSE,
  ) |>
  arrange(district) |>
  knitr::kable()
```

#### Distance to river mouth

Sea operations split in two (see above): inland ones are snapped to the
river network and get channel-length distance, exactly like river
stations. Sea-side ones aren't on the river network, and a straight line
can cut across land, so they use a coast-aware sea mesh instead.

##### Inland sea operations: snapping quality

Same check as for river stations, now for inland-flagged sea operations
(`nurse`/`pomet`), snapped to the network for the first time. Unlike river
ASPE stations, which sit directly on a gauged channel, `nurse`/`pomet`
stations sample across the width of estuary mouths - and the network
models only a single channel line there, not the full width. So a much
looser threshold is expected and appropriate: 2100m, not 150m.
Investigating individual operations up to that distance confirmed
they sit exactly where estuaries widen (Gironde, Seine), not on data
errors - except one, `61837633`, which turned out to be miscoded inland
right at the coast and is now classified sea-side directly in
`classify_operation_location()`.

```{r}
#| fig-width: 7
#| fig-height: 9

threshold_sea <- 2100
sea_stations_selected <- tar_read(inland_operations_snapped) |>
  filter(survey != "river")

ggplot(sea_stations_selected, aes(x = snap_dist_m)) +
  geom_histogram(fill = "#a9c9e0", color = "white", bins = 40) +
  geom_vline(xintercept = threshold_sea) +
  annotate(
    "text",
    x = threshold_sea, y = Inf, label = "2100m", hjust = -0.2, vjust = 1.5
  ) +
  scale_x_continuous(trans = "log10", labels = scales::label_number()) +
  labs(x = "Snapping distance (m)", y = "Number of operations")
```

```{r}
n_bad_snap_sea <- sum(sea_stations_selected$snap_dist_m > threshold_sea)
```

`r n_bad_snap_sea` inland sea operations, out of
`r nrow(sea_stations_selected)`, have a snapping distance greater than
`r threshold_sea` meters.

##### Sea-side operations: mesh

`build_sea_mesh()` builds a water-only mesh over the sea (using
[rnaturalearth's land polygons](https://rnaturalearth.com/) rather than
`basin`, which only partially covers islands), `build_sea_graph()` turns
it into a graph, and `compute_sea_distance_to_mouth()` runs the same
per-district Dijkstra as `compute_distance_to_mouth()`, over that graph.

Two checkpoints: does the mesh respect the coastline, and do the
resulting distances actually account for land where a straight line
wouldn't?

```{r}
#| fig-width: 12
#| fig-height: 13

sea_mesh <- tar_read(sea_mesh)
sea_graph <- tar_read(sea_graph)
land_raw <- tar_read(land)
mesh_edges <- sea_mesh_edges(sea_mesh)

zoom_islands <- list(xlim = c(-2.6, -0.9), ylim = c(45.7, 47.1))
zoom_gironde <- list(xlim = c(-1.3, -0.3), ylim = c(44.8, 45.7))

zoom_box <- function(xlim, ylim) {
  st_as_sfc(st_bbox(
    c(xmin = xlim[1], xmax = xlim[2], ymin = ylim[1], ymax = ylim[2]),
    crs = 4326
  ))
}
zoom_boxes <- st_sf(geometry = c(
  zoom_box(zoom_islands$xlim, zoom_islands$ylim),
  zoom_box(zoom_gironde$xlim, zoom_gironde$ylim)
))

p_overview <- ggplot() +
  geom_sf(data = land_raw, fill = "grey75", color = NA) +
  geom_sf(data = mesh_edges, color = "steelblue", linewidth = 0.1) +
  geom_sf(data = zoom_boxes, fill = NA, color = "firebrick", linewidth = 0.6) +
  coord_sf(xlim = c(-5.6, 2.9), ylim = c(42.8, 51.6)) +
  labs(title = "Sea mesh domain", x = NULL, y = NULL) +
  theme(panel.grid = element_blank())

p_islands <- plot_sea_mesh(
  sea_mesh, mesh_edges, zoom_islands$xlim, zoom_islands$ylim,
  "Ile de Re"
)
p_gironde <- plot_sea_mesh(
  sea_mesh, mesh_edges, zoom_gironde$xlim, zoom_gironde$ylim,
  "Gironde"
)

p_overview / (p_islands + p_gironde)
```

The mesh covers water only - islands come through as untriangulated
holes, and the Gironde's channel is followed without cutting across the
peninsula on its western side.

Second checkpoint: per zoomed region, pick one operation whose straight
line to its nearest same-district mouth crosses land, and one that
doesn't, and compare straight-line to mesh-path distance. A crossing only
counts past 100m of overlap - `st_intersects()` alone also flags a line
that merely touches a boundary vertex, not a real crossing.

```{r}
#| fig-width: 9
#| fig-height: 28

sea_dist <- tar_read(sea_offshore_distance_to_mouth)
mouth_kept <- river_mouth_district |>
  filter(matched, district %in% district_kept) |>
  st_as_sf() |>
  st_transform(4326)
mouth_kept$node_id <- snap_to_sea_graph(mouth_kept, sea_graph)$node_id

sea_dist_sf <- sea_dist |>
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326, remove = FALSE)
crossing_len <- straight_line_land_crossing(
  sea_dist_sf, mouth_kept, sea_mesh$land, district_kept
)
sea_dist_checked <- sea_dist |> left_join(crossing_len, by = "operation_id")

pick_pair <- function(xlim, ylim) {
  in_zoom <- sea_dist_checked |>
    filter(
      longitude >= xlim[1], longitude <= xlim[2],
      latitude >= ylim[1], latitude <= ylim[2]
    )
  crossing <- in_zoom |>
    filter(crossing_len_m > 100) |>
    arrange(desc(crossing_len_m)) |>
    slice(1)
  clean <- in_zoom |> filter(crossing_len_m == 0) |> slice(1)
  list(crossing = crossing, clean = clean)
}

pair_islands <- pick_pair(zoom_islands$xlim, zoom_islands$ylim)
pair_gironde <- pick_pair(zoom_gironde$xlim, zoom_gironde$ylim)

plot_pair <- function(op_row, label) {
  path <- sea_mouth_path(op_row, mouth_kept, sea_graph)
  plot_sea_path_checkpoint(op_row, path, mesh_edges, sea_mesh$land, label)
}

p1 <- plot_pair(
  pair_islands$crossing, "Ile de Re - line crosses land"
)
p2 <- plot_pair(
  pair_islands$clean, "Ile de Re - line doesn't cross land"
)
p3 <- plot_pair(
  pair_gironde$crossing, "Gironde - line crosses land"
)
p4 <- plot_pair(
  pair_gironde$clean, "Gironde - line doesn't cross land"
)

p1 / p2 / p3 / p4
```

Where the straight line would cut across land, the mesh path detours
around it and comes out longer; where it wouldn't, the mesh path stays
close to the straight-line distance.

## Rephy interpolation

River mouth points come from the AMOBIO river network (Lambert-93), so we
reproject to lon/lat before plotting alongside the coastline.

```{r}
#| fig-width: 6
#| fig-height: 9

river_mouth_sf <- river_mouth |>
  sf::st_as_sf() |>
  sf::st_transform(4326)

ggplot() +
  geom_sf(data = coastline(), linewidth = 0.3) +
  geom_sf(data = river_mouth_sf, size = 0.8) +
  coord_sf(
    xlim = sf::st_bbox(river_mouth_sf)[c("xmin", "xmax")],
    ylim = sf::st_bbox(river_mouth_sf)[c("ymin", "ymax")]
  ) +
  labs(x = NULL, y = NULL)
```

### Sea mesh and river network

Before building a combined graph, we look at the sea mesh (built the same way
as in the interpolation notebook) and the AMOBIO river network nodes overlaid
on the same map. The river network has ~1M nodes, so we plot with base R
(much faster than ggplot for this many points) and use tiny markers.

```{r}
rephy_filtered <- tar_read(rephy_data_sample)
mesh <- build_rephy_mesh(rephy_filtered)

network <- tar_read(river_network_full)
river_coords <- network$nodes |>
  sf::st_as_sf() |>
  sf::st_transform(4326) |>
  sf::st_coordinates()
```

```{r}
#| fig-width: 6
#| fig-height: 9

plot(mesh, main = "")
points(river_coords[, "X"], river_coords[, "Y"], col = "steelblue", pch = ".")
points(
  river_mouth_sf |> sf::st_coordinates(),
  col = "firebrick", pch = 19, cex = 0.6
)
legend(
  "bottomright",
  legend = c("River network node", "River mouth"),
  col = c("steelblue", "firebrick"),
  pch = 19,
  bty = "n"
)
```

### River mouths vs. hydrographic area

We match each river mouth to a district (of interest for our study) and
highlight the ones that still don't match. These points will be removed.

```{r}
#| fig-width: 6
#| fig-height: 9

coastal_water <- tar_read(coastal_water)
district_kept <- c("Loire", "Ardour-Garonne", "Seine", "Escaut-Somme")
hydro_kept <- coastal_water |>
  filter(district %in% district_kept)

river_mouth_district <- tar_read(river_mouth_district)

ggplot() +
  geom_sf(data = hydro_kept, aes(fill = district), color = NA, alpha = 0.4) +
  geom_sf(data = coastline(), linewidth = 0.3) +
  geom_sf(data = river_mouth_district, aes(color = matched), size = 1) +
  scale_color_manual(values = c("TRUE" = "grey20", "FALSE" = "firebrick")) +
  coord_sf(
    xlim = sf::st_bbox(river_mouth_sf)[c("xmin", "xmax")],
    ylim = sf::st_bbox(river_mouth_sf)[c("ymin", "ymax")]
  ) +
  labs(x = NULL, y = NULL, fill = "District", color = "Matched (≤ 10 km)")
```

### Restricted river network

We grow the network outward (channel length) from the matched river mouths,
using a single Dijkstra run from a virtual source connected to every matched
mouth at once - much cheaper than one search per mouth. This reuses
`river_network` (restricted to `river_station_network_max_dist`, the river
station radius) rather than a separate 100km REPHY-only cutoff - the
dedicated REPHY-scope target was a duplicate of this one and has been
removed, since `river_network`'s radius is a safe superset.

```{r}
#| fig-width: 6
#| fig-height: 9

network_restricted <- tar_read(river_network)

river_restricted_nodes <- network_restricted$nodes |>
  sf::st_as_sf() |>
  sf::st_transform(4326)

ggplot() +
  geom_sf(data = coastline(), linewidth = 0.3) +
  geom_sf(
    data = river_restricted_nodes,
    aes(color = "River network node"),
    size = 0.1,
    alpha = 0.1
  ) +
  geom_sf(
    data = river_mouth_district |> filter(matched),
    aes(color = "River mouth"),
    size = 1
  ) +
  geom_point(
    data = rephy_filtered,
    aes(x = longitude, y = latitude, color = "REPHY station"),
    size = 0.8
  ) +
  scale_color_manual(
    values = c(
      "River network node" = "steelblue",
      "River mouth" = "firebrick",
      "REPHY station" = "darkorange"
    )
  ) +
  coord_sf(
    xlim = sf::st_bbox(river_restricted_nodes)[c("xmin", "xmax")],
    ylim = sf::st_bbox(river_restricted_nodes)[c("ymin", "ymax")]
  ) +
  labs(x = NULL, y = NULL, color = NULL)
```
