---
title: "Compute distance to river mouth"
bibliography: ../references.bib
format:
  html:
    theme: cosmo
    code-fold: true
    toc: true
    embed-resources: true
execute:
    message: false
    warning: false
---

The goal is to compute the distance to river mouth REPHY data and fishing
operations at sea.

## Todo

- [x] Extract river mouth points
- [x] Plot river mouth points
- [x] Simplify AMOBIO network
- [x] Filter river mouths
- [x] Restrict river network
- [x] Check rephy data overlap with restricted river network
- [x] Reconstruct full river network for `river` survey stations
- [x] Snap `river` survey stations to the network
- [x] Classify `river` survey stations by true catchment
- [x] Close the `Ardour-Garonne` under-classification gap
- [x] Investigate why the Vendée cluster is disconnected
- [x] Patch the missing link (Nantes/Sèvre Nantaise) and confirm the
cluster recovers
- [x] Snap directly against the restricted network now that it's fixed
- [x] Check quality of snapping, how many non-snapped stations and where?
- [x] Find about Somme classification, as it doesn't appear in the current
cateogries
- [x] Keep edge linestring geometry in `extract_river_network()`
(currently dropped for straight node-to-node lines) - check if it
recovers some non-snapped stations
- [ ] Check connected components vs. catchments - is one catchment one
connected component?
- [ ] Compute distance to river mouth for `river` survey stations
- [ ] Clean notebook, reorganise, snapping and rephy are mixed

## Setup

```{r}
library(renv)
renv::load()
library(targets)
library(here)
tar_config_set(
  store = here::here("_targets")
)
library(ggplot2)
library(dplyr)
library(sf)
devtools::load_all()
riversea::nice_theme()
```

```{r}
river_mouth <- tar_read(amobio_river_mouth)
```

River mouth points come from the AMOBIO river network (Lambert-93), so we
reproject to lon/lat before plotting alongside the coastline.

```{r}
#| fig-width: 6
#| fig-height: 9

river_mouth_sf <- river_mouth |>
  sf::st_as_sf() |>
  sf::st_transform(4326)

ggplot() +
  geom_sf(data = coastline(), linewidth = 0.3) +
  geom_sf(data = river_mouth_sf, size = 0.8) +
  coord_sf(
    xlim = sf::st_bbox(river_mouth_sf)[c("xmin", "xmax")],
    ylim = sf::st_bbox(river_mouth_sf)[c("ymin", "ymax")]
  ) +
  labs(x = NULL, y = NULL)
```

### Sea mesh and river network

Before building a combined graph, we look at the sea mesh (built the same way
as in the interpolation notebook) and the AMOBIO river network nodes overlaid
on the same map. The river network has ~1M nodes, so we plot with base R
(much faster than ggplot for this many points) and use tiny markers.

```{r}
rephy_filtered <- tar_read(rephy_data_sample)
mesh <- build_rephy_mesh(rephy_filtered)

network <- tar_read(river_network_full)
river_coords <- network$nodes |>
  sf::st_as_sf() |>
  sf::st_transform(4326) |>
  sf::st_coordinates()
```

```{r}
#| fig-width: 6
#| fig-height: 9

plot(mesh, main = "")
points(river_coords[, "X"], river_coords[, "Y"], col = "steelblue", pch = ".")
points(
  river_mouth_sf |> sf::st_coordinates(),
  col = "firebrick", pch = 19, cex = 0.6
)
legend(
  "bottomright",
  legend = c("River network node", "River mouth"),
  col = c("steelblue", "firebrick"),
  pch = 19,
  bty = "n"
)
```

### River mouths vs. hydrographic area

We match each river mouth to a district (of interest for our study) and
highlight the ones that still don't match. These points will be removed.

```{r}
#| fig-width: 6
#| fig-height: 9

hydro <- tar_read(hydrographic_area)
district_kept <- c("Loire", "Ardour-Garonne", "Seine", "Escaut-Somme")
hydro_kept <- hydro |>
  filter(district %in% district_kept)

river_mouth_district <- tar_read(river_mouth_district)

ggplot() +
  geom_sf(data = hydro_kept, aes(fill = district), color = NA, alpha = 0.4) +
  geom_sf(data = coastline(), linewidth = 0.3) +
  geom_sf(data = river_mouth_district, aes(color = matched), size = 1) +
  scale_color_manual(values = c("TRUE" = "grey20", "FALSE" = "firebrick")) +
  coord_sf(
    xlim = sf::st_bbox(river_mouth_sf)[c("xmin", "xmax")],
    ylim = sf::st_bbox(river_mouth_sf)[c("ymin", "ymax")]
  ) +
  labs(x = NULL, y = NULL, fill = "District", color = "Matched (≤ 10 km)")
```

### Restricted river network

We grow the network outward (channel length) from the matched river mouths,
using a single Dijkstra run from a virtual source connected to every matched
mouth at once - much cheaper than one search per mouth. This reuses
`river_network` (restricted to `river_station_network_max_dist`, the river
station radius) rather than a separate 100km REPHY-only cutoff - the
dedicated REPHY-scope target was a duplicate of this one and has been
removed, since `river_network`'s radius is a safe superset.

```{r}
#| fig-width: 6
#| fig-height: 9

network_restricted <- tar_read(river_network)

river_restricted_nodes <- network_restricted$nodes |>
  sf::st_as_sf() |>
  sf::st_transform(4326)

ggplot() +
  geom_sf(data = coastline(), linewidth = 0.3) +
  geom_sf(
    data = river_restricted_nodes,
    aes(color = "River network node"),
    size = 0.1,
    alpha = 0.1
  ) +
  geom_sf(
    data = river_mouth_district |> filter(matched),
    aes(color = "River mouth"),
    size = 1
  ) +
  geom_point(
    data = rephy_filtered,
    aes(x = longitude, y = latitude, color = "REPHY station"),
    size = 0.8
  ) +
  scale_color_manual(
    values = c(
      "River network node" = "steelblue",
      "River mouth" = "firebrick",
      "REPHY station" = "darkorange"
    )
  ) +
  coord_sf(
    xlim = sf::st_bbox(river_restricted_nodes)[c("xmin", "xmax")],
    ylim = sf::st_bbox(river_restricted_nodes)[c("ymin", "ymax")]
  ) +
  labs(x = NULL, y = NULL, color = NULL)
```

## Snapping river survey stations to the network

Same idea as above, but for river (ASPE) stations instead of REPHY.
Stations reach far upstream, so we reconstruct with a 5000km radius
instead of REPHY's, then snap each station to its nearest node.

```{r}
network_river <- tar_read(river_network)
river_stations <- tar_read(river_station_snapped)
district_kept <- c("Loire", "Ardour-Garonne", "Seine", "Escaut-Somme")

cat("Median snap distance (m):", round(median(river_stations$snap_dist_m), 1), "\n")
cat("N snapping over 1km:", sum(river_stations$snap_dist_m > 1000), "\n")
```

Snapping is straight-line, not channel-length, so a large distance flags a
bad station coordinate rather than a real network gap - and there's
barely any left (see below for why).

Each station's catchment comes from a point-in-polygon join against
`hydrographic_basin` - BD Topage's official river basin district
polygons (SANDRE, full-country coverage, downloaded and formatted the
same way as `hydrographic_area`) - rather than from the network at all.
This replaced an earlier approach that parsed ASPE's own water body code
(`classify_water_body_district()`, since removed): that code-based
classification validated fine where a code existed, but ~2,600 stations
had no code at all - including every single `Escaut-Somme` station (see
"Recovering uncoded stations" below for the full story and why the
polygon approach won over two network-based alternatives).

```{r}
table(river_stations$district, useNA = "always")
```

`Escaut-Somme` now has 519 stations (previously zero). `Rhin` and `Meuse`
come back merged as `Rhin-Meuse` - BD Topage manages them as one basin at
this resolution, which doesn't matter since neither is a kept district.
The remaining 154 NAs are mostly border-precision edge cases (a station
sitting right at a basin boundary, or just outside France near Belgium)
rather than a classification gap.

```{r}
#| fig-width: 7
#| fig-height: 9

in_scope <- river_stations |>
  filter(district %in% district_kept)
river_coords <- network_river$nodes |>
  sf::st_as_sf() |>
  sf::st_transform(4326) |>
  sf::st_coordinates()
cols <- c(
  Loire = "steelblue",
  "Ardour-Garonne" = "firebrick",
  Seine = "darkgreen",
  "Escaut-Somme" = "orange"
)

plot(
  river_coords[, "X"],
  river_coords[, "Y"],
  col = "grey85",
  pch = ".",
  asp = 1.5,
  xlab = NULL,
  ylab = NULL
)
points(
  in_scope$longitude,
  in_scope$latitude,
  col = cols[in_scope$district],
  pch = 19,
  cex = 0.3
)
legend("bottomright", legend = names(cols), col = cols, pch = 19, bty = "n")
```

Clean catchment boundaries, no stray points in the wrong basin.


### Fixing a network gap (Vendée / Nantes)

While investigating, one ~6,000-node component near the Vendée coast had
no station snapping to it despite sitting inside Loire territory - and no
node in it was flagged a river mouth. Rather than assume real
hydrological separation, checking the gap to the rest of the network:

```{r}
river_network_full <- tar_read(river_network_full)
network_before <- river_network_full
network_before$edges <- network_before$edges |>
  filter(!(from == 614458 & to == 614552))

comp_before <- igraph::components(igraph::graph_from_edgelist(
  as.matrix(network_before$edges[c("from", "to")]),
  directed = FALSE
))
vendee_comp_nodes <- which(comp_before$membership == comp_before$membership[595001])
main_sf <- river_network_full$nodes |>
  filter(node_id %in% which(comp_before$membership == which.max(comp_before$csize))) |>
  sf::st_as_sf()
vendee_sf <- river_network_full$nodes |>
  filter(node_id %in% vendee_comp_nodes) |>
  sf::st_as_sf()

gap_dist_m <- as.numeric(sf::st_distance(
  vendee_sf, main_sf[sf::st_nearest_feature(vendee_sf, main_sf), ],
  by_element = TRUE
))
cat("Component size:", length(vendee_comp_nodes), "\n")
cat("Closest gap:", round(min(gap_dist_m), 1), "m -", sum(gap_dist_m < 500), "nodes within 500m.\n")
```

A gap under 100m is within GPS error, and several nodes are within 500m -
this reads as a missing digitized link, not a real separation. The
clearest of these gaps sits at Nantes, where the Sèvre Nantaise joins the
Loire - checked against a basemap that shows the rivers:

```{r}
gap_pts <- river_network_full$nodes |>
  filter(node_id %in% c(614458, 614552)) |>
  sf::st_as_sf() |>
  sf::st_transform(4326) |>
  mutate(label = c("Vendée-side node", "Main-network node"))

leaflet::leaflet(gap_pts) |>
  leaflet::addProviderTiles(leaflet::providers$Esri.WorldTopoMap) |>
  leaflet::addCircleMarkers(
    color = c("firebrick", "steelblue"),
    radius = 8,
    label = ~label
  ) |>
  leaflet::setView(
    lng = mean(sf::st_coordinates(gap_pts)[, 1]),
    lat = mean(sf::st_coordinates(gap_pts)[, 2]),
    zoom = 15
  )
```

Both points sit right on the river, either side of the confluence, 162m
apart. `patch_river_network()` adds this edge (length computed from the
node geometries, not hardcoded) and is already active in `river_network_full`
above - snapping and the map/table earlier already reflect the fix:

```{r}
comp_after <- igraph::components(igraph::graph_from_edgelist(
  as.matrix(river_network_full$edges[c("from", "to")]),
  directed = FALSE
))
cat(
  "Same component as main network now?",
  comp_after$membership[595001] == comp_after$membership[main_sf$node_id[1]], "\n"
)
recovered <- in_scope |> filter(node_id %in% vendee_comp_nodes)
cat(
  "Recovered:", nrow(recovered), "stations, snap distance now median",
  round(median(recovered$snap_dist_m), 1), "m\n"
)
```

### Snapping quality

The median above (7.7m) hides a long right tail - `river_network` is
grown outward only from river mouths in our four catchments
(`district_kept`), so a station outside them has essentially nowhere
nearby to snap to, by construction. Splitting by district separates that
expected effect from real bad coordinates:

```{r}
river_stations |>
  mutate(group = ifelse(district %in% district_kept, "in scope", district)) |>
  mutate(group = ifelse(is.na(group), "NA", group)) |>
  group_by(group) |>
  summarise(
    n = n(),
    n_over_1km = sum(snap_dist_m > 1000),
    median_km = round(median(snap_dist_m) / 1000, 3),
    max_km = round(max(snap_dist_m) / 1000, 1)
  ) |>
  arrange(desc(max_km)) |>
  knitr::kable()
```

Within scope, snapping is essentially perfect: only 7 of 18,246 stations
land more than 1km away. Three are newly-recovered `Escaut-Somme`
stations at 5.1km, right at/across the Belgian border where the
France-only AMOBIO network doesn't reach - a real coverage gap, not a
bad coordinate. The rest (two `Seine`, two `Loire`) are under 1.4km,
plausibly imprecise GPS logging.

```{r}
in_scope |>
  filter(snap_dist_m > 1000) |>
  select(operation_id, longitude, latitude, district, snap_dist_m) |>
  knitr::kable(digits = 0)
```

Tightening the threshold to 100m or 500m barely moves the needle either -
still under 0.8% of in-scope stations:

```{r}
tibble::tibble(threshold_m = c(100, 500)) |>
  mutate(
    n_not_snapped = purrr::map_int(threshold_m, \(t) sum(in_scope$snap_dist_m > t)),
    n_tot = nrow(in_scope)
  ) |>
  knitr::kable()
```

Out of scope, distances top out at a few hundred km and track
geography, not data quality - `Corse` sits 189-283km away because the
mainland network never reaches the island, and `Rhin-Meuse`/`Rhone`
stations sit on the far edge of the 5000km growth radius from our
catchments' mouths. Plotting them against the network extent confirms
it: every out-of-scope cluster falls outside the blue river network,
exactly where it should.

```{r}
#| fig-width: 7
#| fig-height: 9

river_nodes_sf <- network_river$nodes |>
  sf::st_as_sf() |>
  sf::st_transform(4326)
bbox <- sf::st_bbox(river_nodes_sf)

flagged <- river_stations |>
  filter(snap_dist_m > 1000, snap_dist_m < 500000) |>
  mutate(group = ifelse(district %in% district_kept, "in scope", district)) |>
  mutate(group = ifelse(is.na(group), "NA", group))

ggplot() +
  geom_sf(data = coastline(), linewidth = 0.3) +
  geom_sf(data = river_nodes_sf, size = 0.05, alpha = 0.05, color = "steelblue") +
  geom_point(data = flagged, aes(x = longitude, y = latitude, color = group), size = 1.2) +
  coord_sf(xlim = bbox[c("xmin", "xmax")], ylim = bbox[c("ymin", "ymax")]) +
  labs(
    x = NULL, y = NULL, color = "Group",
    title = "Stations snapping >1km from network (excl. one 6,187km outlier)"
  )
```

One station doesn't fit this pattern: `operation_id` 12640 snaps 6,187km
away, at (lon -1.36, lat -5.98). A latitude south of the equator can't be
a real French river station - this is a genuine bad coordinate in the
source data, not a network artifact. It falls outside every basin
polygon too (`district` = `NA`), so it doesn't affect anything
downstream, but it's worth flagging rather than silently ignoring if
`river_operation` is reused elsewhere.

### Recovering uncoded stations

`district` was `NA` for 2,587 stations under the old
`classify_water_body_district()` (ASPE's water body code parser) - worth
double checking, since a station with no code isn't the same as a
station outside our four catchments. Two things stood out:

- Almost 90% of these uncoded stations snapped to the network just as
  well as coded ones do - too many to be coincidental overlap.
- Every one of the 2,587 had `body_water_code == NA` outright (not an
  unrecognized format) - not the `Ardour-Garonne` regex gap again, just
  stations that were never coded at all. Assigning each to its nearest
  matched river mouth (straight-line) pointed heavily at `Escaut-Somme`
  (793 of them) - suspicious, since `Escaut-Somme` had exactly zero
  *coded* stations.

**Two network-based fixes, both with real problems.** Nearest-mouth
assignment is exactly the kind of thing the network should be good at,
so two approaches were validated against the 16,561 stations that *did*
have a code:

- **Straight-line nearest mouth**: only 77% agreement, badly wrong for
  `Loire` (3,335 of 7,481 stations misassigned) - straight-line distance
  ignores how far inland a large catchment reaches.
- **Channel-length nearest mouth** (multi-source Dijkstra from each
  district's matched mouths): much better, 91.5% agreement - except
  1,403 of 3,748 `Seine`-coded stations routed to `Escaut-Somme`. The
  shortest path for these runs through **Canal du Nord** and **la Somme
  Canalisée**, real navigation canals linking the Oise (a Seine
  tributary) to the Somme - the network doesn't distinguish natural
  rivers from artificial canals, so channel-routing "shortcuts" across
  basin boundaries wherever one exists. Stripping every edge with
  "canal" in its name fixed it (100% agreement) but disconnected 16% of
  all nodes, since most "Canal..." names are channelized natural rivers
  (e.g. "la Somme Canalisée" *is* the Somme), not shortcuts - too blunt
  an instrument, and the ~2,200 distinct canal-ish names aren't
  curatable by hand.

**The fix: skip the network entirely.** `hydrographic_basin` (BD
Topage's official river basin district polygons, SANDRE, full-country
coverage - see `format_hydrographic_basin()`) turns this into a plain
point-in-polygon lookup: no routing, so no canal artifact, and no
straight-line distortion for large basins either. Validated against all
16,561 previously-coded stations: **100% agreement**. It's now
`river_operation`'s only source for `district`
(`classify_water_body_district()` has been removed - see
`get_river_operation()`), and it recovered 1,685 of the 2,587 uncoded
stations into our four catchments - `Seine` +946, `Escaut-Somme` +519,
`Ardour-Garonne` +111, `Loire` +109 - resolving both the uncoded-station
problem and the `Escaut-Somme` gap at once (0 -> 519 stations, see the
table above).

### Node-only vs. edge-linestring snapping precision

`river_network`'s edges drop geometry (that's what `extract_river_network()`
does, to keep the Dijkstra restriction cheap), so snapping only ever
measures distance to the nearest *node* - a station sitting midway along
a straight stretch between two nodes looks farther away than it really
is. `river_network_geo` re-attaches the original edge linestrings
(`add_edge_geometry()`, joined back onto the already-restricted edges by
`from`/`to` - much cheaper than redoing the restriction with geometry
kept throughout). Snapping to the nearest point on the nearest edge
instead of the nearest node tests whether that explains some of the
remaining non-snapped in-scope stations.

```{r}
network_river_geo <- tar_read(river_network_geo)
edges_geo <- sf::st_as_sf(network_river_geo$edges)

stations_sf <- in_scope |>
  sf::st_as_sf(coords = c("longitude", "latitude"), crs = 4326, remove = FALSE) |>
  sf::st_transform(sf::st_crs(edges_geo))

nearest_edge <- sf::st_nearest_feature(stations_sf, edges_geo)
dist_line <- as.numeric(sf::st_distance(stations_sf, edges_geo[nearest_edge, ], by_element = TRUE))
```

By construction, `dist_line` can only be smaller or equal to the current
node-based `snap_dist_m` (a point on a line is always at least as close
as its nearest endpoint) - the question is how much smaller, and whether
it flips non-snapped stations into snapped ones:

```{r}
thresholds <- c(50, 100, 150, 200, 500)
tibble::tibble(threshold_m = thresholds) |>
  mutate(
    n_not_snapped_node = purrr::map_int(threshold_m, \(t) sum(in_scope$snap_dist_m > t)),
    n_not_snapped_line = purrr::map_int(threshold_m, \(t) sum(dist_line > t))
  ) |>
  knitr::kable()
```

```{r}
#| fig-width: 7
#| fig-height: 5

thresholds_fine <- c(seq(0, 100, by = 2), seq(105, 500, by = 5))
compare_df <- dplyr::bind_rows(
  tibble::tibble(
    threshold_m = thresholds_fine,
    n_not_snapped = purrr::map_int(thresholds_fine, \(t) sum(in_scope$snap_dist_m > t)),
    method = "node only (current)"
  ),
  tibble::tibble(
    threshold_m = thresholds_fine,
    n_not_snapped = purrr::map_int(thresholds_fine, \(t) sum(dist_line > t)),
    method = "edge linestring"
  )
)

ggplot(compare_df, aes(x = threshold_m, y = n_not_snapped, color = method)) +
  geom_step() +
  scale_y_log10() +
  labs(
    x = "Snap distance threshold (m)",
    y = "N in-scope stations not snapped (log scale)", color = NULL
  )
```

Keeping the edge geometry roughly halves-to-tenths the non-snapped count
at every threshold from 50m to 500m (e.g. 117 -> 20 at 150m). It's a real,
consistent improvement, not noise - worth adopting if we end up caring
about snap precision below ~100m (e.g. for anything joining stations to
specific river reaches), though it doesn't change the "snapping quality
is good" conclusion above, since the node-only numbers were already
well under 1%.

