---
title: "Compute distance to river mouth"
bibliography: ../references.bib
format:
  html:
    theme: cosmo
    code-fold: true
    toc: true
    embed-resources: true
execute:
    message: false
    warning: false
---

The goal is to compute the distance to river mouth REPHY data and fishing
operations at sea.

## Todo

- [x] Extract river mouth points
- [x] Plot river mouth points
- [x] Simplify AMOBIO network
- [x] Filter river mouths
- [x] Restrict river network
- [x] Check rephy data overlap with restricted river network
- [ ] Compute distance to river mouth for rephy stations
- [ ] Update INLA model with distance to river mouth as new covariate

## Setup

```{r}
library(renv)
renv::load()
library(targets)
library(here)
tar_config_set(
  store = here::here("_targets")
)
library(ggplot2)
library(dplyr)
library(sf)
devtools::load_all()
riversea::nice_theme()
```

```{r}
river_mouth <- tar_read(amobio_river_mouth)
```

River mouth points come from the AMOBIO river network (Lambert-93), so we
reproject to lon/lat before plotting alongside the coastline.

```{r}
#| fig-width: 6
#| fig-height: 9

river_mouth_sf <- river_mouth |>
  sf::st_as_sf() |>
  sf::st_transform(4326)

ggplot() +
  geom_sf(data = coastline(), linewidth = 0.3) +
  geom_sf(data = river_mouth_sf, size = 0.8) +
  coord_sf(
    xlim = sf::st_bbox(river_mouth_sf)[c("xmin", "xmax")],
    ylim = sf::st_bbox(river_mouth_sf)[c("ymin", "ymax")]
  ) +
  labs(x = NULL, y = NULL)
```

### Sea mesh and river network

Before building a combined graph, we look at the sea mesh (built the same way
as in the interpolation notebook) and the AMOBIO river network nodes overlaid
on the same map. The river network has ~1M nodes, so we plot with base R
(much faster than ggplot for this many points) and use tiny markers.

```{r}
rephy_filtered <- tar_read(rephy_data_sample)
mesh <- build_rephy_mesh(rephy_filtered)

network <- tar_read(amobio_network)
river_coords <- network$nodes |>
  sf::st_as_sf() |>
  sf::st_transform(4326) |>
  sf::st_coordinates()
```

```{r}
#| fig-width: 6
#| fig-height: 9

plot(mesh, main = "")
points(river_coords[, "X"], river_coords[, "Y"], col = "steelblue", pch = ".")
points(
  river_mouth_sf |> sf::st_coordinates(),
  col = "firebrick", pch = 19, cex = 0.6
)
legend(
  "bottomright",
  legend = c("River network node", "River mouth"),
  col = c("steelblue", "firebrick"),
  pch = 19,
  bty = "n"
)
```

### River mouths vs. hydrographic area

We match each river mouth to a district (of interest for our study) and
highlight the ones that still don't match. These points will be removed.

```{r}
#| fig-width: 6
#| fig-height: 9

hydro <- tar_read(hydrographic_area)
district_kept <- c("Loire", "Ardour-Garonne", "Seine", "Escaut-Somme")
hydro_kept <- hydro |>
  filter(district %in% district_kept)

river_mouth_district <- tar_read(river_mouth_district)

ggplot() +
  geom_sf(data = hydro_kept, aes(fill = district), color = NA, alpha = 0.4) +
  geom_sf(data = coastline(), linewidth = 0.3) +
  geom_sf(data = river_mouth_district, aes(color = matched), size = 1) +
  scale_color_manual(values = c("TRUE" = "grey20", "FALSE" = "firebrick")) +
  coord_sf(
    xlim = sf::st_bbox(river_mouth_sf)[c("xmin", "xmax")],
    ylim = sf::st_bbox(river_mouth_sf)[c("ymin", "ymax")]
  ) +
  labs(x = NULL, y = NULL, fill = "District", color = "Matched (≤ 10 km)")
```

### Restricted river network

We grow the network outward (channel length) from the matched river mouths,
up to 100 km, using a single Dijkstra run from a virtual source connected to
every matched mouth at once - much cheaper than one search per mouth.


```{r}
#| fig-width: 6
#| fig-height: 9

network_restricted <- tar_read(amobio_network_restricted)

river_restricted_nodes <- network_restricted$nodes |>
  sf::st_as_sf() |>
  sf::st_transform(4326)

ggplot() +
  geom_sf(data = coastline(), linewidth = 0.3) +
  geom_sf(
    data = river_restricted_nodes,
    aes(color = "River network node"),
    size = 0.1,
    alpha = 0.1
  ) +
  geom_sf(
    data = river_mouth_district |> filter(matched),
    aes(color = "River mouth"),
    size = 1
  ) +
  geom_point(
    data = rephy_filtered,
    aes(x = longitude, y = latitude, color = "REPHY station"),
    size = 0.8
  ) +
  scale_color_manual(
    values = c(
      "River network node" = "steelblue",
      "River mouth" = "firebrick",
      "REPHY station" = "darkorange"
    )
  ) +
  coord_sf(
    xlim = sf::st_bbox(river_restricted_nodes)[c("xmin", "xmax")],
    ylim = sf::st_bbox(river_restricted_nodes)[c("ymin", "ymax")]
  ) +
  labs(x = NULL, y = NULL, color = NULL)
```

