---
title: "Compute distance to river mouth"
bibliography: ../references.bib
format:
  html:
    theme: cosmo
    code-fold: true
    toc: true
    embed-resources: true
execute:
    message: false
    warning: false
---

The goal is to compute the distance to river mouth REPHY data and fishing
operations at sea.

## Todo

- [x] Extract river mouth points
- [x] Plot river mouth points
- [x] Simplify AMOBIO network
- [x] Filter river mouths
- [x] Restrict river network
- [x] Check rephy data overlap with restricted river network
- [x] Reconstruct full river network for `river` survey stations
- [x] Snap `river` survey stations to the network
- [x] Classify `river` survey stations by true catchment
- [x] Close the `Ardour-Garonne` under-classification gap
- [x] Investigate why the Vendée cluster is disconnected
- [x] Patch the missing link (Nantes/Sèvre Nantaise) and confirm the
cluster recovers
- [x] Snap directly against the restricted network now that it's fixed
- [x] Check quality of snapping, how many non-snapped stations and where?
- [x] Find about Somme classification, as it doesn't appear in the current
cateogries
- [x] Keep edge linestring geometry in `extract_river_network()`
(currently dropped for straight node-to-node lines) - check if it
recovers some non-snapped stations
- [ ] Clean notebook, reorganise, snapping and rephy are mixed
- [ ] Plot catchments
- [ ] Merge catchments (if needed) to cover land and coast
- [ ] Check connected components vs. catchments - is one catchment one
connected component?
- [ ] Compute distance to river mouth for `river` survey stations

## Setup

```{r}
library(renv)
renv::load()
library(targets)
library(here)
tar_config_set(
  store = here::here("_targets")
)
library(ggplot2)
library(dplyr)
library(sf)
library(patchwork)
devtools::load_all()
riversea::nice_theme()
```

```{r}
river_mouth <- tar_read(amobio_river_mouth)
```

## Snapping river survey stations to the network

### Catchments

We focus on four catchments: Loire, Ardour-Garonne, Seine and Escaut-Somme. We
attribute stations to catchment using the BD Topage's polygones delimiting
hydrographic basins.

```{r}
network_river <- tar_read(river_network)
river_stations <- tar_read(river_station_snapped)
basin <- tar_read(hydrographic_basin)
district_kept <- c("Loire", "Ardour-Garonne", "Seine", "Escaut-Somme")

cols <- c(
  Loire = "steelblue",
  "Ardour-Garonne" = "firebrick",
  Seine = "darkgreen",
  "Escaut-Somme" = "orange"
)

basin_kept <- basin |> filter(district %in% district_kept)
in_scope <- river_stations |> filter(district %in% district_kept)
river_coords <- network_river$nodes |>
  sf::st_as_sf() |>
  sf::st_transform(4326) |>
  sf::st_coordinates()
```

We can visualise the catchments of interest as wells the corresponding stations.

```{r}
#| fig-width: 7
#| fig-height: 9

plot(st_geometry(basin), col = NA, border = "grey80")
plot(
  st_geometry(basin_kept),
  col = scales::alpha(cols[basin_kept$district], 0.15),
  border = "grey70",
  add = TRUE
)
points(
  river_coords[, "X"],
  river_coords[, "Y"],
  col = scales::alpha("grey60", 0.25),
  pch = ".",
  cex = 0.5
)
points(
  in_scope$longitude,
  in_scope$latitude,
  col = cols[in_scope$district],
  pch = 19,
  cex = 0.3
)
legend("bottomright", legend = names(cols), col = cols, pch = 19, bty = "n")
```

We can have a look at how many stations we have per catchments

```{r}
count(river_stations, district) |>
  filter(district %in% district_kept) |>
  arrange(n) |>
  knitr::kable()
```

Next, we evaluate the snapping quality for those catchments.

### Evaluate snapping quality

We snap fish-monitoring stations to the network hydrographic network. Because
snapping requires precision, we use the network with line geometry
`river_network_geo`.  We can evaluate the snapping quality through the snapping
distance, that is, how far the station was from the closest point in the river
network. Short distances indicate confident snapping.

```{r}
stations_selected <- river_stations |>
  filter(district %in% district_kept)

ggplot(stations_selected, aes(x = snap_dist_m)) +
  geom_histogram(fill = "#a9c9e0", color = "white", bins = 40) +
  geom_vline(xintercept = 150) +
  annotate("text", x = 150, y = Inf, label = "150m", hjust = -0.2, vjust = 1.5) +
  scale_x_continuous(trans = "log10", labels = scales::label_number()) +
  labs(x = "Snapping distance (m)", y = "Number of stations")
```

```{r}
threshold <- 150
n_bad_snap <- sum(stations_selected$snap_dist_m > threshold)
```

We see that the vast majority of stations are well-snapped to the river network.
In particular, we have only with `r n_bad_snap` a snapping distance greater than
`r threshold` meters.

### Adding a network missing edge (Nantes)

While investigating, one component near the Vendée coast appeared disconnected
from the rest of the Loire basin, despite sitting inside it. That led this
component to be disconnected from any river mouth. After checking, we found
two points that were unlinked but suspiciously close to one another, so we
decided to investigate where they were.

```{r}
river_network_full <- tar_read(river_network_full)
network_before <- river_network_full
network_before$edges <- network_before$edges |>
  filter(!(from == 614458 & to == 614552))

comp_before <- igraph::components(igraph::graph_from_edgelist(
  as.matrix(network_before$edges[c("from", "to")]),
  directed = FALSE
))
vendee_comp_nodes <- which(comp_before$membership == comp_before$membership[595001])
main_sf <- river_network_full$nodes |>
  filter(node_id %in% which(comp_before$membership == which.max(comp_before$csize))) |>
  sf::st_as_sf()
vendee_sf <- river_network_full$nodes |>
  filter(node_id %in% vendee_comp_nodes) |>
  sf::st_as_sf()

gap_dist_m <- as.numeric(sf::st_distance(
  vendee_sf, main_sf[sf::st_nearest_feature(vendee_sf, main_sf), ],
  by_element = TRUE
))

gap_pts <- river_network_full$nodes |>
  filter(node_id %in% c(614458, 614552)) |>
  sf::st_as_sf() |>
  sf::st_transform(4326) |>
  mutate(label = c("Vendée-side node", "Main-network node"))

leaflet::leaflet(gap_pts) |>
  leaflet::addProviderTiles(leaflet::providers$Esri.WorldTopoMap) |>
  leaflet::addCircleMarkers(
    color = c("firebrick", "steelblue"),
    radius = 8,
    label = ~label
  ) |>
  leaflet::setView(
    lng = mean(sf::st_coordinates(gap_pts)[, 1]),
    lat = mean(sf::st_coordinates(gap_pts)[, 2]),
    zoom = 15
  )
```


These points represent the confluence between the Loire and the Sèvre Nantaise.
Restoring this missing link allowed us to recover the entire connected
component. This operation is documented in `patch_river_network()`.

## Distance to river mouth

Here we have everything to compute the distance to the river mouth for river
stations at least.

### River

We group our stations per catchment, and identify for each catchment transitions
point between freshwater and marine areas.

```{r}
river_mouth_district <- tar_read(river_mouth_district)
mouths_kept <- river_mouth_district |>
  filter(matched, district %in% district_kept)
mouth_coords <- sf::st_coordinates(mouths_kept)
```

```{r}
#| fig-width: 7
#| fig-height: 9

plot(st_geometry(basin), col = NA, border = "grey80")
plot(
  st_geometry(basin_kept),
  col = scales::alpha(cols[basin_kept$district], 0.15),
  border = "grey70",
  add = TRUE
)
points(
  river_coords[, "X"],
  river_coords[, "Y"],
  col = scales::alpha("grey60", 0.25),
  pch = ".",
  cex = 0.5
)
points(
  in_scope$longitude,
  in_scope$latitude,
  col = cols[in_scope$district],
  pch = 19,
  cex = 0.3
)
points(mouth_coords[, "X"], mouth_coords[, "Y"], col = "black", pch = 17, cex = 0.8)
legend(
  "bottomright",
  legend = c(names(cols), "River mouth"),
  col = c(cols, "black"),
  pch = c(rep(19, length(cols)), 17),
  bty = "n"
)
```

We then compute for each station the channel-length distance to the closest
river mouth of its own catchment (so a station can only be matched to a mouth of
its own catchment - see `compute_distance_to_mouth()`).

```{r}
river_distance <- tar_read(river_distance_to_mouth)

ggplot(river_distance, aes(x = dist_mouth_m / 1000, fill = district)) +
  geom_histogram(bins = 30, color = "white") +
  scale_fill_manual(values = cols, guide = "none") +
  facet_wrap(~district, scales = "free_y") +
  labs(x = "Distance to river mouth (km)", y = "Number of stations")
```

```{r}
river_distance |>
  group_by(district) |>
  summarise(
    n = n(),
    median_km = round(median(dist_mouth_m) / 1000, 1),
    max_km = round(max(dist_mouth_m) / 1000, 1)
  ) |>
  arrange(median_km) |>
  knitr::kable()
```

```{r}
#| fig-width: 7
#| fig-height: 9.5

river_coords_df <- as.data.frame(river_coords)
mouth_coords_df <- as.data.frame(mouth_coords)

ggplot() +
  geom_sf(data = basin, fill = NA, color = "grey80") +
  geom_sf(data = basin_kept, aes(fill = district), color = "grey70", alpha = 0.10) +
  scale_fill_manual(values = cols, name = NULL) +
  geom_point(
    data = river_distance,
    aes(longitude, latitude, color = dist_mouth_m / 1000),
    size = 0.4
  ) +
  scale_color_viridis_c(name = "Distance to river mouth (km)") +
  geom_point(data = mouth_coords_df, aes(X, Y), shape = 17, size = 1) +
  coord_sf() +
  labs(x = NULL, y = NULL) +
  guides(
    fill = guide_legend(nrow = 1, override.aes = list(alpha = 0.3)),
    color = guide_colorbar(barwidth = 10)
  ) +
  theme(legend.box = "vertical", panel.grid = element_blank())
```

Distances range from a few tens of meters (stations right at the mouth) to
under 950km. Escaut-Somme has the shortest distances overall (small,
coastal catchment), while Seine stations sit farthest inland on average.

### Sea operations: inland vs. sea-side

For sea operations, we first need to know whether an operation sits
upstream or downstream of the river mouth before a distance is even
meaningful. One idea: use `hydrographic_basin` (land catchment) vs.
`hydrographic_area` (coastal/transitional water body) - if a point falls
in the basin polygon, it's inland; if in the area polygon, sea-side.

To validate our choice let's have a look at the Gironde and Loire estuaries.

```{r}
#| fig-width: 12
#| fig-height: 7

hydro_area <- tar_read(hydrographic_area)

plot_estuary <- function(name, xlim, ylim) {
  bbox <- st_as_sfc(st_bbox(
    c(xmin = xlim[1], xmax = xlim[2], ymin = ylim[1], ymax = ylim[2]),
    crs = 4326
  ))
  basin_e <- st_intersection(st_make_valid(basin), bbox)
  area_e <- st_intersection(st_make_valid(hydro_area), bbox)

  ggplot() +
    geom_sf(data = basin_e, aes(fill = "Basin (catchment)"), color = NA, alpha = 0.4) +
    geom_sf(data = area_e, aes(fill = "Area (coastal water body)"), color = NA, alpha = 0.4) +
    geom_point(data = mouth_coords_df, aes(X, Y), shape = 17, size = 1) +
    scale_fill_manual(
      name = NULL,
      values = c("Basin (catchment)" = "steelblue", "Area (coastal water body)" = "firebrick")
    ) +
    coord_sf(xlim = xlim, ylim = ylim) +
    labs(x = NULL, y = NULL, title = name) +
    theme(panel.grid = element_blank())
}

p_gironde <- plot_estuary("Gironde", c(-1.3, -0.3), c(44.8, 45.7))
p_loire <- plot_estuary("Loire", c(-2.3, -1.0), c(47.0, 47.5))

p_gironde + p_loire + plot_layout(guides = "collect") & theme(legend.position = "bottom")
```

From what we see there it is quite clear the the natural delimitation is the
basin polygon whose boundaries fit well with our river mouths.

So we use `hydrographic_basin` alone as the inland test: an operation
falling inside a basin polygon is inland, everything else is sea-side (no
fallback nearest-match, unlike `match_sea_district()` - a point just
offshore should stay "sea", not get pulled into the nearest basin). This is
now target `operation_location` (`classify_operation_location()`, applied
to every operation, `river` included, not just sea ones). `district` is
kept too (not just `inland`), so it can later be joined against
`river_distance_to_mouth` once we can compute the same distance for
sea-side operations. `river` operations outside our four catchments of
interest are dropped here - the same filter also drops the border-precision
edge cases found earlier (no matched `district` at all, so they can't be in
`district_kept` either).

```{r}
operation_location <- tar_read(operation_location)

operation_location |>
  mutate(district = tidyr::replace_na(district, "(sea, no catchment)")) |>
  count(district, survey, inland) |>
  tidyr::pivot_wider(
    names_from = inland,
    values_from = n,
    values_fill = 0,
    names_prefix = "inland_"
  ) |>
  arrange(district, survey) |>
  knitr::kable()
```

One row per (district, survey) makes the consistency check easy: every real
catchment row has `inland_FALSE = 0` and the "(sea, no catchment)" rows have
`inland_TRUE = 0` - a `district` label and `inland` always agree, exactly as
the classification should behave. Every surviving `river` operation is
inland now (that's what the filter guarantees) and split across our four
catchments - `operation_year_filtered` had 21,761 `river` operations before
this filter, 14,359 after: 7,402 dropped, of which 90 were the
border-precision misses and the rest sat in catchments outside our scope
(Rhone, Corse, Rhin-Meuse). `nurse` stays almost entirely sea-side, `pomet`
splits between sea-side and inland, and none of the sea surveys' inland
operations fall outside our four catchments either.

### Choosing a sea-match threshold

Sea operations that miss every basin polygon are candidates for a
nearest-match fallback against `hydrographic_area`, within a distance
threshold to be chosen. Restricting to `nurse`/`pomet` (the `river`
border-precision misses are already explained, and aren't a coastal
question):

```{r}
operation_year_filtered <- tar_read(operation_year_filtered)

outside_sea <- operation_year_filtered |>
  filter(survey != "river") |>
  sf::st_as_sf(coords = c("longitude", "latitude"), crs = 4326, remove = FALSE) |>
  sf::st_join(basin["district"], join = sf::st_within) |>
  filter(is.na(district)) |>
  select(-district)

nearest_idx <- sf::st_nearest_feature(outside_sea, hydro_area)
outside_sea <- outside_sea |>
  mutate(dist_km = as.numeric(sf::st_distance(
    outside_sea, hydro_area[nearest_idx, ],
    by_element = TRUE
  )) / 1000)

thresholds <- c(1, 2, 5, 10, 15, 20)
tibble::tibble(threshold_km = thresholds) |>
  mutate(
    pct_matched = round(100 * purrr::map_dbl(threshold_km, \(t) mean(outside_sea$dist_km <= t)), 1),
    n_unmatched = purrr::map_int(threshold_km, \(t) sum(outside_sea$dist_km > t))
  ) |>
  knitr::kable()
```

10km (`params$distance_match_max`, already used for river mouth matching)
leaves 7 `nurse` operations unmatched, all clustered right at the Gironde
mouth, 10.0-11.3km out - past 15km nothing more is gained, coverage is
already 100%.

```{r}
#| fig-width: 7
#| fig-height: 9

threshold_candidate <- 10
outside_sea <- outside_sea |>
  mutate(matched = dist_km <= threshold_candidate)

ggplot() +
  geom_sf(data = basin, fill = NA, color = "grey80") +
  geom_sf(data = hydro_area, fill = "firebrick", alpha = 0.1, color = NA) +
  geom_sf(data = outside_sea, aes(color = matched), size = 1.3) +
  scale_color_manual(
    values = c(`TRUE` = "steelblue", `FALSE` = "firebrick"),
    name = paste0("Matched within ", threshold_candidate, "km")
  ) +
  coord_sf(xlim = c(-1.45, -0.9), ylim = c(45.4, 45.75)) +
  theme(panel.grid = element_blank())
```

Zoomed on the Gironde mouth, where all 7 unmatched points sit: they're
just outside the coastal polygon's western edge, right at the estuary
mouth - nothing suspicious, just borderline. 10km vs. 15km is the real
call to make (100% coverage vs. reusing the existing `distance_match_max`
param) - over to you before I wire this into
`classify_operation_location()`.

## Rephy interpolation

River mouth points come from the AMOBIO river network (Lambert-93), so we
reproject to lon/lat before plotting alongside the coastline.

```{r}
#| fig-width: 6
#| fig-height: 9

river_mouth_sf <- river_mouth |>
  sf::st_as_sf() |>
  sf::st_transform(4326)

ggplot() +
  geom_sf(data = coastline(), linewidth = 0.3) +
  geom_sf(data = river_mouth_sf, size = 0.8) +
  coord_sf(
    xlim = sf::st_bbox(river_mouth_sf)[c("xmin", "xmax")],
    ylim = sf::st_bbox(river_mouth_sf)[c("ymin", "ymax")]
  ) +
  labs(x = NULL, y = NULL)
```

### Sea mesh and river network

Before building a combined graph, we look at the sea mesh (built the same way
as in the interpolation notebook) and the AMOBIO river network nodes overlaid
on the same map. The river network has ~1M nodes, so we plot with base R
(much faster than ggplot for this many points) and use tiny markers.

```{r}
rephy_filtered <- tar_read(rephy_data_sample)
mesh <- build_rephy_mesh(rephy_filtered)

network <- tar_read(river_network_full)
river_coords <- network$nodes |>
  sf::st_as_sf() |>
  sf::st_transform(4326) |>
  sf::st_coordinates()
```

```{r}
#| fig-width: 6
#| fig-height: 9

plot(mesh, main = "")
points(river_coords[, "X"], river_coords[, "Y"], col = "steelblue", pch = ".")
points(
  river_mouth_sf |> sf::st_coordinates(),
  col = "firebrick", pch = 19, cex = 0.6
)
legend(
  "bottomright",
  legend = c("River network node", "River mouth"),
  col = c("steelblue", "firebrick"),
  pch = 19,
  bty = "n"
)
```

### River mouths vs. hydrographic area

We match each river mouth to a district (of interest for our study) and
highlight the ones that still don't match. These points will be removed.

```{r}
#| fig-width: 6
#| fig-height: 9

hydro <- tar_read(hydrographic_area)
district_kept <- c("Loire", "Ardour-Garonne", "Seine", "Escaut-Somme")
hydro_kept <- hydro |>
  filter(district %in% district_kept)

river_mouth_district <- tar_read(river_mouth_district)

ggplot() +
  geom_sf(data = hydro_kept, aes(fill = district), color = NA, alpha = 0.4) +
  geom_sf(data = coastline(), linewidth = 0.3) +
  geom_sf(data = river_mouth_district, aes(color = matched), size = 1) +
  scale_color_manual(values = c("TRUE" = "grey20", "FALSE" = "firebrick")) +
  coord_sf(
    xlim = sf::st_bbox(river_mouth_sf)[c("xmin", "xmax")],
    ylim = sf::st_bbox(river_mouth_sf)[c("ymin", "ymax")]
  ) +
  labs(x = NULL, y = NULL, fill = "District", color = "Matched (≤ 10 km)")
```

### Restricted river network

We grow the network outward (channel length) from the matched river mouths,
using a single Dijkstra run from a virtual source connected to every matched
mouth at once - much cheaper than one search per mouth. This reuses
`river_network` (restricted to `river_station_network_max_dist`, the river
station radius) rather than a separate 100km REPHY-only cutoff - the
dedicated REPHY-scope target was a duplicate of this one and has been
removed, since `river_network`'s radius is a safe superset.

```{r}
#| fig-width: 6
#| fig-height: 9

network_restricted <- tar_read(river_network)

river_restricted_nodes <- network_restricted$nodes |>
  sf::st_as_sf() |>
  sf::st_transform(4326)

ggplot() +
  geom_sf(data = coastline(), linewidth = 0.3) +
  geom_sf(
    data = river_restricted_nodes,
    aes(color = "River network node"),
    size = 0.1,
    alpha = 0.1
  ) +
  geom_sf(
    data = river_mouth_district |> filter(matched),
    aes(color = "River mouth"),
    size = 1
  ) +
  geom_point(
    data = rephy_filtered,
    aes(x = longitude, y = latitude, color = "REPHY station"),
    size = 0.8
  ) +
  scale_color_manual(
    values = c(
      "River network node" = "steelblue",
      "River mouth" = "firebrick",
      "REPHY station" = "darkorange"
    )
  ) +
  coord_sf(
    xlim = sf::st_bbox(river_restricted_nodes)[c("xmin", "xmax")],
    ylim = sf::st_bbox(river_restricted_nodes)[c("ymin", "ymax")]
  ) +
  labs(x = NULL, y = NULL, color = NULL)
```
