---
title: "Compute distance to river mouth"
bibliography: ../references.bib
format:
  html:
    theme: cosmo
    code-fold: true
    toc: true
    embed-resources: true
execute:
    message: false
    warning: false
---

The goal is to compute the distance to river mouth REPHY data and fishing
operations at sea.

## Todo

- [x] Extract river mouth points
- [x] Plot river mouth points
- [x] Simplify AMOBIO network
- [x] Filter river mouths
- [x] Restrict river network
- [x] Check rephy data overlap with restricted river network
- [ ] Compute distance to river mouth for rephy stations
- [ ] Update INLA model with distance to river mouth as new covariate
- [x] Reconstruct full river network for `river` survey stations
- [x] Snap `river` survey stations to the network
- [x] Classify `river` survey stations by true catchment (from ASPE's own
water body code, not guessed from the network - see caveat below)
- [x] Close the `Ardour-Garonne` under-classification gap (its own
pre-harmonization water body numbering, no basin letter)
- [x] Investigate why the Vendée cluster is disconnected (no mouth flagged,
~77m from the main network - a broken digitized link, not real
separation)
- [x] Patch the missing link (Nantes/Sèvre Nantaise) and confirm the
cluster recovers
- [x] Snap directly against the restricted network now that it's fixed
- [ ] Compute distance to river mouth for `river` survey stations

## Setup

```{r}
library(renv)
renv::load()
library(targets)
library(here)
tar_config_set(
  store = here::here("_targets")
)
library(ggplot2)
library(dplyr)
library(sf)
devtools::load_all()
riversea::nice_theme()
```

```{r}
river_mouth <- tar_read(amobio_river_mouth)
```

River mouth points come from the AMOBIO river network (Lambert-93), so we
reproject to lon/lat before plotting alongside the coastline.

```{r}
#| fig-width: 6
#| fig-height: 9

river_mouth_sf <- river_mouth |>
  sf::st_as_sf() |>
  sf::st_transform(4326)

ggplot() +
  geom_sf(data = coastline(), linewidth = 0.3) +
  geom_sf(data = river_mouth_sf, size = 0.8) +
  coord_sf(
    xlim = sf::st_bbox(river_mouth_sf)[c("xmin", "xmax")],
    ylim = sf::st_bbox(river_mouth_sf)[c("ymin", "ymax")]
  ) +
  labs(x = NULL, y = NULL)
```

### Sea mesh and river network

Before building a combined graph, we look at the sea mesh (built the same way
as in the interpolation notebook) and the AMOBIO river network nodes overlaid
on the same map. The river network has ~1M nodes, so we plot with base R
(much faster than ggplot for this many points) and use tiny markers.

```{r}
rephy_filtered <- tar_read(rephy_data_sample)
mesh <- build_rephy_mesh(rephy_filtered)

network <- tar_read(amobio_network)
river_coords <- network$nodes |>
  sf::st_as_sf() |>
  sf::st_transform(4326) |>
  sf::st_coordinates()
```

```{r}
#| fig-width: 6
#| fig-height: 9

plot(mesh, main = "")
points(river_coords[, "X"], river_coords[, "Y"], col = "steelblue", pch = ".")
points(
  river_mouth_sf |> sf::st_coordinates(),
  col = "firebrick", pch = 19, cex = 0.6
)
legend(
  "bottomright",
  legend = c("River network node", "River mouth"),
  col = c("steelblue", "firebrick"),
  pch = 19,
  bty = "n"
)
```

### River mouths vs. hydrographic area

We match each river mouth to a district (of interest for our study) and
highlight the ones that still don't match. These points will be removed.

```{r}
#| fig-width: 6
#| fig-height: 9

hydro <- tar_read(hydrographic_area)
district_kept <- c("Loire", "Ardour-Garonne", "Seine", "Escaut-Somme")
hydro_kept <- hydro |>
  filter(district %in% district_kept)

river_mouth_district <- tar_read(river_mouth_district)

ggplot() +
  geom_sf(data = hydro_kept, aes(fill = district), color = NA, alpha = 0.4) +
  geom_sf(data = coastline(), linewidth = 0.3) +
  geom_sf(data = river_mouth_district, aes(color = matched), size = 1) +
  scale_color_manual(values = c("TRUE" = "grey20", "FALSE" = "firebrick")) +
  coord_sf(
    xlim = sf::st_bbox(river_mouth_sf)[c("xmin", "xmax")],
    ylim = sf::st_bbox(river_mouth_sf)[c("ymin", "ymax")]
  ) +
  labs(x = NULL, y = NULL, fill = "District", color = "Matched (≤ 10 km)")
```

### Restricted river network

We grow the network outward (channel length) from the matched river mouths,
up to 100 km, using a single Dijkstra run from a virtual source connected to
every matched mouth at once - much cheaper than one search per mouth.


```{r}
#| fig-width: 6
#| fig-height: 9

network_restricted <- tar_read(amobio_network_restricted)

river_restricted_nodes <- network_restricted$nodes |>
  sf::st_as_sf() |>
  sf::st_transform(4326)

ggplot() +
  geom_sf(data = coastline(), linewidth = 0.3) +
  geom_sf(
    data = river_restricted_nodes,
    aes(color = "River network node"),
    size = 0.1,
    alpha = 0.1
  ) +
  geom_sf(
    data = river_mouth_district |> filter(matched),
    aes(color = "River mouth"),
    size = 1
  ) +
  geom_point(
    data = rephy_filtered,
    aes(x = longitude, y = latitude, color = "REPHY station"),
    size = 0.8
  ) +
  scale_color_manual(
    values = c(
      "River network node" = "steelblue",
      "River mouth" = "firebrick",
      "REPHY station" = "darkorange"
    )
  ) +
  coord_sf(
    xlim = sf::st_bbox(river_restricted_nodes)[c("xmin", "xmax")],
    ylim = sf::st_bbox(river_restricted_nodes)[c("ymin", "ymax")]
  ) +
  labs(x = NULL, y = NULL, color = NULL)
```

## Snapping river survey stations to the network

Same idea as above, but for river (ASPE) stations instead of REPHY.
Stations reach far upstream, so we reconstruct with a 2000km radius
instead of REPHY's 100km, then snap each station to its nearest node.

```{r}
network_river <- tar_read(amobio_network_river)
river_stations <- tar_read(river_station_snapped)
district_kept <- c("Loire", "Ardour-Garonne", "Seine", "Escaut-Somme")

cat("Median snap distance (m):", round(median(river_stations$snap_dist_m), 1), "\n")
cat("N snapping over 1km:", sum(river_stations$snap_dist_m > 1000), "\n")
```

Snapping is straight-line, not channel-length, so a large distance flags a
bad station coordinate rather than a real network gap - and there's
barely any left (see below for why).

Each station's catchment comes from ASPE's own water body code, not from
the network: `hydrographic_area` only covers the coast (median 150km from
an inland station), so it can't tell a station outside our four catchments
apart from a genuine network gap. The code's basin-letter prefix gives the
district directly, once two quirks are handled: the country code (`FR`)
sometimes comes before the basin letter, sometimes not, and Ardour-Garonne
used its own pre-harmonization numbering (no basin letter at all - fixed
in `classify_water_body_district()`).

```{r}
table(river_stations$district, useNA = "always")
```

Escaut-Somme has zero stations (matches it having the fewest matched
mouths, 15 of 274). The remaining NAs are genuinely missing codes, mostly
outside our catchments anyway - checked geographically, none of them
reveal a hidden classification gap like Ardour-Garonne did.

```{r}
#| fig-width: 7
#| fig-height: 9

in_scope <- river_stations |> filter(district %in% district_kept)
river_coords <- network_river$nodes |> sf::st_as_sf() |> sf::st_transform(4326) |> sf::st_coordinates()
cols <- c(Loire = "steelblue", "Ardour-Garonne" = "firebrick", Seine = "darkgreen", "Escaut-Somme" = "orange")

plot(river_coords[, "X"], river_coords[, "Y"], col = "grey85", pch = ".", asp = 1.5, xlab = NULL, ylab = NULL)
points(in_scope$longitude, in_scope$latitude, col = cols[in_scope$district], pch = 19, cex = 0.3)
legend("bottomright", legend = names(cols), col = cols, pch = 19, bty = "n")
```

Clean catchment boundaries, no stray points in the wrong basin.

### Fixing a network gap (Vendée / Nantes)

While investigating, one ~6,000-node component near the Vendée coast had
no station snapping to it despite sitting inside Loire territory - and no
node in it was flagged a river mouth. Rather than assume real
hydrological separation, checking the gap to the rest of the network:

```{r}
amobio_network <- tar_read(amobio_network)
network_before <- amobio_network
network_before$edges <- network_before$edges |> filter(!(from == 614458 & to == 614552))

comp_before <- igraph::components(igraph::graph_from_edgelist(
  as.matrix(network_before$edges[c("from", "to")]), directed = FALSE
))
vendee_comp_nodes <- which(comp_before$membership == comp_before$membership[595001])
main_sf <- amobio_network$nodes |>
  filter(node_id %in% which(comp_before$membership == which.max(comp_before$csize))) |>
  sf::st_as_sf()
vendee_sf <- amobio_network$nodes |> filter(node_id %in% vendee_comp_nodes) |> sf::st_as_sf()

gap_dist_m <- as.numeric(sf::st_distance(
  vendee_sf, main_sf[sf::st_nearest_feature(vendee_sf, main_sf), ], by_element = TRUE
))
cat("Component size:", length(vendee_comp_nodes), "\n")
cat("Closest gap:", round(min(gap_dist_m), 1), "m -", sum(gap_dist_m < 500), "nodes within 500m.\n")
```

A gap under 100m is within GPS error, and several nodes are within 500m -
this reads as a missing digitized link, not a real separation. The
clearest of these gaps sits at Nantes, where the Sèvre Nantaise joins the
Loire - checked against a basemap that shows the rivers:

```{r}
gap_pts <- amobio_network$nodes |>
  filter(node_id %in% c(614458, 614552)) |>
  sf::st_as_sf() |>
  sf::st_transform(4326) |>
  mutate(label = c("Vendée-side node", "Main-network node"))

leaflet::leaflet(gap_pts) |>
  leaflet::addProviderTiles(leaflet::providers$Esri.WorldTopoMap) |>
  leaflet::addCircleMarkers(color = c("firebrick", "steelblue"), radius = 8, label = ~label) |>
  leaflet::setView(
    lng = mean(sf::st_coordinates(gap_pts)[, 1]),
    lat = mean(sf::st_coordinates(gap_pts)[, 2]),
    zoom = 15
  )
```

Both points sit right on the river, either side of the confluence, 162m
apart. `patch_amobio_network()` adds this edge (length computed from the
node geometries, not hardcoded) and is already active in `amobio_network`
above - snapping and the map/table earlier already reflect the fix:

```{r}
comp_after <- igraph::components(igraph::graph_from_edgelist(
  as.matrix(amobio_network$edges[c("from", "to")]), directed = FALSE
))
cat(
  "Same component as main network now?",
  comp_after$membership[595001] == comp_after$membership[main_sf$node_id[1]], "\n"
)
recovered <- in_scope |> filter(node_id %in% vendee_comp_nodes)
cat("Recovered:", nrow(recovered), "stations, snap distance now median",
    round(median(recovered$snap_dist_m), 1), "m\n")
```

