---
title: "Food web structures along the river-sea continuum"
bibliography: ../references.bib
format:
  html:
    theme: cosmo
    code-fold: true
    toc: true
    embed-resources: true
execute:
    message: false
    warning: false
---

## Todo

- [x] Species set difference and intersection between river and sea dataset
- [x] Plot metaweb
- [x] Plot local webs along the continuum
- [x] Refine food web plots
- [x] Add map with operations to local food web plot
- [x] Diet category analysis along the continuum
- [x] Add information to the metaweb (e.g. migratory status of species, invasive
or native, freshwater, marine, estuarine, etc.)
- [x] General information at the species level, such as, mean size, count,
status, etc.
- [x] Most common invasive species
- [x] Location of invasive species
- [ ] Impact of invasive/diadromous species on food web structure
(richness-controlled)
- [ ] **(blocked)** Structure metrics versus distance to river mouth

## Setup

```{r}
library(renv)
renv::load()
library(targets)
library(here)
tar_config_set(
  store = here::here("_targets")
)
library(ggplot2)
library(dplyr)
library(sf)
devtools::load_all()
nice_theme()
```

## Species composition along the continuum

We start by looking at how much species overlap there is between the three
surveys that make up the river-estuary-sea continuum: `river` (freshwater,
ASPE), `nurse` (estuary and sea) and `pomet` (estuary and sea).

```{r}
#| fig-width: 6
#| fig-height: 5.5

size <- tar_read(size_year_filtered)
plot_venn_diagram(size)
```

Three species are shared by all three surveys: *Anguilla anguilla* (European
eel, catadromous), *Chelon ramada* (grey mullet, euryhaline
estuarine migrant) and *Platichthys flesus* (European flounder, catadromous
flatfish).

More surprising is the `river`/`pomet` overlap (16 species), which excludes
`nurse` entirely. These 16 are all strict freshwater residents (e.g. roach,
bream, perch, pikeperch, barbel, chub, three-spined stickleback), not diadromous
species. These species most probably come from `pomet` fishing operation that
occurs deep into the estuaries (Girond and Loire).

So `river` and open sea share species through because of two reasons.
First, through diadromous species that live across the whole continuum.
Second, through residents caught by `pomet` operations deep into estuaries.

## Metaweb

The metaweb represents every possible predator-prey interaction across the
whole continuum, aggregated over all size classes so each species appears as a
single node. Species are placed vertically by trophic level, from resources
(grey, bottom) up to fish (top), colored by whether they prey on other fish at
any life stage, and sized by each species' mean observed body length.

```{r}
#| fig-width: 14
#| fig-height: 9

web_list <- tar_read(web_list)
resource_list <- tar_read(resource_list)
diet <- tar_read(diet)
size <- tar_read(size_year_filtered)
plot_metaweb(web_list$metaweb, resource_list, diet, size)
```

`r nrow(aggregate_species_web(web_list$metaweb)) - length(resource_list)`
fish species and `r length(resource_list)` basal resources make up the
metaweb, for a total of `r sum(aggregate_species_web(web_list$metaweb))`
possible interactions at the species level.

Here is the interactive version of the static plot above, so we can zoom in and
inspect in more details the metaweb.

```{r}
#| fig-width: 11
#| fig-height: 9

species_traits <- tar_read(species_traits)
plot_metaweb_interactive(
  web_list$metaweb,
  resource_list,
  diet,
  size,
  species_traits
)
```

## Species-level information

Environment (freshwater/marine/both), demersal/pelagic position and
diadromous status come from FishBase's `species()` table; native status is
FishBase's `country()` table filtered to France. Two caveats:

- **Diadromous status** (`is_diadromous`) is `TRUE` for FishBase migratory
categories that cross the fresh/salt water boundary (`amphidromous`,
`anadromous`, `catadromous`), `FALSE` for every other documented category
(`non-migratory`, `potamodromous`, `oceanodromous`), and `NA` only when
FishBase leaves the category genuinely blank rather than coding it
"non-migratory". In practice, almost all `NA` species are strict freshwater
residents (roach, bream, perch, etc.), so read `NA` as "presumed `FALSE`"
rather than a real unknown.

- **Native status** (`is_native`) is `TRUE` or `FALSE` ("invasive"). The
three gobies *Neogobius melanostomus*, *Ponticola kessleri* and
*Proterorhinus semilunaris* have no France record in FishBase's `country()`
table, even though they are documented recent invaders in French rivers in
the literature, so they are hardcoded to `FALSE` rather than left `NA`.

```{r}
species_summary <- size |>
  group_by(species_valid) |>
  summarise(
    n_individuals = n(),
    n_operations = n_distinct(operation_id),
    mean_length = mean(length, na.rm = TRUE),
    .groups = "drop"
  ) |>
  left_join(species_traits, by = c("species_valid" = "species")) |>
  arrange(desc(n_individuals))
```

```{r}
#| fig-width: 10
#| fig-height: 3.5

count_bar <- function(data, var, title) {
  data |>
    count(.data[[var]]) |>
    mutate(
      val = forcats::fct_reorder(
        tidyr::replace_na(as.character(.data[[var]]), "NA"), n
      )
    ) |>
    ggplot(aes(n, val)) +
    geom_col() +
    labs(x = "Number of species", y = NULL, title = title)
}

patchwork::wrap_plots(
  count_bar(species_summary, "environment", "Environment"),
  count_bar(species_summary, "is_diadromous", "Diadromous"),
  count_bar(species_summary, "is_native", "Native (France)"),
  nrow = 1
)
```

The full species-level information table

```{r}
species_summary |>
  select(
    species = species_valid, environment, is_diadromous, is_native,
    demers_pelag, n_individuals, n_operations, mean_length
  ) |>
  arrange(species) |>
  knitr::kable(digits = 1)
```

## Invasive and diadromous species

We now look specifically at two groups: **invasive** species (`!is_native`)
and **diadromous** species (`is_diadromous`), i.e. those that cross the
fresh/salt water boundary during their life cycle.

For each species we also report its diet, taken from the literature-curated
diet table and listing every resource category referenced.

```{r}
diet_list <- c("fish", resource_list)

species_diet_summary <- diet |>
  select(species, all_of(diet_list)) |>
  group_by(species) |>
  summarise(across(everything(), max), .groups = "drop") |>
  rowwise() |>
  mutate(
    diet_summary = paste(
      diet_list[c_across(all_of(diet_list)) > 0],
      collapse = ", "
    )
  ) |>
  ungroup() |>
  select(species, diet_summary)

operation_year_filtered <- tar_read(operation_year_filtered)

sp_status <- size |>
  left_join(species_traits, by = c("species_valid" = "species")) |>
  left_join(
    operation_year_filtered |> select(operation_id, year, longitude, latitude),
    by = "operation_id"
  ) |>
  mutate(is_invasive = !is_native)

top_species_per_survey <- function(data, filter_col, n = 6) {
  data |>
    filter(.data[[filter_col]]) |>
    group_by(survey, species_valid) |>
    summarise(
      n_individuals = n(), mean_length = mean(length, na.rm = TRUE),
      .groups = "drop"
    ) |>
    group_by(survey) |>
    slice_max(n_individuals, n = n) |>
    ungroup() |>
    left_join(species_diet_summary, by = c("species_valid" = "species")) |>
    arrange(survey, desc(n_individuals))
}

# Points are per-operation totals (summed across years) for the map, and
# per-year totals (summed across operations) for the time series.
plot_species_map <- function(data, species_list) {
  map_data <- data |>
    filter(species_valid %in% species_list) |>
    group_by(species_valid, operation_id, longitude, latitude) |>
    summarise(n = n(), .groups = "drop")
  ggplot(map_data) +
    geom_sf(data = coastline(), linewidth = 0.3, inherit.aes = FALSE) +
    geom_point(aes(longitude, latitude, size = n, color = n), alpha = 0.6) +
    coord_sf(xlim = c(-5, 9), ylim = c(41, 51)) +
    scale_color_viridis_c(trans = "log10") +
    scale_size_continuous(range = c(0.3, 3), guide = "none") +
    facet_wrap(~species_valid, ncol = 4) +
    labs(x = NULL, y = NULL, color = "N caught") +
    theme(
      plot.background = element_rect(fill = "white", color = NA),
      panel.background = element_rect(fill = "white", color = NA),
      panel.grid = element_blank(),
      strip.text = element_text(face = "italic", size = 8)
    )
}

n_ops_survey_year <- size |>
  distinct(operation_id, survey) |>
  left_join(operation_year_filtered |> select(operation_id, year), by = "operation_id") |>
  count(survey, year, name = "n_ops")

plot_species_occupancy <- function(data, species_list) {
  present <- data |>
    filter(species_valid %in% species_list) |>
    group_by(species_valid, survey, year) |>
    summarise(n_present = n_distinct(operation_id), .groups = "drop")
  occ_data <- tidyr::expand_grid(species_valid = species_list, n_ops_survey_year) |>
    left_join(present, by = c("species_valid", "survey", "year")) |>
    mutate(n_present = tidyr::replace_na(n_present, 0), frac = n_present / n_ops) |>
    filter(n_ops > 0)
  ggplot(occ_data, aes(year, frac, color = survey)) +
    geom_line() +
    geom_point(size = 1) +
    facet_wrap(~species_valid, ncol = 4, scales = "free_y") +
    labs(x = NULL, y = "Fraction of operations\nwith species present") +
    theme(strip.text = element_text(face = "italic", size = 8))
}
```

### Invasive species

The six most caught invasive species per survey, with mean observed
length (cm) and diet. `nurse` has no invasive species recorded at all.

```{r}
invasive_table <- top_species_per_survey(sp_status, "is_invasive")
invasive_table |>
  rename(species = species_valid, diet = diet_summary) |>
  knitr::kable(digits = 1)
```

Invasive species are essentially a `river` phenomenon: they're present in
half of river operations, versus a handful of `pomet` operations and none
in `nurse`.

Where each is caught (per-operation total, summed across years):

```{r}
#| fig-width: 10
#| fig-height: 6

plot_species_map(sp_status, unique(invasive_table$species_valid))
```

*Neogobius melanostomus* stands out with a tight cluster in the northeast
(Moselle/Rhine basin) rather than the wide spread the other introduced
species show - consistent with it being a recent, still-localized invader.

And how each has evolved over the years, as the fraction of operations
where it was caught, so this reflects abundance rather than sampling
effort, split by survey since the fraction isn't comparable across surveys
with different gear:

```{r}
#| fig-width: 10
#| fig-height: 6

plot_species_occupancy(sp_status, unique(invasive_table$species_valid))
```

*Neogobius melanostomus* shows a clear, sustained increase since its first
detection around 2011.

```{r}
#| fig-width: 6
#| fig-height: 3

# is_diadromous can be NA (undocumented) - any(..., na.rm = TRUE) treats
# that as "not confirmed present" rather than letting a single NA make the
# whole operation's presence flag NA.
op_prop <- sp_status |>
  group_by(operation_id, survey) |>
  summarise(
    has_invasive = any(is_invasive),
    has_diadromous = any(is_diadromous, na.rm = TRUE),
    .groups = "drop"
  ) |>
  group_by(survey) |>
  summarise(
    n_operations = n(),
    pct_with_invasive = 100 * mean(has_invasive),
    pct_with_diadromous = 100 * mean(has_diadromous),
    .groups = "drop"
  )

ggplot(op_prop, aes(survey, pct_with_invasive)) +
  geom_col(width = 0.6) +
  labs(x = NULL, y = "% operations with\n≥1 invasive species")
```

### Diadromous species

Same view for diadromous species: the six most caught per survey, with
mean observed length and diet.

```{r}
diadromous_table <- top_species_per_survey(sp_status, "is_diadromous")
diadromous_table |>
  rename(species = species_valid, diet = diet_summary) |>
  knitr::kable(digits = 1)
```

Unlike invasive species, diadromous species are present across the whole
continuum, though slightly less so in `nurse` - consistent with `nurse`
being a nursery habitat rather than a migration corridor.

Where each is caught:

```{r}
#| fig-width: 10
#| fig-height: 8

plot_species_map(sp_status, unique(diadromous_table$species_valid))
```

The coastal/estuarine specialists (*Pomatoschistus* gobies, *Platichthys
flesus*, *Chelon ramada*, *Alosa fallax*) hug the coastline and estuary
mouths, while *Anguilla anguilla* and *Salmo trutta* spread across the
entire river network, as expected for species that migrate far upstream.

And their occupancy time series, split by survey:

```{r}
#| fig-width: 10
#| fig-height: 8

plot_species_occupancy(sp_status, unique(diadromous_table$species_valid))
```

```{r}
#| fig-width: 6
#| fig-height: 3

ggplot(op_prop, aes(survey, pct_with_diadromous)) +
  geom_col(width = 0.6) +
  labs(x = NULL, y = "% operations with\n≥1 diadromous species")
```

## Local food webs along the continuum

To see how food web structure evolves along the continuum, we pick three
individual operations along Gironde: one on the coast (`pomet`), one within the
Gironde estuary itself (`nurse`), and one upstream on the Dordogne (`river`).

```{r}
foodweb_structure <- tar_read(foodweb_structure)
operation_year_filtered <- tar_read(operation_year_filtered)

ops <- c(
  pomet = "61832307", nurse = "2022_1_GIR12_14_S", river = "83544"
)
titles <- c(
  pomet = "Coast",
  nurse = "Estuary",
  river = "River"
)
```

First, a map showing where each of the three operations sits along the
coastline/estuary/river.

```{r}
#| fig-width: 9
#| fig-height: 6

continuum_pts <- operation_year_filtered |>
  filter(operation_id %in% ops) |>
  mutate(
    label = titles[names(ops)[match(operation_id, ops)]]
  )

france <- rnaturalearth::ne_countries(
  geounit = "france", type = "map_units", scale = "medium", returnclass = "sf"
)
rivers_fr <- rnaturalearth::ne_download(
  scale = 10,
  type = "rivers_lake_centerlines",
  category = "physical",
  returnclass = "sf"
) |>
  sf::st_intersection(france)

e <- 0.4
ggplot() +
  geom_sf(data = rivers_fr, color = "grey60", linewidth = 0.3) +
  geom_sf(data = coastline(), linewidth = 0.3) +
  geom_point(
    data = continuum_pts,
    aes(longitude, latitude),
    color = "black",
    size = 3
  ) +
  ggrepel::geom_label_repel(
    data = continuum_pts,
    aes(longitude, latitude, label = label),
    size = 3.5
  ) +
  coord_sf(
    xlim = range(continuum_pts$longitude) + c(-e, e),
    ylim = range(continuum_pts$latitude) + c(-e, e)
  ) +
  labs(x = "Longitude", y = "Latitude") +
  theme(
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    panel.grid = element_blank()
  )
```

```{r}
#| fig-width: 16
#| fig-height: 6

# Shared size scale across all three panels, so `guides = "collect"` below
# merges them into a single legend instead of one per panel.
size_range <- (size |>
  filter(operation_id %in% ops) |>
  group_by(species_valid) |>
  summarise(v = mean(length, na.rm = TRUE), .groups = "drop"))$v |>
  range()

local_web_plots <- lapply(names(ops), function(nm) {
  fw_row <- foodweb_structure |> filter(operation_id == ops[nm])
  web <- fw_row$foodweb[[1]]
  op_size <- size |> filter(operation_id == ops[nm])
  subtitle <- sprintf(
    "Richness %d · Connectance %.2f · Mean TL %.1f",
    round(exp(fw_row$log_species_richness)), fw_row$connectance,
    fw_row$trophic_length
  )
  plot_local_foodweb(
    web, resource_list, diet, op_size,
    title = titles[nm], size_limits = size_range
  ) +
    labs(subtitle = subtitle) +
    theme(plot.subtitle = element_text(size = 9, color = "grey40"))
})

patchwork::wrap_plots(local_web_plots, nrow = 1) +
  patchwork::plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

The three webs illustrate a clear shift in composition: the coastal web is built
around marine flatfish and with several piscivorous predators; the estuarine web
is sparser, dominated by basal resources with only a couple of connected fish;
and the river web is dominated by small non-piscivorous cyprinids alongside a
handful of large piscivorous predators.

## Structure metrics along the continuum

Rather than comparing surveys as discrete boxplots, we map each operation's
structure metrics at its actual location, restricted to the Gironde
estuary/coast and the Dordogne river, to see how food web structure changes
continuously along the gradient instead of as three discrete groups.

We look at trophic length, diet overlap, connectance and species richness.
Diet overlap is the mean pairwise Jaccard similarity between predators' prey
sets (low = predators partition prey between them/complementarity, high =
predators eat largely the same things/redundancy) - unlike connectance, it
is close to independent of species richness
(`r round(cor(foodweb_structure$diet_overlap, foodweb_structure$log_species_richness), 2)`
here vs.
`r round(cor(foodweb_structure$connectance, foodweb_structure$log_species_richness), 2)`
for connectance), so it adds information richness alone doesn't already
give us.

```{r}
continuum_box <- foodweb_structure |>
  filter(longitude > -1.8, longitude < 1.3, latitude > 44.3, latitude < 46.0)

plot_metric_map <- function(data, metric, title) {
  ggplot(data) +
    geom_sf(
      data = rivers_fr, color = "grey70", linewidth = 0.3, inherit.aes = FALSE
    ) +
    geom_sf(data = coastline(), linewidth = 0.3, inherit.aes = FALSE) +
    geom_point(
      aes(longitude, latitude, color = .data[[metric]]), size = 1.8, alpha = 0.8
    ) +
    coord_sf(xlim = c(-1.8, 1.3), ylim = c(44.3, 46.0)) +
    scale_color_viridis_c() +
    labs(title = title, x = NULL, y = NULL, color = NULL) +
    theme(
      plot.background = element_rect(fill = "white", color = NA),
      panel.background = element_rect(fill = "white", color = NA),
      panel.grid = element_blank()
    )
}
```

```{r}
#| fig-width: 10
#| fig-height: 8

plot_metric_map(continuum_box, "trophic_length", "Trophic length")
```

```{r}
#| fig-width: 10
#| fig-height: 8

plot_metric_map(continuum_box, "diet_overlap", "Diet overlap")
```

```{r}
#| fig-width: 10
#| fig-height: 8

plot_metric_map(continuum_box, "connectance", "Connectance")
```

```{r}
#| fig-width: 10
#| fig-height: 8

plot_metric_map(continuum_box, "log_species_richness", "Log species richness")
```

## Diet composition by fish origin

We split species into three groups, reusing the same survey-membership
logic as the venn diagram above: `freshwater` (found only in `river`),
`marine` (found only in `nurse`/`pomet`), and `migratory` (found in both).
For each group, we look at the average diet composition across resource
categories (from the literature-curated diet table, not the constructed
food webs), normalised so each group's bars sum to 1.

```{r}
#| fig-width: 8
#| fig-height: 6.5

species_origin <- size |>
  distinct(species_valid, survey) |>
  group_by(species_valid) |>
  summarise(
    surveys = paste(sort(unique(survey)), collapse = ","), .groups = "drop"
  ) |>
  mutate(
    has_river = grepl("river", surveys),
    has_marine = grepl("nurse|pomet", surveys),
    origin = case_when(
      has_river & !has_marine ~ "freshwater",
      !has_river & has_marine ~ "marine",
      has_river & has_marine ~ "migratory"
    )
  )

species_diet <- diet |>
  select(species, all_of(resource_list)) |>
  group_by(species) |>
  summarise(across(everything(), mean), .groups = "drop") |>
  inner_join(
    species_origin |> select(species_valid, origin),
    by = c("species" = "species_valid")
  )

origin_diet <- species_diet |>
  group_by(origin) |>
  summarise(across(all_of(resource_list), mean), .groups = "drop") |>
  tidyr::pivot_longer(
    -origin, names_to = "resource_cat", values_to = "value"
  ) |>
  group_by(origin) |>
  mutate(share = value / sum(value)) |>
  ungroup()

ggplot(
  origin_diet,
  aes(x = factor(origin, levels = c("freshwater", "migratory", "marine")),
      y = share, fill = resource_cat)
) +
  geom_col(width = 0.6) +
  scale_fill_brewer(palette = "Set3") +
  labs(x = NULL, y = "Share of diet (resource categories)", fill = "Resource")
```

Freshwater and migratory species share a similar diet profile, both with a
substantial insect component; marine species show almost no insect
consumption (aquatic insects being a specifically freshwater resource) and
rely more on zooplankton and phytoplankton instead.

## Demersal/pelagic position by survey

Same idea, but for `demers_pelag` and grouped directly by `survey` rather
than by the freshwater/marine/migratory origin above: for each survey, the
share of distinct species falling into each demersal/pelagic category.

```{r}
#| fig-width: 7
#| fig-height: 5.5

species_survey <- size |>
  distinct(species_valid, survey)

survey_demers_pelag <- species_survey |>
  left_join(
    species_traits |> select(species, demers_pelag),
    by = c("species_valid" = "species")
  ) |>
  count(survey, demers_pelag) |>
  group_by(survey) |>
  mutate(share = n / sum(n)) |>
  ungroup()

ggplot(survey_demers_pelag, aes(x = survey, y = share, fill = demers_pelag)) +
  geom_col(width = 0.6) +
  scale_fill_brewer(palette = "Set3") +
  labs(x = NULL, y = "Share of species", fill = "Position")
```

`nurse` and `pomet` look alike: both dominated by demersal species (~60%),
with the rest split across pelagic-neritic and reef-associated. `river` is
the opposite - mostly benthopelagic (61%) and demersal (31%), with almost no
pelagic-neritic and, unsurprisingly, no reef-associated or bathydemersal
species at all.
