---
title: "Food web structures along the river-sea continuum"
bibliography: ../references.bib
format:
  html:
    theme: cosmo
    code-fold: true
    toc: true
    embed-resources: true
execute:
    message: false
    warning: false
---

## Todo

- [x] Species set difference and intersection between river and sea dataset
- [x] Plot metaweb
- [x] Plot local webs along the continuum
- [x] Refine food web plots
- [x] Add map with operations to local food web plot
- [x] Diet category analysis along the continuum
- [x] Add information to the metaweb (e.g. migratory status of species, invasive
or native, freshwater, marine, estuarine, etc.)
- [x] General information at the species level, such as, mean size, count,
status, etc.
- [ ] Most common invasive species
- [ ] Location of invasive species
- [ ] **(blocked)** Structure metrics versus distance to river mouth

## Setup

```{r}
library(renv)
renv::load()
library(targets)
library(here)
tar_config_set(
  store = here::here("_targets")
)
library(ggplot2)
library(dplyr)
library(sf)
devtools::load_all()
nice_theme()
```

## Species composition along the continuum

We start by looking at how much species overlap there is between the three
surveys that make up the river-estuary-sea continuum: `river` (freshwater,
ASPE), `nurse` (estuarine nursery grounds) and `pomet` (open coastal/sea).

```{r}
#| fig-width: 6
#| fig-height: 5.5

size <- tar_read(size_year_filtered)
plot_venn_diagram(size)
```

Three species are shared by all three surveys: *Anguilla anguilla* (European
eel, catadromous), *Chelon ramada* (grey mullet, euryhaline
estuarine migrant) and *Platichthys flesus* (European flounder, catadromous
flatfish).

More surprising is the `river`/`pomet` overlap (16 species), which excludes
`nurse` entirely. These 16 are all strict freshwater residents (e.g. roach,
bream, perch, pikeperch, barbel, chub, three-spined stickleback), not diadromous
species. These species most probably come from `pomet` fishing operation that
occurs deep into the estuaries (Girond and Loire).

So `river` and open sea share species through because of two reasons.
First, through diadromous species that live across the whole continuum.
Second, through residents caught by `pomet` operations deep into estuaries.

## Metaweb

The metaweb represents every possible predator-prey interaction across the
whole continuum, aggregated over all size classes so each species appears as
a single node. Species are placed vertically by trophic level, from basal
resources (grey, bottom) up to fish (top), colored by whether they prey on
other fish at any life stage, and sized by each species' mean observed body
length.

```{r}
#| fig-width: 14
#| fig-height: 9

web_list <- tar_read(web_list)
resource_list <- tar_read(resource_list)
diet <- tar_read(diet)
size <- tar_read(size_year_filtered)
plot_metaweb(web_list$metaweb, resource_list, diet, size)
```

`r nrow(aggregate_species_web(web_list$metaweb)) - length(resource_list)`
fish species and `r length(resource_list)` basal resources make up the
metaweb, for a total of `r sum(aggregate_species_web(web_list$metaweb))`
possible interactions at the species level.

The static plot above omits species labels for readability. For exploring
individual species, here is the same network as an interactive plot: hover
over a node to see its name, and zoom/pan to inspect dense areas. Fish nodes'
hover text also shows each species' environment, migratory category and
native/introduced status in France (see next section for where these come
from and how missing values are handled).

```{r}
#| fig-width: 11
#| fig-height: 9

species_traits <- tar_read(species_traits)
plot_metaweb_interactive(
  web_list$metaweb,
  resource_list,
  diet,
  size,
  species_traits
)
```

## Species-level information

Environment (freshwater/estuarine/marine/diadromous), demersal/pelagic
position and migratory category come from FishBase's `species()` table;
native/introduced status is FishBase's `country()` table filtered to France.
Two caveats:

- **Migratory category** (`migratory_category`) is unset for species FishBase
  doesn't code as diadromous, rather than explicitly labelled
  "non-migratory" - so `NA` here is ambiguous between "documented as
  non-migratory" and "not documented". In practice, almost all of these are
  strict freshwater residents (roach, bream, perch, etc.), so we read `NA`
  as "presumed non-migratory" without asserting it as a FishBase-sourced
  fact. Kept as `NA` rather than imputed.
- **Native status** (`native_status`) is collapsed to just "native" or
  "introduced": FishBase's finer-grained "endemic" and "not established"
  categories are folded into "native" for simplicity. The three Ponto-Caspian
  gobies *Neogobius melanostomus*, *Ponticola kessleri* and *Proterorhinus
  semilunaris* have no France record in FishBase's `country()` table, even
  though they are documented recent invaders in French rivers in the
  literature, so they are hardcoded to "introduced" rather than left `NA`.

```{r}
species_summary <- size |>
  group_by(species_valid) |>
  summarise(
    n_individuals = n(),
    n_operations = n_distinct(operation_id),
    mean_length = mean(length, na.rm = TRUE),
    .groups = "drop"
  ) |>
  left_join(species_traits, by = c("species_valid" = "species")) |>
  arrange(desc(n_individuals))
```

```{r}
#| fig-width: 10
#| fig-height: 3.5

count_bar <- function(data, var, title) {
  data |>
    count(.data[[var]]) |>
    mutate(val = forcats::fct_reorder(tidyr::replace_na(.data[[var]], "NA"), n)) |>
    ggplot(aes(n, val)) +
    geom_col() +
    labs(x = "Number of species", y = NULL, title = title)
}

patchwork::wrap_plots(
  count_bar(species_summary, "environment", "Environment"),
  count_bar(species_summary, "migratory_category", "Migratory category"),
  count_bar(species_summary, "native_status", "Native status (France)"),
  nrow = 1
)
```

The full species-level table: FishBase traits alongside observed count
(number of individuals and operations) and mean length in the combined
river-estuary-sea dataset.

```{r}
species_summary |>
  select(
    species = species_valid, environment, migratory_category, native_status,
    demers_pelag, n_individuals, n_operations, mean_length
  ) |>
  arrange(species) |>
  knitr::kable(digits = 1)
```

## Local food webs along the continuum

To see how food web structure evolves along the continuum, we pick three
individual operations along Gironde: one on the coast (`pomet`), one within the
Gironde estuary itself (`nurse`), and one upstream on the Dordogne (`river`).

```{r}
foodweb_structure <- tar_read(foodweb_structure)
operation_year_filtered <- tar_read(operation_year_filtered)

ops <- c(
  pomet = "61832307", nurse = "2022_1_GIR12_14_S", river = "83544"
)
titles <- c(
  pomet = "Coast",
  nurse = "Estuary",
  river = "River"
)
```

First, a map showing where each of the three operations sits along the
coastline/estuary/river.

```{r}
#| fig-width: 9
#| fig-height: 6

continuum_pts <- operation_year_filtered |>
  filter(operation_id %in% ops) |>
  mutate(
    label = titles[names(ops)[match(operation_id, ops)]]
  )

france <- rnaturalearth::ne_countries(
  geounit = "france", type = "map_units", scale = "medium", returnclass = "sf"
)
rivers_fr <- rnaturalearth::ne_download(
  scale = 10,
  type = "rivers_lake_centerlines",
  category = "physical",
  returnclass = "sf"
) |>
  sf::st_intersection(france)

e <- 0.4
ggplot() +
  geom_sf(data = rivers_fr, color = "grey60", linewidth = 0.3) +
  geom_sf(data = coastline(), linewidth = 0.3) +
  geom_point(
    data = continuum_pts,
    aes(longitude, latitude),
    color = "black",
    size = 3
  ) +
  ggrepel::geom_label_repel(
    data = continuum_pts,
    aes(longitude, latitude, label = label),
    size = 3.5
  ) +
  coord_sf(
    xlim = range(continuum_pts$longitude) + c(-e, e),
    ylim = range(continuum_pts$latitude) + c(-e, e)
  ) +
  labs(x = "Longitude", y = "Latitude") +
  theme(
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    panel.grid = element_blank()
  )
```

```{r}
#| fig-width: 16
#| fig-height: 6

# Shared size scale across all three panels, so `guides = "collect"` below
# merges them into a single legend instead of one per panel.
size_range <- (size |>
  filter(operation_id %in% ops) |>
  group_by(species_valid) |>
  summarise(v = mean(length, na.rm = TRUE), .groups = "drop"))$v |>
  range()

local_web_plots <- lapply(names(ops), function(nm) {
  fw_row <- foodweb_structure |> filter(operation_id == ops[nm])
  web <- fw_row$foodweb[[1]]
  op_size <- size |> filter(operation_id == ops[nm])
  subtitle <- sprintf(
    "Richness %d · Connectance %.2f · Mean TL %.1f",
    round(exp(fw_row$log_species_richness)), fw_row$connectance,
    fw_row$trophic_length
  )
  plot_local_foodweb(
    web, resource_list, diet, op_size,
    title = titles[nm], size_limits = size_range
  ) +
    labs(subtitle = subtitle) +
    theme(plot.subtitle = element_text(size = 9, color = "grey40"))
})

patchwork::wrap_plots(local_web_plots, nrow = 1) +
  patchwork::plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

The three webs illustrate a clear shift in composition: the coastal web is built
around marine flatfish and with several piscivorous predators; the estuarine web
is sparser, dominated by basal resources with only a couple of connected fish;
and the river web is dominated by small non-piscivorous cyprinids alongside a
handful of large piscivorous predators.

## Structure metrics along the continuum

Rather than comparing surveys as discrete boxplots, we map each operation's
structure metrics at its actual location, restricted to the Gironde
estuary/coast and the Dordogne river, to see how food web structure changes
continuously along the gradient instead of as three discrete groups.

We look at trophic length, diet overlap, connectance and species richness.
Diet overlap is the mean pairwise Jaccard similarity between predators' prey
sets (low = predators partition prey between them/complementarity, high =
predators eat largely the same things/redundancy) - unlike connectance, it
is close to independent of species richness
(`r round(cor(foodweb_structure$diet_overlap, foodweb_structure$log_species_richness), 2)`
here vs.
`r round(cor(foodweb_structure$connectance, foodweb_structure$log_species_richness), 2)`
for connectance), so it adds information richness alone doesn't already
give us.

```{r}
continuum_box <- foodweb_structure |>
  filter(longitude > -1.8, longitude < 1.3, latitude > 44.3, latitude < 46.0)

plot_metric_map <- function(data, metric, title) {
  ggplot(data) +
    geom_sf(
      data = rivers_fr, color = "grey70", linewidth = 0.3, inherit.aes = FALSE
    ) +
    geom_sf(data = coastline(), linewidth = 0.3, inherit.aes = FALSE) +
    geom_point(
      aes(longitude, latitude, color = .data[[metric]]), size = 1.8, alpha = 0.8
    ) +
    coord_sf(xlim = c(-1.8, 1.3), ylim = c(44.3, 46.0)) +
    scale_color_viridis_c() +
    labs(title = title, x = NULL, y = NULL, color = NULL) +
    theme(
      plot.background = element_rect(fill = "white", color = NA),
      panel.background = element_rect(fill = "white", color = NA),
      panel.grid = element_blank()
    )
}
```

```{r}
#| fig-width: 10
#| fig-height: 8

plot_metric_map(continuum_box, "trophic_length", "Trophic length")
```

```{r}
#| fig-width: 10
#| fig-height: 8

plot_metric_map(continuum_box, "diet_overlap", "Diet overlap")
```

```{r}
#| fig-width: 10
#| fig-height: 8

plot_metric_map(continuum_box, "connectance", "Connectance")
```

```{r}
#| fig-width: 10
#| fig-height: 8

plot_metric_map(continuum_box, "log_species_richness", "Log species richness")
```

## Diet composition by fish origin

We split species into three groups, reusing the same survey-membership
logic as the venn diagram above: `freshwater` (found only in `river`),
`marine` (found only in `nurse`/`pomet`), and `migratory` (found in both).
For each group, we look at the average diet composition across resource
categories (from the literature-curated diet table, not the constructed
food webs), normalised so each group's bars sum to 1.

```{r}
#| fig-width: 8
#| fig-height: 6.5

species_origin <- size |>
  distinct(species_valid, survey) |>
  group_by(species_valid) |>
  summarise(
    surveys = paste(sort(unique(survey)), collapse = ","), .groups = "drop"
  ) |>
  mutate(
    has_river = grepl("river", surveys),
    has_marine = grepl("nurse|pomet", surveys),
    origin = case_when(
      has_river & !has_marine ~ "freshwater",
      !has_river & has_marine ~ "marine",
      has_river & has_marine ~ "migratory"
    )
  )

species_diet <- diet |>
  select(species, all_of(resource_list)) |>
  group_by(species) |>
  summarise(across(everything(), mean), .groups = "drop") |>
  inner_join(
    species_origin |> select(species_valid, origin),
    by = c("species" = "species_valid")
  )

origin_diet <- species_diet |>
  group_by(origin) |>
  summarise(across(all_of(resource_list), mean), .groups = "drop") |>
  tidyr::pivot_longer(
    -origin, names_to = "resource_cat", values_to = "value"
  ) |>
  group_by(origin) |>
  mutate(share = value / sum(value)) |>
  ungroup()

ggplot(
  origin_diet,
  aes(x = factor(origin, levels = c("freshwater", "migratory", "marine")),
      y = share, fill = resource_cat)
) +
  geom_col(width = 0.6) +
  scale_fill_brewer(palette = "Set3") +
  labs(x = NULL, y = "Share of diet (resource categories)", fill = "Resource")
```

Freshwater and migratory species share a similar diet profile, both with a
substantial insect component; marine species show almost no insect
consumption (aquatic insects being a specifically freshwater resource) and
rely more on zooplankton and phytoplankton instead.
